### 2020-06-04 presentation/presentation_job_market.Rmd
initialize repo with extant files

---
title: "Government through Patronage"
subtitle: "Bargaining for Education in Decentralized Brazil"
author: "Galileu Kim"
institute: "Princeton University"
output:
  beamer_presentation:
    includes:
      in_header: "presentation_template.txt"
    incremental: false
    keep_tex: true
    slide_level: 2
---

```{r setup, include = F}
library(knitr)
library(here)
source(
  here("scripts/thesis_setup.R")
)

update_geom_defaults(
  "point",
  list(color = "#375b7c", size = 1)
)

update_geom_defaults(
  "line",
  list(color = "steelblue3", size = 0.7)
)

path_figs <- "papers/paper_cp_2020/figs/"

opts_chunk$set(echo = FALSE, out.width = "230px", cache = F, root.dir = here())
```

## Education: a global perspective

- Children across the world rely on governments for education.
- Access improved significantly over the past decades.
- Improvements in quality remain a challenge.
- "Schooling does not always lead to learning. Worldwide, there are more non-learners in school than out of school."
    - \source{UNICEF.}

## Global education

```{r out.width = '240px'}
include_graphics(
  'figs/plot_global_1.pdf'
)
```

## Education in Brazil

```{r out.width = '240px'}
include_graphics(
  'figs/plot_global_2.pdf'
)
```

## A domestic challenge

- Each country faces a particular set of challenges.
    - Widespread institutional reforms make cases comparable.
- Decentralization: subnational politicians have direct impact on public services. 
    - \source{Falleti 2010, Arretche 2017.}
- Quality of education varies along subnational lines.

## Mapping domestic inequality

```{r, fig.cap= "Municipal variation in the quality of education (2015). Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores.", out.width = "180px"}
include_graphics(
  "figs/saeb_map.pdf"
)
```

## Findings

- Decentralization delegates management over personnel.
    - Mayors and legislators have competing claims.
- Mayors allocate public sector jobs to gain legislative support.
    - Increased control by the executive reduces patronages.
- Bureaucratic turnover has negative effects on education.
    - Students perform worse on standardized exams.

## Contribution

- Growing literature on the politics of personnel and public services.
    - \source{Pepinsky 2013, Finan et al. 2015, Akhtari et al. 2017.}
- Interaction between political elites in local government.
    - \source{Klasnja and Titiunik 2017, Novaes 2019.}
- Effect of competition in weakly institutionalized countries.
    - \source{Gottlieb and Kosec 2019.}

## Outline

\singlespacing

\tableofcontents

# Related literature

## Bureaucracy and public goods provision

- Initial emphasis on industrialization and the developmental state.
    - \source{Johnson 1982, Kohli 2004.}
- Political control can make bureaucracies deliver better public services.
    -  \source{Toral 2018, Raffler 2016, Gulzar and Pasquale 2017.}
- Analyze who are the politicians managing bureaucracies.
    - And who do they govern with.

## Presidential coalitionism and patronage

- Executive and legislators reshape policy and bureaucratic institutions. 
    - \source{Cameron and McCarty 2004, McCarty 2004.}
- Presidential coalitionism.
    -  Mayor (president) builds a coalition by allocating cabinet positions to city councilors (Congress members
    - \source{Raile et al. 2011, Amorim Neto 2000.}
- Jobs allocated to city councilors.
    - "City councilors knocked on my door with...people they wanted me to hire." \source{- former mayor of S.}

## Bureaucratic turnover and inexperienced education

- Staff rotation induces negative productivity shock.
    - \source{Gailmard and Patty 2007.}
- Students taught by inexperienced teachers and principals learn less.
    - \source{Clotfelter et al. 2007, Akhtari et al. 2017.}
- "I am aware that the position is temporary. Especially because it is a political position, decided by the administration." \source{- school principal in municipality I.}

# Theory

## Theory

- Model the executive decision over patronage allocation during government.
    - Mayors need to coopt a winning coalition in the legislature.
- Trade-off between patronage and quality of public services.
    - Probability of reelection vs. governance.
- From pre-electoral bargains with brokers to administrative term.
    - \source{Stokes et al. 2013, Robinson and Verdier 2013.}

## Prediction
- Mayors with a larger coalition engage in less patronage.
    -  Executive prioritizes cheap votes.
- Legislators who already formed an electoral alliance with the government.
    - \source{Groseclose and Snyder 1996, Banks 2000.}
- Controlling less seats makes buying a majority more expensive.
    - Indifferent or opposition legislators demand more jobs.

# Institutional context

## Public education in Brazil

- Over 20 million students enrolled in municipal schools.
    - 1.2 million public teachers and 100 thousand school principals.
- Primary education managed by municipal governments.
    - Personnel decisions centralized by the executive, subject to legislative influence.
- "We are invited to work at the school by the department of education, with the city councilor...appointing people they think have the necessary qualifications." \source{- school principal A.}

## Enrollment by administrative level

```{r fig.cap = "A decentralized system. Lower school education is primarily under the responsibility of municipalities."}
include_graphics(
  "figs/enrollment_dep.pdf"
)
```

## Local government

- Local politicians elected for four year terms, with reelection.
    - 5 thousand mayors and over 50 thousand city councilors.
- Executive responsible for delivering public services.
    - City council gives budgetary approval and legal oversight.
- Executive coalitions declared during electoral campaign.
    - Political alliance for electoral campaign and during mayoral term.

# Research design and data

## Research design

- Bureaucratic turnover $\rightarrow$ student learning.
\begin{align*}
\text{test score}_{isjt} &= \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
&\psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
\end{align*}

- Model specification: hierarchical linear modeling with heterogeneous effect by grade and state-year fixed effects.
    - Outcome of interest: average student test scores at the classroom level.
    - Treatment: (1) teacher and school principal turnover, (2) teacher turnover index.
    - Unit of analysis: classroom.

## Research design

- Executive coalition $\rightarrow$ bureaucratic turnover.
$$\text{turnover}_{sjt} = \gamma \text{coalition seats}_{jt} + \mu P_{jt} + \zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{jt}$$

- Model specification: (1) logistic regression, (2) linear model with state-year fixed effects.
    - Outcome of interest: (1) hiring/dismissal of teacher/school principal, (2) proportion of dismissal and hiring.
    - Treatment: share of legislative seats held by executive coalition.
    - Unit of analysis: (1) individual bureaucrat, (2) municipal bureaucracy.

## Data

- Student learning.
    - SAEB (2007 - 2015): national test scores for 5 million students (5th and 9th grades).
    - SPAECE (2007 - 2018): state test scores for 400 thousand students (2nd, 5th and 9th grades).
- Educational staff.
    - National school census (1995 - 2016): annual census of 100 thousand schools.
    - RAIS (1985 - 2015): annual employment data on school principals and teachers (over 1 million per year).
- Fieldwork.
    - Interviews conducted in summmers of 2018 and 2019 in Brazil.
    - Mayors, secretaries of education, school principals and teachers.

## Bureaucratic turnover by department

```{r staff, fig.cap = 'Municipal bureaucracies. Total number of bureaucrats and hiring/dismissal rates for bureaucratic staff over time. Disaggregated by department.'}
include_graphics(
  'figs/turnover_category.pdf'
)
```

## Results

- Staff turnover has negative effects on student learning.
    - Robust to different turnover and standardized exam specifications.
- Increase in patronage in municipalities with lower share of executive-held seats.
    - Similar pattern for different turnover specifications (hiring, dismissal and turnover).

## Staff turnover and student test scores: SAEB and SPAECE

\begin{table}[t]
  \centering
  \resizebox*{!}{\height}{%
  \input{tables/student_learning}%
  }
  \caption{\textbf{Bureaucratic turnover and student learning.} Higher turnover in teachers and principals reduce student learning. All models include year and state fixed effects.}
  \label{tbl:hlm_mods}
\end{table}

## Staff turnover and student test scores: SAEB and SPAECE

```{r, fig.cap= "Relative impact of bureaucratic turnover. Point estimates and confidence intervals retrieved from models with controls.", out.width = "210px"}
include_graphics(
  "figs/turnover_fit.pdf"
)
```

## Executive coalition and staff turnover

\begin{table}[t]
  \centering
  \resizebox*{!}{\height}{%
  \input{tables/turnover_result}%
  }
  \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals. Models 1, 3, and 5 include year and state fixed effects.}
  \label{tbl:turnover}
\end{table}

## Executive coalition and staff turnover

```{r fig.cap = "Predicted and actual values for bureaucratic turnover. The line graph plots the predicted proportion of educational staff hired or firede. Circles' size is proportional to the number of observations contained in the data."}
include_graphics(
  "figs/plot_pred.pdf"
)
```

# Conclusion

## Conclusion

- Subnational political actors manage public services.
    - Political elites interact and negotiate with one another.
- Bureaucratic turnover results from executives coopting legislators.
    - Negative downstream consequences for the quality of education.
- Is political competition welfare enhancing for citizen?
    - Future research can theorize on how heterogeneous political elites reshape governments.

## Thank you!

If you have any comments/suggestions:

\fontfamily{cmtt}\selectfont galileuk@princeton.edu

\appendix

### 2020-06-04 presentation/presentation_job_market.tex
initialize repo with extant files

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[ignorenonframetext,]{beamer}
\usepackage{pgfpages}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
% Prevent slide breaks in the middle of a paragraph:
\widowpenalties 1 10000
\raggedbottom
\setbeamertemplate{part page}{
\centering
\begin{beamercolorbox}[sep=16pt,center]{part title}
  \usebeamerfont{part title}\insertpart\par
\end{beamercolorbox}
}
\setbeamertemplate{section page}{
\centering
\begin{beamercolorbox}[sep=12pt,center]{part title}
  \usebeamerfont{section title}\insertsection\par
\end{beamercolorbox}
}
\setbeamertemplate{subsection page}{
\centering
\begin{beamercolorbox}[sep=8pt,center]{part title}
  \usebeamerfont{subsection title}\insertsubsection\par
\end{beamercolorbox}
}
\AtBeginPart{
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \frame{\subsectionpage}
}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Government through Patronage},
            pdfauthor={Galileu Kim},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\newif\ifbibliography
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usetheme{metropolis}
\setbeamertemplate{footline}[page number]{}

% packages
\usepackage{xcolor}
\usepackage{ragged2e}
\usepackage{setspace}
\usepackage{etoolbox}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{blindtext}
\captionsetup[figure]{font=scriptsize} % set size of caption
\AtBeginEnvironment{knitrout}{\singlespacing}
\useoutertheme{infolines} % Alternatively: miniframes, infolines, split
\useinnertheme{circles}
\date[]{\today}

% colors
\definecolor{princeton}{rgb}{1.0, 0.6, 0.2} % princeton (primary)
\definecolor{arsenic}{rgb}{0.23, 0.27, 0.29}

\setbeamercolor{title}{fg=black}
\setbeamercolor{frametitle}{fg=white,bg=arsenic}
\setbeamercolor{section in head/foot}{fg=arsenic,bg=princeton}
\setbeamerfont{frametitle}{size=\Large}
\setbeamercolor{palette primary}{fg=white,bg=arsenic}
\setbeamercolor{palette secondary}{fg=arsenic,bg=princeton}
\setbeamercolor{palette tertiary}{fg=princeton,bg=arsenic}
\setbeamercolor{palette quaternary}{fg=princeton,bg=arsenic}
\setbeamercolor{structure}{fg=princeton} % itemize, enumerate, etc
\setbeamercolor{section in toc}{fg=arsenic} % TOC sections
\setbeamerfont{subtitle}{size=\small}

% override palette coloring with secondary
\setbeamercolor{subsection in head/foot}{bg=arsenic,fg=white}

% defining font
\usepackage[T1]{fontenc}

% Cabin
\usepackage[sfdefault,condensed]{cabin}

% Helvetica
%\usepackage[scaled]{helvet}
%\renewcommand\familydefault{\sfdefault}

% Inria Sans
%\usepackage[lining]{InriaSans}
%\renewcommand*\oldstylenums[1]{{\fontfamily{InriaSans-OsF}\selectfont #1}}
%% The font package uses mweights.sty which has som issues with the
%% \normalfont command. The following two lines fixes this issue.
%\let\oldnormalfont\normalfont
%\def\normalfont{\oldnormalfont\mdseries}

% FiraSans
%\usepackage[sfdefault]{FiraSans}
%\renewcommand*\oldstylenums[1]{{\firaoldstyle #1}}
%\usepackage{ragged2e}

% aesthetics
\setbeamersize{text margin left=10mm,text margin right=10mm}

% source command
\newcommand{\source}{\footnotesize\textcolor{black!25}}

% suppress section slides
\AtBeginPart{}
\AtBeginSubsection{}
\AtBeginSection{}
\setlength{\emergencystretch}{0em}
\setlength{\parskip}{0pt}

\title{Government through Patronage}
\providecommand{\subtitle}[1]{}
\subtitle{Bargaining for Education in Decentralized Brazil}
\author{Galileu Kim}
\providecommand{\institute}[1]{}
\institute{Princeton University}
\date{}

\begin{document}
\frame{\titlepage}

\begin{frame}{Education: a global perspective}
\protect\hypertarget{education-a-global-perspective}{}

\begin{itemize}
\tightlist
\item
  Children across the world rely on governments for education.
\item
  Access improved significantly over the past decades.
\item
  Improvements in quality remain a challenge.
\item
  ``Schooling does not always lead to learning. Worldwide, there are
  more non-learners in school than out of school.''

  \begin{itemize}
  \item
    \source{UNICEF.}
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Global education}
\protect\hypertarget{global-education}{}

\begin{center}\includegraphics[width=240px]{figs/plot_global_1} \end{center}

\end{frame}

\begin{frame}{Education in Brazil}
\protect\hypertarget{education-in-brazil}{}

\begin{center}\includegraphics[width=240px]{figs/plot_global_2} \end{center}

\end{frame}

\begin{frame}{A domestic challenge}
\protect\hypertarget{a-domestic-challenge}{}

\begin{itemize}
\tightlist
\item
  Each country faces a particular set of challenges.

  \begin{itemize}
  \tightlist
  \item
    Widespread institutional reforms make cases comparable.
  \end{itemize}
\item
  Decentralization: subnational politicians have direct impact on public
  services.

  \begin{itemize}
  \item
    \source{Falleti 2010, Arretche 2017.}
  \end{itemize}
\item
  Quality of education varies along subnational lines.
\end{itemize}

\end{frame}

\begin{frame}{Mapping domestic inequality}
\protect\hypertarget{mapping-domestic-inequality}{}

\begin{figure}

{\centering \includegraphics[width=180px]{figs/saeb_map}

}

\caption{Municipal variation in the quality of education (2015). Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores.}\label{fig:unnamed-chunk-3}
\end{figure}

\end{frame}

\begin{frame}{Findings}
\protect\hypertarget{findings}{}

\begin{itemize}
\tightlist
\item
  Decentralization delegates management over personnel.

  \begin{itemize}
  \tightlist
  \item
    Mayors and legislators have competing claims.
  \end{itemize}
\item
  Mayors allocate public sector jobs to gain legislative support.

  \begin{itemize}
  \tightlist
  \item
    Increased control by the executive reduces patronages.
  \end{itemize}
\item
  Bureaucratic turnover has negative effects on education.

  \begin{itemize}
  \tightlist
  \item
    Students perform worse on standardized exams.
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Contribution}
\protect\hypertarget{contribution}{}

\begin{itemize}
\tightlist
\item
  Growing literature on the politics of personnel and public services.

  \begin{itemize}
  \item
    \source{Pepinsky 2013, Finan et al. 2015, Akhtari et al. 2017.}
  \end{itemize}
\item
  Interaction between political elites in local government.

  \begin{itemize}
  \item
    \source{Klasnja and Titiunik 2017, Novaes 2019.}
  \end{itemize}
\item
  Effect of competition in weakly institutionalized countries.

  \begin{itemize}
  \item
    \source{Gottlieb and Kosec 2019.}
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Outline}
\protect\hypertarget{outline}{}

\singlespacing

\tableofcontents

\end{frame}

\hypertarget{related-literature}{%
\section{Related literature}\label{related-literature}}

\begin{frame}{Bureaucracy and public goods provision}
\protect\hypertarget{bureaucracy-and-public-goods-provision}{}

\begin{itemize}
\tightlist
\item
  Initial emphasis on industrialization and the developmental state.

  \begin{itemize}
  \item
    \source{Johnson 1982, Kohli 2004.}
  \end{itemize}
\item
  Political control can make bureaucracies deliver better public
  services.

  \begin{itemize}
  \item
    \source{Toral 2018, Shaffler 2016, Gulzar and Pasquale 2017.}
  \end{itemize}
\item
  Analyze who are the politicians managing bureaucracies.

  \begin{itemize}
  \tightlist
  \item
    And who do they govern with.
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Presidential coalitionism and patronage}
\protect\hypertarget{presidential-coalitionism-and-patronage}{}

\begin{itemize}
\tightlist
\item
  Executive and legislators reshape policy and bureaucratic
  institutions.

  \begin{itemize}
  \item
    \source{Cameron and McCarty 2004, McCarty 2004.}
  \end{itemize}
\item
  Presidential coalitionism.

  \begin{itemize}
  \item
    Mayor (president) builds a coalition by allocating cabinet positions
    to city councilors (Congress members
  \item
    \source{Raile et al. 2011, Amorim Neto 2000.}
  \end{itemize}
\item
  Jobs allocated to city councilors.

  \begin{itemize}
  \tightlist
  \item
    ``City councilors knocked on my door with\ldots{}people they wanted
    me to hire.'' \source{- former mayor of S.}
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Bureaucratic turnover and inexperienced education}
\protect\hypertarget{bureaucratic-turnover-and-inexperienced-education}{}

\begin{itemize}
\tightlist
\item
  Staff rotation induces negative productivity shock.

  \begin{itemize}
  \item
    \source{Gailmard and Patty 2007.}
  \end{itemize}
\item
  Students taught by inexperienced teachers and principals learn less.

  \begin{itemize}
  \item
    \source{Clotfelter et al. 2007, Akhtari et al. 2017.}
  \end{itemize}
\item
  ``I am aware that the position is temporary. Especially because it is
  a political position, decided by the administration.''
  \source{- school principal in municipality I.}
\end{itemize}

\end{frame}

\hypertarget{theory}{%
\section{Theory}\label{theory}}

\begin{frame}{Theory}
\protect\hypertarget{theory-1}{}

\begin{itemize}
\tightlist
\item
  Model the executive decision over patronage allocation during
  government.

  \begin{itemize}
  \tightlist
  \item
    Mayors need to coopt a winning coalition in the legislature.
  \end{itemize}
\item
  Trade-off between patronage and quality of public services.

  \begin{itemize}
  \tightlist
  \item
    Probability of reelection vs.~governance.
  \end{itemize}
\item
  From pre-electoral bargains with brokers to administrative term.

  \begin{itemize}
  \item
    \source{Stokes et al. 2013, Robinson and Verdier 2013.}
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Prediction}
\protect\hypertarget{prediction}{}

\begin{itemize}
\tightlist
\item
  Mayors with a larger coalition engage in less patronage.

  \begin{itemize}
  \tightlist
  \item
    Executive prioritizes cheap votes.
  \end{itemize}
\item
  Legislators who already formed an electoral alliance with the
  government.

  \begin{itemize}
  \item
    \source{Groseclose and Snyder 1996, Banks 2000.}
  \end{itemize}
\item
  Controlling less seats makes buying a majority more expensive.

  \begin{itemize}
  \tightlist
  \item
    Indifferent or opposition legislators demand more jobs.
  \end{itemize}
\end{itemize}

\end{frame}

\hypertarget{institutional-context}{%
\section{Institutional context}\label{institutional-context}}

\begin{frame}{Public education in Brazil}
\protect\hypertarget{public-education-in-brazil}{}

\begin{itemize}
\tightlist
\item
  Over 20 million students enrolled in municipal schools.

  \begin{itemize}
  \tightlist
  \item
    1.2 million public teachers and 100 thousand school principals.
  \end{itemize}
\item
  Primary education managed by municipal governments.

  \begin{itemize}
  \tightlist
  \item
    Personnel decisions centralized by the executive, subject to
    legislative influence.
  \end{itemize}
\item
  ``We are invited to work at the school by the department of education,
  with the city councilor\ldots{}appointing people they think have the
  necessary qualifications.'' \source{- school principal A.}
\end{itemize}

\end{frame}

\begin{frame}{Enrollment by administrative level}
\protect\hypertarget{enrollment-by-administrative-level}{}

\begin{figure}

{\centering \includegraphics[width=230px]{figs/enrollment_dep}

}

\caption{A decentralized system. Lower school education is primarily under the responsibility of municipalities.}\label{fig:unnamed-chunk-4}
\end{figure}

\end{frame}

\begin{frame}{Local government}
\protect\hypertarget{local-government}{}

\begin{itemize}
\tightlist
\item
  Local politicians elected for four year terms, with reelection.

  \begin{itemize}
  \tightlist
  \item
    5 thousand mayors and over 50 thousand city councilors.
  \end{itemize}
\item
  Executive responsible for delivering public services.

  \begin{itemize}
  \tightlist
  \item
    City council gives budgetary approval and legal oversight.
  \end{itemize}
\item
  Executive coalitions declared during electoral campaign.

  \begin{itemize}
  \tightlist
  \item
    Political alliance for electoral campaign and during mayoral term.
  \end{itemize}
\end{itemize}

\end{frame}

\hypertarget{research-design-and-data}{%
\section{Research design and data}\label{research-design-and-data}}

\begin{frame}{Research design}
\protect\hypertarget{research-design}{}

\begin{itemize}
\item
  Bureaucratic turnover \(\rightarrow\) student learning. \begin{align*}
  \text{test score}_{isjt} &= \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
  &\psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
  \end{align*}
\item
  Model specification: hierarchical linear modeling with heterogeneous
  effect by grade and state-year fixed effects.

  \begin{itemize}
  \tightlist
  \item
    Outcome of interest: average student test scores at the classroom
    level.
  \item
    Treatment: (1) teacher and school principal turnover, (2) teacher
    turnover index.
  \item
    Unit of analysis: classroom.
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Research design}
\protect\hypertarget{research-design-1}{}

\begin{itemize}
\item
  Executive coalition \(\rightarrow\) bureaucratic turnover.
  \[\text{turnover}_{sjt} = \gamma \text{coalition seats}_{jt} + \mu P_{jt} + \zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{jt}\]
\item
  Model specification: (1) logistic regression, (2) linear model with
  state-year fixed effects.

  \begin{itemize}
  \tightlist
  \item
    Outcome of interest: (1) hiring/dismissal of teacher/school
    principal, (2) proportion of dismissal and hiring.
  \item
    Treatment: share of legislative seats held by executive coalition.
  \item
    Unit of analysis: (1) individual bureaucrat, (2) municipal
    bureaucracy.
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Data}
\protect\hypertarget{data}{}

\begin{itemize}
\tightlist
\item
  Student learning.

  \begin{itemize}
  \tightlist
  \item
    SAEB (2007 - 2015): national test scores for 5 million students (5th
    and 9th grades).
  \item
    SPAECE (2007 - 2018): state test scores for 400 thousand students
    (2nd, 5th and 9th grades).
  \end{itemize}
\item
  Educational staff.

  \begin{itemize}
  \tightlist
  \item
    National school census (1995 - 2016): annual census of 100 thousand
    schools.
  \item
    RAIS (1985 - 2015): annual employment data on school principals and
    teachers (over 1 million per year).
  \end{itemize}
\item
  Fieldwork.

  \begin{itemize}
  \tightlist
  \item
    Interviews conducted in summmers of 2018 and 2019 in Brazil.
  \item
    Mayors, secretaries of education, school principals and teachers.
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Bureaucratic turnover by department}
\protect\hypertarget{bureaucratic-turnover-by-department}{}

\begin{figure}

{\centering \includegraphics[width=230px]{figs/turnover_category}

}

\caption{Municipal bureaucracies. Total number of bureaucrats and hiring/dismissal rates for bureaucratic staff over time. Disaggregated by department.}\label{fig:staff}
\end{figure}

\end{frame}

\begin{frame}{Results}
\protect\hypertarget{results}{}

\begin{itemize}
\tightlist
\item
  Staff turnover has negative effects on student learning.

  \begin{itemize}
  \tightlist
  \item
    Robust to different turnover and standardized exam specifications.
  \end{itemize}
\item
  Increase in patronage in municipalities with lower share of
  executive-held seats.

  \begin{itemize}
  \tightlist
  \item
    Similar pattern for different turnover specifications (hiring,
    dismissal and turnover).
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Staff turnover and student test scores: SAEB and SPAECE}
\protect\hypertarget{staff-turnover-and-student-test-scores-saeb-and-spaece}{}

\begin{table}[t]
  \centering
  \resizebox*{!}{\height}{%
  \input{tables/student_learning}%
  }
  \caption{\textbf{Bureaucratic turnover and student learning.} Higher turnover in teachers and principals reduce student learning. All models include year and state fixed effects.}
  \label{tbl:hlm_mods}
\end{table}

\end{frame}

\begin{frame}{Staff turnover and student test scores: SAEB and SPAECE}
\protect\hypertarget{staff-turnover-and-student-test-scores-saeb-and-spaece-1}{}

\begin{figure}

{\centering \includegraphics[width=210px]{figs/turnover_fit}

}

\caption{Relative impact of bureaucratic turnover. Point estimates and confidence intervals retrieved from models with controls.}\label{fig:unnamed-chunk-5}
\end{figure}

\end{frame}

\begin{frame}{Executive coalition and staff turnover}
\protect\hypertarget{executive-coalition-and-staff-turnover}{}

\begin{table}[t]
  \centering
  \resizebox*{!}{\height}{%
  \input{tables/turnover_result}%
  }
  \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals. Models 1, 3, and 5 include year and state fixed effects.}
  \label{tbl:turnover}
\end{table}

\end{frame}

\begin{frame}{Executive coalition and staff turnover}
\protect\hypertarget{executive-coalition-and-staff-turnover-1}{}

\begin{figure}

{\centering \includegraphics[width=230px]{figs/plot_pred}

}

\caption{Predicted and actual values for bureaucratic turnover. The line graph plots the predicted proportion of educational staff hired or firede. Circles' size is proportional to the number of observations contained in the data.}\label{fig:unnamed-chunk-6}
\end{figure}

\end{frame}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\begin{frame}{Conclusion}
\protect\hypertarget{conclusion-1}{}

\begin{itemize}
\tightlist
\item
  Subnational political actors manage public services.

  \begin{itemize}
  \tightlist
  \item
    Political elites interact and negotiate with one another.
  \end{itemize}
\item
  Bureaucratic turnover results from executives coopting legislators.

  \begin{itemize}
  \tightlist
  \item
    Negative downstream consequences for the quality of education.
  \end{itemize}
\item
  Is political competition welfare enhancing for citizen?

  \begin{itemize}
  \tightlist
  \item
    Future research can theorize on how heterogeneous political elites
    reshape governments.
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Thank you!}
\protect\hypertarget{thank-you}{}

If you have any comments/suggestions:

\fontfamily{cmtt}\selectfont \href{mailto:galileuk@princeton.edu}{\nolinkurl{galileuk@princeton.edu}}

\appendix

\end{frame}

\end{document}

### 2020-06-04 publication/paper_job_market.Rmd
initialize repo with extant files

---
title: "Government through Patronage"
subtitle: "Bargaining for Education in Decentralized Brazil"
author: Galileu Kim
date: "`r gsub('^0', '', format(Sys.time(), '%d %B, %Y'))`"
documentclass: article
fontsize: 12pt
geometry: margin=1.5in
abstract: "Why do politicians engage in patronage? What are its consequences for public service delivery? I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators. Leveraging fieldwork evidence and administrative data, I demonstrate that mayors with a weaker hold on the local legislature resort to patronage to co-opt legislators. Educational staff turnover has negative downstream effects on student learning, but limited electoral backlash. Weak electoral accountability channels suggests that patronage is primarily a political elite game, with little involvement by the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare."
header-includes:
  - \setlength\parindent{20pt}
  - \usepackage{setspace}
  - \usepackage{multirow}
  - \usepackage{caption}
  - \usepackage{titling}
  - \usepackage{amsmath}
  - \DeclareMathOperator*{\argmax}{arg\,max}
  - \onehalfspacing
  # - \pretitle{\centering\Large\textbf}
  # - \preauthor{\centering\large}
  # - \predate{\centering\normalsize}
  - \usepackage[hang,flushmargin]{footmisc}
bibliography: dissertation.bib
output: 
  pdf_document:
    citation_package: biblatex
keep_tex: TRUE
biblio-style: authoryear
---
\pagebreak

```{r setup, include=FALSE}
source(
  here::here("scripts/thesis_setup.R")
)

opts_chunk$set(echo = FALSE, fig.width = 6, fig.height = 4, cache = F, root.dir = here())
```

Across the world, vulnerable populations rely on governments for access to basic public services such as education. Over the past two decades, despite setbacks during neoliberal reforms, enrollment rates in public education have improved dramatically.^[See @haggard_development_2008, @kaufman_globalization_2001.] Improvements in access have not been matched, for many, with a better quality education. From a global perspective, educational quality indicators such as the PISA (Programme for International Student Assessment) score have broadly stagnated.^[For coverage, see [here](https://www.economist.com/international/2019/12/05/pisa-results-can-lead-policymakers-astray).]

How can we improve the quality of education? In large part, the answers are domestic. While each country faces a particular set of challenges, broad institutional reforms in the developing world make their cases comparable. Decentralization has delegated responsibility for public education to subnational governments, as well as induced variation in the quality of education along subnational lines (@falleti_decentralization_2010, @arretche_mitos_1996). Understanding how this level of government manages these services, and in particular how local actors strategically shape them can help find answers to that question. (@gulzar_politicians_2017, @min_power_2015).

```{r fig.cap = "Global enrollment rates and PISA test scores. Net enrollment rates calculated for primary education. PISA test scores based on a sample design."}
include_graphics(
  "figs/plot_global.pdf"
)
```

In this paper, I highlight the role of the strategic interaction between political elites and its downstream effects on the provision of educational services. Similar to coalition building, local executive leaders secure broader support for their policy agenda through the allocation of public sector positions to city councilors.^[@laver_coalitions_1990, @power_optimism_2010.] As highlighted by previous research, teacher turnover can lead to decreased morale and inexperienced staff entering public service (@ingersoll_teacher_2001, @ronfeldt_how_2013). Patronage, by inducing new hires or dismissals, can therefore cause instability in the educational sector, with negative effects for student learning.

<!-- mention the growing investment in education, public sector jobs, etc. -->

To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education, both at a national and subnational level. I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory, and a school-grade-specific turnover index. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage, in particular during the first year of administration. Mayors therefore resort to patronage in order to shore up legislative support, with negative consequences for student learning. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

This study contributes to an emerging literature on the politics of personnel and public services (@pepinsky_bureaucracy_2017, @finan_personnel_2015 and @gulzar_politicians_2017). Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services (@ferraz_corrupting_2012). It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy (@kohli_state-directed_2004, @evans_embedded_1995).

The paper is structured as follows. Section 1 provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. In section 2 I present the main theoretical argument, with a formal treatment of the subject. Section 3 presents the institutional context and data, followed in section 4 by the research design and main results. Section 5 concludes.

# Related literature and debate

In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions. I also highlight how my research incorporates multiplicity in political actors and how this affects bargaining over public sector jobs. I address this gap by adapting previous analyses of executive-legislative bargaining to bureaucratic control at the local level.

## Bureaucratic personnel and public goods provision

Bureaucracies have a clear impact on the delivery of public services. A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (@centeno_unpacking_2017, @kohli_state-directed_2004). A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (@johnson_miti_1982, kohli_state-directed_2004). A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (@evans_bureaucracy_1999).

Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. @gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. @akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

This recent wave of studies shed light on the intersection between politicians and bureaucracies. However, few of these studies explicitly model the multiple actors involved in managing bureaucracies. Understanding their diverse goals and action space provides a firm theoretical foundation to how different politicians can reshape bureaucracies. To do so I turn to the well-established literature on executive-legislative bargaining, applying its insights to the analysis of local government and administration.

## Presidential coalitionism and patronage

For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (@grindle_jobs_2012, @mccarty_appointments_2004). Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. (@raile_executive_2011, @power_optimism_2010).

In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. (@raile_executive_2011). In a setting characterized by weak party cohesion and programmatic commitments (@ames_electoral_1995, @lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (@batista_o_2013, @neto_presidential_2006, @figueiredo_executivo_1999).

<!-- Something about coalitions. -->

In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.^[Interview with C, August and September 2019.] Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."^[Interview with C, August 2019.] These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

## Bureaucratic turnover and inexperienced education

Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (@gailmard_slackers_2007). Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (@clotfelter_teacher_2007.) @akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (@colonnelli_patronage_2017).

> "I am aware that the position is temporary. Especially because it is a political position, decided by the administration. If the current administration is out of power, we are automatically dismissed." - Interview with school principal A, August 2019.

Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level. In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers. In the following section I describe the institutional context for public education in Brazil, as well as the data employed for the the estimation.

# The argument

## The executive dilemma

I model the executive decision over patronage as a product of political career concerns and strategic interaction with the local legislative power. In particular, mayors are assumed to be career-oriented and re-election seeking (@besley_principled_2006). This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (@stokes_brokers_2013, @robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.

The mayor's strategy centers around the redistribution public sector jobs. When the mayor takes office, a set of bureaucratic positions become available for allocation to city councilors. In a municipal parallel to presidential coalitionism, jobs are allocated to legislators in order to buy their vote. I assume, for now, that job allocations have a negative effect on public services. The mayor can refrain from distributing jobs for vote buying, since reducing turnover helps retain expertise and improves the quality of public services. However, doing so raises the threat of defection by coalition members, which can lead to the withholding of budgetary approval. This is the executive's dilemma.

The main proposition is that mayors who control a smaller (larger) share of legislative seats allocate more (less) public sector job to city councilors. The intuition is as follows. Mayors need to secure a simple majority of legislative votes to have the budget approved. @groseclose_buying_1996 and @banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor. Having exhausted these votes, the executive has to resort to buying votes from the opposition, which demand more patronage. The less seats occupied by members of the mayor's electoral coalition, the more expensive vote-buying becomes, and the observable implication is increased patronage in the municipality.

The model builds on rich theoretical treatment of electoral accountability and majority-buying in legislatures. Starting with @barro_control_1973 and @ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (@coate_form_1995; @besley_principled_2006). The possibility of reelection can motivate them to exert effort and convince voters to reelect them. This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (@groseclose_buying_1996, @banks_buying_2000).

## The model

There is one mayor $m$, a city council $N$ and representative median voter $v$. The goal of the mayor is to be reelected into office, yet for any given time period $m$ needs to garner legislative support. The city council plays that key role by approving or not the budget for the executive. The legislative chamber is composed by $N = {1,..,n}$ legislators, where $S \subseteq N$ denote the subset of legislators who belong to the electoral coalition of mayor $m$. For each $i \in N$, let $\textbf{v} = \{v_1,...,v_n\}$ describe legislator i's preference to support the mayor in the budgetary process. Let $v_i > v_j$ for all $i \in S$ and $j \in \{N \setminus S\}$. Therefore, members of the mayor's electoral coalition are more likely to approve the budget than outside members.

I introduce a set of simplifying assumptions. First, mayors are one of two types: a technocratic politician who delivers high quality public goods, and a pragmatic one, who fails to do so. Note that this is an informational heuristic. Voters have beliefs about how competent a politician is and strictly prefer them over the alternative. Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (@coate_form_1995). Additionally, I assume that the voter is sincere, excluding pivotality concerns and strategic voting from voters' utility calculations. Finally, I normalize the politician's exit option to 0: the politician always prefers to be re-elected.

The game has two periods. Voter $v$ does not observe the amount of patronage handed out by mayor $m$ nor her type, but observes and derives utility from the realization of public good quality $\omega_t$. An important feature in the model is that high quality public services $\omega_{h}$ are produced probabilistically as a function of $\theta$, the quality of the bureaucracy. The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See @kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.

The mayor's optimal strategy is pinned down by the share of patronage which maximizes expected utility from holding onto office, conditional on her type and expected benefits from reelection. Mayor $m$ has to offer public sector jobs to coopt the opposition while ensuring that the voter is satisfied enough to reelect her. The executive dilemma is how to reconcile the incentive for reelection and the pressure imposed by the opposition.

The timing of the game is as follows.

1) Nature draws the politician's type.
2) Politician observes her type and allocates public sector jobs to legislators, setting $\theta_1$.
4) Nature realizes public good $\omega_1(\theta_1)$.
5) Voter observes $\omega_1$ and casts vote to retain or fire the incumbent.
6) If incumbent is retained, she sets $\theta_2$ and nature draws $\omega_2(\theta_2)$. If incumbent is deposed, a challenger takes office, sets $\theta_2$ and the game ends.

## The executive dilemma

The mayor's choice set is the bureaucratic quality parameter $\theta_t$. As quality increases - the mayor allocates a smaller proportion of jobs to legislators - the less votes become secure in the legislature. This reduces the likelihood of securing the necessary resources to provide public goods for a given fiscal year. Adapting @gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic. Technocratic politicians invariably deliver high quality public goods to the voter. Pragmatic politicians, on the other hand, are susceptible to legislative pressure and at times fail to deliver them.

The probability of a politician being technocratic is common knowledge: $\xi \in [0,1]$. Conversely, the probability of her being pragmatic is simply $1 - \xi$. Both types derive utility from public goods provision $\omega$, where for simplicity $u(\omega_l)=0$ and $u(\omega_h) = 1$. We do not model the decision of the competent mayor: she simply delivers high quality goods and is inevitably reelected. We focus on the more interesting case of the pragmatic mayor.

\begin{align*}
Eu_m \left(\theta_t, s_1 \middle|\lambda \right) &= \textbf{E}(\omega_1) + \delta \text{Pr}(\text{reelection})\left[\psi + \textbf{E}(\omega_2)  \right] \\
&= \frac{1-s_1\theta_1}{1+e^{-(10\theta_1-5)}} + \delta \text{Pr}(\text{reelection})\left[\psi + \frac{1-s_2\theta_2}{1+e^{-(10\theta_2-5)}} \right] \\
\end{align*}

The equation above formalizes the mayor's dilemma. The probability of reelection decreases with the share of jobs allocated to the opposition, since it lowers the quality of the bureaucracy. However, if the mayor fails to appease them through patronage, it is increasingly costly to "govern". While technocrats are immune to these political pressures, pragmatists ultimately cave in. The need to coopt legislators decreases the quality of the bureaucracy and public goods, as mayors are forced to allocate positions within the bureaucracy to garner support in the legislative. Details on deriving the equilibrium strategy are in the appendix. I now describe Brazil as a case for testing the theory.

# Institutional context and data

## Education in Brazil

In Brazil, responsibility for primary education is delegated to local governments (@paschoal_historia_2009). The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments. Figure \ref{fig:enrollment} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

```{r enrollment, fig.cap = "A decentralized system. Lower school education is primarily under the responsibility of municipalities, with important consequences for early cognitive development."}
include_graphics(
  "figs/enrollment_dep.pdf"
)
```

The local executive branch has sole prerogative over the hiring and firing of teachers and school principals. School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (@brollo_political_2017). Teachers usually enter the public service through a public exam and are eligible for tenure after two years (@gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount. Municipal teachers and school principals are overseen by a local department of education (@militao_educacao_2014).

Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education. At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (@ferraz_corrupting_2012).

## Local governance and mayoral coalitions

Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council. Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016. These take place ] The executive is responsible for the implementation of social policies such as education, as well as drafting the municipal budget. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by @souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.

In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (@dantas_eleicoes_2013). These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber. Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}). While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.^[Interview with C, former chief of staff of municipality J. September 2018.]

Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. In the words of a set of school administrators in the municipality of I:

> School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.

> Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]

These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff. To verify these claims in a broader set of municipalities, as well as outlining the research design, I employ extensive administrative and electoral data collected by the Brazilian federal government. In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.

## Data

Brazil collects fine-grained data on education and makes it publicly available for research. For this study, data on education are collected from two main sources: the SAEB and School Census.^[These can be accessed (here)[http://portal.inep.gov.br/web/guest/dados].] The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

```{r map, fig.cap="Uneven quality. Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores. Averages are weighted by participation rate."}
include_graphics(
  "figs/saeb_map.pdf"
) 
```

Figure \ref{fig:map} presents a map of Brazil with the municipal averages for the SAEB of 2015. Warmer colors such as red and orange indicate higher average scores, with the converse denoted by colder, blue shades. There is clear variation in average test scores, with the Southeast and Midwest outperforming poorer regions in the North and Northeast. Even within regions, however, there is wide heterogeneity. In particular, note that in the northern part of Brazil, high-performing municipalities neighbor low-performing ones. This indicates that despite spatial clustering, municipal factors can lead to variation in the quality of education.

Employment data on educational staff, including teachers and school principals, are extracted from the RAIS, SAEB and the School Census.^[RAIS data, along with additional Brazilian employment data, can be accessed (here)[http://trabalho.gov.br/dados-abertos].] The \emph{Relatório Anual de Informações Sociais} (RAIS) is an annual census administered by the now Ministry of Finance to all formal employers in Brazil. Employers are mandated to accurately report data on employees, subject to fines for misreporting. Subnational governments, including municipalities, also report to the RAIS. The dataset therefore contains micro-level information on all municipal personnel, including when they were hired/fired, as well as wages, type of contract and education levels.

Figure \ref{fig:staff} provides descriptive statistics on bureaucratic personnel in Brazil, segmented by department. Educational staff, in this case school principals and teachers, comprise approximately a third of municipal public sector jobs, totaling around 2 million in 2015. While comparable to administrative staff, this total significantly exceeds that for healthcare services. Focusing on turnover, we note that new hiring and dismissals in education staff is relatively high. While lower than that for administrative staff, new hires can represent over 10 percent of extant staff. The past two decades was a moment of rapid expansion of municipal staff, and hiring has far exceeded dismissals in that time period.

```{r staff, fig.cap = 'Municipal bureaucracies. Total number of bureaucrats and hiring/dismissal rates for bureaucratic staff over time. Disaggregated by department.'}
include_graphics(
  'figs/turnover_category.pdf'
)
```

I propose three different specifications for measuring bureaucratic turnover. From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by @pereira_junior_indicadores_2016. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

$$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$

Information on mayor, city councilors and mayoral coalitions are made available by the Supreme Electoral Court (TSE), the national authority responsible for overseeing elections.^[Available (here)[http://www.tse.jus.br/eleicoes/estatisticas/repositorio-de-dados-eleitorais-1].] In order to calculate the share of legislative seats held by the mayoral coalition, I match the partisanship of each city councilor to the mayoral electoral coalition. The distribution of share of coalition seats is depicted in figure \ref{fig:coalition}. Note that although most city councils are controlled by the mayor, in over 40 percent of municipalities the mayoral coalition is a minority government.

```{r coalition, fig.cap = 'Empirical distribution of share of legislative seats in the mayoral coalition. The unit of analysis is the city council.'}
include_graphics(
  "figs/hist_coalition.pdf"
)
```

Additional data has been collected to supplement the estimation exercise. Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.^[Respectively, data can be found here: (1)[https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html], (2)[https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf], (3)[https://www.seduc.ce.gov.br/spaece/].]

# Estimation

## Empirical strategy

This study presents two sets of estimations. The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by SAEB and SPAECE to estimate the effect of turnover on student test scores. To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.

For the first estimation, I employ a hierarchical linear model to estimate the effect of teacher turnover on average test scores for grade $i$ at school $s$, in municipality $j$ and year $t$. This estimation strategy has been used in multiple studies to analyze educational, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation (@diprete_multilevel_1994, @lee_using_2000.) Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise. The main specification is as follows:
\begin{align*}
\text{test score}_{isjt} = \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
\psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
\end{align*}
We are interested in $\beta_1$ and $\beta_3$, the change in test scores associated with teacher or school principal turnover. In this set-up, $\beta_3$ is the additional effect of staff turnover on test scores if students are in 9th grade. The first level is the grade, with associated characteristics $X_isjt$ at the grade level: among others, percentage of students who have not graduated last year and share of children with a fridge in their house. The second is the school $s$, with characteristics $V_{sjt}$, such as access to internet or the presence of a cafeteria for students. The third level is the municipality $j$, including municipal sociodemographic characteristics ($W_{jt}$) such as population size and median wages, as well as per pupil budgetary expenditures on eduation. Finally, I include state ($\alpha_kj$) and year ($\delta_t$) fixed effects to account for unobserved time-invariant state characteristics and annual seasonality.

The second set of estimations tests the hypothesis that mayors who control less seats in the city council engage in more patronage. To do so I leverage employment data from RAIS and school census data to measure teacher turnover rates. I deploy three sets of models. For the second and third estimation I turn to the micro, bureaucrat-level and a then to aggregated turnover at the municipal level. The first model is a logistic regression where the outcome of interest is whether a particular teacher or school principal is hired/fired for any given year. The second model is a linear model with fixed effects, where the outcome of interest is the proportion of educational staff hired/fired for a muncipality at any given year. For the second set of estimations, the main specification is:
$$\text{turnover}_{jt} = \gamma \text{coalition seats}_{jst} + \mu P_{jt} + \zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{jt}$$
The share of coalition seats held by the mayoral coalition is the treatment in this set-up. We are interested in $\gamma$, the change in bureaucratic turnover associated with variation in the share of legislative seats held by the mayoral coalition. Municipal characteristics $W_{jt}$ are similar to the ones used for the first set of estimation. I add political variables to the estimation in order to control for differences in mayor partisanship, incumbency status, and individual characteristics of the mayor: professional background, education, and age.

Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in @imai_covariate_2014 which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

## Results

Teacher and school principal turnover have significant, negative effects on average test scores, highlighting the potential dangers of bureaucratic turnover on student learning. These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations. Table \ref{tbl:hlm_mods} presents the results of the estimation on SAEB test scores. We present two specifications for turnover. Turnover index measures the amount of turnover in teachers at the school level. Work experience for teachers and school principals serve as an alternative measure of turnover. Personnel who has only recently entered the school recently amass a shorter amount of work experience (less than two years).

\begin{landscape}
\begin{table}[t]
  \centering
  \footnotesize
  \input{tables/hlm_result}
  \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning. Models 1 and 2 present results for teacher turnover index constructed at the school level. Models 3 and 4 estimate the effect of new teachers and school principals entering the school (less than two years). All models include year and state fixed effects.}
  \label{tbl:hlm_mods}
\end{table}
\end{landscape}

Additionally, I present results for the SPAECE standardized exam in table \ref{tbl:spaece}. Differently than the SAEB, the SPAECE takes place annually and the average test scores are calculated at the grade, rather than the classroom level. The annual frequency of this exam allows us to retrieve more precise estiamtes with respect to the effect of teacher turnover on student test scores, as opposed to the biannual frequency of the SAEB. Note that for both estimations, we standardize the outcome of interest. Coefficients are therefore interpretable as a standard deviation change in student test with respect to the covariate.

\begin{table}[t]
  \centering
  \footnotesize
  \input{tables/spaece_result}
  \caption{{\bf Turnover and student learning in Ceará.} Teacher turnover have a negative effect on student learning. Models 1 and 2 present results for teacher turnover index constructed at the school level.}
  \label{tbl:spaece}
\end{table}

I now turn to testing the main proposition of this paper: that turnover in educational staff stems indeed from coalitional dynamics. There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage. Table \ref{tbl:turnover} presents the results of the estimation. The first model regresses turnover index on coalition share, where the unit of analysis is a grade-level per school. Models 2 and 3 estimate the effect of the share of legislative seats on hiring or dismissal of individual bureaucrats. The third set of models (4 and 5) estimate the change in the share of new hires and dismissals at the municipal level. All models suggest that an increase in the executive share of legislative seats decrease turnover.

\begin{landscape}
\begin{table}[t]
  \centering
  \footnotesize
  \input{tables/turnover_result}
  \caption{{\bf Executive coalitions and staff turnover.} An incrase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals. Model 1 presents results on the turnover index. Models 2 and 3 are logistic regressions at the individual, bureaucrat level. Models 4 and 5 are aggregated at the municipal level. where the outcome of interest is the proportion of staff either hired or dismissed at a given year. Models 1, 3, and 5 include year and state fixed effects.}
  \label{tbl:turnover}
\end{table}
\end{landscape}

For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council. Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature. Overall, these results suggest that weaker executive control over the legislature increases patronage, with potentially negative effects for public service delivery in municipalities in Brazil.

\begin{figure}[H]
    \centering
    \includegraphics{figs/plot_pred}
    \caption{Predicted values for bureaucratic turnover. The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive. The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
    \label{fig:turnover}
\end{figure}

As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity. It is in these circumstances that validating causal mechanisms through fieldwork and careful data analysis can increase the validity of a causal claim. In-depth interviews with local actors involved in managing education, as well as the use of different measurements for both educational staff turnover and student learning, provides strong evidence that staff turnover has a negative impact on student learning. Bureaucratic turnover stems from political considerations.

## Mechanisms

The question remains: why do city councilors engage in patronage. Previous studies have found that citizens directly access

# Conclusion

Improving the quality of education received by children across the world remains a challenge. This paper proposes an analytical framework and estimation strategy to understand the decision-making process behind the administration of educational services in Brazil. In a decentralized context, subnational political actors have a direct say on how educational services are managed, with profound implications for the quality of public services delivered to citizens. These actors interact with bureaucracies and other local elites in complex ways that are only beginning to be mapped.

In this study I theorize and demonstrate that staff turnover stems from the executive's need to garner support from legislators in the city council. This process of coalition building is consolidated through employment offers to city councilors and their constituencies. As the share of seats held by the executive coalition decreases, the costlier it becomes to coopt legislators to support the executive. As a result, the amount of patronage we observe should increase. That is precisely what the data indicates, with teacher turnover increasing in schools, as well as increased hiring and dismissals of teachers and school principals.

This bureaucratic turnover has important consequences for the quality of education delivered in municipalities. I find that turnover has a negative effect on student learning, across a set of specifications for turnover and different evaluation metrics for student learning. The evidence therefore seems to point out that mayors with a weaker hold on the city council resort to greater patronage, with negative consequences for student learning. This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts (@gottlieb_countervailing_2019).

Future research on patronage and public service delivery would benefit from a clearer treatment of the institutional context in which local political actors operate. Important insights have been derived on executive leaders, but these actors seldom govern alone. Incorporating other local elites paints a more complex and accurate understanding of the strategic considerations taking place in the political management of education, and public services more broadly.

\pagebreak

# Appendix

## Model

### Solving the dilemma (compromise):

Mayors solve the the game through backward induction. If the mayor is a technocrat, she simply and effectively delivers, and the voter consumes $\omega_h$. The average politician myopically maximizes expected utility in the second period.

\begin{align*}
\frac{\partial Eu_{m}(\theta_1, \theta_2|\lambda = 1, s_2)}{\partial \theta_2} &= 0 \\
\frac{\partial}{\partial \theta_2} \frac{1-s_2\theta_2}{1+e^{-(10\theta_2-5)}} &= 0 \\
\frac{1}{e^{-(10\theta_2^*-5)}}+10\theta_2^* &= \frac{10}{s_2} - 1\\
\theta_2^* &= \frac{1}{10s_1}\left[10-s_2-\boldsymbol{W}\left(e^{\frac{1}{s_2}-6}\right)\right]
\end{align*}

The interpretation of this result is straightforward: increases in the legislative share of the opposition reduces incentives to invest in bureaucratic quality.^[Note that for $s=0$, this expression is undefined. Also, $\boldsymbol{W}$ denotes the Lambert W function.] We can visualize how the optimal bureaucratic quality in the second period changes as a function of $s_2$ by plotting changes in the probability of high quality public goods.

```{r visualization}
fun_1 <- function(x){ (1-x)/(1+exp(-(10*x - 5)))}
fun_2 <- function(x){ (1-0.5*x)/(1+exp(-(10*x - 5)))}
fun_3 <- function(x){ (1)/(1+exp(-(10*x - 5)))}

ggplot(data.frame(x = c(0, 1)), aes(x)) +
  stat_function(
    fun = fun_1
  ) +
  stat_function(
    fun = fun_2
  ) +
  stat_function(
    fun = fun_3
  ) +
  annotate(
    geom = "text",
    x = 0.6,
    y = 0.34,
    label = "s = 1"
  ) +
  annotate(
    geom = "text",
    x = 0.75,
    y = 0.62,
    label = "s = 0.5"
  ) +
  annotate(
    geom = "text",
    x = 0.70,
    y = 0.95,
    label = "s = 0"
  ) +
  xlab("Bureaucratic Quality") +
  ylab("Probability") +
  theme_clean +
  ggtitle(
    "Impact of opposition legislative share on public goods provision"
  )
```

The quality of the bureaucracy $\theta_2$ which maximizes the mayor's expected utility in the second period is strictly decreasing in $s_2$, the share of seats controlled by the opposition. As the power of the opposition increases, its ability to veto and punish the mayor for withholding from patronage increases. The greater the threat of veto power, the more concessions politicians need to make in order to deliver high quality public goods $\omega_h$. Conversely, when the mayor controls the legislative chamber, the benefits of bureaucratic quality are fully internalized and it is in turn maximized. We now turn to the electoral problem faced by the voters.

<!-- ## Opposition's problem: -->

<!-- The opposition decides whether or not to support the mayor's policies, where $a_o=\left\{ 0,1 \right\}$. If the opposition supports the executive, they receive an $\alpha \in [0,1]$ proportion of public sector jobs, a continuum normalized to 1. The concept of a public sector pie stems from interviews with former and current executive secretaries, who describe a "quota" of jobs allocated to the legislative.^[This echoes the concept of islands of excellence described by @geddes_politicians_1994: small pockets of bureaucratic insulation reserved by the executive for core functions such as macroeconomic management. A similar argument is made in @loureiro_burocracia_2010.] If she refuses, the mayor allocates no pork. I apply the Rubinstein bargaining model solution concept (perfect equilibrium) to solve this subgame with minor adaptations.^[@rubinstein_perfect_1982.] -->

<!-- Mayor's and opposition's utilities are identical to the original setup with two players. I conceptualize the bargain over public sectors as a sequence of offers and counter-offers made by the mayor and her legislative opposition. Following @rubinstein_perfect_1982, players differ in $\delta_i$, a parameter capturing their impatience across sequences in the bargaining game. Respectively: -->

<!-- $$u_m = \prod^{T}_{t=1}\delta_m^t\theta_{1|2}$$  -->
<!-- $$u_o = \prod^{T}_{t=1}\delta_i^t\theta_{1|2}$$ -->

<!-- Applying stationarity restrictions and solving for the unique subgame perfect equilibrium, we have: -->

### State capacity and democratic accountability:

Let $v$ denote the median voter in the municipal electorate. Voters derive utility from the quality of public goods at time $t$, with a discount factor $\delta$.

$$u_v(\omega_t) = \omega_1 + \delta \omega_2$$

Because voters cannot observe directly the type of the politician nor her hiring decisions, voters update beliefs about the politician's type after observing first period public goods output. If a voter observes a low public goods quality, he knows with certainty that the incumbent is incompetent, and therefore $\Pr(\text{competent}|\omega_{1l}) = 0$. Let $\mu(\omega_1)$ denote voter's belief that the politician is a competent after observing the first period realization of public goods.

$$\mu(\omega_h) = \text{Pr}(\text{competent}| \omega_h) = \frac{\xi}{\xi + \text{Pr}(\omega_1|s)(1-\xi)}>\xi$$

Note that for any $\omega_h$, the probability that the incumbent is competent is higher than the probability of a random challenger being so. As a result, whenever the voter observes a high quality output in the first period, the incumbent gets reelected. This simplifies our analysis. After observing the realization of public good quality $\omega_1$, he retains the incumbent if and only if $\omega_1 = \omega_h$, and fires her otherwise. The probability of re-election is simply the probability that public goods provided are of high quality. This is conditional on $\theta_1$, the amount of investment in state capacity in the first period and $s_1$, the opposition share in the first period.^[Note that the quality of the second period bureaucracy is independent of that in the first period. This is a strong restriction that I will relax in future iterations of this project.]

### Initial investments in state capacity:

We turn to the initial decision of the politician. The executive maximizes her expected utility with respect to first term hiring decisions, factoring in the optimal decision and payoff for the second period. To reiterate, the probability of reelection is simply the probability of high quality public goods provision in the first period. Taking the first order condition with respect to $\theta_1$:

\begin{align*}
\argmax_{\theta_1} \frac{1-s_1\theta_1}{1+e^{-(10\theta_1-5)}}\left[1+\delta\left(\psi + \frac{1-s_2\theta_2^*}{1+e^{-(10\theta_2^*-5)}}\right)\right] &= 0 \\
\end{align*}

The optimal amount of investment in state capacity in the first term is analogous to the myopic optimization in the second period, weighed by the expected payoff of reelection. The intuition behind this result is straightforward. As politicians value the future term more, they will invest more in state capacity. This is captured by the parameter $\delta$ and $\psi$, the former being the discount factor and the latter, office benefits. However, in a clear illustration of the trade-off implied between investing in state capacity and resorting to patronage to ensure legislative support, the dominant effect is imposed by $s_1$, the share of legislative seats held by the opposition in the first period.

\pagebreak

## Covariate balance

### Education quality vs. staff turnover

```{r}
include_graphics(
  "figs/balance_hlm.pdf"
)
```

### Staff turnover vs. executive share of legislative seats

```{r}
include_graphics(
  "figs/balance_felm.pdf"
)
```

\pagebreak

### 2020-06-04 build.R

@@ -1,0 +1,3 @@
+# data wrangle
+# replication
+# publication

### 2020-06-04 replication/replication.R

@@ -1,0 +1,1988 @@
+# set-up ------------------------------------------------------------------
+library(here)
+
+source(
+  here("scripts", "thesis_setup.R")
+)
+
+source(
+  here("scripts", "thesis_funs.R")
+)
+
+# source(
+#   here("scripts/thesis_map.R")
+# )
+
+p_file_here <- partial(
+  file_here,
+  paper = 'job_market'
+)
+
+ggsave <- partial(
+  ggplot2::ggsave,
+  width = 5,
+  height = 3
+)
+
+controls <- c(
+  "rais_permanent", "rais_size", "I(budget_education/censo_pop)", 
+  "censo_median_wage", "censo_log_pop", "mayor_edu", 
+  "mayor_coalition_size", "as.factor(mayor_party)"
+)
+
+rais_main_categories <- c("education", "administration", "services", "healthcare")
+
+# data --------------------------------------------------------------------
+edu_global <- fread(
+  here("data/worldbank/global_edu_indicators.csv")
+) %>% 
+  filter(
+    between(year, 2000, 2016)
+  )
+
+edu_global %<>%
+  mutate(
+    pisa = rowMeans(select(., starts_with("pisa_")), na.rm = T)
+  ) %>% 
+  select(
+    -starts_with("pisa_")
+  )
+
+pisa_global <- edu_global %>%
+  group_by(year) %>% 
+  summarise_at(
+    vars(pisa),
+    mean, na.rm = T
+  ) %>% 
+  mutate(
+    type = "global"
+  )
+
+edu_global %<>% 
+  mutate(
+    type = case_when(
+      # country_code == "lcn" ~ "la",
+      country_code == "bra" ~ "br",
+      country_name == "world" ~ "global",
+      T ~ NA_character_
+    )
+  ) %>% 
+  filter(
+    !is.na(type)
+  )
+
+edu_global %<>% 
+  group_by(
+    type,
+    year
+  ) %>% 
+  select_if(is.numeric) %>% 
+  summarise_all(
+    mean,
+    na.rm = T
+  ) %>% 
+  ungroup()
+
+edu_global %<>% 
+  left_join(
+    pisa_global,
+    by = c("type", "year")
+  ) %>% 
+  mutate(
+    pisa = coalesce(pisa.x, pisa.y)
+  ) %>% 
+  select(-starts_with("pisa."))
+
+# saeb
[…]

### 2020-06-04 replication/scripts/thesis_eda.R

@@ -1,0 +1,64 @@
+# set-up ------------------------------------------------------------------
+lapply(
+  c("tidyverse", "data.table", "here"),
+  require,
+  character.only = T
+)
+
+set.seed(1789)
+
+source(here("scripts/thesis_setup.R"))
+# eda ---------------------------------------------------------------------
+mods <- list()
+
+mods$patronage <- lm(
+  log(budget_total/censo_pop + 0.1) ~ rais_hired*coalition_share + as.factor(year) + as.factor(state) +
+    censo_median_wage + chamber_size,
+  data = cabinet_mun
+)
+
+mods$coalition <- lm(
+  log(rais_hired + 0.1) ~ coalition_share + share_vote_mayor + censo_rural + censo_log_pop +
+    as.factor(year) + as.factor(state) + as.factor(mayor_party) + as.factor(mayor_edu_desc) +
+    effective_parties + mayor_coalition_size + party_turnover + 
+    mayor_reelected,
+  data = cabinet_mun %>% 
+    mutate(
+      share_vote_mayor = total_vote_mayor/total_vote_legislative
+    )
+)
+
+map(
+  mods,
+  summary
+)
+
+# ideology ----------------------------------------------------------------
+ideology <- fread(
+  "~/Dropbox/Value_new_Candidates/data/ideology/candidate_ideology_local.csv"
+) %>% 
+  select(
+    cpf_candidate,
+    cfscore
+  )
+
+# join election with ideology
+election <- election %>% 
+  left_join(
+    ideology,
+    by = "cpf_candidate"
+  )
+
+# join
+rais_mun <- rais_mun %>% 
+  left_join(
+    election,
+    by = c("cod_ibge_6", "election_year")
+  )
+
+lm(
+  rais_hired ~ as.factor(rais_category) + coalition_share + cfscore +
+    as.factor(year) + as.factor(str_sub(cod_ibge_6, 1, 2)) ,
+  data = rais_mun
+) %>% 
+  summary

### 2020-06-04 replication/scripts/thesis_funs.R

@@ -1,0 +1,667 @@
+# functions ---------------------------------------------------------------
+`%<>%` <- magrittr::`%<>%`
+ivreg <- AER::ivreg
+between <- data.table::between
+
+# data io -----------------------------------------------------------------
+fread <- partial(
+  data.table::fread,
+  nThread = parallel::detectCores(),
+  integer64 = c("character")
+)
+
+list.files <- partial(
+  base::list.files,
+  full.names = T
+)
+
+list_files <- function(path, pattern){
+  files <- map2(
+    path,
+    pattern,
+    list.files
+  ) %>% 
+    flatten_chr
+  
+  return(files)
+}
+
+import_data <- function(path, pattern, names = F){
+  files <- list_files(
+    path,
+    pattern
+  )
+  
+  if(names == F){
+    names <- basename(files) %>% 
+      str_remove_all(
+        "\\.csv|\\.gz"
+      )
+  }
+  
+  data <- map(
+    files,
+    fread
+  )
+  
+  names(data) <- names
+  
+  return(data)
+}
+
+assign_data <- function(data, rm = T){
+  walk2(
+    names(data),
+    data,
+    assign,
+    envir = globalenv()
+  )
+}
+
+append_db <- function(path, pattern, conn, names = F){
+  data <- import(
+    files,
+    names
+  )
+  
+  pwalk(
+    list(
+      name = names,
+      value = data
+    ),
+    RSQLite::dbWriteTable,
+    conn = con,
+    overwrite = T
+  )
+}
+
+fwrite_gz <- function(data, filename, overwrite = T) {
+  data.table::fwrite(
+    data,  
+    filename
+  )
+  
+  R.utils::gzip(
+    filename,
+    overwrite = overwrite
+  )
+}
+
+file_here <- function(paper = NULL, folder = NULL, file = NULL){
+  if(is.null(file)){
+    stop('no file name.')
+  }
+  
+  path <- here(paste0('papers/paper_', paper), folder, file)
+  return(path)
+}
+
+object_size <- function(object){
+  size <- object.size(object)
+  
+  print(size, units = "MB")
+}
+
+# data manipulation -------------------------------------------------------
+# sample groups
+sample_group <- function(data, n, ...){
+  grouping_vars <- enquos(...)
+  
+  data %<>%
+    nest(
+      cols = -c(!!!grouping_vars)
+    ) %>% 
[…]

### 2020-06-04 replication/scripts/thesis_import_data.R

@@ -1,0 +1,236 @@
+# saeb
+import_data(
+  here("data/saeb"),
+  "exam_mun|hierarchical|exam_school"
+) %>%
+  assign_data
+
+saeb_dep <- fread(
+  here("data/saeb/saeb_exam_dep.csv")
+)
+
+# rais
+rais_edu <- fread(
+  here("data/rais/rais_edu.csv.gz")
+) %>% 
+  filter(
+    year >= 2005
+  )
+
+rais_edu_mun <- fread(
+  here('data/rais/rais_edu_mun.csv')
+) %>% 
+  join_covariate() %>% 
+  mutate(
+    state = str_sub(cod_ibge_6, 1, 2)
+  ) %>% 
+  mutate_if(
+    is.double,
+    rescale
+  )
+
+rais_mun <- fread(
+  here("data/rais/rais_mun.csv.gz")
+)
+
+rais_fit <- rais_edu %>% 
+  join_covariate() %>%
+  mutate(
+    state = str_sub(cod_ibge_6, 1, 2),
+    coalition_share = rescale(coalition_share),
+    rais_category = fct_relevel(rais_category, 'teacher')
+  )
+
+# censo escolar
+censo_class <- fread(
+  here("data/censo_escolar/censo_class.csv.gz")
+) %>% 
+  group_by(
+    year,
+    dep,
+    grade_level
+  ) %>%
+  summarise(
+    num_enroll = sum(num_enroll)
+  ) %>% 
+  mutate(
+    freq = num_enroll/sum(num_enroll)
+  )
+
+censo_school <- fread(
+  here('data/censo_escolar/censo_school.csv.gz')
+)
+
+censo_school_turnover <- fread(
+  here("data/censo_escolar/censo_school_turnover.csv.gz")
+)
+
+censo_turnover_ce <- censo_school_turnover %>% 
+  filter(
+    state == "23",
+    grade_level %in% c(2, 5, 9)
+  )
+
+censo_mun_turnover <- fread(
+  here("data/censo_escolar/censo_mun_turnover.csv.gz")
+) %>% 
+  group_by(
+    cod_ibge_6, year
+  ) %>% 
+  summarise_stats(
+    turnover_index
+  ) %>% 
+  join_covariate() %>% 
+  mutate(
+    state = str_sub(cod_ibge_6, 1, 2)
+  ) %>% 
+  mutate_if(
+    is.double,
+    rescale
+  )
+
+spaece <- fread(
+  here("data/spaece/spaece.csv")
+)
+
[…]

### 2020-06-04 replication/scripts/thesis_map.R

@@ -1,0 +1,40 @@
+pacman::p_load(
+  "rgdal",
+  "rgeos",
+  "RColorBrewer",
+  "viridis",
+  "scales",
+  "rmapshaper",
+  "naniar"
+)
+
+map_br <- readOGR(
+  here("data/maps/"),
+  "municipio"
+)
+
+map_br <- SpatialPolygonsDataFrame(
+  gSimplify(map_br, tol = 0.01, topologyPreserve = T),
+  data = map_br@data
+)
+
+map_br@data <- map_br@data %>%
+  mutate(id = row.names(map_br))
+
+map_br <- map_br %>%
+  fortify %>%
+  left_join(., map_br@data, by = "id") %>%
+  transmute(
+    long,
+    lat,
+    group,
+    cod_ibge_6 = as.numeric(levels(cod_ibge_6))[cod_ibge_6],
+    mun_name = sem_acento
+  )
+
+# state boundaries
+# map_state <- readOGR(
+#   here("data/maps/"),
+#   "estados_2010"
+# ) %>%
+#   fortify

### 2020-06-04 replication/scripts/thesis_models.R

@@ -1,0 +1,210 @@
+# set-up ------------------------------------------------------------------
+source(
+  here::here("scripts/thesis_setup.R")
+)
+
+source(
+  here("scripts/thesis_funs.R")
+)
+
+ggsave <- partial(
+  ggplot2::ggsave,
+  width = 4,
+  height = 2.5
+)
+
+# data --------------------------------------------------------------------
+saeb_hierarchical <- fread(
+  here("data/saeb/saeb_hierarchical.csv.gz")
+) 
+
+censo_turnover <- fread(
+  here('data/censo_escolar/censo_school_turnover.csv.gz')
+) %>% 
+  filter(
+    grade_level %in% c(5, 9)
+  )
+
+censo_turnover %<>% 
+  join_covariate
+
+# remove small schools
+saeb_hierarchical %<>% 
+  filter(n_teacher >= 5)
+
+saeb_hierarchical %<>%
+  mutate_if(
+    is.double,
+    ~scale(.) %>% 
+      as.vector
+  )
+
+# prep data ---------------------------------------------------------------
+formula_model <- as.formula(
+  turnover_index ~ grade_exam + grade + saeb_principal_female + saeb_principal_education + 
+    saeb_principal_experience +  saeb_principal_appointment + education_teacher + type_contract_teacher + 
+    saeb_wage_teacher + year_as_teacher + parent_attend_college_student + 
+    parents_school_meeting_student + toilet + meal +
+    censo_median_wage + censo_log_pop + censo_rural + censo_lit_rate + year + state
+)
+
+model_hierarchical <- model.frame(
+  formula_model,
+  saeb_hierarchical
+) %>% 
+  as_tibble
+
+
+# visualization -----------------------------------------------------------
+model_hierarchical %>% 
+  sample_n(2e3) %>% 
+  ggplot(
+    aes(
+      turnover_index,
+      grade_exam,
+      color = grade,
+      group = grade
+    )
+  ) +
+  geom_point(
+    alpha = 0.1
+  ) +
+  geom_smooth(
+    method = "lm",
+    color = "orange"
+  ) 
+
+model_hierarchical %>% 
+  gg_histogram(
+    aes(turnover_index)
+  )
+
+# covariate balance -------------------------------------------------------
+formula_weight <- as.formula(
+  turnover_index ~ grade + saeb_principal_female + saeb_principal_education + 
[…]

### 2020-06-04 replication/scripts/thesis_setup.R

@@ -1,0 +1,65 @@
+# set up ------------------------------------------------------------------
+pacman::p_load(
+  tidyverse,
+  data.table,
+  rgdal,
+  here,
+  R.utils,
+  GGally,
+  lfe,
+  lme4,
+  plm,
+  broom,
+  extrafont,
+  rprojroot,
+  tmap,
+  gridExtra,
+  cowplot,
+  ggdag,
+  scales,
+  kableExtra,
+  RColorBrewer,
+  gghighlight,
+  parallel,
+  foreach,
+  gridExtra,
+  naniar,
+  knitr,
+  gghighlight,
+  egg,
+  WeightIt,
+  cobalt,
+  boot,
+  rdrobust,
+  magrittr,
+  sjPlot
+)
+
+knitr::opts_chunk$set(
+  eval = T,
+  cache = T,
+  cache.lazy = F,
+  message = F,
+  warning = F,
+  echo = F,
+  fig.height = 3,
+  fig.width = 4,
+  fig.align = "center"
+  # dev = "pdf"
+)
+
+source(
+  here("scripts/thesis_funs.R")
+)
+
+# connect to database
+con <- DBI::dbConnect(
+  odbc::odbc(),
+  driver = "PostgreSQL Unicode",
+  database = "rais",
+  UID = "gali",
+  PWD = 'gali1789!',
+  port = 5432
+)
+
+set.seed(1789)

### 2020-06-11 build/data/build.R
finish hierarchical estimation saeb

@@ -1,4 +1,6 @@
-# set-up ------------------------------------------------------------------
+# ==============================================================================
+# set-up
+# ==============================================================================
 pacman::p_load(
   "tidyverse",
   "purrr",
@@ -15,7 +17,9 @@ source(
   here("build", "data", "functions.R")
 )
 
+# ==============================================================================
 # build data repos
+# ==============================================================================
 repos <- c(
   "censo_br",
   "censo_escolar",

### 2020-06-11 build/data/modules/datagen_censo_escolar.R

@@ -4,6 +4,8 @@
 # 3) municipal classrooms.
 # 4) all classrooms.
 # 5) teacher turnover (municipal and all)
+init_env <- ls()
+
 censo_school <- read_data(
   "raw",
   "censo_escolar",
@@ -95,6 +97,27 @@ censo_school_mun <- censo_school_mun %>%
 censo_school_mun %>%
   write_data(
     "censo_escolar", "censo_school_mun.rds"
+  )
+
+censo_school %>%
+  transmute(
+    cod_ibge_6,
+    year,
+    dep,
+    school_id,
+    school_name,
+    active,
+    association_parents_tchrs,
+    access_electricity = 1 - electricity_inexistent,
+    access_water = 1 - water_inexistent,
+    library,
+    lab_info,
+    meal,
+    toilet
+  ) %>%
+  write_data(
+    "censo_escolar",
+    "censo_school.rds"
   )
 
 reset_env(init_env)

### 2020-06-11 build/replication/modules/model_hierarchical.R

@@ -1,7 +1,7 @@
 # hierarchical model ------------------------------------------------------
 f <- stats::as.formula
 
-saeb_hierarchical <- read_data(
+saeb <- read_data(
   "saeb",
   "saeb_hierarchical.rds"
 )
@@ -11,99 +11,55 @@ finbra <- read_data(
   "finbra.rds"
 )
 
-# censo_school <- read_data(
-#   "censo_escolar",
-#   "censo_school_mun.rds"
-# )
+censo_school <- read_data(
+  "censo_escolar",
+  "censo_school.rds"
+) %>%
+  mutate(
+    school_id = as.integer(school_id)
+  )
 
 censo_school_turnover <- read_data(
   "censo_escolar",
   "censo_school_turnover.rds"
-)
+) %>%
+  select(
+    cod_ibge_6,
+    year,
+    school_id,
+    grade_level,
+    turnover_index,
+    starts_with("percent"),
+    n_teacher = n
+  ) %>%
+  filter(
+    n_teacher >= 5
+  )
 
 # prepare data for estimation
-saeb_hierarchical <- list(
-  saeb_hierarchical,
+saeb_hierarchical <- lst(
+  saeb,
   finbra,
+  censo_school,
+  censo_school_turnover
 ) %>%
   reduce(
     left_join
   ) %>%
   mutate(
+    saeb_teacher_work_school = fct_relevel(saeb_teacher_work_school, "2 to 10"),
+    censo_log_pop = log(censo_pop),
     budget_education_capita = budget_education / censo_pop
   )
 
-# join school covariates
-# censo_school %<>% 
-#   filter(
-#     dep == "municipal", year >= 2001
-#   ) %>%
-#   select(
-#     state,
-#     cod_ibge_6,
-#     year,
-#     location,
-#     school_id,
-#     school_name,
-#     internet,
-#     kitchen,
-#     library,
-#     lab_info,
-#     meal,
-#     starts_with('principal'),
-#     sewer_grid,
-#     toilet,
-#     water_grid
-#   )
-
-# join_cols <- c("state", "cod_ibge_6", "school_id", "year")
-
-# saeb_hlm %<>%
-#   mutate(state = str_sub(cod_ibge_6, 1, 2)) %>% 
-#   rename(school_id = cod_school) %>% 
-#   mutate_at(join_cols, as.character) %>% 
-#   left_join(
-#     censo_school %>%
-#       mutate_at(join_cols, as.character) %>% 
-#       select(-location),
[…]

### 2020-06-11 build/replication/build.R
beta test of visualization of model results for spaece and saeb

@@ -83,7 +83,9 @@ modules <- c(
   "visual_map"
   "visual_global_edu",
   "visual_budget",
-  "visual_bureaucracy"
+  "visual_bureaucracy",
+  "model_hierarchical",
+  "model_saeb"
   )
 
 walk(

### 2020-06-11 build/replication/functions.R

@@ -38,13 +38,17 @@ read_data <- function(type, dir, file) {
     data <- fread(file_path)
   }
   
-  obj_size <- object.size(data)
-  print(obj_size, units = "MB")
+  print_obj_size(data)
 
   return(data)
 }
 
-ls_data <- function(env = .GlobalEnv){
+print_obj_size <- function(obj) {
+  obj_size <- object.size(obj)
+  print(obj_size, units = "MB")
+}
+
+ls_data <- function(env = .GlobalEnv) {
   obj <- ls(env)
 
   is_data <- map_lgl(
@@ -55,6 +59,25 @@ ls_data <- function(env = .GlobalEnv){
   data_obj <- obj[is_data]
 
   return(data_obj)
+}
+
+read_model <- function(file){
+  fit <- read_rds(
+    here("replication", "models", file)
+  )
+
+  print_obj_size(fit)
+
+  return(fit)
+}
+
+write_model <- function(model, file) {
+  print_obj_size(model)
+  
+  write_rds(
+    model,
+    here("replication", "models", file)
+  )
 }
 
 load_p <- function(packages){
@@ -263,7 +286,7 @@ round_integer <- function(number, digits) {
 
 # tidy and ggcoef
 tidyfit <- function(fit, vars = '.'){
-  tidy(fit) %>%
+  broom::tidy(fit) %>%
     filter(
       !str_detect(term, "Intercept|as.factor|Observation.Residual"),
       str_detect(term, vars)

### 2020-06-11 build/replication/modules/model_hierarchical.R

@@ -138,6 +138,11 @@ fit_lmer <- map(
   )
 )
 
+fit_lmer %>%
+  write_model(
+    "fit_saeb_hierarchical.rds"
+  )
+
 sink(
   here("replication", "results", "model_hierarchical.tex")
 )
@@ -154,203 +159,3 @@ mstar(
   dep.var.labels = "SAEB average test scores"
 )
 sink()
-
-# ==============================================================================
-# spaece: annual standardized test data for the state of ceara
-# ==============================================================================
-spaece <- fread(
-  here("data/spaece/spaece.csv")
-)
-
-censo_turnover_ce <- censo_school_turnover %>% 
-  filter(
-    str_sub(cod_ibge_6, 1, 2) == "23",
-    grade_level %in% c(2, 5, 9)
-  )
-
-censo_turnover_score <- censo_turnover_ce %>% 
-  left_join(
-    censo_school %>% 
-      filter(dep == 'municipal', year >= 2007) %>% 
-      transmute(school_id = as.integer(school_id), year, toilet, meal),
-    by = c('school_id', 'year')
-  ) %>% 
-  left_join(
-    spaece,
-    by = c("cod_ibge_6", "school_id", "year", "grade_level" = "grade")
-  ) %>% 
-  left_join(
-    saeb_exam_school,
-    by = c("cod_ibge_6", "school_id", "year", "grade_level" = "grade")
-  ) %>% 
-  join_covariate() %>% 
-  mutate(
-    mandate_year = if_else(
-      year %in% seq(2005, 2013, 4), 1, 0
-    )
-  )
-
-censo_turnover_score <- censo_turnover_score %>% 
-  mutate(
-    state = str_sub(cod_ibge_6, 1, 2)
-  ) %>% 
-  mutate_at(
-    vars(grade_level, year, state),
-    as.factor
-  ) %>% 
-  mutate_if(
-    is.double,
-    scale
-  )
-
-formulae_spaece <- c(
-  as.formula(
-    spaece_mean ~ turnover_index * grade + as.factor(year)
-  ),
-  as.formula(
-    spaece_mean ~ turnover_index * grade + as.factor(year) + toilet + meal +
-      censo_rural + censo_log_pop + censo_lit_rate
-  )
-)
-
-fit_spaece <- map(
-  formulae_spaece,
-  ~ lm(
-    formula = .,
-    data = censo_turnover_score %>%
-      mutate_if(is.double, rescale) %>%
[…]

### 2020-06-11 build/replication/modules/model_spaece.R

@@ -1,0 +1,106 @@
+# ==============================================================================
+# spaece: annual standardized test data for the state of ceara
+# ==============================================================================
+spaece <- read_data(
+  "spaece",
+  "spaece.rds"
+)
+
+censo_school <- read_data(
+  "censo_escolar",
+  "censo_school.rds"
+) %>%
+  mutate(
+    school_id = as.integer(school_id)
+  )
+
+censo_school_turnover <- read_data(
+  "censo_escolar",
+  "censo_school_turnover.rds"
+) %>%
+  select(
+    cod_ibge_6,
+    year,
+    school_id,
+    grade_level,
+    turnover_index,
+    starts_with("percent"),
+    n_teacher = n
+  ) %>%
+  filter(
+    n_teacher >= 5
+  )
+
+censo_school_ceara <- censo_school_turnover %>% 
+  filter(
+    str_sub(cod_ibge_6, 1, 2) == "23",
+    grade_level %in% c(2, 5, 9)
+  )
+
+spaece_turnover <- censo_school_ceara %>% 
+  left_join(
+    censo_school %>% 
+      filter(dep == 'municipal', year >= 2007) %>% 
+      transmute(school_id = as.integer(school_id), year, toilet, meal),
+    by = c('school_id', 'year')
+  ) %>% 
+  left_join(
+    spaece,
+    by = c("cod_ibge_6", "school_id", "year", "grade_level")
+  ) %>% 
+  join_covariate() %>% 
+  mutate(
+    mandate_year = if_else(
+      year %in% seq(2005, 2013, 4), 1, 0
+    )
+  )
+
+spaece_turnover <- spaece_turnover %>% 
+  mutate(
+    state = str_sub(cod_ibge_6, 1, 2)
+  ) %>% 
+  mutate_at(
+    vars(grade_level, year, state),
+    as.factor
+  ) %>% 
+  mutate_if(
+    is.double,
+    scale
+  )
+
+formulae_spaece <- c(
+  as.formula(
+    spaece_mean ~ turnover_index * grade + as.factor(year)
+  ),
+  as.formula(
+    spaece_mean ~ turnover_index * grade + as.factor(year) + toilet + meal +
+      censo_rural + censo_log_pop + censo_lit_rate
+  )
+)
+
+fit_spaece <- map(
+  formulae_spaece,
+  ~ lm(
+    formula = .,
+    data = spaece_turnover %>%
[…]

### 2020-06-11 build/replication/modules/visual_student_learning.R

@@ -1,0 +1,126 @@
+# ==============================================================================
+# visualize combined results from saeb hierarchical and spaece models
+# ==============================================================================
+fit_lmer <- read_model("fit_saeb_hierarchical.rds")
+fit_spaece <- read_model("fit_spaece.rds")
+
+sink(here("replication", "results", "student_learning.tex"))
+mstar(
+  c(fit_lmer, fit_spaece),
+  keep = c("turnover_index", "saeb_principal_experience", "saeb_teacher_work_school"),
+  add.lines = list(c("Controls", rep(c("\\_", "\\checkmark"), 3))),
+  dep.var.labels.include = F,
+  covariate.labels = c(
+    "Turnover index", "Turnover index $\\times$ Grade 9",
+    "Teacher experience (2 years)", "Teacher experience (10 years)",
+    "School principal experience (2 years)", "School principal experience (10 years)"
+  ),
+  dep.var.caption = "Student learning",
+  column.labels = c("SAEB test score", "SPAECE test score"),
+  column.separate = c(4, 2),
+  model.names = FALSE,
+  keep.stat = "n"
+)
+sink()
+
+# visualization
+mods <- c(
+  fit_lmer[c(2, 4)],
+  fit_spaece[2]
+)
+
+names(mods) <- c("model turnover index", "model teacher and principal", "model spaece")
+
+estimate_hlm <- map2_dfr(
+  mods,
+  names(mods),
+  ~ tidyfit(
+    .x
+  ) %>%
+    mutate(
+      type = .y
+    )
+)
+
+estimate_hlm <- map2_dfr(
+  mods,
+  names(mods),
+  ~ tidyfit(
+    .x
+  ) %>%
+    mutate(
+      type = .y
+    )
+)
+
+plot_hlm <- estimate_hlm %>%
+  ggplot(
+    aes(
+      y = term,
+      x = estimate,
+      color = type,
+      group = type
+    )
+  ) +
+  geom_point(
+    position = ggstance::position_dodgev(height = 0.3),
+    size = 2
+  ) +
+  geom_errorbar_tidy +
+  theme(
+    legend.position = "bottom",
+    legend.title = element_blank()
+  ) +
+  geom_vline(xintercept = 0) +
+  scale_y_discrete(
+    limits = rev(c(
+      "turnover_index",
+      "saeb_principal_experienceless than 2 years",
+      "saeb_teacher_work_schoolless than 2",
+      "censo_rural",
[…]

### 2020-06-12 build/replication/build.R
create new turnover files

@@ -1,83 +1,6 @@
 # ==============================================================================
-# set-up
-# ==============================================================================
-pacman::p_load(
-  tidyverse,
-  data.table,
-  here
-  # GGally,
-  # lfe,
-  # lme4,
-  # plm,
-  # broom,
-  # extrafont,
-  # rprojroot,
-  # tmap,
-  # gridExtra,
-  # cowplot,
-  # ggdag,
-  # scales,
-  # kableExtra,
-  # RColorBrewer,
-  # gghighlight,
-  # parallel,
-  # foreach,
-  # gridExtra,
-  # naniar,
-  # knitr,
-  # gghighlight,
-  # egg,
-  # WeightIt,
-  # cobalt,
-  # boot,
-  # rdrobust,
-  # magrittr,
-  # sjPlot
-)
-
-knitr::opts_chunk$set(
-  eval = T,
-  cache = T,
-  cache.lazy = F,
-  message = F,
-  warning = F,
-  echo = F,
-  fig.height = 3,
-  fig.width = 4,
-  fig.align = "center"
-  # dev = "pdf"
-)
-
-# connect to database
-con <- DBI::dbConnect(
-  odbc::odbc(),
-  driver = "PostgreSQL Unicode",
-  database = "rais",
-  UID = "gali",
-  PWD = 'gali1789!',
-  port = 5432
-)
-
-set.seed(1789)
-
-source(
-  here("build", "replication", "functions.R")
-)
-
-run_module <- partial(
-  run_module,
-  domain = "replication"
-)
-
-read_data <- partial(
-  read_data,
-  type = "clean"
-)
-
-# ==============================================================================
 # run modules
 # ==============================================================================
-
 # run tasks
 modules <- c(
   "visual_map"

### 2020-06-12 build/replication/functions.R

@@ -103,6 +103,8 @@ run_module <- function(domain, module) {
     paste0("running module ", module, "...")
   )
 
+  source_setup(domain)
+
   if(domain == "data"){
     build_repo(module)
   }
@@ -118,6 +120,10 @@ run_module <- function(domain, module) {
   print(
     paste("module", module, "complete!")
   )
+}
+
+source_setup <- function(domain){
+  source(here("build", domain, "setup.R"))
 }
 
 reset_env <- function(init_env, packages){
@@ -148,16 +154,16 @@ fread <- partial(
   integer64 = c("character")
 )
 
-# list_files <- function(path, pattern){
-#   files <- map2(
-#     path,
-#     pattern,
-#     list.files
-#   ) %>% 
-#     flatten_chr
+list_files <- function(path, pattern){
+  files <- map2(
+    path,
+    pattern,
+    list.files
+  ) %>% 
+    flatten_chr
   
-#   return(files)
-# }
+  return(files)
+}
 
 # import_data <- function(path, pattern, names = F){
 #   files <- list_files(
@@ -231,6 +237,15 @@ round_integer <- function(number, digits) {
   number %>% 
     as.integer %>% 
     round(digits)
+}
+
+select_starts <- function(data, pattern){
+  out <- data %>%
+    select(
+      starts_with(pattern)
+    )
+
+  return(out)
 }
 
 # calc_turnover <- function(data, group_vars){
@@ -496,40 +511,17 @@ summarise_fun <- function(data, var){
 # ==============================================================================
 ivreg <- AER::ivreg
 
-fm <- function(dv, predictor,...){
-  covariate <- fm_cov(...)
-  
-  fm_main <- as.formula(
-    paste(substitute(dv), substitute(predictor), sep = ' ~ ')
-  )
-
-  fm <- update(
-    fm_main,
-    covariate
-  )
-  
-  return(fm)
-}
-
-fm_cov <- function(...){
-  controls <- enquos(...) %>% 
-    purrr::map_chr(
-    rlang::as_label
-  )
-
-  covariate <- paste0(
-    controls, 
-    collapse = ' + '
-  )
-
-  formula <- as.formula(
-    paste0(
-      ". ~ . + ",
-      covariate
+formulate_models <- function(response, predictor, fe, controls) {
+  formulae <- c(
+    reformulate(
+      c(predictor, fe), response
+    ),
+    reformulate(
+      c(predictor, controls), response
     )
   )
 
-  return(formula)
+  return(formulae)
 }
 
 # ols

### 2020-06-12 build/replication/modules/model_turnover_saeb.R

@@ -47,6 +47,7 @@ saeb_hierarchical <- lst(
     left_join
   ) %>%
   mutate(
+    saeb_principal_experience = fct_relevel(saeb_principal_experience, "2 to 10"),
     saeb_teacher_work_school = fct_relevel(saeb_teacher_work_school, "2 to 10"),
     censo_log_pop = log(censo_pop),
     budget_education_capita = budget_education / censo_pop
@@ -110,25 +111,21 @@ controls <- c(
   fe
 )
 
-fm_hierarchical <- purrr::partial(
-  stats::reformulate,
-  response = "grade_exam"
+predictors <- paste0(
+  c("turnover_index", "saeb_principal_experience", "saeb_teacher_work_school"),
+  " * grade_level"
 )
 
-formulae <- c(
-  fm_hierarchical(
-    c("turnover_index*grade_level", fe)
-  ),
-  fm_hierarchical(
-    c("turnover_index*grade_level", controls)
-  ),
-  fm_hierarchical(
-    c("saeb_teacher_work_school*grade_level", fe)
-  ),
-  fm_hierarchical(
-    c("saeb_teacher_work_school*grade_level", controls)
+formulae <- map(
+  predictors,
+  ~ formulate_models(
+    .,
+    response = "grade_exam",
+    fe = fe,
+    controls = controls
   )
-)
+) %>%
+flatten
 
 fit_lmer <- map(
   formulae,
@@ -137,6 +134,12 @@ fit_lmer <- map(
     data = saeb_hierarchical
   )
 )
+
+names(fit_lmer) <- map(
+  c("turnover", "principal", "teacher"),
+  ~paste(., c("baseline", "controls"), sep = "_")
+) %>%
+flatten_chr
 
 fit_lmer %>%
   write_model(

### 2020-06-12 build/replication/modules/model_turnover_spaece.R

@@ -1,6 +1,10 @@
 # ==============================================================================
 # spaece: annual standardized test data for the state of ceara
 # ==============================================================================
+source(
+  here::here("build", "replication", "setup.R")
+)
+
 spaece <- read_data(
   "spaece",
   "spaece.rds"
@@ -40,8 +44,7 @@ censo_school_ceara <- censo_school_turnover %>%
 spaece_turnover <- censo_school_ceara %>% 
   left_join(
     censo_school %>% 
-      filter(dep == 'municipal', year >= 2007) %>% 
-      transmute(school_id = as.integer(school_id), year, toilet, meal),
+      filter(dep == 'municipal', year >= 2007),
     by = c('school_id', 'year')
   ) %>% 
   left_join(
@@ -68,14 +71,28 @@ spaece_turnover <- spaece_turnover %>%
     scale
   )
 
-formulae_spaece <- c(
-  as.formula(
-    spaece_mean ~ turnover_index * grade + as.factor(year)
-  ),
-  as.formula(
-    spaece_mean ~ turnover_index * grade + as.factor(year) + toilet + meal +
-      censo_rural + censo_log_pop + censo_lit_rate
-  )
+school_cov <- c(
+  "access_water",
+  "access_electricity",
+  "library",
+  "meal"
+)
+
+mun_cov <- c(
+  "censo_median_wage",
+  "censo_log_pop",
+  "censo_rural",
+  "censo_lit_rate",
+  "budget_education_capita"
+)
+
+controls <- c(school_cov, mun_cov)
+
+formulae_spaece <- formulate_models(
+  "spaece_mean",
+  "turnover_index * grade",
+  fe = "as.factor(year)",
+  controls
 )
 
 fit_spaece <- map(
@@ -89,18 +106,23 @@ fit_spaece <- map(
   )
 )
 
+names(fit_spaece) <- c(
+  "turnover_baseline",
+  "turnover_controls"
+)
+
 fit_spaece %>%
     write_model(
         "fit_spaece.rds"
     )
 
-sink(here("replication", "results", "spaece_result.tex"))
-mstar(
-  fit_spaece,
-  keep = c("turnover_index"),
-  add.lines = list(c("Controls", rep(c("\\_", "\\checkmark"), 1))),
-  covariate.labels = c("Turnover index", "Turnover index $\\times$ Grade 9"),
-  dep.var.caption = "Student learning",
[…]

### 2020-06-12 build/replication/modules/visual_student_learning.R

@@ -1,6 +1,8 @@
 # ==============================================================================
 # visualize combined results from saeb hierarchical and spaece models
 # ==============================================================================
+source_setup("replication")
+
 fit_lmer <- read_model("fit_saeb_hierarchical.rds")
 fit_spaece <- read_model("fit_spaece.rds")
 
@@ -24,9 +26,11 @@ mstar(
 sink()
 
 # visualization
+specs <- paste0(c("turnover", "principal", "teacher"), "_controls")
+
 mods <- c(
-  fit_lmer[c(2, 4)],
-  fit_spaece[2]
+  fit_lmer[specs],
+  fit_spaece["turnover_controls"]
 )
 
 names(mods) <- c("model turnover index", "model teacher and principal", "model spaece")
@@ -50,7 +54,8 @@ estimate_hlm <- map2_dfr(
   ) %>%
     mutate(
       type = .y
-    )
+    ),
+  .id = "model"
 )
 
 plot_hlm <- estimate_hlm %>%
@@ -91,11 +96,10 @@ plot_hlm <- estimate_hlm %>%
       "education_teacherhigher education" = "Teacher: higher education",
       "saeb_wage_teachermore than 3000" = "Teacher: wage more than $750"
     )
-  ) 
-#   +
-#   coord_cartesian(
-#     xlim = c(-.1, .1)
-#   )
+  ) +
+  coord_cartesian(
+    xlim = c(-.25, .25)
+  )
 
 ggsave(
   plot_hlm,

### 2020-06-12 build/replication/setup.R

@@ -1,0 +1,75 @@
+# ==============================================================================
+# set-up
+# ==============================================================================
+pacman::p_load(
+  tidyverse,
+  data.table,
+  here
+  # GGally,
+  # lfe,
+  # lme4,
+  # plm,
+  # broom,
+  # extrafont,
+  # rprojroot,
+  # tmap,
+  # gridExtra,
+  # cowplot,
+  # ggdag,
+  # scales,
+  # kableExtra,
+  # RColorBrewer,
+  # gghighlight,
+  # parallel,
+  # foreach,
+  # gridExtra,
+  # naniar,
+  # knitr,
+  # gghighlight,
+  # egg,
+  # WeightIt,
+  # cobalt,
+  # boot,
+  # rdrobust,
+  # magrittr,
+  # sjPlot
+)
+
+knitr::opts_chunk$set(
+  eval = T,
+  cache = T,
+  cache.lazy = F,
+  message = F,
+  warning = F,
+  echo = F,
+  fig.height = 3,
+  fig.width = 4,
+  fig.align = "center"
+  # dev = "pdf"
+)
+
+# connect to database
+con <- DBI::dbConnect(
+  odbc::odbc(),
+  driver = "PostgreSQL Unicode",
+  database = "rais",
+  UID = "gali",
+  PWD = 'gali1789!',
+  port = 5432
+)
+
+set.seed(1789)
+
+source(
+  here("build", "replication", "functions.R")
+)
+
+run_module <- partial(
+  run_module,
+  domain = "replication"
+)
+
+read_data <- partial(
+  read_data,
+  type = "clean"
+)

### 2020-06-12 build/replication/build.R
finalize visualization of staff turnover

@@ -1,6 +1,10 @@
 # ==============================================================================
 # run modules
 # ==============================================================================
+source(
+  here("build", domain, "setup.R")
+)
+
 # run tasks
 modules <- c(
   "visual_map"

### 2020-06-12 build/replication/functions.R

@@ -122,10 +122,6 @@ run_module <- function(domain, module) {
   )
 }
 
-source_setup <- function(domain){
-  source(here("build", domain, "setup.R"))
-}
-
 reset_env <- function(init_env, packages){
   final_env <- ls(.GlobalEnv)
 

### 2020-06-12 build/replication/modules/model_turnover_saeb.R

@@ -1,4 +1,6 @@
-# hierarchical model ------------------------------------------------------
+# ==============================================================================
+# hierarchical estimation
+# ==============================================================================
 f <- stats::as.formula
 
 saeb <- read_data(
@@ -78,8 +80,8 @@ teacher_cov <- c(
 
 principal_cov <- c(
   "saeb_principal_female",
-  "saeb_principal_education"
-  # "saeb_principal_appointment"
+  "saeb_principal_education",
+  "saeb_principal_appointment"
 )
 
 student_cov <- c(
@@ -111,21 +113,20 @@ controls <- c(
   fe
 )
 
-predictors <- paste0(
-  c("turnover_index", "saeb_principal_experience", "saeb_teacher_work_school"),
-  " * grade_level"
-)
-
-formulae <- map(
-  predictors,
-  ~ formulate_models(
-    .,
+formulae <- c(
+  formulate_models(
+    "turnover_index * grade_level",
+    response = "grade_exam",
+    fe = fe,
+    controls = controls
+  ),
+  formulate_models(
+    "saeb_principal_experience * grade_level + saeb_teacher_work_school * grade_level",
     response = "grade_exam",
     fe = fe,
     controls = controls
   )
-) %>%
-flatten
+)
 
 fit_lmer <- map(
   formulae,
@@ -136,7 +137,7 @@ fit_lmer <- map(
 )
 
 names(fit_lmer) <- map(
-  c("turnover", "principal", "teacher"),
+  c("turnover", "principal_teacher"),
   ~paste(., c("baseline", "controls"), sep = "_")
 ) %>%
 flatten_chr
@@ -146,19 +147,19 @@ fit_lmer %>%
     "fit_saeb_hierarchical.rds"
   )
 
-sink(
-  here("replication", "results", "model_hierarchical.tex")
-)
-mstar(
-  fit_lmer,
-  keep = c("turnover_index", "saeb_principal_experience", "saeb_teacher_work_school"),
-  add.lines = list(c("Controls", rep(c("\\_", "\\checkmark"), 2))),
-  covariate.labels = c(
-    "Turnover index", "Turnover index $\\times$ Grade 9",
-    "Teacher experience (2 years)", "Teacher experience (10 years)",
-    "School principal experience (2 years)", "School principal experience (10 years)"
-  ),
-  dep.var.caption = "Student learning",
-  dep.var.labels = "SAEB average test scores"
-)
-sink()
+# sink(
[…]

### 2020-06-12 build/replication/modules/model_turnover_spaece.R

@@ -10,6 +10,11 @@ spaece <- read_data(
   "spaece.rds"
 )
 
+finbra <- read_data(
+  "finbra",
+  "finbra.rds"
+)
+
 censo_school <- read_data(
   "censo_escolar",
   "censo_school.rds"
@@ -45,7 +50,7 @@ spaece_turnover <- censo_school_ceara %>%
   left_join(
     censo_school %>% 
       filter(dep == 'municipal', year >= 2007),
-    by = c('school_id', 'year')
+    by = c('cod_ibge_6', 'school_id', 'year')
   ) %>% 
   left_join(
     spaece,
@@ -60,7 +65,8 @@ spaece_turnover <- censo_school_ceara %>%
 
 spaece_turnover <- spaece_turnover %>% 
   mutate(
-    state = str_sub(cod_ibge_6, 1, 2)
+    state = str_sub(cod_ibge_6, 1, 2),
+    budget_education_capita = budget_education / censo_pop
   ) %>% 
   mutate_at(
     vars(grade_level, year, state),

### 2020-06-12 build/replication/modules/visual_model_turnover.R

@@ -1,7 +1,9 @@
 # ==============================================================================
 # visualize combined results from saeb hierarchical and spaece models
 # ==============================================================================
-source_setup("replication")
+source(
+  here::here("build", "replication", "setup.R")
+)
 
 fit_lmer <- read_model("fit_saeb_hierarchical.rds")
 fit_spaece <- read_model("fit_spaece.rds")
@@ -26,14 +28,12 @@ mstar(
 sink()
 
 # visualization
-specs <- paste0(c("turnover", "principal", "teacher"), "_controls")
-
 mods <- c(
-  fit_lmer[specs],
+  fit_lmer[c("turnover_controls", "principal_teacher_controls")],
   fit_spaece["turnover_controls"]
 )
 
-names(mods) <- c("model turnover index", "model teacher and principal", "model spaece")
+names(mods) <- c("model turnover saeb", "model principal_teacher", "model turnover_spaece")
 
 estimate_hlm <- map2_dfr(
   mods,
@@ -54,11 +54,10 @@ estimate_hlm <- map2_dfr(
   ) %>%
     mutate(
       type = .y
-    ),
-  .id = "model"
+    )
 )
 
-plot_hlm <- estimate_hlm %>%
+plot_model_turnover <- estimate_hlm %>%
   ggplot(
     aes(
       y = term,
@@ -78,33 +77,34 @@ plot_hlm <- estimate_hlm %>%
   ) +
   geom_vline(xintercept = 0) +
   scale_y_discrete(
-    limits = rev(c(
+    limits = c(
       "turnover_index",
       "saeb_principal_experienceless than 2 years",
       "saeb_teacher_work_schoolless than 2",
-      "censo_rural",
-      "censo_log_pop",
+      "saeb_principal_appointmentpublic exam",
+      "saeb_principal_educationhigher education",
       "education_teacherhigher education",
       "saeb_wage_teachermore than 3000"
-    )),
+    ),
     labels = c(
-      "turnover_index" = "Teacher turnover index",
-      "saeb_principal_experienceless than 2 years" = "Principal: less than 2 years of exp.",
-      "saeb_teacher_work_schoolless than 2" = "Teacher: less than 2 years in school",
-      "censo_rural" = "Rural population",
-      "censo_log_pop" = "Logged population",
-      "education_teacherhigher education" = "Teacher: higher education",
[…]

### 2020-06-13 build/replication/build.R
fix replication files

@@ -2,22 +2,25 @@
 # run modules
 # ==============================================================================
 source(
-  here("build", domain, "setup.R")
+  here::here("build", "replication", "setup.R")
 )
 
 # run tasks
 modules <- c(
-  "visual_map"
+  "visual_map",
   "visual_global_edu",
   "visual_budget",
   "visual_bureaucracy",
-  "model_hierarchical",
-  "model_saeb"
+  "visual_electoral",
+  "visual_turnover",
+  "model_turnover_saeb",
+  "model_turnover_spaece",
+  "visual_model_turnover"
   )
-
+  
 walk(
   modules,
-  ~run_module(.)
+  ~run_module("replication", .)
 )
 
 # turnover spaece ---------------------------------------------------------

### 2020-06-13 build/replication/functions.R

@@ -103,8 +103,6 @@ run_module <- function(domain, module) {
     paste0("running module ", module, "...")
   )
 
-  source_setup(domain)
-
   if(domain == "data"){
     build_repo(module)
   }

### 2020-06-13 build/replication/modules/visual_global_edu.R

@@ -74,9 +74,9 @@ plot_global <- cowplot::plot_grid(
   ncol = 2
 )
 
-ggsave(
+save_fig(
   plot_global,
-  filename = p_file_here('figs', 'plot_global.pdf')
+  "plot_global.pdf"
 )
 
 plot_enroll <- edu_global %>%
@@ -150,9 +150,9 @@ plot_edu_1 <- cowplot::plot_grid(
   ncol = 2
 )
 
-ggsave(
+save_fig(
   plot_edu_1,
-  filename = p_file_here('figs', 'plot_global_1.pdf')
+   'plot_global_1.pdf'
 )
 
 plot_enroll_type <- edu_global %>%
@@ -245,7 +245,7 @@ plot_edu_2 <- cowplot::plot_grid(
   ncol = 2
 )
 
-ggsave(
+save_fig(
   plot_edu_2,
-  filename = p_file_here('figs', 'plot_global_2.pdf')
+  "plot_global_2.pdf"
 )

### 2020-06-13 build/replication/modules/visual_map.R

@@ -120,5 +120,3 @@ save_fig(
     width = 5,
     height = 5
 )
-
-detach_p(packages)

### 2020-06-13 build/replication/modules/visual_turnover.R

@@ -85,79 +85,79 @@ save_fig(
 )
 
 
-patronage_category <- patronage %>%
-    mutate(
-        year = as.integer(year)
-    ) %>%
-    filter(
-        cbo_category %in% c("administration", "education", "healthcare", "services")
-    )
+# patronage_category <- patronage %>%
+#     mutate(
+#         year = as.integer(year)
+#     ) %>%
+#     filter(
+#         cbo_category %in% c("administration", "education", "healthcare", "services")
+#     )
 
-patronage_category %>%
-    group_by(year, cbo_category) %>%
-    summarise(
-        prop_hired = sum(prop_hired * total) / sum(total)
-    ) %>%
-    ungroup() %>%
-    ggplot(
-        aes(year, prop_hired, group = cbo_category, color = cbo_category)
-    ) +
-    labs(x = "year", y = "proportion of hires") +
-    geom_line() +
-    geom_point(
-        size = 2
-    ) +
-    geom_vline(
-        xintercept = seq(2005, 2013, 4),
-        linetype = "dotted"
-    ) +
-    scale_x_continuous(
-        breaks = seq(2003, 2016, 2)
-    ) +
-    ggsave(
-        p_file_here(
-            "figs", "hire_by_category"
-        )
-    )
+# patronage_category %>%
+#     group_by(year, cbo_category) %>%
+#     summarise(
+#         prop_hired = sum(prop_hired * total) / sum(total)
+#     ) %>%
+#     ungroup() %>%
+#     ggplot(
+#         aes(year, prop_hired, group = cbo_category, color = cbo_category)
+#     ) +
+#     labs(x = "year", y = "proportion of hires") +
+#     geom_line() +
+#     geom_point(
+#         size = 2
+#     ) +
+#     geom_vline(
+#         xintercept = seq(2005, 2013, 4),
+#         linetype = "dotted"
+#     ) +
+#     scale_x_continuous(
+#         breaks = seq(2003, 2016, 2)
+#     ) +
+#     ggsave(
+#         p_file_here(
+#             "figs", "hire_by_category"
+#         )
+#     )
 
 
-# address alternative theories
-# population
-patronage_category %>%
-    gg_summary(
-        censo_pop,
-        prop_hired
-    ) +
-    scale_x_log10(
-        breaks = scales::trans_breaks("log10", function(x) 10^x),
[…]

### 2020-06-13 build/replication/build.R
finalize refactoring

@@ -17,10 +17,10 @@ modules <- c(
   "model_turnover_spaece",
   "visual_model_turnover"
   )
-  
+
 walk(
   modules,
-  ~run_module("replication", .)
+  ~run_module(.)
 )
 
 # turnover spaece ---------------------------------------------------------

### 2020-06-15 build/data/modules/datagen_saeb.R
fix saeb bug: incorrect grade level

@@ -73,6 +73,9 @@ saeb_hierarchical <- saeb_hierarchical %>%
   rename(
     school_id = cod_school,
     grade_level = grade
+  ) %>%
+  mutate(
+    grade_level = grade + 1
   )
   
 saeb_hierarchical %>%

### 2020-06-15 build/replication/build.R

@@ -7,74 +7,21 @@ source(
 
 # run tasks
 modules <- c(
-  "visual_map",
-  "visual_global_edu",
-  "visual_budget",
-  "visual_bureaucracy",
-  "visual_electoral",
-  "visual_turnover",
-  "model_turnover_saeb",
-  "model_turnover_spaece",
-  "visual_model_turnover"
+  # "visual_map",
+  # "visual_global_edu",
+  # "visual_budget",
+  # "visual_bureaucracy",
+  # "visual_electoral",
+  # "visual_turnover",
+  "visual_turnover_score"
+  # "model_turnover_saeb",
+  # "model_turnover_spaece",
+  # "visual_model_turnover"
   )
 
 walk(
   modules,
   ~run_module(.)
-)
-
-# turnover spaece ---------------------------------------------------------
-plot_spaece <- censo_turnover_score %>% 
-  filter(
-    n >= 5
-  ) %>% 
-  mutate(
-    grade_level = as.factor(grade_level)
-  ) %>% 
-  ggplot(
-    aes(
-      turnover_index,
-      spaece_mean,
-      group = as.factor(grade_level),
-      col = as.factor(grade_level)
-    )
-  ) +
-  geom_smooth(
-    method = lm, formula = y ~ splines::bs(x, 3),
-    se = F
-  ) +
-  annotate(
-    "text",
-    label = "2nd grade",
-    size = 2,
-    col = "grey65",
-    0.45, 165
-  ) +
-  annotate(
-    "text",
-    label = "5th grade",
-    size = 2,
-    col = "grey65",
-    0.6, 200
-  ) +
-  annotate(
-    "text",
-    label = "9th grade",
-    size = 2,
-    col = "grey65",
-    0.7, 225
-  ) +
-  labs(
-    x = "Teacher turnover (index)",
-    y = "Test scores (SPAECE)"
-  ) +
-  theme(
-    legend.position = "none"
-  )
-
-ggsave(
-  plot_spaece,
-  filename = p_file_here('figs', "turnover_spaece.pdf")
 )
 
 # explaining turnover -----------------------------------------------------

### 2020-06-15 build/replication/functions.R

@@ -132,6 +132,8 @@ reset_env <- function(init_env, packages){
 }
 
 save_fig <- function(pl, file, width = 5, height = 3, ...){
+  print(pl)
+
   ggplot2::ggsave(
     filename = here("replication", "figs", file),
     pl,

### 2020-06-15 build/replication/modules/model_turnover_saeb.R

@@ -6,7 +6,10 @@ f <- stats::as.formula
 saeb <- read_data(
   "saeb",
   "saeb_hierarchical.rds"
-)
+) %>%
+  mutate(
+    grade_level = grade_level + 1
+  )
 
 finbra <- read_data(
   "finbra",
@@ -38,6 +41,38 @@ censo_school_turnover <- read_data(
     n_teacher >= 5
   )
 
+# # fix problem with turnover index: why is there so much missingness
+# turnover_test <- censo_school_turnover %>%
+#   filter(year == 2007) %>%
+#   select(
+#     cod_ibge_6,
+#     school_id,
+#     year,
+#     grade_level
+#   )
+
+# saeb_test  <- saeb %>%
+#   filter(year == 2007) %>%
+#   select(
+#     cod_ibge_6,
+#     school_id,
+#     year,
+#     grade_level
+#   )
+
+# list(turnover_test, saeb_test) %>%
+# map(
+#   ~filter(., cod_ibge_6 == 110002) %>%
+#   distinct(school_id) %>% 
+#   arrange(school_id)%>%
+#   head
+# )
+
+# # check how many school ids are present
+# turnover_test %>% 
+#   distinct(school_id) %>% 
+#   anti_join(saeb_test %>% distinct(school_id))
+
 # prepare data for estimation
 saeb_hierarchical <- lst(
   saeb,
@@ -54,6 +89,9 @@ saeb_hierarchical <- lst(
     censo_log_pop = log(censo_pop),
     budget_education_capita = budget_education / censo_pop
   )
+
+saeb_hierarchical %>%
+  gg_miss_var()
 
 # fix blank strings
 saeb_hierarchical <- saeb_hierarchical %>%

### 2020-06-15 build/replication/modules/model_turnover_spaece.R

@@ -8,7 +8,11 @@ source(
 spaece <- read_data(
   "spaece",
   "spaece.rds"
-)
+) %>%
+  filter(
+    dep == "municipal",
+    grade_level != 2
+  )
 
 finbra <- read_data(
   "finbra",
@@ -42,7 +46,7 @@ censo_school_turnover <- read_data(
 
 censo_school_ceara <- censo_school_turnover %>% 
   filter(
-    str_sub(cod_ibge_6, 1, 2) == "23",
+    str_sub(cod_ibge_6, 1, 2) == "23", # state of ceara
     grade_level %in% c(2, 5, 9)
   )
 
@@ -75,6 +79,9 @@ spaece_turnover <- spaece_turnover %>%
   mutate_if(
     is.double,
     scale
+  ) %>%
+  filter(
+    grade_level != 2
   )
 
 school_cov <- c(
@@ -105,10 +112,7 @@ fit_spaece <- map(
   formulae_spaece,
   ~ lm(
     formula = .,
-    data = spaece_turnover %>%
-      mutate_if(is.double, scale) %>%
-      rename(grade = grade_level) %>%
-      filter(grade != 2)
+    data = spaece_turnover
   )
 )
 

### 2020-06-15 build/replication/modules/visual_model_turnover.R

@@ -1,10 +1,6 @@
 # ==============================================================================
 # visualize combined results from saeb hierarchical and spaece models
 # ==============================================================================
-source(
-  here::here("build", "replication", "setup.R")
-)
-
 fit_lmer <- read_model("fit_saeb_hierarchical.rds")
 fit_spaece <- read_model("fit_spaece.rds")
 
@@ -104,27 +100,3 @@ save_fig(
   plot_model_turnover,
   file = "model_turnover_learning.pdf"
 )
-
-#
-# fit_hlm <- invoke_map(
-#   list(
-#     lm,
-#     lmer
-#   ),
-#   list(
-#     list(
-#       formula = update.formula(
-#         formula_hlm,
-#         . ~ percent_entry + percent_exit + as.factor(uf) + as.factor(year) + .
-#       ),
-#     data = saeb_hierarchical
-#     ),
-#     list(
-#       formula = update.formula(
-#         formula_hlm,
-#         . ~ percent_entry + percent_exit + (1|uf) + (1|year) + .
-#     ),
-#     data = saeb_hierarchical
-#     )
-#   )
-# )

### 2020-06-15 build/replication/modules/visual_turnover_score.R

@@ -1,0 +1,108 @@
+# turnover spaece ---------------------------------------------------------
+saeb <- read_data(
+  "saeb",
+  "saeb_hierarchical.rds"
+) %>%
+  select(
+    cod_ibge_6,
+    school_id,
+    year,
+    grade_level,
+    grade_exam
+  )
+
+spaece <- read_data(
+  "spaece",
+  "spaece.rds"
+)
+
+censo_school_turnover <- read_data(
+  "censo_escolar",
+  "censo_school_turnover.rds"
+) %>%
+  select(
+    cod_ibge_6,
+    year,
+    school_id,
+    grade_level,
+    turnover_index,
+    starts_with("percent"),
+    n_teacher = n
+  ) %>%
+  filter(
+    n_teacher >= 5
+  )
+
+saeb_turnover <- saeb %>%
+  left_join(
+    censo_school_turnover,
+    by = c("cod_ibge_6", "year", "school_id", "grade_level")
+  )
+
+spaece_turnover <- spaece %>%
+  filter(
+    dep == "municipal",
+    grade_level != 2
+  ) %>%
+  left_join(
+    censo_school_turnover,
+    by = c("cod_ibge_6", "year", "school_id", "grade_level")
+  )
+
+plot_turnover_saeb <- censo_school_turnover %>%
+  left_join(
+    by = c("")
+  )
+
+plot_turnover_spaece <- censo_turnover_score %>% 
+  filter(
+    n >= 5
+  ) %>% 
+  mutate(
+    grade_level = as.factor(grade_level)
+  ) %>% 
+  ggplot(
+    aes(
+      turnover_index,
+      spaece_mean,
+      group = as.factor(grade_level),
+      col = as.factor(grade_level)
+    )
+  ) +
+  geom_smooth(
+    method = lm, formula = y ~ splines::bs(x, 3),
+    se = F
+  ) +
+  annotate(
+    "text",
+    label = "2nd grade",
+    size = 2,
+    col = "grey65",
+    0.45, 165
+  ) +
+  annotate(
+    "text",
+    label = "5th grade",
+    size = 2,
+    col = "grey65",
+    0.6, 200
+  ) +
+  annotate(
+    "text",
+    label = "9th grade",
+    size = 2,
+    col = "grey65",
+    0.7, 225
+  ) +
+  labs(
+    x = "Teacher turnover (index)",
+    y = "Test scores (SPAECE)"
+  ) +
+  theme(
[…]

### 2020-06-18 source/models/modules/model_coalition_turnover.R
finish refactoring and replicating turnover models

@@ -33,7 +33,7 @@ model_school_turnover <- censo_school_turnover %>%
 # estimation of the effect of coalition share on staff turnover
 # ==============================================================================
 controls_turnover <- c(
-  mun_covariates school_covariates
+  mun_covariates, school_covariates
 )
 
 formulae_turnover <- formulate_models(

### 2020-06-18 source/models/modules/model_turnover_saeb.R

@@ -8,10 +8,23 @@ f <- stats::as.formula
 saeb <- read_data(
   "saeb",
   "saeb_hierarchical.rds"
-) %>%
+)
+
+saeb <- saeb %>%
   filter(
     year >= 2007
-  )
+  ) %>%
+  mutate(
+    saeb_principal_higher_education = if_else(
+      saeb_principal_education == "higher education", 1L, 0L
+    ) %>% as.factor(),
+    saeb_principal_experience = fct_relevel(
+      saeb_principal_experience, "2 to 10"
+    ),
+    saeb_teacher_work_school = fct_relevel(
+      saeb_teacher_work_school, "2 to 10"
+    )
+  ) 
 
 finbra <- read_data(
   "finbra",
@@ -24,6 +37,12 @@ censo_school <- read_data(
 ) %>%
   mutate(
     school_id = as.integer(school_id)
+  ) %>%
+  select(
+    cod_ibge_6,
+    school_id,
+    year,
+    all_of(school_covariates)
   )
 
 censo_school_turnover <- read_data(
@@ -34,13 +53,14 @@ censo_school_turnover <- read_data(
     cod_ibge_6,
     year,
     school_id,
+    grade_level,
     turnover_index,
     starts_with("percent"),
     n_teacher = n
   )
 
-# prepare data for estimation
-print("joining datasets")
+print("joining datasets for estimation")
+
 saeb_hierarchical <- list(
   saeb,
   finbra,
@@ -49,38 +69,30 @@ saeb_hierarchical <- list(
 ) %>%
   reduce(
     left_join
-  ) %>%
+  )
+
+saeb_hierarchical <- saeb_hierarchical %>%
   mutate_at(
     vars(
       state, year, grade_level,
       starts_with("access")
-      ),
-      as.factor
+    ),
+    as.factor
   ) %>%
   mutate(
-    saeb_principal_higher_education = if_else(
-      saeb_principal_education == "higher education", 1L, 0L
-    ) %>% as.factor,
-    saeb_principal_experience = fct_relevel(
-      saeb_principal_experience, "2 to 10"
-    ),
-    saeb_teacher_work_school = fct_relevel(
-      saeb_teacher_work_school, "2 to 10"
-    ),
     censo_log_pop = log(censo_pop),
     budget_education_capita = budget_education / censo_pop
-  )
-
-# fix blank strings
-saeb_hierarchical <- saeb_hierarchical %>%
+  ) %>%
[…]

### 2020-06-18 source/models/modules/model_turnover_spaece.R

@@ -5,6 +5,8 @@ source(
   here::here("source", "models", "setup.R")
 )
 
+print("pre-processing data")
+
 spaece <- read_data(
   "spaece",
   "spaece.rds"
@@ -74,16 +76,18 @@ spaece_turnover <- spaece_turnover %>%
   ) %>% 
   mutate_if(
     is.double,
-    scale
+    scale_z
   ) %>%
   filter(
     grade_level != 2
   )
 
+  print("pre-processing complete!")
+
 # ==============================================================================
 # estimate effect of turnover on mean student test scores
 # ==============================================================================
-controls <- c(school_covariates mun_cov)
+controls <- c(school_covariates, mun_covariates)
 
 formulae_spaece <- formulate_models(
   "spaece_mean",
@@ -92,7 +96,7 @@ formulae_spaece <- formulate_models(
   controls
 ) %>%
   map(
-    ~add_felm(., fe = "year", cluster = "cod_ibge_6" )
+    ~add_felm(., fe = "year", cluster = "0" )
   )
 
 fit_spaece <- map(

### 2020-06-18 source/models/modules/visual_model_coalition.R

@@ -1,126 +1,0 @@
-# table of results
-sink(p_file_here("results", "turnover_result.tex"))
-mstar(
-  list(
-    fit_turnover[[2]],
-    fit_hired,
-    fit_fired,
-    fit_hired_mun,
-    fit_fired_mun
-  ),
-  keep = c("coalition_share", "rais_category"),
-  add.lines = list(c("Controls", rep("\\checkmark", 5))),
-  covariate.labels = c("Executive share of seats", "School principal", "Executive share of seats X School principal"),
-  dep.var.labels.include = F,
-  column.labels = c("Turnover index", "Hired (Logit)", "Fired (Logit)", "Hired (FE)", "Fired (FE)"),
-  column.separate = rep(1, 5),
-  model.names = FALSE
-)
-sink()
-
-plot_mun <- fit_hired_mun%>% 
-  tidycoef(
-    vars = "coalition|mayor_reelected|censo_log_pop|censo_median_wage"
-  ) +
-  # scale_y_discrete(
-  #   labels = rev(c("Share of legislative seats", "Population size (logged)", "Median wage", "Size of coalition", "Reelected", "Share of legislative seats (sq.)"))
-  # ) +
-  theme(
-    axis.title = element_blank()
-  )
-
-ggsave(
-  plot_mun,
-  filename = p_file_here("figs", "fit_mun.pdf")
-)
-
-plot_hired <- plot_grid(
-  plot_logit,
-  plot_mun,
-  nrow = 2,
-  labels = c("Logit", "OLS"),
-  label_size = 7.5
-)
-
-ggsave(
-  plot_hired,
-  filename = p_file_here("figs", "hired_fit.pdf")
-)
-
-# predicted vals
-model_fit_hired_mun <- model.frame(
-  fit_hired_mun
-) 
-
-plot_hired <- model_fit_hired_mun %>% 
-  mutate(
-    predict = fit_hired_mun$fitted.values
-  ) %>% 
-  group_by(coalition_share) %>% 
-  mutate(predict_mean = mean(rais_hired, na.rm = T)) %>% 
-  ggplot(
-    aes(x = coalition_share)
-  ) +
-  geom_count(
-    aes(y = predict_mean),
-    alpha = 0.1
-  ) +
-  geom_smooth(
-    aes(y = predict),
-    method = lm,
-    formula = y ~ x,
-    col = "coral3"
-  ) +
-  coord_cartesian(
[…]

### 2020-06-18 source/models/source_models.R

@@ -7,9 +7,10 @@ source(
 
 # run tasks
 modules <- c(
+  "model_turnover_saeb",
+  "model_turnover_spaece",
+  "visual_model_turnover",
   "model_coalition_turnover"
-  # "model_turnover_spaece",
-  # "model_coalition_turnover"
   )
 
 walk(

### 2020-06-18 source/models/modules/visual_model_coalition.R
finalize visual_model_turnover and ensure replicability

@@ -1,0 +1,126 @@
+# table of results
+sink(p_file_here('results', 'turnover_result.tex'))
+mstar(
+  list(
+    fit_turnover[[2]],
+    fit_hired,
+    fit_fired,
+    fit_hired_mun,
+    fit_fired_mun
+  ),
+  keep = c('coalition_share', 'rais_category'),
+  add.lines = list(c("Controls", rep("\\checkmark", 5))),
+  covariate.labels = c('Executive share of seats', 'School principal', 'Executive share of seats X School principal'),
+  dep.var.labels.include = F,
+  column.labels = c('Turnover index', 'Hired (Logit)', 'Fired (Logit)', 'Hired (FE)', 'Fired (FE)'),
+  column.separate = rep(1, 5),
+  model.names = FALSE
+)
+sink()
+
+plot_mun <- fit_hired_mun%>% 
+  tidycoef(
+    vars = "coalition|mayor_reelected|censo_log_pop|censo_median_wage"
+  ) +
+  # scale_y_discrete(
+  #   labels = rev(c("Share of legislative seats", "Population size (logged)", "Median wage", "Size of coalition", "Reelected", "Share of legislative seats (sq.)"))
+  # ) +
+  theme(
+    axis.title = element_blank()
+  )
+
+ggsave(
+  plot_mun,
+  filename = p_file_here('figs', "fit_mun.pdf")
+)
+
+plot_hired <- plot_grid(
+  plot_logit,
+  plot_mun,
+  nrow = 2,
+  labels = c("Logit", "OLS"),
+  label_size = 7.5
+)
+
+ggsave(
+  plot_hired,
+  filename = p_file_here('figs', "hired_fit.pdf")
+)
+
+# predicted vals
+model_fit_hired_mun <- model.frame(
+  fit_hired_mun
+) 
+
+plot_hired <- model_fit_hired_mun %>% 
+  mutate(
+    predict = fit_hired_mun$fitted.values
+  ) %>% 
+  group_by(coalition_share) %>% 
+  mutate(predict_mean = mean(rais_hired, na.rm = T)) %>% 
+  ggplot(
+    aes(x = coalition_share)
+  ) +
+  geom_count(
+    aes(y = predict_mean),
+    alpha = 0.1
+  ) +
+  geom_smooth(
+    aes(y = predict),
+    method = lm,
+    formula = y ~ x,
+    col = "coral3"
+  ) +
[…]

### 2020-06-18 source/models/setup.R

@@ -39,6 +39,11 @@ read_data <- partial(
   type = "clean"
 )
 
+save_fig <- partial(
+  save_fig,
+  domain = "models"
+)
+
 theme_set(
   theme_minimal() +
     theme_clean

### 2020-06-18 source/models/utils.R

@@ -111,11 +111,11 @@ reset_env <- function(init_env, packages){
   gc()
 }
 
-save_fig <- function(pl, file, width = 5, height = 3, ...){
+save_fig <- function(pl, domain, file, width = 5, height = 3, ...){
   print(pl)
 
   ggplot2::ggsave(
-    filename = here("replication", "figs", file),
+    filename = here(domain, "figs", file),
     pl,
     width = width,
     height = height,

### 2020-06-22 theory/notes_groseclose_et_snyder.tex
begin analysis of groseclose and snyder

\documentclass[a4paper,12pt]{article}
\usepackage[english]{babel}
\usepackage{amsmath}
\setlength{\parindent}{0pt}

\begin{document}
\section{In brief}
Objective is to redact notes on Groseclose and Snyder (1991) and provide comparative statics on the key hypotheses of the model. Summarize how these partial derivatives translate into observable implications for reduced form findings.

\section{Set-up}

\subsection*{Players:}
\begin{enumerate}
    \item Legislator: \textit{i}.
    \item Vote buyers: \{\textit{A}, \textit{B}\}
\end{enumerate}

\subsection*{Outcomes:}
\begin{enumerate}
    \item $y = \{s, x\}$
\end{enumerate}

\subsection*{Preferences:}
For legislator $i$, denote the intensity of $i$'s preference of voting for $x$ over $s$ as

$$v(i) = u_i(x) - u_i(s)$$

For vote buyer \{\textit{A}, \textit{B}\}

\end{document}

### 2020-06-24 source/descriptive/modules/visual_turnover_score.R
finalize incorporating new iv regression into table

@@ -84,7 +84,7 @@ plot_turnover <- map2_dfr(
 ) %>%
   mutate(
     state = str_sub(cod_ibge_6, 1, 2),
-    test = recode(test, `1` = "spaece", `2` = "saeb")
+    test = dplyr::recode(test, `1` = "spaece", `2` = "saeb")
   ) %>%
   group_by(state, year, test) %>%
   mutate(

### 2020-06-24 source/descriptive/scripts_old/saeb_hierarchical.R

@@ -68,7 +68,7 @@ saeb_hlm %<>%
 saeb_hlm %<>%
   mutate(
     saeb_teacher_work_school = if_else(year_working_school_teacher == "", NA_character_, year_working_school_teacher),
-    grade = recode(grade, `4` = "5", `8` = "9")
+    grade = dplyr::recode(grade, `4` = "5", `8` = "9")
   )
 
 saeb_hlm %>% 

### 2020-06-24 source/models/modules/model_turnover_ivreg.R

@@ -3,10 +3,12 @@
 # note that this relies on the exclusion restriction, which is fundamentally
 # unobservable.
 # ==============================================================================
+source(here::here("source/models/setup.R"))
+
 print("importing pre-processed saeb hierarchical")
 
-run_module(
-    "preprocess_saeb_hierarchical.R"
+source(
+    here("source/models/modules/preprocess_saeb_hierarchical.R")
 )
 
 # add dummy for first year of the electoral term
@@ -34,7 +36,7 @@ controls <- c(
     fe
 )
 
-formula_iv <- reformulate(
+formula_controls <- reformulate(
     c(predictor, controls),
     response
 )
@@ -51,12 +53,12 @@ formula_iv_first_term <- formula_iv %>%
         instrument = "mayor_reelected"
     )
 
-fit_iv_coalition <- AER::ivreg(
+fit_iv_coalition <- ivreg(
     formula_iv_coalition,
     data = saeb_hierarchical
 )
 
-fit_iv_first_term <- AER::ivreg(
+fit_iv_first_term <- ivreg(
     formula_iv_first_term,
     data = saeb_hierarchical
 )

### 2020-06-24 source/models/modules/model_turnover_saeb.R

@@ -5,8 +5,8 @@ print("loading datasets")
 
 f <- stats::as.formula
 
-run_module(
-  "preprocess_saeb_hierarchical.R"
+source(
+    here("source/models/modules/preprocess_saeb_hierarchical.R")
 )
 
 # ==============================================================================
@@ -57,7 +57,7 @@ names(fit_lmer) <- map(
 ) %>%
   flatten_chr()
 
-print("wriing out model output")
+print("writing out model output")
 
 fit_lmer %>%
   write_model(

### 2020-06-24 source/models/modules/visual_covariate_balance.R

@@ -1,6 +1,5 @@
 # present set of covariance balance visualizations to ensure that the treatment var
 # is not strongly correlated with other independent variables
-
 source(here::here("source/models/setup.R"))
 
 print("loading data")
@@ -128,4 +127,4 @@ love_plot_education <- generate_love_plot(
     )
 )
 
-save_fig(love_plot_education, "love_plot_education.pdf")
+save_fig(love_plot_education, "love_plot_education_test.pdf")

### 2020-06-24 source/models/modules/visual_model_turnover.R

@@ -1,14 +1,20 @@
 # ==============================================================================
 # visualize combined results from saeb hierarchical and spaece models
 # ==============================================================================
+library(stargazer)
+
 fit_lmer <- read_model("fit_saeb_hierarchical.rds")
 fit_spaece <- read_model("fit_spaece.rds")
+fit_ivreg <- map(
+  c("fit_ivreg_coalition.rds", "fit_ivreg_first_term.rds"),
+  read_model
+)
 
-sink(here("replication", "results", "student_learning.tex"))
+sink(here("replication", "tables", "student_learning.tex"))
 mstar(
-  c(fit_lmer, fit_spaece),
+  fit_lmer, fit_spaece, fit_spaece, fit_ivreg,
   keep = c("turnover_index", "saeb_principal_experience", "saeb_teacher_work_school"),
-  add.lines = list(c("Controls", rep(c("\\_", "\\checkmark"), 3))),
+  add.lines = list(c("Controls", rep(c("\\_", "\\checkmark"), 4))),
   dep.var.labels.include = F,
   covariate.labels = c(
     "Turnover index", "Turnover index $\\times$ Grade 9",
@@ -17,7 +23,7 @@ mstar(
   ),
   dep.var.caption = "Student learning",
   column.labels = c("SAEB test score", "SPAECE test score"),
-  column.separate = c(4, 2),
+  column.separate = c(6, 2),
   model.names = FALSE,
   keep.stat = "n"
 )

### 2020-06-24 source/models/setup.R

@@ -41,7 +41,7 @@ read_data <- partial(
 
 save_fig <- partial(
   save_fig,
-  domain = "models"
+  domain = "replication"
 )
 
 theme_set(

### 2020-06-24 source/models/source_models.R

@@ -9,7 +9,7 @@ source(
 modules <- c(
   "model_turnover_saeb",
   "model_turnover_spaece",
-  "visual_model_turnover",
+  "model_turnover_ivreg.R",
   "model_coalition_turnover",
   "visual_regression_discontinuity"
   )

### 2020-07-02 source/models/modules/model_coalition_rais_felm.R
incorporate new predicted prob

@@ -1,0 +1,57 @@
+# ==============================================================================
+# construct data to estimate effect of coalition on edu staff turnover
+# ==============================================================================
+rais_edu_mun <- read_data(
+  "rais",
+  "rais_mun_edu.rds"
+)
+
+# ==============================================================================
+print("model specification")
+# ==============================================================================
+controls <- c(
+  mun_covariates, rais_covariates, mayor_covariates, chamber_covariates
+)
+
+formulae_felm <- map(
+  c("rais_hired", "rais_fired"),
+  ~formulate(., "coalition_share*rais_category", fe = NULL, controls) %>%
+    update(. ~ . - rais_edu + rais_higher_edu)
+) %>%
+  map(
+    ~add_felm(., fe = "state + year", cluster = "cod_ibge_6")
+  )
+
+# ==============================================================================
+print("reshape data for estimation")
+# ==============================================================================
+model_rais_mun <- rais_edu_mun %>% 
+  join_covariate() %>%
+  mutate(
+    state = str_sub(cod_ibge_6, 1, 2),
+    rais_category = fct_relevel(rais_category, "teacher"),
+    budget_education_capita = budget_education / censo_pop,
+    across(c(state, year), as.factor)
+  ) %>%
+  model.frame(
+    formula = update(
+      f_logit, . ~ . + cod_ibge_6 - rais_edu + rais_higher_edu + rais_fired
+    ),
+    data = .
+  ) %>%
+  fix_scale()
+
+# ==============================================================================
+# estimation of effect of coalition share on staff turnover
+# ============================================================================== 
+# municipal fe model
+fit_felm <- map(
+  formulae_felm,
+  ~lfe::felm(., data = model_rais_mun)
+)
+
+# ==============================================================================
+print("write-out data")
+# ==============================================================================
+fit_felm %>%
+  write_model("fit_felm_turnover_coalition.rds")

### 2020-07-02 source/models/modules/model_coalition_rais_logit.R

@@ -6,11 +6,6 @@ rais_edu <- read_data(
   "rais_edu.rds"
 )
 
-rais_edu_mun <- read_data(
-  "rais",
-  "rais_mun_edu.rds"
-)
-
 # ==============================================================================
 print("model specification")
 # ==============================================================================
@@ -29,21 +24,11 @@ formulae_logit <- c(
   f_logit,
   update(f_logit, rais_fired ~ .)
 )
-
-formulae_felm <- map(
-  c("rais_hired", "rais_fired"),
-  ~formulate(., "coalition_share*rais_category", fe = NULL, controls) %>%
-    update(. ~ . - rais_edu + rais_higher_edu)
-) %>%
-  map(
-    ~add_felm(., fe = "state + year", cluster = "cod_ibge_6")
-  )
-
 # ==============================================================================
 print("reshape data for estimation")
 # ==============================================================================
 model_rais_micro <- rais_edu %>%
-  sample_frac(1/5) %>%
+  sample_frac(1/3) %>%
   join_covariate() %>%
   mutate(
     state = str_sub(cod_ibge_6, 1, 2),
@@ -57,22 +42,6 @@ model_rais_micro <- rais_edu %>%
   ) %>%
   fix_scale()
 
-model_rais_mun <- rais_edu_mun %>% 
-  join_covariate() %>%
-  mutate(
-    state = str_sub(cod_ibge_6, 1, 2),
-    rais_category = fct_relevel(rais_category, "teacher"),
-    budget_education_capita = budget_education / censo_pop,
-    across(c(state, year), as.factor)
-  ) %>%
-  model.frame(
-    formula = update(
-      f_logit, . ~ . + cod_ibge_6 - rais_edu + rais_higher_edu + rais_fired
-    ),
-    data = .
-  ) %>%
-  fix_scale()
-
 # ==============================================================================
 # estimation of effect of coalition share on staff turnover
 # ============================================================================== 
@@ -81,17 +50,8 @@ fit_turnover <- map(
     ~logit(., data = model_rais_micro)
   )
 
-# municipal fe model
-fit_felm <- map(
-  formulae_felm,
-  ~lfe::felm(., data = model_rais_mun)
-)
-
 # ==============================================================================
 print("write-out data")
 # ==============================================================================
 fit_turnover %>%
   write_model("fit_logit_turnover_coalition.rds")
-
-fit_felm %>%
-  write_model("fit_felm_turnover_coalition.rds")

### 2020-07-02 source/models/modules/visual_model_coalition.R

@@ -41,14 +41,32 @@ mstar(
 )
 sink()
 
-plot_interactions_logit <- map(
+plot_interactions_logit <- map2(
   fit_logit,
+  c("hired", "fired"),
   ~ sjPlot::plot_model(
     .,
     type = "pred",
     terms = c("coalition_share [all]", "rais_category"),
-    title = NULL
+    title = ""
+  ) +
+  labs(
+    x = "share of legislative seats",
+    y = "percentage"
   )
+)
+
+n <- 100
+predict_logit <- tibble(
+  coalition_share = seq(0, 1, length.out = n * 2)
+  rais_category = 
+)
+
+fit_logit[[1]]$model %>% ggplot(aes(coalition_share, rais_hired)) + geom_smooth(method = "glm", method.args = list(family = "binomial")) + facet_wrap(. ~ rais_category)
+
+gridExtra::grid.arrange(
+  grobs = plot_interactions_logit,
+  nrow = 2
 )
 
 save_fig(

### 2020-07-02 source/models/source_models.R

@@ -15,7 +15,8 @@ model_modules <- c(
   "model_turnover_saeb",
   "model_turnover_spaece",
   "model_turnover_ivreg",
-  "model_coalition_rais",
+  "model_coalition_rais_logit",
+  "model_coalition_rais_felm",
   "model_coalition_turnover",
   "model_accountability",
   "model_patronage_reelection"

### 2020-07-03 source/models/modules/model_coalition_rais_felm.R
incorporate changes to visualization

@@ -9,15 +9,21 @@ rais_edu_mun <- read_data(
 # ==============================================================================
 print("model specification")
 # ==============================================================================
+rais_covariates <- c("rais_higher_edu", "rais_wage", "rais_permanent")
+municipal_covariates <- c("cod_ibge_6", mun_covariates)
+
 controls <- c(
-  mun_covariates, rais_covariates, mayor_covariates, chamber_covariates
+  municipal_covariates, rais_covariates, mayor_covariates, chamber_covariates
 )
 
-formulae_felm <- map(
-  c("rais_hired", "rais_fired"),
-  ~formulate(., "coalition_share*rais_category", fe = NULL, controls) %>%
-    update(. ~ . - rais_edu + rais_higher_edu)
-) %>%
+formulae_no_fe <- formulate_models(
+  "rais_hired",
+  "coalition_share*rais_category",
+  controls,
+  fe = NULL
+)
+
+formulae_felm <- formulae_no_fe %>%
   map(
     ~add_felm(., fe = "state + year", cluster = "cod_ibge_6")
   )
@@ -35,7 +41,8 @@ model_rais_mun <- rais_edu_mun %>%
   ) %>%
   model.frame(
     formula = update(
-      f_logit, . ~ . + cod_ibge_6 - rais_edu + rais_higher_edu + rais_fired
+      formulae_no_fe[["controls"]],
+      . ~ . + state + year
     ),
     data = .
   ) %>%

### 2020-07-03 source/models/modules/model_coalition_rais_logit.R

@@ -13,22 +13,21 @@ controls <- c(
   mun_covariates, rais_covariates, mayor_covariates, chamber_covariates
 )
 
-f_logit <- formulate(
+formulae_logit <- formulate_models(
   "rais_hired",
   "coalition_share*rais_category",
   controls,
   fe = c("state", "year")
-)
+) %>%
+  set_names(
+    c("baseline", "controls")
+  )
 
-formulae_logit <- c(
-  f_logit,
-  update(f_logit, rais_fired ~ .)
-)
 # ==============================================================================
 print("reshape data for estimation")
 # ==============================================================================
 model_rais_micro <- rais_edu %>%
-  sample_frac(1/3) %>%
+  sample_frac(1/5) %>%
   join_covariate() %>%
   mutate(
     state = str_sub(cod_ibge_6, 1, 2),
@@ -37,7 +36,7 @@ model_rais_micro <- rais_edu %>%
     across(c(state, year), as.factor)
     ) %>%
   model.frame(
-    formula = update(f_logit, . ~ . + rais_fired),
+    formula = formulae_logit[["controls"]],
     data = .
   ) %>%
   fix_scale()

### 2020-07-03 source/models/modules/visual_model_coalition.R

@@ -11,40 +11,40 @@ fit_felm <- read_model(
 
 fit_turnover <- read_model(
   "fit_coalition_turnover.rds"
-) %>%
-pluck(2) # extract model with controls
+) # extract model with controls
 
 # turn into debugging
 fit_logit <- fit_logit %>%
   map(
     . %>%
-      modify_at(c("model"), ~sample_frac(., 0.01))
+      modify_at(c("model"), ~ sample_frac(., 0.01))
   )
 
 models_table <- list(
-    fit_turnover,
-    fit_logit,
-    fit_felm
-  )
+  fit_turnover,
+  fit_felm,
+  fit_logit
+)
 
 sink(here("replication", "results", "turnover_result.tex"))
 mstar(
-  type = "text",
   models_table,
-  keep = c('coalition_share', 'rais_category'),
-  add.lines = list(c("Controls", rep("\\checkmark", 5))),
-  covariate.labels = c('Executive share of seats', 'School principal', 'Executive share of seats X School principal'),
+  keep = c("coalition_share", "rais_category"),
+  add.lines = list(
+    c("Controls", rep(c("", "\\checkmark"), 3)),
+    c("State and year FE", rep("\\checkmarck", 6))
+  ),
+  dep.var.caption = "Outcome",
+  covariate.labels = c("Share of legislative seats", "School principal", "Executive share of seats X School principal"),
   dep.var.labels.include = F,
-  column.labels = c('Turnover index', 'Hired (Logit)', 'Fired (Logit)', 'Hired (FE)', 'Fired (FE)'),
-  column.separate = rep(1, 5),
+  column.labels = c("Turnover index (municipal)", "Hires (municipal)", "Hires (individual)"),
+  column.separate = rep(2, 3),
   model.names = FALSE
 )
 sink()
 
-plot_interactions_logit <- map2(
-  fit_logit,
-  c("hired", "fired"),
-  ~ sjPlot::plot_model(
+plot_interactions_logit <- fit_logit[["controls"]] %>%
+  sjPlot::plot_model(
     .,
     type = "pred",
     terms = c("coalition_share [all]", "rais_category"),
@@ -52,22 +52,10 @@ plot_interactions_logit <- map2(
   ) +
   labs(
     x = "share of legislative seats",
-    y = "percentage"
-  )
-)
-
-n <- 100
-predict_logit <- tibble(
-  coalition_share = seq(0, 1, length.out = n * 2)
-  rais_category = 
-)
-
[…]

### 2020-07-03 source/models/utils.R

@@ -574,11 +574,11 @@ update_formula_iv <- function(formula, endogenous, instrument) {
 
 formulate_models <- function(response, predictor, fe, controls) {
   formulae <- c(
-    formulate(
+    baseline = formulate(
       response, predictor,
       controls = NULL, fe
     ),
-    formulate(
+    controls = formulate(
       response, predictor, controls, fe
     )
   )

### 2020-07-06 presentation/presentation_job_market.Rmd
improve file naming

@@
 
-path_figs <- "papers/paper_cp_2020/figs/"
+path_figs <- "papers/paper_cp_2020/plots/"
 
@@
 include_graphics(
-  'figs/plot_global_1.pdf'
+  'plots/plot_global_1.pdf'
 )
@@
 include_graphics(
-  'figs/plot_global_2.pdf'
+  'plots/plot_global_2.pdf'
 )
@@
 include_graphics(
-  "figs/saeb_map.pdf"
+  "plots/saeb_map.pdf"
 )
@@
 include_graphics(
-  "figs/enrollment_dep.pdf"
+  "plots/enrollment_dep.pdf"
 )
@@
 include_graphics(
-  'figs/turnover_category.pdf'
+  'plots/turnover_category.pdf'
 )
@@
 include_graphics(
-  "figs/turnover_fit.pdf"
+  "plots/turnover_fit.pdf"
 )
@@
 include_graphics(
-  "figs/plot_pred.pdf"
+  "plots/plot_pred.pdf"
 )

### 2020-07-06 presentation/presentation_job_market.tex
improve file naming

@@
 
-\begin{center}\includegraphics[width=240px]{figs/plot_global_1} \end{center}
+\begin{center}\includegraphics[width=240px]{plots/plot_global_1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=240px]{figs/plot_global_2} \end{center}
+\begin{center}\includegraphics[width=240px]{plots/plot_global_2} \end{center}
 
@@
 
-{\centering \includegraphics[width=180px]{figs/saeb_map}
+{\centering \includegraphics[width=180px]{plots/saeb_map}
 
@@
 
-{\centering \includegraphics[width=230px]{figs/enrollment_dep}
+{\centering \includegraphics[width=230px]{plots/enrollment_dep}
 
@@
 
-{\centering \includegraphics[width=230px]{figs/turnover_category}
+{\centering \includegraphics[width=230px]{plots/turnover_category}
 
@@
 
-{\centering \includegraphics[width=210px]{figs/turnover_fit}
+{\centering \includegraphics[width=210px]{plots/turnover_fit}
 
@@
 
-{\centering \includegraphics[width=230px]{figs/plot_pred}
+{\centering \includegraphics[width=230px]{plots/plot_pred}
 

### 2020-07-06 publication/paper_job_market.Rmd
improve file naming

@@
 include_graphics(
-  "figs/plot_global.pdf"
+  "plots/plot_global.pdf"
 )
@@
 include_graphics(
-  "figs/saeb_map.pdf"
+  "plots/saeb_map.pdf"
 ) 
@@
 include_graphics(
-  "figs/hist_coalition.pdf"
+  "plots/hist_coalition.pdf"
 )
@@
     \centering
-    \includegraphics{figs/plot_pred}
+    \includegraphics{plots/plot_pred}
     \caption{Predicted values for bureaucratic turnover.
@@
 include_graphics(
-  "figs/balance_hlm.pdf"
+  "plots/balance_hlm.pdf"
 )
@@
 include_graphics(
-  "figs/balance_felm.pdf"
+  "plots/balance_felm.pdf"
 )

### 2020-07-06 source/descriptive/utils.R

@@ -63,7 +63,7 @@ ls_data <- function(env = .GlobalEnv) {
 
 read_model <- function(file){
   fit <- read_rds(
-    here("replication", "models", file)
+    here("figures", "models", file)
   )
 
   print_obj_size(fit)
@@ -76,7 +76,7 @@ write_model <- function(model, file) {
   
   write_rds(
     model,
-    here("replication", "models", file)
+    here("figures", "models", file)
   )
 }
 
@@ -115,7 +115,7 @@ reset_env <- function(init_env, packages){
 
 save_fig <- function(pl, file, width = 5, height = 3, ...){
   ggplot2::ggsave(
-    filename = here("replication", "figs", file),
+    filename = here("figures", "figs", file),
     pl,
     width = width,
     height = height,

### 2020-07-06 source/models/modules/visual_model_coalition.R

@@ -26,7 +26,7 @@ models_table <- list(
   fit_logit
 )
 
-sink(here("replication", "results", "turnover_result.tex"))
+sink(here("figures", "results", "turnover_result.tex"))
 mstar(
   models_table,
   keep = c("coalition_share", "rais_category"),

### 2020-07-06 source/models/modules/visual_model_turnover.R

@@ -10,7 +10,7 @@ fit_ivreg <- map(
   read_model
 )
 
-sink(here("replication", "tables", "student_learning.tex"))
+sink(here("figures", "tables", "student_learning.tex"))
 mstar(
   fit_lmer, fit_spaece, fit_spaece, fit_ivreg,
   keep = c("turnover_index", "saeb_principal_experience", "saeb_teacher_work_school"),

### 2020-07-06 source/models/setup.R

@@ -41,7 +41,7 @@ read_data <- partial(
 
 save_fig <- partial(
   save_fig,
-  domain = "replication"
+  domain = "figures"
 )
 
 theme_set(

### 2020-07-06 source/setup.R

@@ -1,4 +1,4 @@
 # data cleaning
-# replication (estimation, visualization)
+# figures (estimation, visualization)
 # publication (relatorio)
 # presentation

### 2020-07-06 source/descriptive/modules/visual_staff_turnover.R
change domain names

@@ -116,7 +116,7 @@ save_fig(
 #     ) +
 #     ggsave(
 #         p_file_here(
-#             "figs", "hire_by_category"
+#             "plots", "hire_by_category"
 #         )
 #     )
 

### 2020-07-06 source/descriptive/scripts_old/replication.R

@@ -360,7 +360,7 @@ patronage_reelection <- fread(
 #   limits = c(100, 250)
 # ) +
 #   ggplot2::ggsave(
-#     p_file_here('figs', "saeb_map.pdf"),
+#     p_file_here('plots', "saeb_map.pdf"),
 #     width = 5,
 #     height = 5
 #   )
@@ -438,7 +438,7 @@ plot_global <- cowplot::plot_grid(
 
 ggsave(
   plot_global,
-  filename = p_file_here('figs', 'plot_global.pdf')
+  filename = p_file_here('plots', 'plot_global.pdf')
 )
 
 plot_enroll <- edu_global %>%
@@ -514,7 +514,7 @@ plot_edu_1 <- cowplot::plot_grid(
 
 ggsave(
   plot_edu_1,
-  filename = p_file_here('figs', 'plot_global_1.pdf')
+  filename = p_file_here('plots', 'plot_global_1.pdf')
 )
 
 plot_enroll_type <- edu_global %>%
@@ -609,7 +609,7 @@ plot_edu_2 <- cowplot::plot_grid(
 
 ggsave(
   plot_edu_2,
-  filename = p_file_here('figs', 'plot_global_2.pdf')
+  filename = p_file_here('plots', 'plot_global_2.pdf')
 )
 
 # descriptive statistics --------------------------------------------------
@@ -642,7 +642,7 @@ plot_budget <- finbra %>%
 
 plot_budget %>% 
   ggsave(
-    filename = p_file_here('figs', "budget.pdf")
+    filename = p_file_here('plots', "budget.pdf")
   )
 
 # student coverage by administratioon
@@ -690,7 +690,7 @@ plot_dep <- censo_class %>%
   )
 
 ggsave(
-  p_file_here('figs', 'enrollment_dep.pdf')
+  p_file_here('plots', 'enrollment_dep.pdf')
 )
 
 # rais by category
@@ -754,7 +754,7 @@ plot_cat_total <- rais_category %>%
   ) +
   theme(legend.title = element_blank()) +
   ggsave(
-    filename = p_file_here('figs', 'rais_cat_total.pdf')
+    filename = p_file_here('plots', 'rais_cat_total.pdf')
   )
 
 plot_cat_turnover <- rais_category %>% 
@@ -790,7 +790,7 @@ plot_cat_turnover <- rais_category %>%
     legend.title = element_blank()
   ) +
   ggsave(
-    filename = p_file_here('figs', "turnover_cat.pdf")
[…]

### 2020-07-06 source/descriptive/scripts_old/thesis_models.R

@@ -206,5 +206,5 @@ boot_lm <- boot(
 
 # ggsave(
 #   plot_hlm,
-#   filename = here(path_figs, "turnover_fit.pdf")
+#   filename = here(path_plots, "turnover_fit.pdf")
 # )

### 2020-07-06 source/descriptive/utils.R

@@ -115,7 +115,7 @@ reset_env <- function(init_env, packages){
 
 save_fig <- function(pl, file, width = 5, height = 3, ...){
   ggplot2::ggsave(
-    filename = here("figures", "figs", file),
+    filename = here("figures", "plots", file),
     pl,
     width = width,
     height = height,

### 2020-07-06 source/models/modules/model_accountability.R

@@ -139,5 +139,5 @@ plot_accountability <- ggplot(
 
 ggsave(
   plot_accountability,
-  filename = p_file_here('figs', "accountability_fit.pdf")
+  filename = p_file_here('plots', "accountability_fit.pdf")
 )

### 2020-07-06 source/models/modules/visual_model_coalition.R

@@ -84,7 +84,7 @@ save_fig(
 #   names(plot_logit_int),
 #   ~ ggsave(
 #     .x,
-#     filename = p_file_here("figs", paste0("plot_logit_", .y, ".pdf"))
+#     filename = p_file_here("plots", paste0("plot_logit_", .y, ".pdf"))
 #   )
 # )
 

### 2020-07-06 source/models/modules/visual_regression_discontinuity.R

@@ -101,7 +101,7 @@ rdd_plots <- pmap(
 
 
         ggsave(
-            here("models", "figs", paste0("visual_rdd_", names, ".pdf"))
+            here("models", "plots", paste0("visual_rdd_", names, ".pdf"))
         )
     }
 )

### 2020-07-06 source/models/utils.R

@@ -136,7 +136,7 @@ save_fig <- function(pl, domain, file, width = 5, height = 3, ...) {
   print(pl)
 
   ggplot2::ggsave(
-    filename = here(domain, "figs", file),
+    filename = here(domain, "plots", file),
     pl,
     width = width,
     height = height,

### 2020-07-22 theory/notes_banks_buyingfinitemajorities.tex
begin redacting banks 2000 model

\documentclass[12pt,a4paper]{article}
\usepackage{setspace} \onehalfspacing
\usepackage[top = 1in, bottom = 1in, left = 1in, right = 1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}

\newcommand{\real}{\mathbb{R}_+^n}

\begin{document}

\section{The model}

Two states of the world: policies $x$ and $y$. Party $A$ and $B$ prefer $x$ and $y$ respectively. We can make this more explicit with
\begin{align*}
    U_A(x) > U_A(y)\\
    U_B(y) > U_B(x)
\end{align*}
For voters, $v_i > 0$ means voter $i$ prefers $x$ to $y$ All $i \in N$ vote for $x$ or $y$. Simple majority rule determines which policy gets implemented. For each voter, each party sets a bribe schedule
\begin{align*}
    a \in (a_1, ..., a_n) \in \mathbb{R}^n_+\\
    b \in (b_1, ..., b_n) \in \mathbb{R}^n_+
\end{align*}
Solving through backward induction, given bribe schedules $(a,b)$, voter $i$ prefers to vote for $x$ if $a_i + v_i > b_i$ and for $y$ otherwise. Since indifferent voters choose $y$, party $B$ needs to only match bribes from $A$, adjusting for individual voters' preferences: $b_i = a_i + v_i$. Therefore, $B$ solves

$$\min_C \left\{\sum_{i \in C} \max\{0, a_i + v_i\} : |C| > \frac{n}{2} \right\}$$

As long as this sum is strictly less than $W_B$; otherwise party $B$ chooses to set $b_i = 0 : \forall i \in N$.

Following Banks (2000), we restrict our analysis to the set of equilibria in which party $A$ wins, i.e. $W_A$ is sufficiently large relative to $\mathbf{v}$ and $W_B$ so that policy $x$ prevails over $y$. In other words, the folloing inequality must hold:

$$ \sum_{i \in C} \max{0, a_i + v_i} \geq W_B$$

Let $U(v, W_b) \subseteq \mathbb{R}^n_+$ denote the set of unbeatable bribe schedules. Additionally, let $S(a) = \sum^n_i a_i$ denote the bribe schedule for party $A$. The above assumptions on $W_A$, $W_B$ and $v$ guarantee that there is an

$$\tilde{a} \in U(\mathbf{v}, W_B) : S(\tilde{a}) \leq W_A$$

For party $A$, the solution is

\begin{equation}
\label{eqn:solution}
min\{S(a) : a \in U(\mathbf{v}, W_B) \}
\end{equation}

To fully describe the solution to equation \ref{eqn:solution}, we note the following: for any $a \in \mathbb{R}_+^n$, let $C(a) : {i \in N : a_i > 0}$ denote the set of individuals who receive a bribe from $A$. One can show that there is a bribe schedule $a'$ such that for any $i,j \in C(a)$, $a'_i + v_i = a'_j + v_j$. The intuition is that $A$ has no incentive to make voters differentially bribed, because $B$ will simply ignore the more expensive voters and target the weakest rings in the chain. Following Groseclose and Snyder (1996) we refer to this as a leveling schedule.

Let $U^l(\mathbf{v}, W_B) \subseteq U(\mathbf{v}, W_B)$ denote the set of unbeatable leveling schedules. These are bribe schedules such that $a_i + v_i = a_j + v_j  \equiv t(a)$. The bribe $a_i = t(a) - v_i$ is the sum of two terms. The first is the common "transfer" among all voters in $C(a)$, the second ($-v_i$) is individual specific. The latter term makes voters indifferent between $x$ and $y$ absent any bribe from $B$; the former represents the per capita amount necessary to make $C(a)$, together with any unbribed voters, unaffordable for $B$.

To further simplify the analysis, Banks introduces the folowing sets of assumption:

\begin{align*}
    A_1 &: v_{(n+1)/2} < 0\\
    A_2 &: v_1 < 2W_B/(n+1) 
\end{align*}

$A_1$ implies that absent any bribes by $A$, $y$ will defeat $x$. Therefore $A$ must bribe at least one voter. $A_2$ further implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win.

Banks then proceeds to show that there are monotonic bribing schedules contained within the solution for equation \ref{eqn:solution}. For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$. Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$. Then, under $A_2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$. Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e. $a'_i \leq a_j$. $A_2$ guarantees that $a'_i \geq 0$, in other words every $i \in C(a)$ is receiving a non-negative bribe.

Generalizing, and recalling that $v_1 \geq ... \geq v_n$, we see that for all $a \in U^l(\mathbf{v}, W_b)$ there exists a bribe schedule $a' \in U^l(\mathbf{v}, W_b)$ such that $S(a') \leq S(a)$ and $C(a') = \{1, ..., k(a)\}$. Therefore, we can without loss of generality restrict attention to schedules $a$ by $A$ which bribe the first $k(a)$ voters. Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.

Therefore, when $A_2$ holds,

$$\min\{S(a): a \in U(\mathbf{v}, W_B)\} = \min\{S(a): a \in U^l_m(\mathbf{v}, W_B)\}$$

\end{document}

### 2020-07-23 theory/notes_banks_buyingfinitemajorities.tex
incorporate minimization problem to theoretical section

@@
 Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$.
-Then, under $A_2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$.
-Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
+Then, under $A_2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
 $a'_i \leq a_j$.
-$A_2$ guarantees that $a'_i \geq 0$, in other words every $i \in C(a)$ is receiving a non-negative bribe.
+$A_2$ guarantees that $a'_i$ or $a_j$ are non-negative.}
 
@@
 
+We can now further simplify the total expenditure $S(a)$
+
+\begin{equation*}
+    S(a) = \sum_{i \in C(a)}a_i = k(a) \cdot t(a) - \sum_{i \leq k(a)}v_i
+\end{equation*}
+
+Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B$. We can thus fully characterize the optimization problem of $A$ as
+
+\begin{equation*}
+    \min_{k,t} k \cdot t- \sum_{i \leq k} v_i
+\end{equation*}
+
+subject to the constraint that the induced schedule $a \in U_m^l$.
+Banks then reformulates this as an unconstrained problem by noting the following.
+First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A_1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$.
+Therefore, $B$ receives all non-bribed voters for free.
+For $a(k, t, \mathbf{v})$ to be unbeatable, then, it must be that $B$ cannot afford the remaining $(n + 1)/2 - (n - k)$
+
 \end{document}

### 2020-07-27 theory/notes_banks_buyingfinitemajorities.tex
incorporate additional notes on banks

@@
 
-$$ \sum_{i \in C} \max{0, a_i + v_i} \geq W_B$$
+$$ \sum_{i \in C} \max\{0, a_i + v_i\} \geq W_B$$
 
@@
 \label{eqn:solution}
-min\{S(a) : a \in U(\mathbf{v}, W_B) \}
+\min_a\{S(a) : a \in U(\mathbf{v}, W_B) \}
 \end{equation}
@@
 $a'_i \leq a_j$.
-$A_2$ guarantees that $a'_i$ or $a_j$ are non-negative.}
+$A_2$ guarantees that $a'_i$ and $a_j$ are non-negative.}
 
@@
 
-Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B$. We can thus fully characterize the optimization problem of $A$ as
+Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B)$. We can thus fully characterize the optimization problem of $A$ as
 
@@
 Therefore, $B$ receives all non-bribed voters for free.
-For $a(k, t, \mathbf{v})$ to be unbeatable, then, it must be that $B$ cannot afford the remaining $(n + 1)/2 - (n - k)$
+For $a(k, t, \mathbf{v})$ to be unbeatable, then, it must be that $B$ cannot afford the remaining $(n + 1)/2 - (n - k) = k - (n - 1)/2$ voters, or
+
+$$t * (k - (n - 1)/2) \geq W_B$$
+
+Solving this for equality yields the optimal transfer from $A$ to members of $C(A) = \{1, ..., k\}$, conditional on k:
+
+\begin{align}
+    \label{eqn:common_transfer}
+    t(k, W_B) = \frac{W_B}{k - (n - 1)/2}
+\end{align}
+
+Defining minimal winning expenditures as
+
+\begin{align}
+    \label{eqn:total_expenditure}
+    E(k, \mathbf{v}, W_B) = k \cdot t(k, W_B) - \sum_{i \leq k} v_i
+\end{align}
+
+we can state $A$'s problem as
+
+\begin{align}
+\label{eqn:minimal_expenditure}
+\min_k \{E(k, \mathbf{v}, W_B) : k \in {(n + 1/2), ..., n}\}
+\end{align}
+
+Denote the solution to expression \ref{eqn:minimal_expenditure} as $k^*(\mathbf{v}, W_B)$. This solution genenerates a solution to expression \ref{eqn:solution}
+
+\section{Results}
+
+First, characterize a solution for $k^*$.
+Because $k$ is finite, calculus cannot be employed.
+Instead, we deploy a discrete version of these techniques.
+First let's define $\Delta(k) = E(k + 1) - E(k)$ as the difference in expenditure from adding another coalition member.
+Note that if $\Delta(k) \geq 0$ then $A$ does not want to add another member to the coalition.
+Conversely, if $\Delta(k) < 0$, then $A$ is strictly better off by adding the $k + 1$th member of the coalition.
+
+Next, suppose that $\Delta(k)$ is increasing in $k$.
+The following algorithm can then be used to identify $k^*$: if $\Delta((n + 1)/2) \geq 0$, then we know from $]\Delta(k)$ increasing that $A$ is better off by setting $k^*$ to $(n + 1)/2$.
+If $\Delta((n + 1)/2) < 0$, then we know that $k^*$ must be greater than $(n + 1)/2$, so we next solve for $\Delta((n + 3)/2)$, and so on.
+
+We can therefore search for the optimal $k^*$ with the following algorithm:
+
+\begin{align}
+    k^* = 
+    \begin{cases}
+        (n + 1)/2 & \text{if } \Delta((n + 1)/2) \geq 0 \\
+        \max\{k : \Delta(k - 1) < 0\} & \text{otherwise}
+    \end{cases}
+\end{align}
+
+Banks also shows that $\Delta(k)$ is indeed nondecreasing.
+
+\begin{align}
+    \Delta(k)  & = \left[\frac{(k + 1)W_B}{k + 1 - (n - 1)/2} - \sum_{i \leq k + 1}v_i \right]\\
+    & = \frac{-W_B (n - 1)}{2(k + 1 - (n - 1)/2)(k - (n - 1)/2)} - v_{k + 1}\\
+    & \equiv T(k, W_B) - v_{k + 1}
+\end{align}
 

### 2020-07-28 theory/notes_banks_buyingfinitemajorities.tex
improve notes on banks

@@
 
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}{Proposition}
+\renewcommand{\theproposition}{\arabic{proposition}}
+\newtheorem{corollary}[theorem]{Corollary}
 \newcommand{\real}{\mathbb{R}_+^n}
@@
 \begin{align}
+    \label{eqn:optimal_k}
     k^* = 
@@
     \Delta(k)  & = \left[\frac{(k + 1)W_B}{k + 1 - (n - 1)/2} - \sum_{i \leq k + 1}v_i \right]\\
+    \label{eqn:delta_k}
     & = \frac{-W_B (n - 1)}{2(k + 1 - (n - 1)/2)(k - (n - 1)/2)} - v_{k + 1}\\
+    \label{eqn:def_delta_k}
     & \equiv T(k, W_B) - v_{k + 1}
@@
 
+Note that because $v_1 \geq ...
+\geq v_n$, $-v_{k + 1}$ is non-decreasing in $k$.
+Also, taking the first term as a continuous variable and differentiating w.r.t.
+$k$, we note that it is also increasing in $k$.
+
+Generating an explicit characterization of $k$ is not straightforward.
+Instead, Banks opts to show the conditions which need to hold in order for the coalition to be universal or minimal.
+Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:delta_k} we have the following.
+
+\begin{proposition}
+    (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
+\end{proposition}
+
+Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity.
+Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v}, W_B)$.
+If $k' = (n + 1)/2$, then we know that $k' \leq k*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.
+
+From equation \ref{eqn:optimal_k} we infer that $\Delta(k - 1, \mathbf{v}', W_B) < 0)$, which from equations \ref{eqn:delta_k} and \ref{eqn:def_delta_k} is equivalent to $v_k' > T(k' - 1, W_B)$.
+Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$.
+Then, $v_{k'} > T(k' - 1, W_B)$, and hence $\Delta(k' - 1, \mathbf{v}, W_B) < 0$.
+But then from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$.
+Therefore, the following holds:
+
+\begin{proposition}
+    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
+\end{proposition}
+
+In words, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases.
+Substantively, the number of voters bribed by $A$ weakly increases as the voter who receives the largest bribe finds $A$'s preferred alternative, $x$, more attractive.
+Similarly
+
+\begin{proposition}
+    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k' + 1} \leq v'_{k' + 1}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \leq k^*(\mathbf{v}', W_B)$
+\end{proposition}
+
 \end{document}

### 2020-07-29 theory/notes_banks_buyingfinitemajorities.tex
incorporate additional notes on banks

@@
 \usepackage{setspace} \onehalfspacing
-\usepackage[top = 1in, bottom = 1in, left = 1in, right = 1in]{geometry}
+\usepackage[top = 1.25in, bottom = 1.25in, left = 1.25in, right = 1.25in]{geometry}
 \usepackage[utf8]{inputenc}
@@
 \newtheorem{proposition}{Proposition}
+\newtheorem{corollary}{Corollary}
 \renewcommand{\theproposition}{\arabic{proposition}}
-\newtheorem{corollary}[theorem]{Corollary}
 \newcommand{\real}{\mathbb{R}_+^n}
@@
 
+The "convexity" of $E$ guarantees that local information is sufficient to generate comparative statistics regarding changes in preferences $\mathbf{v}' \rightarrow \mathbf{v}$.
+Here, this local information is summarized by the preferences of the marginal voter $v_k$ and non-bribed voter $v_{k + 1}$.
+In order to identify these voters, one needs to solve for the optimal coalition size.
+Instead, Banks takes an alternative route and characterizes a weaker comparative statistic result which holds globally.
+Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v} \geq \mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$.
+
+\begin{corollary}
+    \label{cor:change_k}
+    For all $\mathrm{W_B}$, if $\mathbf{v} \geq \mathbf{v}'$ then $\mathrm{k}^*(\mathbf{v}, \mathrm{W_B}) \geq \mathrm{k}^*(v', \mathrm{W_B})$
+\end{corollary}
+
+Thus, the optimal number of voters bribed by $A$ and hence the optimal size of $A$'s optimal coalition, weakly increases as voters find $A$'s preferred alternative, $x$, more attractive.
+Although $A$'s total expenditure will decrease as $x$ becomes more attractive to the legislature, the optimal way to allocate this lower amount is to spread it more widely among voters.
+
+We can also characterize the change in total expenditures as a result of a shift in voter preferences. From equation \ref{eqn:total_expenditure} we have
+
+\begin{align*}
+    E(k, \mathbf{v}, W_B) - E(k, \mathbf{v}', W_B) & = \\
+    &= k \cdot t(k, W_B) - \sum_{i \leq k} v_i - \left[k \cdot t(k, W_B) - \sum_{i \leq k} v'_i\right]\\
+    &= \sum_{i \leq k}(v'_i - v_i)
+\end{align*}
+
+Since $v_i - v'_i \geq 0$, the difference in expenditure between a more favorable and unfavorable legislature is always non-positive.
+
 \end{document}

### 2020-07-30 theory/notes_banks_buyingfinitemajorities.tex
finish redacting notes for banks

@@
 \end{align*}
-For voters, $v_i > 0$ means voter $i$ prefers $x$ to $y$ All $i \in N$ vote for $x$ or $y$.
+For voters, $v_i > 0$ means voter $i$ prefers $x$ to $y$.
+All $i \in N$ vote for $x$ or $y$.
 Simple majority rule determines which policy gets implemented.
@@
 \begin{align*}
-    a \in (a_1, ..., a_n) \in \mathbb{R}^n_+\\
+    a \in (A1, ..., An) \in \mathbb{R}^n_+\\
     b \in (b_1, ..., b_n) \in \mathbb{R}^n_+
@@
 
-To fully describe the solution to equation \ref{eqn:solution}, we note the following: for any $a \in \mathbb{R}_+^n$, let $C(a) : {i \in N : a_i > 0}$ denote the set of individuals who receive a bribe from $A$.
+To fully describe the solution to equation \ref{eqn:solution}, we note the following: for any $a \in \mathbb{R}_+^n$, let $C(a) = \{i \in N : a_i > 0\}$ denote the set of individuals who receive a bribe from $A$.
 One can show that there is a bribe schedule $a'$ such that for any $i,j \in C(a)$, $a'_i + v_i = a'_j + v_j$.
@@
 Let $U^l(\mathbf{v}, W_B) \subseteq U(\mathbf{v}, W_B)$ denote the set of unbeatable leveling schedules.
-These are bribe schedules such that $a_i + v_i = a_j + v_j  \equiv t(a)$.
+These are bribe schedules such that $a_i + v_i = Aj + v_j  \equiv t(a)$.
 The bribe $a_i = t(a) - v_i$ is the sum of two terms.
@@
 \begin{align*}
-    A_1 &: v_{(n+1)/2} < 0\\
-    A_2 &: v_1 < 2W_B/(n+1) 
+    A1 &: v_{(n+1)/2} < 0\\
+    A2 &: v_1 < 2W_B/(n+1) 
 \end{align*}
 
-$A_1$ implies that absent any bribes by $A$, $y$ will defeat $x$.
+$A1$ implies that absent any bribes by $A$, $y$ will defeat $x$.
 Therefore $A$ must bribe at least one voter.
-$A_2$ further implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win.
+$A2$ implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win.
+$A2$ also implies that for all $a \in U^l(\mathbf{v}, W_B)$, it must be that $t(a) \geq 2W_B/(n + 1)$, otherwise $B$ can bribe a majority from $C(a)$ itself and win the vote.
 
@@
 Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$.
-Then, under $A_2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
+Then, under $A2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$: $a'_i = t(a) - v_i$, $a'_j = 0$ and $a'_m = a_m$ for all $m \notin i, j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
 $a'_i \leq a_j$.
-$A_2$ guarantees that $a'_i$ and $a_j$ are non-negative.}
+$A2$ guarantees that $a'_i$ and $a_j$ are non-negative.}
 
@@
 
-Therefore, when $A_2$ holds,
+Therefore, when $A2$ holds,
 
@@
 Banks then reformulates this as an unconstrained problem by noting the following.
-First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A_1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$.
+First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$.
 Therefore, $B$ receives all non-bribed voters for free.
@@
 
-$$t * (k - (n - 1)/2) \geq W_B$$
+$$t \cdot (k - (n - 1)/2) \geq W_B$$
 
@@
 
-Denote the solution to expression \ref{eqn:minimal_expenditure} as $k^*(\mathbf{v}, W_B)$. This solution genenerates a solution to expression \ref{eqn:solution}
+Denote the solution to expression \ref{eqn:minimal_expenditure} as $k^*(\mathbf{v}, W_B)$.
+This solution implicitly generates a solution to expression \ref{eqn:solution} through expression \ref{eqn:total_expenditure} and the induced bribe schedule above.
 
@@
 Next, suppose that $\Delta(k)$ is increasing in $k$.
-The following algorithm can then be used to identify $k^*$: if $\Delta((n + 1)/2) \geq 0$, then we know from $]\Delta(k)$ increasing that $A$ is better off by setting $k^*$ to $(n + 1)/2$.
+The following algorithm can then be used to identify $k^*$: if $\Delta((n + 1)/2) \geq 0$, then we know from $\Delta(k)$ increasing that $A$ is better off by setting $k^*$ to $(n + 1)/2$.
 If $\Delta((n + 1)/2) < 0$, then we know that $k^*$ must be greater than $(n + 1)/2$, so we next solve for $\Delta((n + 3)/2)$, and so on.
@@
     \Delta(k)  & = \left[\frac{(k + 1)W_B}{k + 1 - (n - 1)/2} - \sum_{i \leq k + 1}v_i \right]\\
-    \label{eqn:delta_k}
+    \label{eqn:deltAk}
     & = \frac{-W_B (n - 1)}{2(k + 1 - (n - 1)/2)(k - (n - 1)/2)} - v_{k + 1}\\
-    \label{eqn:def_delta_k}
+    \label{eqn:def_deltAk}
     & \equiv T(k, W_B) - v_{k + 1}
@@
 Instead, Banks opts to show the conditions which need to hold in order for the coalition to be universal or minimal.
-Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:delta_k} we have the following.
+Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.
 
@@
 
-From equation \ref{eqn:optimal_k} we infer that $\Delta(k - 1, \mathbf{v}', W_B) < 0)$, which from equations \ref{eqn:delta_k} and \ref{eqn:def_delta_k} is equivalent to $v_k' > T(k' - 1, W_B)$.
+From equation \ref{eqn:optimal_k} we infer that $\Delta(k - 1, \mathbf{v}', W_B) < 0)$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$.
 Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$.

### 2020-07-31 theory/model_executive_bargain.tex
start redacting theory section

\documentclass[12pt,a4paper]{article}
\usepackage{setspace} \onehalfspacing
\usepackage[top = 1.25in, bottom = 1.25in, left = 1.25in, right = 1.25in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\renewcommand{\theproposition}{\arabic{proposition}}
\newcommand{\real}{\mathbb{R}_+^n}

\begin{document}

The following model is adapted, with little to no modifications, from Banks (2000). We can model the executive bargain with the legislature as follows. A mayor wants to pass her proposal $x$. We can include under this category a set of policy proposals: a new legislation, an infrastructure project or budgetary approval. In order to implement her proposal, she must pass it in the city council through a simple majority rule. However, the mayor does not govern alone. An opposition wants to block the mayor's proposal and maintain the status quo, denoted as $y$.

Each city councilor is characterized by a proposal valuation parameter $v_i$ for all $i \in N$, where $v_i > 0$ means that the mayor's proposal $x$ is preferred. Payoff gets realized when city councilor $i$ votes, independent of the outcome. For each councilor, the mayor and opposition set a bribe schedule

\begin{align*}
    a \in (a_1, ..., a_n) \in \real\\
    b \in (b_1, ..., b_n) \in \real+
\end{align*}

Solving through backward induction, given a bribe schedule $(a, b)$, councilor $i$ prefers to vote for the mayor's proposal ($x$) if $a_i + v_i > b_i$ and to maintain the status quo ($y$) otherwise. Since indifferent councilors vote for the status quo, party $b$ needs to only match bribes from $A$, adjusting for individual preferences, i.e. $b_i = a_i + v_i$. Therefore, the opposition solves

\end{document}

### 2020-08-12 theory/model_executive_bargain.tex
reshape paper

@@
 
-The following model is adapted, with little to no modifications, from Banks (2000).
-We can model the executive bargain with the legislature as follows.
-A mayor wants to pass her proposal $x$.
-We can include under this category a set of policy proposals: a new legislation, an infrastructure project or budgetary approval.
-In order to implement her proposal, she must pass it in the city council through a simple majority rule.
-However, the mayor does not govern alone.
-An opposition wants to block the mayor's proposal and maintain the status quo, denoted as $y$.
+\section*{Theory}
 
-Each city councilor is characterized by a proposal valuation parameter $v_i$ for all $i \in N$, where $v_i > 0$ means that the mayor's proposal $x$ is preferred.
-Payoff gets realized when city councilor $i$ votes, independent of the outcome.
-For each councilor, the mayor and opposition set a bribe schedule
+\begin{itemize}
+    \item Patronage is a political currency to buy legislative votes. 
+    \item Mayor has to compete with an opposition that wants to block her proposals.
+    \item Variation in the amount of support by the city councilors.
+    \item City councils that are more favorable to the mayor are cheaper to buy.
+    \item This is independent of which term the mayor is in. Mechanism sld still operate in the second term. That is exactly what we find.
+\end{itemize}
 
-\begin{align*}
-    a \in (a_1, ..., a_n) \in \real\\
-    b \in (b_1, ..., b_n) \in \real+
-\end{align*}
+In this section, I use the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to analyze how mayors engage in patronage in Brazil.
+For any given term, a mayor wants to enact her preferred policy, which requires winning over potential resistance by the city council.
+The preferences of these legislators may vary: in some municipalities a mayor enjoys widespread political support ("base política"), while in others, the opposition ("oposição") garners more seats.
+In this battle over votes, patronage is the currency used to buy legislative votes.
+
+I show that, in equilibrium, the cost of passing the mayor's preferred policy decreases monotonically as the legislative support for the mayor increases.
+Substantively, the model predicts that municipalities in which the mayor has more (less) councilors favorable to her, we should observe less (more) patronage.
+Below, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
+
+\subsection*{The Setting}
+
+A mayor $M$ wants to pass her preferred policy proposal $x$.
+The definition of said proposal is flexible: it can include a new law, a development project or approval for that fiscal year's budget.
+In order to implement her policy she must gain the approval of the city council, comprised of an odd $N$ number of voters, through a simple majority rule.
+Blocking her efforts, an opposition $O$ wants to maintain the status quo, denoted as $y$.
+
+Each city councilor is characterized by a policy preference parameter $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
+Note that there are no sharp boundaries between pro-mayor and opposition legislators.
+Payoff gets realized when city councilor $i$ votes, independent of the outcome of the voting procedure.
+For each councilor, the mayor and opposition set a bribe schedule $a \in (a_1, ..., a_n) \in \real$, $b \in (b_1, ..., b_n) \in \real+$.
 

### 2020-08-12 theory/model_executive_bargain.tex
continue building the theoryu section

@@
 \newcommand{\real}{\mathbb{R}_+^n}
+\newcommand{\bfv}{\mathbf{v}}
 
@@
 
-\begin{itemize}
-    \item Patronage is a political currency to buy legislative votes. 
-    \item Mayor has to compete with an opposition that wants to block her proposals.
-    \item Variation in the amount of support by the city councilors.
-    \item City councils that are more favorable to the mayor are cheaper to buy.
-    \item This is independent of which term the mayor is in. Mechanism sld still operate in the second term. That is exactly what we find.
-\end{itemize}
-
-In this section, I use the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to analyze how mayors engage in patronage in Brazil.
+In this section, I use the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to model how mayors engage in patronage in Brazil.
 For any given term, a mayor wants to enact her preferred policy, which requires winning over potential resistance by the city council.
@@
 
-A mayor $M$ wants to pass her preferred policy proposal $x$.
-The definition of said proposal is flexible: it can include a new law, a development project or approval for that fiscal year's budget.
-In order to implement her policy she must gain the approval of the city council, comprised of an odd $N$ number of voters, through a simple majority rule.
-Blocking her efforts, an opposition $O$ wants to maintain the status quo, denoted as $y$.
+The mayor $M$ and opposition $O$ compete over legislative votes to enact their preferred policies.
+There are two possible outcomes: a policy $x$ favored by the mayor, and the status quo, denote as $y$, preferred by the opposition.
+In order to implement her policy the mayor must gain the approval of the city council, comprised of an odd $N$ number of voters, through a simple majority rule.
+The mayor and opposition spend political resources $W_M$ and $W_O$ to win over votes, which for the mayor includes patronage appointments into the public sector.
 
-Each city councilor is characterized by a policy preference parameter $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
-Note that there are no sharp boundaries between pro-mayor and opposition legislators.
+Each city councilor is characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
+Let $\mathbf{v} = (v_1, ..., v_n)$ denote a preference profile for the city council.
+Let $v_i$ measure the degree to which an individual city councilor supports the mayor, with higher values denoting stronger support for the mayor and vice versa.
 Payoff gets realized when city councilor $i$ votes, independent of the outcome of the voting procedure.
-For each councilor, the mayor and opposition set a bribe schedule $a \in (a_1, ..., a_n) \in \real$, $b \in (b_1, ..., b_n) \in \real+$.
 
-Solving through backward induction, given a bribe schedule $(a, b)$, councilor $i$ prefers to vote for the mayor's proposal ($x$) if $a_i + v_i > b_i$ and to maintain the status quo ($y$) otherwise.
-Since indifferent councilors vote for the status quo, party $b$ needs to only match bribes from $A$, adjusting for individual preferences, i.e.
-$b_i = a_i + v_i$.
-Therefore, the opposition solves
+We solve the game through backward induction. The timing of the game is as follows:
+
+\begin{enumerate}
+    \item Mayor $M$ offers a bribe schedule $m \in (m_1, ..., m_n) \in \real$.
+    \item Opposition $O$ observes the bribe schedule $m$ and makes a counter-offer $o \in (o_1, ..., o_n) \in \real$.
+    \item City councilors cast their votes and payoffs are realized.
+\end{enumerate}
+
+Given a bribe schedule $(a, b)$, councilor $i$ prefers to vote for the mayor's proposal $x$ if $a_i + v_i > b_i$ and the status quo $y$ otherwise.
+Since indifferent councilors vote for the status quo, the opposition needs to only match bribes from $M$, adjusting for individual preferences, i.e.
+$o_i = m_i + v_i$.
+For the mayor, she needs to construct the cheapest winning coalition in order to beat the opposition.
+
+Following Groseclose and Snyder (1996) and Banks we focus our analysis on the set of equilibria in which the mayor wins.
+In this context, the amount of patronage resources $W_M$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$.
+Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(m) = \sum_{i = 1}^n m_i$ be the total amount of patronage disbursed.
+The mayor then solves
+
+\begin{equation}
+    \label{eqn:solution}
+    \min_a\{S(a) : a \in U(\mathbf{v}, W_B) \}
+\end{equation}
+
+Note that for any equilibrium strategy, it must be the case that mayor $M$ uses a leveling schedule: every city councilor in her coalition $C$ is equally expensive for the opposition $O$ to bribe.
 

### 2020-08-14 paper/theory/model_executive_bargain.tex
incorporate changes in folder and add theory section

--- 2020-08-12 theory/model_executive_bargain.tex
@@
 
-In this section, I use the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to model how mayors engage in patronage in Brazil.
-For any given term, a mayor wants to enact her preferred policy, which requires winning over potential resistance by the city council.
-The preferences of these legislators may vary: in some municipalities a mayor enjoys widespread political support ("base política"), while in others, the opposition ("oposição") garners more seats.
-In this battle over votes, patronage is the currency used to buy legislative votes.
+In this section, I adapt the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to model how executive control over the city council determines patronage.
+The theoretical framework centers on two main actors, a government (\emph{governo}) and opposition (\emph{oposição}) who compete for votes in the city council.\footnote{I use the terms mayor and government interchangeably for the rest of this section.} The mayor seeks to enact her preferred policy, but can only do so by passing a simple majority vote in the council.
+The opposition, on the other hand, wants to block her efforts and maintain the status quo.
+In order to build a winning coalition, the mayor must spend a key political resource: patronage appointments into the municipal bureaucracy.
+
+Ultimately, two forces shape the mayor's strategy: the amount of resources controlled by the opposition and the underlying preferences of the legislators.
+The key parameter of interest is the underlying preferences -- i.e.
+support for the mayor -- of city councilors.
+In some municipalities a mayor has widespread political support ("base política"), while in others, the opposition ("oposição") garners more seats.
+In the actual empirical estimation, we approximate these preferences by looking into the councilor's partisan alignments in the mayoral election.
+For the second determinant, the opposition's resource, it is clear that an opposition with more resources can better compete with the mayor, forcing the latter to spend more to deter them.
 
@@
 Substantively, the model predicts that municipalities in which the mayor has more (less) councilors favorable to her, we should observe less (more) patronage.
-Below, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
+In the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
+I conclude with a discussion of the model and how it applies to the Brazilian municipal context.
 
@@
 
-The mayor $M$ and opposition $O$ compete over legislative votes to enact their preferred policies.
-There are two possible outcomes: a policy $x$ favored by the mayor, and the status quo, denote as $y$, preferred by the opposition.
+The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.
+There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
 In order to implement her policy the mayor must gain the approval of the city council, comprised of an odd $N$ number of voters, through a simple majority rule.

### 2020-08-14 paper/theory/notes_groseclose_et_snyder.tex
incorporate changes in folder and add theory section

--- 2020-06-22 theory/notes_groseclose_et_snyder.tex
@@
     \item Legislator: \textit{i}.
-    \item Vote buyers: \{\textit{A}, \textit{B}\}
+    \item Vote buyers: $p \in \{A, B\}$
 \end{enumerate}
@@
 \subsection*{Preferences:}
-For legislator $i$, denote the intensity of $i$'s preference of voting for $x$ over $s$ as
+For legislator $i$, denote the intensity of $i$'s preference of voting for $x$ over $s$ as:
 
@@
 
-For vote buyer \{\textit{A}, \textit{B}\}
+For vote buyer \{\textit{A}, \textit{B}\}, simply denote preferences over outcomes as:
+
+$$W_{p \in \{A, B\}} = U_p(x) - U_p(s)$$
+
+\subsection*{Strategies:}
+Let $a(\cdot)$ and $b(\cdot)$ denote the bribe offer functions.
 

### 2020-08-17 paper/theory/model_executive_bargain.tex
incorporate rest of banks theory

@@
 There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
-In order to implement her policy the mayor must gain the approval of the city council, comprised of an odd $N$ number of voters, through a simple majority rule.
-The mayor and opposition spend political resources $W_M$ and $W_O$ to win over votes, which for the mayor includes patronage appointments into the public sector.
+In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators.
+The total amount of political resources available is $W_G$ and $W_O$ to win over votes, which for the mayor includes patronage appointments into the public sector.
 
-Each city councilor is characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
-Let $\mathbf{v} = (v_1, ..., v_n)$ denote a preference profile for the city council.
-Let $v_i$ measure the degree to which an individual city councilor supports the mayor, with higher values denoting stronger support for the mayor and vice versa.
+Each city councilor is fully characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
+Let $\mathbf{v} = (v_1, ..., v_n)$ describe the preference profile for the city council.
+In this setting, $v_i$ measures the degree to which an individual city councilor supports the mayor, with higher values of $v$ denoting stronger support for the mayor and vice versa.
 Payoff gets realized when city councilor $i$ votes, independent of the outcome of the voting procedure.
@@
 \begin{enumerate}
-    \item Mayor $M$ offers a bribe schedule $m \in (m_1, ..., m_n) \in \real$.
-    \item Opposition $O$ observes the bribe schedule $m$ and makes a counter-offer $o \in (o_1, ..., o_n) \in \real$.
-    \item City councilors cast their votes and payoffs are realized.
+    \item Government $G$ offers a bribe schedule $a \in (a_1, ..., a_n) \in \real$.
+    \item Opposition $O$ observes the bribe schedule $m$ and makes a counter-offer $b \in (b_1, ..., b_n) \in \real$.
+    \item City councilors cast their votes at the end of bribing period.
+    \item Nature gathers votes and payoffs are realized.
 \end{enumerate}
@@
 Since indifferent councilors vote for the status quo, the opposition needs to only match bribes from $M$, adjusting for individual preferences, i.e.
-$o_i = m_i + v_i$.
-For the mayor, she needs to construct the cheapest winning coalition in order to beat the opposition.
+$b_i = a_i + v_i$.
+For the mayor, she needs to construct the cheapest winning coalition in order to defeat the opposition.
 
-Following Groseclose and Snyder (1996) and Banks we focus our analysis on the set of equilibria in which the mayor wins.
-In this context, the amount of patronage resources $W_M$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$.
-Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(m) = \sum_{i = 1}^n m_i$ be the total amount of patronage disbursed.
+Following Groseclose and Snyder (1996) and Banks (2000) we focus our analysis on the set of equilibria in which the mayor wins.\footnote{Since strategies for both players are symmetrical, any set of equilibria in which the mayor loses can be modeled as cases in which the the opposition loses.} In this context, the amount of patronage resources $W_G$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$.
+Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(a) = \sum_{i = 1}^n a_i$ be the total amount of patronage disbursed.
 The mayor then solves
@@
 Note that for any equilibrium strategy, it must be the case that mayor $M$ uses a leveling schedule: every city councilor in her coalition $C$ is equally expensive for the opposition $O$ to bribe.
+More formally, for any $a \in \mathbb{R}_+^n$, let $C(a) = \{i \in N : a_i > 0\}$ denote the set of individuals who receive a bribe from from the government $G$.
+One can show that there is a bribe schedule $a'$ such that for any $i,j \in C(a)$, $a'_i + v_i = a'_j + v_j$.
+The intuition is that the mayor has no incentive to make voters differentially expensive, because the opposition $O$ will simply ignore the more expensive voters and target the least favorable members of the coalition.
+Following G+S and BKS we refer to these strategies as leveling schedules.
+
+We can characterize the set of equilibria in the game by introducing additional notation.
+Let $U^l(\mathbf{v}, W_O) \subseteq U(\mathbf{v}, W_O)$ denote the set of unbeatable leveling schedules.
+These are bribe schedules such that $a_i + v_i = a_j + v_j  \equiv t(a)$.
+The bribe $a_i = t(a) - v_i$ is the sum of two terms.
+The first is the common "transfer" among all voters in $C(a)$, the second ($-v_i$) is individual specific.
+The latter term makes voters indifferent between $x$ and $y$ absent any bribe from $B$; the former represents the per capita amount necessary to make $C(a)$, together with any unbribed voters, unaffordable for $B$.
+
+I impose the following two assumptions:
+\begin{align*}
+    A1 &: v_{(n+1)/2} < 0\\
+    A2 &: v_1 < 2W_B/(n+1) 
+\end{align*}
+$A1$ implies that absent any bribes by $A$, $y$ will defeat $x$.
+Therefore $A$ must bribe at least one voter.
+$A2$ implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win.
+$A2$ also implies that for all $a \in U^l(\mathbf{v}, W_B)$, it must be that $t(a) \geq 2W_B/(n + 1)$, otherwise $B$ can bribe a majority from $C(a)$ itself and win the vote.
+
+For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$.
+Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$.
+Then, under $A2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$: $a'_i = t(a) - v_i$, $a'_j = 0$ and $a'_m = a_m$ for all $m \notin i, j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
+$a'_i \leq a_j$.
+Also, because of $A2$ $a'_j \leq a_j$, since it guarantees $a_j$ is non-negative.} Generalizing, and recalling that $v_1 \geq ...
+\geq v_n$, we see that for all $a \in U^l(\mathbf{v}, W_b)$ there exists a bribe schedule $a' \in U^l(\mathbf{v}, W_b)$ such that $S(a') \leq S(a)$ and $C(a') = \{1, ..., k(a)\}$.
+Therefore, we can without loss of generality restrict attention to schedules $a$ by $A$ which bribe the first $k(a)$ voters.
+Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.
+
+Therefore, when $A2$ holds,
+
+$$\min\{S(a): a \in U(\mathbf{v}, W_B)\} = \min\{S(a): a \in U^l_m(\mathbf{v}, W_B)\}$$
+
+We can now further simplify the total expenditure $S(a)$
+
+\begin{equation*}
+    S(a) = \sum_{i \in C(a)}a_i = k(a) \cdot t(a) - \sum_{i \leq k(a)}v_i
+\end{equation*}
+
+Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B)$. We can thus fully characterize the optimization problem of $A$ as
+
+\begin{equation*}
+    \min_{k,t} k \cdot t- \sum_{i \leq k} v_i
+\end{equation*}
+
+subject to the constraint that the induced schedule $a \in U_m^l$.
+Banks then reformulates this as an unconstrained problem by noting the following.
+First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$.
+Therefore, $B$ receives all non-bribed voters for free.
+For $a(k, t, \mathbf{v})$ to be unbeatable, then, it must be that $B$ cannot afford the remaining $(n + 1)/2 - (n - k) = k - (n - 1)/2$ voters, or
+
+$$t \cdot (k - (n - 1)/2) \geq W_B$$
+
+Solving this for equality yields the optimal transfer from $A$ to members of $C(A) = \{1, ..., k\}$, conditional on k:
+
+\begin{align}
+    \label{eqn:common_transfer}
+    t(k, W_B) = \frac{W_B}{k - (n - 1)/2}
+\end{align}
+
+Defining minimal winning expenditures as
+
+\begin{align}
+    \label{eqn:total_expenditure}
+    E(k, \mathbf{v}, W_B) = k \cdot t(k, W_B) - \sum_{i \leq k} v_i
+\end{align}
+
+we can state $A$'s problem as
+
+\begin{align}
+\label{eqn:minimal_expenditure}
+\min_k \{E(k, \mathbf{v}, W_B) : k \in {(n + 1/2), ..., n}\}
+\end{align}
+
+Denote the solution to expression \ref{eqn:minimal_expenditure} as $k^*(\mathbf{v}, W_B)$.
+This solution implicitly generates a solution to expression \ref{eqn:solution} through expression \ref{eqn:total_expenditure} and the induced bribe schedule above.
+Therefore, characterizing the optimal $k^*$ is sufficient to fully characterize the optimal behavior of the government $G$.
+
+\section{Results}
+
+First, characterize a solution for $k^*$.
+Because $k$ is finite, calculus cannot be employed.
+Instead, we deploy a discrete version of these techniques.
+First let's define $\Delta(k) = E(k + 1) - E(k)$ as the difference in expenditure from adding another coalition member.
+Note that if $\Delta(k) \geq 0$ then $A$ does not want to add another member to the coalition.
+Conversely, if $\Delta(k) < 0$, then $A$ is strictly better off by adding the $k + 1$th member of the coalition.
+
+Next, suppose that $\Delta(k)$ is increasing in $k$.
+The following algorithm can then be used to identify $k^*$: if $\Delta((n + 1)/2) \geq 0$, then we know from $\Delta(k)$ increasing that $A$ is better off by setting $k^*$ to $(n + 1)/2$.
+If $\Delta((n + 1)/2) < 0$, then we know that $k^*$ must be greater than $(n + 1)/2$, so we next solve for $\Delta((n + 3)/2)$, and so on.
+
+We can therefore search for the optimal $k^*$ with the following algorithm:
+
+\begin{align}
+    \label{eqn:optimal_k}
+    k^* = 
+    \begin{cases}
+        (n + 1)/2 & \text{if } \Delta((n + 1)/2) \geq 0 \\
+        \max\{k : \Delta(k - 1) < 0\} & \text{otherwise}
+    \end{cases}
+\end{align}
+
+Banks also shows that $\Delta(k)$ is indeed nondecreasing.
+
+\begin{align}
+    \Delta(k)  & = \left[\frac{(k + 1)W_B}{k + 1 - (n - 1)/2} - \sum_{i \leq k + 1}v_i \right]\\
+    \label{eqn:deltAk}
+    & = \frac{-W_B (n - 1)}{2(k + 1 - (n - 1)/2)(k - (n - 1)/2)} - v_{k + 1}\\
+    \label{eqn:def_deltAk}
+    & \equiv T(k, W_B) - v_{k + 1}
+\end{align}
+
+Note that because $v_1 \geq ...
+\geq v_n$, $-v_{k + 1}$ is non-decreasing in $k$.
+Also, taking the first term as a continuous variable and differentiating w.r.t.
+$k$, we note that it is also increasing in $k$.
+
+Generating an explicit characterization of $k$ is not straightforward.
+Instead, Banks opts to show the conditions which need to hold in order for the coalition to be universal or minimal.
+Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.
+
+\begin{proposition}
+    (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
+\end{proposition}
+
+Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity.
+Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v}, W_B)$.
+If $k' = (n + 1)/2$, then we know that $k' \leq k*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.
+
+From equation \ref{eqn:optimal_k} we infer that $\Delta(k - 1, \mathbf{v}', W_B) < 0)$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$.
+Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$.
+Then, $v_{k'} > T(k' - 1, W_B)$, and hence $\Delta(k' - 1, \mathbf{v}, W_B) < 0$.
+But then from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$.
+Therefore, the following holds:
+
+\begin{proposition}
+    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
+\end{proposition}
+
+In words, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases.
+Substantively, the number of voters bribed by $A$ weakly increases as the voter who receives the largest bribe finds $A$'s preferred alternative, $x$, more attractive.
+Similarly
+
+\begin{proposition}
+    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k' + 1} \leq v'_{k' + 1}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \leq k^*(\mathbf{v}', W_B)$
+\end{proposition}
+
+The "convexity" of $E$ guarantees that local information is sufficient to generate comparative statistics regarding changes in preferences $\mathbf{v}' \rightarrow \mathbf{v}$.
+Here, this local information is summarized by the preferences of the marginal voter $v_k$ and non-bribed voter $v_{k + 1}$.
+In order to identify these voters, one needs to solve for the optimal coalition size.
+Instead, Banks takes an alternative route and characterizes a weaker comparative statistic result which holds globally.
+Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v} \geq \mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$.
+
+\begin{corollary}
+    \label{cor:change_k}
+    For all $\mathrm{W_B}$, if $\mathbf{v} \geq \mathbf{v}'$ then $\mathrm{k}^*(\mathbf{v}, \mathrm{W_B}) \geq \mathrm{k}^*(v', \mathrm{W_B})$
+\end{corollary}
+
+Thus, the optimal number of voters bribed by $A$ and hence the optimal size of $A$'s optimal coalition, weakly increases as voters find $A$'s preferred alternative, $x$, more attractive.
+Although $A$'s total expenditure will decrease as $x$ becomes more attractive to the legislature, the optimal way to allocate this lower amount is to spread it more widely among voters.
+
+We can also characterize the change in total expenditures as a result of a shift in voter preferences. From equation \ref{eqn:total_expenditure} we have
+
+\begin{align*}
+    E(k, \mathbf{v}, W_B) - E(k, \mathbf{v}', W_B) & = \\
+    &= k \cdot t(k, W_B) - \sum_{i \leq k} v_i - \left[k \cdot t(k, W_B) - \sum_{i \leq k} v'_i\right]\\
+    &= \sum_{i \leq k}(v'_i - v_i)
+\end{align*}
+
+Since $v_i - v'_i \geq 0$, the difference in expenditure between a more favorable and unfavorable legislature is always non-positive.
 

### 2020-08-20 theory/model_executive_bargain.tex
improve theory section

@@
 \usepackage{titletoc}
+\bibliographystyle{plain}
 
@@
 
-In this section, I use the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to model how mayors engage in patronage in Brazil.
-For any given term, a mayor wants to enact her preferred policy, which requires winning over potential resistance by the city council.
-The preferences of these legislators may vary: in some municipalities a mayor enjoys widespread political support ("base política"), while in others, the opposition ("oposição") garners more seats.
-In this battle over votes, patronage is the currency used to buy legislative votes.
+In this section, I use the canonical vote-buying model proposed by \citet{groseclose_1996_buying} and \citet{banks_2000_buying} to model mayor's strategy to engage in patronage in Brazil.
+For any given term, a mayor seeks to enact her preferred policy, but to do so needs to garner legislative support in the city council.
+The preferences of these legislators vary: some of these councilors are more inclined to supporting the government (\emph{governista}), while others would rather oppose it.
+These different preferences translate into differential costs to sway their vote: city councilors aligned with the mayor are cheaper than their opposition counterparts.
 
-I show that, in equilibrium, the cost of passing the mayor's preferred policy decreases monotonically as the legislative support for the mayor increases.
+In this set-up, the mayor competes with the opposition to buy votes.
+The strategy of the mayor focuses on how to pre-emptively coopt some of these councilors so that they ultimately vote in favor of the mayor's proposal.
+Building this winning coalition requires side payments of various forms, but my primary focus is on patronage appointments.
+City councilors who vote for the mayor's proposal are rewarded, as loyalists, with positions in government.
+This arrangement is mutually beneificial: as pointed out by \citet{robinson_political_2013}, these appointments tie the councilor's fate to the patron.
+Because patronage appointments are discretionary, in order to maintain these appointments the city councilor must remain loyal to the government and vote in favor of the mayor's proposal.
+
+The main prediction of the model is that, in equilibrium, the cost of passing the mayor's preferred policy decreases monotonically as the legislative support for the mayor increases.
 Substantively, the model predicts that municipalities in which the mayor has more (less) councilors favorable to her, we should observe less (more) patronage.
+Fragmented governments, marked by a strong opposition in the legislature, therefore translate into a greater degree of patronage, with negative consequences for the electorate.
 Below, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
@@
 
+\bibliography{../paper/dissertation}
+
 \end{document}

### 2020-08-20 paper/theory/model_executive_bargain.tex

@@
 
+\bibliographystyle{plain}
 \begin{document}
@@
 
-In this section, I adapt the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to model how executive control over the city council determines patronage.
-The theoretical framework centers on two main actors, a government (\emph{governo}) and opposition (\emph{oposição}) who compete for votes in the city council.\footnote{I use the terms mayor and government interchangeably for the rest of this section.} The mayor seeks to enact her preferred policy, but can only do so by passing a simple majority vote in the council.
-The opposition, on the other hand, wants to block her efforts and maintain the status quo.
-In order to build a winning coalition, the mayor must spend a key political resource: patronage appointments into the municipal bureaucracy.
+In this section, I use the canonical vote-buying model proposed by \citet{groseclose_1996_buying} and \citet{banks_2000_buying} to model mayor's strategy to engage in patronage in Brazil.
+For any given term, a mayor seeks to enact her preferred policy, but to do so needs to garner legislative support in the city council.
+The preferences of these legislators vary: some of these councilors are more inclined to supporting the government (\emph{governista}), while others would rather oppose it.
+These different preferences translate into differential costs to sway their vote: city councilors aligned with the mayor are cheaper than their opposition counterparts.
 
-Ultimately, two forces shape the mayor's strategy: the amount of resources controlled by the opposition and the underlying preferences of the legislators.
-The key parameter of interest is the underlying preferences -- i.e.
-support for the mayor -- of city councilors.
-In some municipalities a mayor has widespread political support ("base política"), while in others, the opposition ("oposição") garners more seats.
-In the actual empirical estimation, we approximate these preferences by looking into the councilor's partisan alignments in the mayoral election.
-For the second determinant, the opposition's resource, it is clear that an opposition with more resources can better compete with the mayor, forcing the latter to spend more to deter them.
+In this set-up, the mayor competes with the opposition to buy votes.
+The strategy of the mayor focuses on how to pre-emptively coopt some of these councilors so that they ultimately vote in favor of the mayor's proposal.
+Building this winning coalition requires side payments of various forms, but my primary focus is on patronage appointments.
+City councilors who vote for the mayor's proposal are rewarded, as loyalists, with positions in government.
+This arrangement is mutually beneificial: as pointed out by \citet{robinson_political_2013}, these appointments tie the councilor's fate to the patron.
+Because patronage appointments are discretionary, in order to maintain these appointments the city councilor must remain loyal to the government and vote in favor of the mayor's proposal.
 
-I show that, in equilibrium, the cost of passing the mayor's preferred policy decreases monotonically as the legislative support for the mayor increases.
+The main prediction of the model is that, in equilibrium, the cost of passing the mayor's preferred policy decreases monotonically as the legislative support for the mayor increases.
 Substantively, the model predicts that municipalities in which the mayor has more (less) councilors favorable to her, we should observe less (more) patronage.
+Fragmented governments, marked by a strong opposition in the legislature, therefore translate into a greater degree of patronage, with negative consequences for the electorate.
 In the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
@@
 
+\bibliography{../paper/dissertation}
 \end{document}

### 2020-08-26 paper/theory/model_executive_bargain.tex
update theory section
incorporate new edits to the theory section

--- 2020-08-17 paper/theory/model_executive_bargain.tex
@@
 
+\bibliographystyle{unsrtnat}
 \begin{document}
@@
 
-In this section, I adapt the canonical vote-buying model proposed by Groseclose and Snyder (1996) and Banks (2000) to model how executive control over the city council determines patronage.
-The theoretical framework centers on two main actors, a government (\emph{governo}) and opposition (\emph{oposição}) who compete for votes in the city council.\footnote{I use the terms mayor and government interchangeably for the rest of this section.} The mayor seeks to enact her preferred policy, but can only do so by passing a simple majority vote in the council.
-The opposition, on the other hand, wants to block her efforts and maintain the status quo.
-In order to build a winning coalition, the mayor must spend a key political resource: patronage appointments into the municipal bureaucracy.
+In this section, I use a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes.
+For any given electoral term, a mayor wants to enact her preferred policy, but to do so needs to pass it through a simple majority in the city council.
+To guarantee her victory, the mayor offers appointment positions to voters, in this case, city councilors, who in exchange vote in favor of the bill.
 
-Ultimately, two forces shape the mayor's strategy: the amount of resources controlled by the opposition and the underlying preferences of the legislators.
-The key parameter of interest is the underlying preferences -- i.e.
-support for the mayor -- of city councilors.
-In some municipalities a mayor has widespread political support ("base política"), while in others, the opposition ("oposição") garners more seats.
-In the actual empirical estimation, we approximate these preferences by looking into the councilor's partisan alignments in the mayoral election.
-For the second determinant, the opposition's resource, it is clear that an opposition with more resources can better compete with the mayor, forcing the latter to spend more to deter them.
+The amount of patronage the mayor offers responds to the composition of the city council.
+When city councilors are predominantly in favor of the mayor's policy, we should expect less patronage, and vice versa.
+The intuition is straightforward: if the city councilors are already likely to support the bill, there is no additional need to offer them additional side payments.
+Note that there is no claim about the presence or complete absence of patronage \citep{stokes_brokers_2013, grindle_jobs_2012}, but rather its degrees: how many appointments must the mayor give out in order to guarantee her victory at the voting of the policy.
 
-I show that, in equilibrium, the cost of passing the mayor's preferred policy decreases monotonically as the legislative support for the mayor increases.
-Substantively, the model predicts that municipalities in which the mayor has more (less) councilors favorable to her, we should observe less (more) patronage.
-In the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
+An important feature of the model is the presence of an opposition party that competes with the mayor for legislative votes.
+In this setting, passing a policy proposal requires outbidding the competitor, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
+In particular, the amount of resources that the opposition controls can force the mayor's hands and increase the amount of patronage necessary to pass the bill.
+We pursue these empirical implicaitons of the model further in our results section.
+
+For the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
 I conclude with a discussion of the model and how it applies to the Brazilian municipal context.
@@
 
-The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.
+The mayor's government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.
 There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
 In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators.
-The total amount of political resources available is $W_G$ and $W_O$ to win over votes, which for the mayor includes patronage appointments into the public sector.
+The total amount of political resources available is $W_G$ and $W_O$, which for the mayor includes patronage appointments into the public sector.
 
@@
 In this setting, $v_i$ measures the degree to which an individual city councilor supports the mayor, with higher values of $v$ denoting stronger support for the mayor and vice versa.
-Payoff gets realized when city councilor $i$ votes, independent of the outcome of the voting procedure.
+Payoff are realized when city councilor $i$ votes, independent of the outcome of the voting procedure.
+This sincere voting preference forecloses the possibility of general equilibria in which $i$'s voting behavior affect $j$.
 
@@
     \item City councilors cast their votes at the end of bribing period.
-    \item Nature gathers votes and payoffs are realized.
+    \item Nature sums legislative votes, legislative outcome is decided and payoffs are realized.
 \end{enumerate}
@@
 
+\bibliography{../dissertation}
+
 \end{document}

### 2020-09-07 paper/theory/model_executive_bargain.tex
redraft intro of theory section
edit further theory section

@@
 For any given electoral term, a mayor wants to enact her preferred policy, but to do so needs to pass it through a simple majority in the city council.
-To guarantee her victory, the mayor offers appointment positions to voters, in this case, city councilors, who in exchange vote in favor of the bill.
-
-The amount of patronage the mayor offers responds to the composition of the city council.
-When city councilors are predominantly in favor of the mayor's policy, we should expect less patronage, and vice versa.
-The intuition is straightforward: if the city councilors are already likely to support the bill, there is no additional need to offer them additional side payments.
-Note that there is no claim about the presence or complete absence of patronage \citep{stokes_brokers_2013, grindle_jobs_2012}, but rather its degrees: how many appointments must the mayor give out in order to guarantee her victory at the voting of the policy.
+To ensure victory, the mayor offers appointment positions to voters, in this case, city councilors, who in exchange vote in favor of the bill.
 
 An important feature of the model is the presence of an opposition party that competes with the mayor for legislative votes.
-In this setting, passing a policy proposal requires outbidding the competitor, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
+In this setting, passing a policy proposal requires outbidding the opposition, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
 In particular, the amount of resources that the opposition controls can force the mayor's hands and increase the amount of patronage necessary to pass the bill.
-We pursue these empirical implicaitons of the model further in our results section.
+We pursue these empirical implications of the model further in our results section.
+
+The amount of public sector positions the mayor offers to these city councilors responds to the composition of the city council.
+The intuition is straightforward: if the city councilors are already likely to support the bill, there is no need to offer them additional side payments.
+Note that there is no claim about a dichotomy of clientelism, in which there is or is not patronage \citep{stokes_brokers_2013, grindle_jobs_2012}, but rather a variation in degrees: how many appointments must the mayor give out in order to govern on her preferred policy.
 
@@
 
-The mayor's government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.
-There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
+The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.\footnote{Note that I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
 In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators.
-The total amount of political resources available is $W_G$ and $W_O$, which for the mayor includes patronage appointments into the public sector.
+The total amount of political resources available is $W_G$ and $W_O$, which for the mayor includes public sector appointments.
 
-Each city councilor is fully characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
+Each city councilor is characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
 Let $\mathbf{v} = (v_1, ..., v_n)$ describe the preference profile for the city council.
@@
 The intuition is that the mayor has no incentive to make voters differentially expensive, because the opposition $O$ will simply ignore the more expensive voters and target the least favorable members of the coalition.
-Following G+S and BKS we refer to these strategies as leveling schedules.
+We refer to these strategies as leveling schedules.
 
@@
 
-For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$.
+% For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$.
 Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$.
@@
 
-Therefore, when $A2$ holds,
-
-$$\min\{S(a): a \in U(\mathbf{v}, W_B)\} = \min\{S(a): a \in U^l_m(\mathbf{v}, W_B)\}$$
-
-We can now further simplify the total expenditure $S(a)$
+These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{Details as to why can be found in the appendix.} We can simplify the total expenditure on patronage by the government, $S(a)$, as
 
@@
 
-Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B)$. We can thus fully characterize the optimization problem of $A$ as
+Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B)$.
+We can thus fully characterize the optimization problem of $A$ in equation \ref{eqn:solution} as
 
@@
 This solution implicitly generates a solution to expression \ref{eqn:solution} through expression \ref{eqn:total_expenditure} and the induced bribe schedule above.
-Therefore, characterizing the optimal $k^*$ is sufficient to fully characterize the optimal behavior of the government $G$.
+Therefore, characterizing the optimal $k^*$ is sufficient to fully characterize the optimal behavior of the mayor.
 
@@
 
-Banks also shows that $\Delta(k)$ is indeed nondecreasing.
+We can also further characterize the change in minimum winning expenditures in equation \ref{eqn:total_expenditure} as
 
@@
 
-Note that because $v_1 \geq ...
+Generating an explicit characterization of $k$ is not straightforward.\footnote{Note that because $v_1 \geq ...
 \geq v_n$, $-v_{k + 1}$ is non-decreasing in $k$.
 Also, taking the first term as a continuous variable and differentiating w.r.t.
-$k$, we note that it is also increasing in $k$.
-
-Generating an explicit characterization of $k$ is not straightforward.
-Instead, Banks opts to show the conditions which need to hold in order for the coalition to be universal or minimal.
+$k$, we note that it is also increasing in $k$.} Instead, Banks opts to show the conditions which need to hold in order for the coalition to be universal or minimal.
 Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.

### 2020-09-07 paper/theory/model_executive_bargain.tex
finalize editing theory section

@@
 subject to the constraint that the induced schedule $a \in U_m^l$.
-Banks then reformulates this as an unconstrained problem by noting the following.
+We can reformulate this as an unconstrained problem by noting the following.
 First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$.
@@
 
-\section{Results}
+\subsection*{Results}
 
@@
 
-Generating an explicit characterization of $k$ is not straightforward.\footnote{Note that because $v_1 \geq ...
-\geq v_n$, $-v_{k + 1}$ is non-decreasing in $k$.
-Also, taking the first term as a continuous variable and differentiating w.r.t.
-$k$, we note that it is also increasing in $k$.} Instead, Banks opts to show the conditions which need to hold in order for the coalition to be universal or minimal.
 Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.
@@
 Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity.
-Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v}, W_B)$.
-If $k' = (n + 1)/2$, then we know that $k' \leq k*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.
+Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v'}, W_B)$.
+If $k' = (n + 1)/2$, then we know that $k' \leq k^*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.
 
-From equation \ref{eqn:optimal_k} we infer that $\Delta(k - 1, \mathbf{v}', W_B) < 0)$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$.
+From equation \ref{eqn:optimal_k} we infer that $\Delta(k' - 1, \mathbf{v}', W_B) < 0$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$.
 Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$.
@@
 
-The "convexity" of $E$ guarantees that local information is sufficient to generate comparative statistics regarding changes in preferences $\mathbf{v}' \rightarrow \mathbf{v}$.
-Here, this local information is summarized by the preferences of the marginal voter $v_k$ and non-bribed voter $v_{k + 1}$.
-In order to identify these voters, one needs to solve for the optimal coalition size.
-Instead, Banks takes an alternative route and characterizes a weaker comparative statistic result which holds globally.
-Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v} \geq \mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$.
-
-\begin{corollary}
-    \label{cor:change_k}
-    For all $\mathrm{W_B}$, if $\mathbf{v} \geq \mathbf{v}'$ then $\mathrm{k}^*(\mathbf{v}, \mathrm{W_B}) \geq \mathrm{k}^*(v', \mathrm{W_B})$
-\end{corollary}
-
-Thus, the optimal number of voters bribed by $A$ and hence the optimal size of $A$'s optimal coalition, weakly increases as voters find $A$'s preferred alternative, $x$, more attractive.
-Although $A$'s total expenditure will decrease as $x$ becomes more attractive to the legislature, the optimal way to allocate this lower amount is to spread it more widely among voters.
-
-We can also characterize the change in total expenditures as a result of a shift in voter preferences. From equation \ref{eqn:total_expenditure} we have
+The ``convexity" of $E$ guarantees that local information is sufficient to generate comparative statistics regarding changes in preferences $\mathbf{v}' \rightarrow \mathbf{v}$.
+We can characterize the change in total expenditures as a result of a shift in voter preferences.
+Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$.
+From equation \ref{eqn:total_expenditure} we have
 
@@
 
-Since $v_i - v'_i \geq 0$, the difference in expenditure between a more favorable and unfavorable legislature is always non-positive.
+Since $v'_i - v_i \leq 0$, the difference in expenditure between moving from a favorable to a less favorable legislature is always non-positive, i.e.
+the government has to spend less resources to pass her preferred policy.
+This holds despite the fact that when these preferences shift there is an increase in the overall size of the coalition.
+This result has a similar flavor to \citet{groseclose_1996_buying}, who motivate their model by stating that it may be optimal to increase the size of the coalition (instead of buying a simple majority) because doing so overall can lead to a reduction in the amount of expenditures by the vote-buyer.
+
+\subsection*{Discussion}
+
+Enacting policy requires the exchange of political currency for votes.
+Whether it be in presidential coalitionism, or in the local city council politics, mayors who wish to govern have to engage in transactions with the legislature.
+What I showed in this section was that political misalignment between the government and the legislature can in fact be counterproductive: more patronage occurs, leading to worse public service outcomes.
+
+The model also highlights a key aspect of clientelism that is often neglected electoral accountability models: voters have a limited voice.
+Ultimately, the exchanges which occur between the legislature and the mayor have little to do with the voter at the end of the pipeline, and more to do with the city councilors.
+The first order requirement for the government is to ensure that it has enough legislative votes in order to enact the very policies that the voter may or may not desire.
+This transactional cost is not illegal: rather, it is necessary for democratic relations between different branches of government.
+
+In the next section, I test whether the empirical implication of the model is correct: does more patronage occur in municipalities with greater political misalignment between the mayors and city councilors.
+I test additional implications of the model, including whether shifts in the resources controlled by the opposition can affect the government's patronage strategy.
 

### 2020-09-08 paper/paper.tex
transfer paper markdown to tex

\documentclass[12pt,a4paper]{article}
\usepackage{setspace}
\usepackage[top = 1.25in, bottom = 1.25in, left = 1.25in, right = 1.25in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}
\setlength\parindent{20pt}
\usepackage{setspace}
\usepackage{multirow}
\usepackage{caption}
\usepackage{titling}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\onehalfspacing
\usepackage[hang,flushmargin]{footmisc}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\renewcommand{\theproposition}{\arabic{proposition}}
\newcommand{\real}{\mathbb{R}_+^n}
\newcommand{\bfv}{\mathbf{v}}
\newcommand{\blue}{\textcolor{blue}}

\bibliographystyle{unsrtnat}
\begin{document}

\section{Introduction}

Across the world, vulnerable populations rely on governments for access to basic public services such as education. Over the past two decades, enrollment rates in public education have improved dramatically.\footnote{See \citet{haggard_development_2008}, \citet{kaufman_globalization_2001}.} Improvements in access have not been matched, for many, with a better quality of education. From a global perspective, educational quality indicators such as the PISA (Programme for International Student Assessment) score have broadly stagnated.\footnote{For coverage, see [here](https://www.economist.com/international/2019/12/05/pisa-results-can-lead-policymakers-astray).}

How can we improve the quality of education? In large part, the answers are domestic. While each country faces a particular set of challenges, broad institutional reforms in the developing world make their cases comparable. Decentralization has delegated responsibility for public education to subnational governments, as well as induced variation in the quality of education along subnational lines \citep{falleti_decentralization_2010, arretche_mitos_1996}. Understanding how this level of government manages these services, and in particular how local actors strategically shape them can help find answers to that question. \citep{gulzar_politicians_2017, min_power_2015}.

% In this paper, I highlight the role of the strategic interaction between political elites and its downstream effects on the provision of educational services. Similar to coalition building, local executive leaders secure broader support for their policy agenda through the allocation of public sector positions to city councilors.^[@laver_coalitions_1990, @power_optimism_2010.] As highlighted by previous research, teacher turnover can lead to decreased morale and inexperienced staff entering public service (@ingersoll_teacher_2001, @ronfeldt_how_2013). Patronage, by inducing new hires or dismissals, can therefore cause instability in the educational sector, with negative effects for student learning.

% To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education, both at a national and subnational level. I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory, and a school-grade-specific turnover index. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

% To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage, in particular during the first year of administration. Mayors therefore resort to patronage in order to shore up legislative support, with negative consequences for student learning. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

% This study contributes to an emerging literature on the politics of personnel and public services (@pepinsky_bureaucracy_2017, @finan_personnel_2015 and @gulzar_politicians_2017). Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services (@ferraz_corrupting_2012). It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy (@kohli_state-directed_2004, @evans_embedded_1995).

% The paper is structured as follows. Section 1 provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. In section 2 I present the main theoretical argument, with a formal treatment of the subject. Section 3 presents the institutional context and data, followed in section 4 by the research design and main results. Section 5 concludes.

% # Related literature and debate

% In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions. I also highlight how my research incorporates multiplicity in political actors and how this affects bargaining over public sector jobs. I address this gap by adapting previous analyses of executive-legislative bargaining to bureaucratic control at the local level.

% ## Bureaucratic personnel and public goods provision

% Bureaucracies have a clear impact on the delivery of public services. A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (@centeno_unpacking_2017, @kohli_state-directed_2004). A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (@johnson_miti_1982, kohli_state-directed_2004). A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (@evans_bureaucracy_1999).

% Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. @gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. @akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

% This recent wave of studies shed light on the intersection between politicians and bureaucracies. However, few of these studies explicitly model the multiple actors involved in managing bureaucracies. Understanding their diverse goals and action space provides a firm theoretical foundation to how different politicians can reshape bureaucracies. To do so I turn to the well-established literature on executive-legislative bargaining, applying its insights to the analysis of local government and administration.

% ## Presidential coalitionism and patronage

% For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (@grindle_jobs_2012, @mccarty_appointments_2004). Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. (@raile_executive_2011, @power_optimism_2010).

% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. (@raile_executive_2011). In a setting characterized by weak party cohesion and programmatic commitments (@ames_electoral_1995, @lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (@batista_o_2013, @neto_presidential_2006, @figueiredo_executivo_1999).

% \blue{something about coalitions}

% In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.^[Interview with C, August and September 2019.] Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."^[Interview with C, August 2019.] These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

% ## Bureaucratic turnover and inexperienced education

% Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (@gailmard_slackers_2007). Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (@clotfelter_teacher_2007.) @akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (@colonnelli_patronage_2017).

% > "I am aware that the position is temporary. Especially because it is a political position, decided by the administration. If the current administration is out of power, we are automatically dismissed." - Interview with school principal A, August 2019.

% Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level. In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers. In the following section I describe the institutional context for public education in Brazil, as well as the data employed for the the estimation.

% # The argument

% ## The executive dilemma

% I model the executive decision over patronage as a product of political career concerns and strategic interaction with the local legislative power. In particular, mayors are assumed to be career-oriented and re-election seeking (@besley_principled_2006). This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (@stokes_brokers_2013, @robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.

% The mayor's strategy centers around the redistribution public sector jobs. When the mayor takes office, a set of bureaucratic positions become available for allocation to city councilors. In a municipal parallel to presidential coalitionism, jobs are allocated to legislators in order to buy their vote. I assume, for now, that job allocations have a negative effect on public services. The mayor can refrain from distributing jobs for vote buying, since reducing turnover helps retain expertise and improves the quality of public services. However, doing so raises the threat of defection by coalition members, which can lead to the withholding of budgetary approval. This is the executive's dilemma.

% The main proposition is that mayors who control a smaller (larger) share of legislative seats allocate more (less) public sector job to city councilors. The intuition is as follows. Mayors need to secure a simple majority of legislative votes to have the budget approved. @groseclose_buying_1996 and @banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor. Having exhausted these votes, the executive has to resort to buying votes from the opposition, which demand more patronage. The less seats occupied by members of the mayor's electoral coalition, the more expensive vote-buying becomes, and the observable implication is increased patronage in the municipality.

% The model builds on rich theoretical treatment of electoral accountability and majority-buying in legislatures. Starting with @barro_control_1973 and @ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (@coate_form_1995; @besley_principled_2006). The possibility of reelection can motivate them to exert effort and convince voters to reelect them. This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (@groseclose_buying_1996, @banks_buying_2000).

% ## The model

% There is one mayor $m$, a city council $N$ and representative median voter $v$. The goal of the mayor is to be reelected into office, yet for any given time period $m$ needs to garner legislative support. The city council plays that key role by approving or not the budget for the executive. The legislative chamber is composed by $N = {1,..,n}$ legislators, where $S \subseteq N$ denote the subset of legislators who belong to the electoral coalition of mayor $m$. For each $i \in N$, let $\textbf{v} = \{v_1,...,v_n\}$ describe legislator i's preference to support the mayor in the budgetary process. Let $v_i > v_j$ for all $i \in S$ and $j \in \{N \setminus S\}$. Therefore, members of the mayor's electoral coalition are more likely to approve the budget than outside members.

% I introduce a set of simplifying assumptions. First, mayors are one of two types: a technocratic politician who delivers high quality public goods, and a pragmatic one, who fails to do so. Note that this is an informational heuristic. Voters have beliefs about how competent a politician is and strictly prefer them over the alternative. Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (@coate_form_1995). Additionally, I assume that the voter is sincere, excluding pivotality concerns and strategic voting from voters' utility calculations. Finally, I normalize the politician's exit option to 0: the politician always prefers to be re-elected.

% The game has two periods. Voter $v$ does not observe the amount of patronage handed out by mayor $m$ nor her type, but observes and derives utility from the realization of public good quality $\omega_t$. An important feature in the model is that high quality public services $\omega_{h}$ are produced probabilistically as a function of $\theta$, the quality of the bureaucracy. The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See @kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.

% The mayor's optimal strategy is pinned down by the share of patronage which maximizes expected utility from holding onto office, conditional on her type and expected benefits from reelection. Mayor $m$ has to offer public sector jobs to coopt the opposition while ensuring that the voter is satisfied enough to reelect her. The executive dilemma is how to reconcile the incentive for reelection and the pressure imposed by the opposition.

% The timing of the game is as follows.

% 1) Nature draws the politician's type.
% 2) Politician observes her type and allocates public sector jobs to legislators, setting $\theta_1$.
% 4) Nature realizes public good $\omega_1(\theta_1)$.
% 5) Voter observes $\omega_1$ and casts vote to retain or fire the incumbent.
% 6) If incumbent is retained, she sets $\theta_2$ and nature draws $\omega_2(\theta_2)$. If incumbent is deposed, a challenger takes office, sets $\theta_2$ and the game ends.

% ## The executive dilemma

% The mayor's choice set is the bureaucratic quality parameter $\theta_t$. As quality increases - the mayor allocates a smaller proportion of jobs to legislators - the less votes become secure in the legislature. This reduces the likelihood of securing the necessary resources to provide public goods for a given fiscal year. Adapting @gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic. Technocratic politicians invariably deliver high quality public goods to the voter. Pragmatic politicians, on the other hand, are susceptible to legislative pressure and at times fail to deliver them.

% The probability of a politician being technocratic is common knowledge: $\xi \in [0,1]$. Conversely, the probability of her being pragmatic is simply $1 - \xi$. Both types derive utility from public goods provision $\omega$, where for simplicity $u(\omega_l)=0$ and $u(\omega_h) = 1$. We do not model the decision of the competent mayor: she simply delivers high quality goods and is inevitably reelected. We focus on the more interesting case of the pragmatic mayor.

% \begin{align*}
% Eu_m \left(\theta_t, s_1 \middle|\lambda \right) &= \textbf{E}(\omega_1) + \delta \text{Pr}(\text{reelection})\left[\psi + \textbf{E}(\omega_2)  \right] \\
% &= \frac{1-s_1\theta_1}{1+e^{-(10\theta_1-5)}} + \delta \text{Pr}(\text{reelection})\left[\psi + \frac{1-s_2\theta_2}{1+e^{-(10\theta_2-5)}} \right] \\
% \end{align*}

% The equation above formalizes the mayor's dilemma. The probability of reelection decreases with the share of jobs allocated to the opposition, since it lowers the quality of the bureaucracy. However, if the mayor fails to appease them through patronage, it is increasingly costly to "govern". While technocrats are immune to these political pressures, pragmatists ultimately cave in. The need to coopt legislators decreases the quality of the bureaucracy and public goods, as mayors are forced to allocate positions within the bureaucracy to garner support in the legislative. Details on deriving the equilibrium strategy are in the appendix. I now describe Brazil as a case for testing the theory.

% # Institutional context and data

% ## Education in Brazil

% In Brazil, responsibility for primary education is delegated to local governments (@paschoal_historia_2009). The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments. Figure \ref{fig:enrollment} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

% \blue{plots/enrollment_dep.pdf}

% The local executive branch has sole prerogative over the hiring and firing of teachers and school principals. School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (@brollo_political_2017). Teachers usually enter the public service through a public exam and are eligible for tenure after two years (@gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount. Municipal teachers and school principals are overseen by a local department of education (@militao_educacao_2014).

% Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education. At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (@ferraz_corrupting_2012).

% ## Local governance and mayoral coalitions

% Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council. Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016. These take place ] The executive is responsible for the implementation of social policies such as education, as well as drafting the municipal budget. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by @souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.

% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (@dantas_eleicoes_2013). These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber. Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}). While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.^[Interview with C, former chief of staff of municipality J. September 2018.]

% Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions. While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves. These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council. In the words of a set of school administrators in the municipality of I:

% > School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.

% > Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]

% These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff. To verify these claims in a broader set of municipalities, as well as outlining the research design, I employ extensive administrative and electoral data collected by the Brazilian federal government. In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.

% ## Data

% Brazil collects fine-grained data on education and makes it publicly available for research. For this study, data on education are collected from two main sources: the SAEB and School Census.^[These can be accessed (here)[http://portal.inep.gov.br/web/guest/dados].] The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

% \blue{"plots/saeb_map.pdf"}

% Figure \ref{fig:map} presents a map of Brazil with the municipal averages for the SAEB of 2015. Warmer colors such as red and orange indicate higher average scores, with the converse denoted by colder, blue shades. There is clear variation in average test scores, with the Southeast and Midwest outperforming poorer regions in the North and Northeast. Even within regions, however, there is wide heterogeneity. In particular, note that in the northern part of Brazil, high-performing municipalities neighbor low-performing ones. This indicates that despite spatial clustering, municipal factors can lead to variation in the quality of education.

% Employment data on educational staff, including teachers and school principals, are extracted from the RAIS, SAEB and the School Census.^[RAIS data, along with additional Brazilian employment data, can be accessed (here)[http://trabalho.gov.br/dados-abertos].] The \emph{Relatório Anual de Informações Sociais} (RAIS) is an annual census administered by the now Ministry of Finance to all formal employers in Brazil. Employers are mandated to accurately report data on employees, subject to fines for misreporting. Subnational governments, including municipalities, also report to the RAIS. The dataset therefore contains micro-level information on all municipal personnel, including when they were hired/fired, as well as wages, type of contract and education levels.

% Figure \ref{fig:staff} provides descriptive statistics on bureaucratic personnel in Brazil, segmented by department. Educational staff, in this case school principals and teachers, comprise approximately a third of municipal public sector jobs, totaling around 2 million in 2015. While comparable to administrative staff, this total significantly exceeds that for healthcare services. Focusing on turnover, we note that new hiring and dismissals in education staff is relatively high. While lower than that for administrative staff, new hires can represent over 10 percent of extant staff. The past two decades was a moment of rapid expansion of municipal staff, and hiring has far exceeded dismissals in that time period.

% \blue{plots/turnover_category.pdf}

% I propose three different specifications for measuring bureaucratic turnover. From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by @pereira_junior_indicadores_2016. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

% $$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$

% Information on mayor, city councilors and mayoral coalitions are made available by the Supreme Electoral Court (TSE), the national authority responsible for overseeing elections.^[Available (here)[http://www.tse.jus.br/eleicoes/estatisticas/repositorio-de-dados-eleitorais-1].] In order to calculate the share of legislative seats held by the mayoral coalition, I match the partisanship of each city councilor to the mayoral electoral coalition. The distribution of share of coalition seats is depicted in figure \ref{fig:coalition}. Note that although most city councils are controlled by the mayor, in over 40 percent of municipalities the mayoral coalition is a minority government.

% \blue{plots of}

% Additional data has been collected to supplement the estimation exercise. Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.^[Respectively, data can be found here: (1)[https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html], (2)[https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf], (3)[https://www.seduc.ce.gov.br/spaece/].]

% # Estimation

% ## Empirical strategy

% This study presents two sets of estimations. The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by SAEB and SPAECE to estimate the effect of turnover on student test scores. To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.

% For the first estimation, I employ a hierarchical linear model to estimate the effect of teacher turnover on average test scores for grade $i$ at school $s$, in municipality $j$ and year $t$. This estimation strategy has been used in multiple studies to analyze educational, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation (@diprete_multilevel_1994, @lee_using_2000.) Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise. The main specification is as follows:
% \begin{align*}
% \text{test score}_{isjt} = \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
% \psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
% \end{align*}
% We are interested in $\beta_1$ and $\beta_3$, the change in test scores associated with teacher or school principal turnover. In this set-up, $\beta_3$ is the additional effect of staff turnover on test scores if students are in 9th grade. The first level is the grade, with associated characteristics $X_isjt$ at the grade level: among others, percentage of students who have not graduated last year and share of children with a fridge in their house. The second is the school $s$, with characteristics $V_{sjt}$, such as access to internet or the presence of a cafeteria for students. The third level is the municipality $j$, including municipal sociodemographic characteristics ($W_{jt}$) such as population size and median wages, as well as per pupil budgetary expenditures on eduation. Finally, I include state ($\alpha_kj$) and year ($\delta_t$) fixed effects to account for unobserved time-invariant state characteristics and annual seasonality.

% The second set of estimations tests the hypothesis that mayors who control less seats in the city council engage in more patronage. To do so I leverage employment data from RAIS and school census data to measure teacher turnover rates. I deploy three sets of models. For the second and third estimation I turn to the micro, bureaucrat-level and a then to aggregated turnover at the municipal level. The first model is a logistic regression where the outcome of interest is whether a particular teacher or school principal is hired/fired for any given year. The second model is a linear model with fixed effects, where the outcome of interest is the proportion of educational staff hired/fired for a muncipality at any given year. For the second set of estimations, the main specification is:
% $$\text{turnover}_{jt} = \gamma \text{coalition seats}_{jst} + \mu P_{jt} + \zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{jt}$$
% The share of coalition seats held by the mayoral coalition is the treatment in this set-up. We are interested in $\gamma$, the change in bureaucratic turnover associated with variation in the share of legislative seats held by the mayoral coalition. Municipal characteristics $W_{jt}$ are similar to the ones used for the first set of estimation. I add political variables to the estimation in order to control for differences in mayor partisanship, incumbency status, and individual characteristics of the mayor: professional background, education, and age.

% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in @imai_covariate_2014 which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

% ## Results

% Teacher and school principal turnover have significant, negative effects on average test scores, highlighting the potential dangers of bureaucratic turnover on student learning. These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations. Table \ref{tbl:hlm_mods} presents the results of the estimation on SAEB test scores. We present two specifications for turnover. Turnover index measures the amount of turnover in teachers at the school level. Work experience for teachers and school principals serve as an alternative measure of turnover. Personnel who has only recently entered the school recently amass a shorter amount of work experience (less than two years).

% \begin{landscape}
% \begin{table}[t]
%   \centering
%   \footnotesize
%   \input{tables/hlm_result}
%   \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning. Models 1 and 2 present results for teacher turnover index constructed at the school level. Models 3 and 4 estimate the effect of new teachers and school principals entering the school (less than two years). All models include year and state fixed effects.}
%   \label{tbl:hlm_mods}
% \end{table}
% \end{landscape}

% Additionally, I present results for the SPAECE standardized exam in table \ref{tbl:spaece}. Differently than the SAEB, the SPAECE takes place annually and the average test scores are calculated at the grade, rather than the classroom level. The annual frequency of this exam allows us to retrieve more precise estiamtes with respect to the effect of teacher turnover on student test scores, as opposed to the biannual frequency of the SAEB. Note that for both estimations, we standardize the outcome of interest. Coefficients are therefore interpretable as a standard deviation change in student test with respect to the covariate.

% \begin{table}[t]
%   \centering
%   \footnotesize
%   \input{tables/spaece_result}
%   \caption{{\bf Turnover and student learning in Ceará.} Teacher turnover have a negative effect on student learning. Models 1 and 2 present results for teacher turnover index constructed at the school level.}
%   \label{tbl:spaece}
% \end{table}

% I now turn to testing the main proposition of this paper: that turnover in educational staff stems indeed from coalitional dynamics. There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage. Table \ref{tbl:turnover} presents the results of the estimation. The first model regresses turnover index on coalition share, where the unit of analysis is a grade-level per school. Models 2 and 3 estimate the effect of the share of legislative seats on hiring or dismissal of individual bureaucrats. The third set of models (4 and 5) estimate the change in the share of new hires and dismissals at the municipal level. All models suggest that an increase in the executive share of legislative seats decrease turnover.

% \begin{landscape}
% \begin{table}[t]
%   \centering
%   \footnotesize
%   \input{tables/turnover_result}
%   \caption{{\bf Executive coalitions and staff turnover.} An incrase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals. Model 1 presents results on the turnover index. Models 2 and 3 are logistic regressions at the individual, bureaucrat level. Models 4 and 5 are aggregated at the municipal level. where the outcome of interest is the proportion of staff either hired or dismissed at a given year. Models 1, 3, and 5 include year and state fixed effects.}
%   \label{tbl:turnover}
% \end{table}
% \end{landscape}

% For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council. Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature. Overall, these results suggest that weaker executive control over the legislature increases patronage, with potentially negative effects for public service delivery in municipalities in Brazil.

% \begin{figure}[H]
%     \centering
%     \includegraphics{plots/plot_pred}
%     \caption{Predicted values for bureaucratic turnover. The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive. The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
%     \label{fig:turnover}
% \end{figure}

% As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity. It is in these circumstances that validating causal mechanisms through fieldwork and careful data analysis can increase the validity of a causal claim. In-depth interviews with local actors involved in managing education, as well as the use of different measurements for both educational staff turnover and student learning, provides strong evidence that staff turnover has a negative impact on student learning. Bureaucratic turnover stems from political considerations.

% ## Mechanisms

% The question remains: why do city councilors engage in patronage. Previous studies have found that citizens directly access

% # Conclusion

% Improving the quality of education received by children across the world remains a challenge. This paper proposes an analytical framework and estimation strategy to understand the decision-making process behind the administration of educational services in Brazil. In a decentralized context, subnational political actors have a direct say on how educational services are managed, with profound implications for the quality of public services delivered to citizens. These actors interact with bureaucracies and other local elites in complex ways that are only beginning to be mapped.

% In this study I theorize and demonstrate that staff turnover stems from the executive's need to garner support from legislators in the city council. This process of coalition building is consolidated through employment offers to city councilors and their constituencies. As the share of seats held by the executive coalition decreases, the costlier it becomes to coopt legislators to support the executive. As a result, the amount of patronage we observe should increase. That is precisely what the data indicates, with teacher turnover increasing in schools, as well as increased hiring and dismissals of teachers and school principals.

% This bureaucratic turnover has important consequences for the quality of education delivered in municipalities. I find that turnover has a negative effect on student learning, across a set of specifications for turnover and different evaluation metrics for student learning. The evidence therefore seems to point out that mayors with a weaker hold on the city council resort to greater patronage, with negative consequences for student learning. This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts (@gottlieb_countervailing_2019).

% Future research on patronage and public service delivery would benefit from a clearer treatment of the institutional context in which local political actors operate. Important insights have been derived on executive leaders, but these actors seldom govern alone. Incorporating other local elites paints a more complex and accurate understanding of the strategic considerations taking place in the political management of education, and public services more broadly.

% \pagebreak

% # Appendix

% \pagebreak

% ## Covariate balance

% ### Education quality vs. staff turnover

% \blue{plots/balance_hlm}

% ### Staff turnover vs. executive share of legislative seats

% \blue{plots/balance_felm.pdf}

% \pagebreak

\bibliography{dissertation}

\end{document}

### 2020-09-08 paper/paper.tex
edit intro

@@
 
-Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-Over the past two decades, enrollment rates in public education have improved dramatically.\footnote{See \citet{haggard_development_2008}, \citet{kaufman_globalization_2001}.} Improvements in access have not been matched, for many, with a better quality of education.
-From a global perspective, educational quality indicators such as the PISA (Programme for International Student Assessment) score have broadly stagnated.\footnote{For coverage, see [here](https://www.economist.com/international/2019/12/05/pisa-results-can-lead-policymakers-astray).}
+% Across the world, vulnerable populations rely on governments for access to basic public services such as education.
+In many countries, particularly in the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, mathematical operations, elusive.
+Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them.
+\citep{cochrane_policies_1983, falleti_decentralization_2010}.
+How can local governments improve the quality the of public services they deliver?
+In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?
 
-How can we improve the quality of education?
-In large part, the answers are domestic.
-While each country faces a particular set of challenges, broad institutional reforms in the developing world make their cases comparable.
-Decentralization has delegated responsibility for public education to subnational governments, as well as induced variation in the quality of education along subnational lines \citep{falleti_decentralization_2010, arretche_mitos_1996}.
-Understanding how this level of government manages these services, and in particular how local actors strategically shape them can help find answers to that question.
-\citep{gulzar_politicians_2017, min_power_2015}.
-
-% In this paper, I highlight the role of the strategic interaction between political elites and its downstream effects on the provision of educational services.
-Similar to coalition building, local executive leaders secure broader support for their policy agenda through the allocation of public sector positions to city councilors.^[@laver_coalitions_1990, @power_optimism_2010.] As highlighted by previous research, teacher turnover can lead to decreased morale and inexperienced staff entering public service (@ingersoll_teacher_2001, @ronfeldt_how_2013).
+In this paper, I highlight the role of the strategic interaction between political elites and its downstream effects on the provision of educational services.
+Similar to coalition building, local executive leaders secure broader support for their policy agenda through the allocation of public sector positions to city councilors.\citep{laver_coalitions_1990, power_optimism_2010} As highlighted by previous research, teacher turnover can lead to decreased morale and inexperienced staff entering public service \citep{ingersoll_teacher_2001, ronfeldt_how_2013}.
 Patronage, by inducing new hires or dismissals, can therefore cause instability in the educational sector, with negative effects for student learning.
 
-% To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence.
+To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence.
 Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students.
@@
 
-% This study contributes to an emerging literature on the politics of personnel and public services (@pepinsky_bureaucracy_2017, @finan_personnel_2015 and @gulzar_politicians_2017).
-Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services (@ferraz_corrupting_2012).
-It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy (@kohli_state-directed_2004, @evans_embedded_1995).
+% This study contributes to an emerging literature on the politics of personnel and public services (pepinsky_bureaucracy_2017, finan_personnel_2015 and gulzar_politicians_2017).
+Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services (ferraz_corrupting_2012).
+It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy (kohli_state-directed_2004, evans_embedded_1995).
 
@@
 % Bureaucracies have a clear impact on the delivery of public services.
-A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (@centeno_unpacking_2017, @kohli_state-directed_2004).
-A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (@johnson_miti_1982, kohli_state-directed_2004).
-A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (@evans_bureaucracy_1999).
+A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (centeno_unpacking_2017, kohli_state-directed_2004).
+A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (johnson_miti_1982, kohli_state-directed_2004).
+A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (evans_bureaucracy_1999).
 
@@
 Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores.
-@gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment.
-@akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.
+gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment.
+akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.
 
@@
 These political actors have competing claims over the local bureaucracy, with important implications for public service delivery.
-This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (@grindle_jobs_2012, @mccarty_appointments_2004).
+This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (grindle_jobs_2012, mccarty_appointments_2004).
 Divergent political interests can lead to strategic interaction between executive and legislative actors.
 A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy.
-(@raile_executive_2011, @power_optimism_2010).
+(raile_executive_2011, power_optimism_2010).
 
@@
 Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
-(@raile_executive_2011).
-In a setting characterized by weak party cohesion and programmatic commitments (@ames_electoral_1995, @lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (@batista_o_2013, @neto_presidential_2006, @figueiredo_executivo_1999).
+(raile_executive_2011).
+In a setting characterized by weak party cohesion and programmatic commitments (ames_electoral_1995, lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999).
 
@@
 % Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects.
-As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (@gailmard_slackers_2007).
-Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (@clotfelter_teacher_2007.) @akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores.
-When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (@colonnelli_patronage_2017).
+As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (gailmard_slackers_2007).
+Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (clotfelter_teacher_2007.) akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores.
+When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (colonnelli_patronage_2017).
 
@@
 % I model the executive decision over patronage as a product of political career concerns and strategic interaction with the local legislative power.
-In particular, mayors are assumed to be career-oriented and re-election seeking (@besley_principled_2006).
-This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (@stokes_brokers_2013, @robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.
+In particular, mayors are assumed to be career-oriented and re-election seeking (besley_principled_2006).
+This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (stokes_brokers_2013, robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.
 
@@
 Mayors need to secure a simple majority of legislative votes to have the budget approved.
-@groseclose_buying_1996 and @banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor.
+groseclose_buying_1996 and banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor.
 Having exhausted these votes, the executive has to resort to buying votes from the opposition, which demand more patronage.
@@
 % The model builds on rich theoretical treatment of electoral accountability and majority-buying in legislatures.
-Starting with @barro_control_1973 and @ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (@coate_form_1995; @besley_principled_2006).
+Starting with barro_control_1973 and ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (coate_form_1995; besley_principled_2006).
 The possibility of reelection can motivate them to exert effort and convince voters to reelect them.
-This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (@groseclose_buying_1996, @banks_buying_2000).
+This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (groseclose_buying_1996, banks_buying_2000).
 
@@
 Voters have beliefs about how competent a politician is and strictly prefer them over the alternative.
-Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (@coate_form_1995).
+Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (coate_form_1995).
 Additionally, I assume that the voter is sincere, excluding pivotality concerns and strategic voting from voters' utility calculations.
@@
 An important feature in the model is that high quality public services $\omega_{h}$ are produced probabilistically as a function of $\theta$, the quality of the bureaucracy.
-The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See @kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.
+The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.
 
@@
 This reduces the likelihood of securing the necessary resources to provide public goods for a given fiscal year.
-Adapting @gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic.
+Adapting gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic.
 Technocratic politicians invariably deliver high quality public goods to the voter.
@@
 
-% In Brazil, responsibility for primary education is delegated to local governments (@paschoal_historia_2009).
+% In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009).
 The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments.
@@
 % The local executive branch has sole prerogative over the hiring and firing of teachers and school principals.
-School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (@brollo_political_2017).
-Teachers usually enter the public service through a public exam and are eligible for tenure after two years (@gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount.
-Municipal teachers and school principals are overseen by a local department of education (@militao_educacao_2014).
+School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (brollo_political_2017).
+Teachers usually enter the public service through a public exam and are eligible for tenure after two years (gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount.
+Municipal teachers and school principals are overseen by a local department of education (militao_educacao_2014).
 
 % Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education.
-At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (@ferraz_corrupting_2012).
+At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).
 
@@
 The city council, on the other hand, oversees legislation and approves the budget for the fiscal year.
-As noted by @souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.
+As noted by souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.
 
-% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (@dantas_eleicoes_2013).
+% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013).
 These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
@@
 Using school census data, it is possible to track individual teachers across time and schools.
-With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by @pereira_junior_indicadores_2016.
+With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by pereira_junior_indicadores_2016.
 The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.
@@
 % For the first estimation, I employ a hierarchical linear model to estimate the effect of teacher turnover on average test scores for grade $i$ at school $s$, in municipality $j$ and year $t$.
-This estimation strategy has been used in multiple studies to analyze educational, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation (@diprete_multilevel_1994, @lee_using_2000.) Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise.
+This estimation strategy has been used in multiple studies to analyze educational, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation (diprete_multilevel_1994, lee_using_2000.) Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise.
 The main specification is as follows:
@@
 In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge.
-To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in @imai_covariate_2014 which is generalized for continuous treatments.
+To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in imai_covariate_2014 which is generalized for continuous treatments.
 A covariate balance test for either estimation is presented in the appendix.
@@
 The evidence therefore seems to point out that mayors with a weaker hold on the city council resort to greater patronage, with negative consequences for student learning.
-This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts (@gottlieb_countervailing_2019).
+This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts (gottlieb_countervailing_2019).
 

### 2020-09-08 paper/paper.tex
finish redrafting intro

@@
 
-In this paper, I highlight the role of the strategic interaction between political elites and its downstream effects on the provision of educational services.
-Similar to coalition building, local executive leaders secure broader support for their policy agenda through the allocation of public sector positions to city councilors.\citep{laver_coalitions_1990, power_optimism_2010} As highlighted by previous research, teacher turnover can lead to decreased morale and inexperienced staff entering public service \citep{ingersoll_teacher_2001, ronfeldt_how_2013}.
-Patronage, by inducing new hires or dismissals, can therefore cause instability in the educational sector, with negative effects for student learning.
+In many countries, particularly in the developing world, the quality of public education remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations.
+How can these educational outcomes be improved?
+Previous studies have found that incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
+Often relying on quasi-experimental research designs, these studies provide compelling evidence that actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers these public services and to whom do they respond?
 
-To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence.
-Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students.
-To validate these accounts I combine administrative data on education, both at a national and subnational level.
-I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory, and a school-grade-specific turnover index.
-A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.
+In this paper, I build on extensive fieldwork research across Brazil to argue that public education is captured by local political elites.
+Similar to coalition building in presidential contexts, mayors secure broader support for their policy agenda through the allocation of public sector positions to city councilors and their loyalists \citep{laver_coalitions_1990, power_optimism_2010}.
+This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}.
+By catering to the city council through patronage appointments, mayors ultimately cause instability in the local educational bureaucracy, with negative downstream consequences for public school students.
 
-% To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council.
+To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council.
 In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage, in particular during the first year of administration.
@@
 
-% This study contributes to an emerging literature on the politics of personnel and public services (pepinsky_bureaucracy_2017, finan_personnel_2015 and gulzar_politicians_2017).
-Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services (ferraz_corrupting_2012).
-It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy (kohli_state-directed_2004, evans_embedded_1995).
+To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence.
+Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students.
+To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning.
+I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory.
+Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff.
+A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.
 
-% The paper is structured as follows.
+This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015, gulzar_politicians_2017}.
+Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services \citep{ferraz_corrupting_2012}.
+It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy \citep{kohli_state-directed_2004, evans_embedded_1995}.
+
+The paper is structured as follows.
 Section 1 provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil.

### 2020-09-08 paper/paper.tex
refactor paper

@@
 
-In many countries, particularly in the developing world, the quality of public education remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations.
+The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations.
 How can these educational outcomes be improved?
@@
 
-% # Related literature and debate
+\section{Related literature}
 
-% In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions.
+In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions.
 I also highlight how my research incorporates multiplicity in political actors and how this affects bargaining over public sector jobs.
@@
 
-% ## Bureaucratic personnel and public goods provision
+\subsection*{Bureaucratic personnel and public goods provision}
 
-% Bureaucracies have a clear impact on the delivery of public services.
-A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (centeno_unpacking_2017, kohli_state-directed_2004).
-A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (johnson_miti_1982, kohli_state-directed_2004).
-A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (evans_bureaucracy_1999).
+Bureaucracies have a clear impact on the delivery of public services.
+A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions \citep{centeno_unpacking_2017, kohli_state-directed_2004}.
+A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy \citep{johnson_miti_1982, kohli_state-directed_2004}.
+A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth \citep{evans_bureaucracy_1999}.
 
-% Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference.
+Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference.
 Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores.
-gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment.
-akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.
+\citet{gulzar_politicians_2017} show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment.
+\citet{akhtari_political_2015}, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.
 
-% This recent wave of studies shed light on the intersection between politicians and bureaucracies.
+This recent wave of studies shed light on the intersection between politicians and bureaucracies.
 However, few of these studies explicitly model the multiple actors involved in managing bureaucracies.
@@
 
-% ## Presidential coalitionism and patronage
+\subsection*{Presidential coalitionism and patronage}
 
-% For every elected mayor in Brazil, a group of legislators are also elected into office.
+For every elected mayor in Brazil, a group of legislators are also elected into office.
 These political actors have competing claims over the local bureaucracy, with important implications for public service delivery.
-This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (grindle_jobs_2012, mccarty_appointments_2004).
+This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}.
 Divergent political interests can lead to strategic interaction between executive and legislative actors.
 A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy.
-(raile_executive_2011, power_optimism_2010).
+\citet{raile_executive_2011, power_optimism_2010}.
 
-% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism.
+In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism.
 Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
-(raile_executive_2011).
-In a setting characterized by weak party cohesion and programmatic commitments (ames_electoral_1995, lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999).
+\citep{raile_executive_2011}.
+In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.
 
-% \blue{something about coalitions}
+\blue{something about coalitions}
 
-% In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
-Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.^[Interview with C, August and September 2019.] Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
-As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."^[Interview with C, August 2019.] These hires induce changes in personnel, with important consequences for bureaucracies and educational services.
+In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
+Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
+As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.
 
-% ## Bureaucratic turnover and inexperienced education
+\subsection*{Bureaucratic turnover and inexperienced education}
 
-% Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects.
-As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (gailmard_slackers_2007).
-Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (clotfelter_teacher_2007.) akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores.
-When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (colonnelli_patronage_2017).
+Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects.
+As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population \citet{gailmard_slackers_2007}.
+Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher \citet{clotfelter_teacher_2007}.
+\citet{akhtari_political_2015} finds that students attending a school with a recently appointed school principal perform worse in standardized test scores.
+When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones \citet{colonnelli_patronage_2017}.
 
-% > "I am aware that the position is temporary.
+> "I am aware that the position is temporary.
 Especially because it is a political position, decided by the administration.
@@
 
-% Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level.
+Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level.
 In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers.
@@
 
-% # The argument
+\section{Institutional context and data}
 
-% ## The executive dilemma
+\subsection{Education in Brazil}
 
-% I model the executive decision over patronage as a product of political career concerns and strategic interaction with the local legislative power.
-In particular, mayors are assumed to be career-oriented and re-election seeking (besley_principled_2006).
-This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (stokes_brokers_2013, robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.
-
-% The mayor's strategy centers around the redistribution public sector jobs.
-When the mayor takes office, a set of bureaucratic positions become available for allocation to city councilors.
-In a municipal parallel to presidential coalitionism, jobs are allocated to legislators in order to buy their vote.
-I assume, for now, that job allocations have a negative effect on public services.
-The mayor can refrain from distributing jobs for vote buying, since reducing turnover helps retain expertise and improves the quality of public services.
-However, doing so raises the threat of defection by coalition members, which can lead to the withholding of budgetary approval.
-This is the executive's dilemma.
-
-% The main proposition is that mayors who control a smaller (larger) share of legislative seats allocate more (less) public sector job to city councilors.
-The intuition is as follows.
-Mayors need to secure a simple majority of legislative votes to have the budget approved.
-groseclose_buying_1996 and banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor.
-Having exhausted these votes, the executive has to resort to buying votes from the opposition, which demand more patronage.
-The less seats occupied by members of the mayor's electoral coalition, the more expensive vote-buying becomes, and the observable implication is increased patronage in the municipality.
-
-% The model builds on rich theoretical treatment of electoral accountability and majority-buying in legislatures.
-Starting with barro_control_1973 and ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (coate_form_1995; besley_principled_2006).
-The possibility of reelection can motivate them to exert effort and convince voters to reelect them.
-This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (groseclose_buying_1996, banks_buying_2000).
-
-% ## The model
-
-% There is one mayor $m$, a city council $N$ and representative median voter $v$.
-The goal of the mayor is to be reelected into office, yet for any given time period $m$ needs to garner legislative support.
-The city council plays that key role by approving or not the budget for the executive.
-The legislative chamber is composed by $N = {1,..,n}$ legislators, where $S \subseteq N$ denote the subset of legislators who belong to the electoral coalition of mayor $m$.
-For each $i \in N$, let $\textbf{v} = \{v_1,...,v_n\}$ describe legislator i's preference to support the mayor in the budgetary process.
-Let $v_i > v_j$ for all $i \in S$ and $j \in \{N \setminus S\}$.
-Therefore, members of the mayor's electoral coalition are more likely to approve the budget than outside members.
-
-% I introduce a set of simplifying assumptions.
-First, mayors are one of two types: a technocratic politician who delivers high quality public goods, and a pragmatic one, who fails to do so.
-Note that this is an informational heuristic.
-Voters have beliefs about how competent a politician is and strictly prefer them over the alternative.
-Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (coate_form_1995).
-Additionally, I assume that the voter is sincere, excluding pivotality concerns and strategic voting from voters' utility calculations.
-Finally, I normalize the politician's exit option to 0: the politician always prefers to be re-elected.
-
-% The game has two periods.
-Voter $v$ does not observe the amount of patronage handed out by mayor $m$ nor her type, but observes and derives utility from the realization of public good quality $\omega_t$.
-An important feature in the model is that high quality public services $\omega_{h}$ are produced probabilistically as a function of $\theta$, the quality of the bureaucracy.
-The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.
-
-% The mayor's optimal strategy is pinned down by the share of patronage which maximizes expected utility from holding onto office, conditional on her type and expected benefits from reelection.
-Mayor $m$ has to offer public sector jobs to coopt the opposition while ensuring that the voter is satisfied enough to reelect her.
-The executive dilemma is how to reconcile the incentive for reelection and the pressure imposed by the opposition.
-
-% The timing of the game is as follows.
-
-% 1) Nature draws the politician's type.
-% 2) Politician observes her type and allocates public sector jobs to legislators, setting $\theta_1$.
-% 4) Nature realizes public good $\omega_1(\theta_1)$.
-% 5) Voter observes $\omega_1$ and casts vote to retain or fire the incumbent.
-% 6) If incumbent is retained, she sets $\theta_2$ and nature draws $\omega_2(\theta_2)$. If incumbent is deposed, a challenger takes office, sets $\theta_2$ and the game ends.
-
-% ## The executive dilemma
-
-% The mayor's choice set is the bureaucratic quality parameter $\theta_t$.
-As quality increases - the mayor allocates a smaller proportion of jobs to legislators - the less votes become secure in the legislature.
-This reduces the likelihood of securing the necessary resources to provide public goods for a given fiscal year.
-Adapting gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic.
-Technocratic politicians invariably deliver high quality public goods to the voter.
-Pragmatic politicians, on the other hand, are susceptible to legislative pressure and at times fail to deliver them.
-
-% The probability of a politician being technocratic is common knowledge: $\xi \in [0,1]$.
-Conversely, the probability of her being pragmatic is simply $1 - \xi$.
-Both types derive utility from public goods provision $\omega$, where for simplicity $u(\omega_l)=0$ and $u(\omega_h) = 1$.
-We do not model the decision of the competent mayor: she simply delivers high quality goods and is inevitably reelected.
-We focus on the more interesting case of the pragmatic mayor.
-
-% \begin{align*}
-% Eu_m \left(\theta_t, s_1 \middle|\lambda \right) &= \textbf{E}(\omega_1) + \delta \text{Pr}(\text{reelection})\left[\psi + \textbf{E}(\omega_2)  \right] \\
-% &= \frac{1-s_1\theta_1}{1+e^{-(10\theta_1-5)}} + \delta \text{Pr}(\text{reelection})\left[\psi + \frac{1-s_2\theta_2}{1+e^{-(10\theta_2-5)}} \right] \\
-% \end{align*}
-
-% The equation above formalizes the mayor's dilemma.
-The probability of reelection decreases with the share of jobs allocated to the opposition, since it lowers the quality of the bureaucracy.
-However, if the mayor fails to appease them through patronage, it is increasingly costly to "govern".
-While technocrats are immune to these political pressures, pragmatists ultimately cave in.
-The need to coopt legislators decreases the quality of the bureaucracy and public goods, as mayors are forced to allocate positions within the bureaucracy to garner support in the legislative.
-Details on deriving the equilibrium strategy are in the appendix.
-I now describe Brazil as a case for testing the theory.
-
-% # Institutional context and data
-
-% ## Education in Brazil
-
-% In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009).
+In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009).
 The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments.
@@
 
-% \blue{plots/enrollment_dep.pdf}
+\blue{plots/enrollment_dep.pdf}
 
-% The local executive branch has sole prerogative over the hiring and firing of teachers and school principals.
+The local executive branch has sole prerogative over the hiring and firing of teachers and school principals.
 School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (brollo_political_2017).
@@
 
-% Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education.
+Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education.
 At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).
 
-% ## Local governance and mayoral coalitions
+\subsection{Local governance and mayoral coalitions}
 
-% Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council.
+Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council.
 Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016.
@@
 
-% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013).
+In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013).
 These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
@@
 
-% Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions.
+Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions.
 While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
@@
 
-% > School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
+> School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
 
-% > Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]
+> Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]
 
-% These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff.
+These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff.
 To verify these claims in a broader set of municipalities, as well as outlining the research design, I employ extensive administrative and electoral data collected by the Brazilian federal government.
 In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.
+
+\section{Theory}
+
+In this section, I apply a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes.
+For any given electoral term, a mayor wants to enact her preferred policy, but to do so needs to pass it through a simple majority in the city council.
+To ensure victory, the mayor offers appointment positions to voters, in this case, city councilors, who in exchange vote in favor of the bill.
+
+In this setting, passing a policy proposal requires outbidding the opposition, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
+In particular, the amount of resources that the opposition controls can force the mayor's hands and increase the amount of patronage necessary to pass the bill.
+We pursue these empirical implications of the model further in our results section.
+
+The amount of public sector positions the mayor offers to these city councilors also responds to the composition of the city council.
+The intuition is straightforward: if the city councilors are already likely to support the bill, there is no need to offer them additional side payments.
+Note that there is no claim about a dichotomy of clientelism, in which there is or is not patronage \citep{stokes_brokers_2013, grindle_jobs_2012}, but rather a variation in degrees: how many appointments must the mayor give out in order to govern on her preferred policy.
+
+For the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
+I conclude with a discussion of the model and how it applies to the Brazilian municipal context.
+
+\subsection*{The Setting}
+
+The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.\footnote{Note that I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
+In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators.
+The total amount of political resources available is $W_G$ and $W_O$, which for the mayor includes public sector appointments.
+
+Each city councilor is characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ denotes that the mayor's proposal $x$ is preferred by legislator $i$.
+Let $\mathbf{v} = (v_1, ..., v_n)$ describe the preference profile for the city councilors.
+In this setting, $v_i$ measures the degree to which an individual city councilor supports the mayor, with higher values of $v$ denoting stronger support for the mayor and vice versa.
+Payoff are realized when city councilor $i$ votes, independent of the outcome of the voting procedure.
+This sincere voting preference closes the possibility of general equilibria in which $i$'s voting behavior affect $j$.
+
+We solve the game through backward induction. The timing of the game is as follows:
+
+\begin{enumerate}
+    \item Government $G$ offers a bribe schedule $a \in (a_1, ..., a_n) \in \real$.
+    \item Opposition $O$ observes the bribe schedule $m$ and makes a counter-offer $b \in (b_1, ..., b_n) \in \real$.
+    \item City councilors cast their votes at the end of bribing period.
+    \item Nature sums legislative votes, legislative outcome is decided and payoffs are realized.
+\end{enumerate}
+
+Given a bribe schedule $(a, b)$, councilor $i$ prefers to vote for the mayor's proposal $x$ if $a_i + v_i > b_i$ and the status quo $y$ otherwise.
+Since indifferent councilors vote for the status quo, the opposition needs to only match bribes from $M$, adjusting for individual preferences, i.e.
+$b_i = a_i + v_i$.
+For the mayor, she needs to construct the cheapest winning coalition in order to defeat the opposition.
+
+Following Groseclose and Snyder (1996) and Banks (2000) we focus our analysis on the set of equilibria in which the mayor wins.\footnote{Since strategies for both players are symmetrical, any set of equilibria in which the mayor loses can be modeled as cases in which the the opposition loses.} In this context, the amount of patronage resources $W_G$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$.
+Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(a) = \sum_{i = 1}^n a_i$ be the total amount of patronage disbursed.
+The mayor then solves
+
+\begin{equation}
+    \label{eqn:solution}
+    \min_a\{S(a) : a \in U(\mathbf{v}, W_B) \}
+\end{equation}
+
+Note that for any equilibrium strategy, it must be the case that mayor $M$ uses a leveling schedule: every city councilor in her coalition $C$ is equally expensive for the opposition $O$ to bribe.
+More formally, for any $a \in \mathbb{R}_+^n$, let $C(a) = \{i \in N : a_i > 0\}$ denote the set of individuals who receive a bribe from from the government $G$.
+One can show that there is a bribe schedule $a'$ such that for any $i,j \in C(a)$, $a'_i + v_i = a'_j + v_j$.
+The intuition is that the mayor has no incentive to make voters differentially expensive, because the opposition $O$ will simply ignore the more expensive voters and target the least favorable members of the coalition.
+We refer to these strategies as leveling schedules.
+
+We can characterize the set of equilibria in the game by introducing additional notation.
+Let $U^l(\mathbf{v}, W_O) \subseteq U(\mathbf{v}, W_O)$ denote the set of unbeatable leveling schedules.
+These are bribe schedules such that $a_i + v_i = a_j + v_j  \equiv t(a)$.
+The bribe $a_i = t(a) - v_i$ is the sum of two terms.
+The first is the common "transfer" among all voters in $C(a)$, the second ($-v_i$) is individual specific.
+The latter term makes voters indifferent between $x$ and $y$ absent any bribe from $B$; the former represents the per capita amount necessary to make $C(a)$, together with any unbribed voters, unaffordable for $B$.
+
+I impose the following two assumptions:
+\begin{align*}
+    A1 &: v_{(n+1)/2} < 0\\
+    A2 &: v_1 < 2W_B/(n+1) 
+\end{align*}
+$A1$ implies that absent any bribes by $A$, $y$ will defeat $x$.
+Therefore $A$ must bribe at least one voter.
+$A2$ implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win.
+$A2$ also implies that for all $a \in U^l(\mathbf{v}, W_B)$, it must be that $t(a) \geq 2W_B/(n + 1)$, otherwise $B$ can bribe a majority from $C(a)$ itself and win the vote.
+
+% For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$.
+Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$.
+Then, under $A2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$: $a'_i = t(a) - v_i$, $a'_j = 0$ and $a'_m = a_m$ for all $m \notin i, j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
+$a'_i \leq a_j$.
+Also, because of $A2$ $a'_j \leq a_j$, since it guarantees $a_j$ is non-negative.} Generalizing, and recalling that $v_1 \geq ...
+\geq v_n$, we see that for all $a \in U^l(\mathbf{v}, W_b)$ there exists a bribe schedule $a' \in U^l(\mathbf{v}, W_b)$ such that $S(a') \leq S(a)$ and $C(a') = \{1, ..., k(a)\}$.
+Therefore, we can without loss of generality restrict attention to schedules $a$ by $A$ which bribe the first $k(a)$ voters.
+Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.
+
+These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{A detailed explanation can be found in the appendix.} We can simplify the total expenditure on patronage by the government, $S(a)$, as
+
+\begin{equation*}
+    S(a) = \sum_{i \in C(a)}a_i = k(a) \cdot t(a) - \sum_{i \leq k(a)}v_i
+\end{equation*}
+
+Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B)$.
+We can thus fully characterize the optimization problem of $A$ in equation \ref{eqn:solution} as
+
+\begin{equation*}
+    \min_{k,t} k \cdot t- \sum_{i \leq k} v_i
+\end{equation*}
+
+subject to the constraint that the induced schedule $a \in U_m^l$.
+We can reformulate this as an unconstrained problem by noting the following.
+First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$.
+Therefore, $B$ receives all non-bribed voters for free.
+For $a(k, t, \mathbf{v})$ to be unbeatable, then, it must be that $B$ cannot afford the remaining $(n + 1)/2 - (n - k) = k - (n - 1)/2$ voters, or
+
+$$t \cdot (k - (n - 1)/2) \geq W_B$$
+
+Solving this for equality yields the optimal transfer from $A$ to members of $C(A) = \{1, ..., k\}$, conditional on k:
+
+\begin{align}
+    \label{eqn:common_transfer}
+    t(k, W_B) = \frac{W_B}{k - (n - 1)/2}
+\end{align}
+
+Defining minimal winning expenditures as
+
+\begin{align}
+    \label{eqn:total_expenditure}
+    E(k, \mathbf{v}, W_B) = k \cdot t(k, W_B) - \sum_{i \leq k} v_i
+\end{align}
+
+we can state $A$'s problem as
+
+\begin{align}
+\label{eqn:minimal_expenditure}
+\min_k \{E(k, \mathbf{v}, W_B) : k \in {(n + 1/2), ..., n}\}
+\end{align}
+
+Denote the solution to expression \ref{eqn:minimal_expenditure} as $k^*(\mathbf{v}, W_B)$.
+This solution implicitly generates a solution to expression \ref{eqn:solution} through expression \ref{eqn:total_expenditure} and the induced bribe schedule above.
+Therefore, characterizing the optimal $k^*$ is sufficient to fully characterize the optimal behavior of the mayor.
+
+\subsection*{Results}
+
+First, characterize a solution for $k^*$.
+Because $k$ is finite, calculus cannot be employed.
+Instead, we deploy a discrete version of these techniques.
+First let's define $\Delta(k) = E(k + 1) - E(k)$ as the difference in expenditure from adding another coalition member.
+Note that if $\Delta(k) \geq 0$ then $A$ does not want to add another member to the coalition.
+Conversely, if $\Delta(k) < 0$, then $A$ is strictly better off by adding the $k + 1$th member of the coalition.
+
+Next, suppose that $\Delta(k)$ is increasing in $k$.
+The following algorithm can then be used to identify $k^*$: if $\Delta((n + 1)/2) \geq 0$, then we know from $\Delta(k)$ increasing that $A$ is better off by setting $k^*$ to $(n + 1)/2$.
+If $\Delta((n + 1)/2) < 0$, then we know that $k^*$ must be greater than $(n + 1)/2$, so we next solve for $\Delta((n + 3)/2)$, and so on.
+
+We can therefore search for the optimal $k^*$ with the following algorithm:
+
+\begin{align}
+    \label{eqn:optimal_k}
+    k^* = 
+    \begin{cases}
+        (n + 1)/2 & \text{if } \Delta((n + 1)/2) \geq 0 \\
+        \max\{k : \Delta(k - 1) < 0\} & \text{otherwise}
+    \end{cases}
+\end{align}
+
+We can also further characterize the change in minimum winning expenditures in equation \ref{eqn:total_expenditure} as
+
+\begin{align}
+    \Delta(k)  & = \left[\frac{(k + 1)W_B}{k + 1 - (n - 1)/2} - \sum_{i \leq k + 1}v_i \right]\\
+    \label{eqn:deltAk}
+    & = \frac{-W_B (n - 1)}{2(k + 1 - (n - 1)/2)(k - (n - 1)/2)} - v_{k + 1}\\
+    \label{eqn:def_deltAk}
+    & \equiv T(k, W_B) - v_{k + 1}
+\end{align}
+
+Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.
+
+\begin{proposition}
+    (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
+\end{proposition}
+
+Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity.
+Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v'}, W_B)$.
+If $k' = (n + 1)/2$, then we know that $k' \leq k^*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.
+
+From equation \ref{eqn:optimal_k} we infer that $\Delta(k' - 1, \mathbf{v}', W_B) < 0$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$.
+Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$.
+Then, $v_{k'} > T(k' - 1, W_B)$, and hence $\Delta(k' - 1, \mathbf{v}, W_B) < 0$.
+But then from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$.
+Therefore, the following holds:
+
+\begin{proposition}
+    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
+\end{proposition}
+
+In words, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases.
+Substantively, the number of voters bribed by $A$ weakly increases as the voter who receives the largest bribe finds $A$'s preferred alternative, $x$, more attractive.
+Similarly
+
+\begin{proposition}
+    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k' + 1} \leq v'_{k' + 1}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \leq k^*(\mathbf{v}', W_B)$
+\end{proposition}
+
+The ``convexity" of $E$ guarantees that local information is sufficient to generate comparative statistics regarding changes in preferences $\mathbf{v}' \rightarrow \mathbf{v}$.
+We can characterize the change in total expenditures as a result of a shift in voter preferences.
+Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$.
+From equation \ref{eqn:total_expenditure} we have
+
+\begin{align*}
+    E(k, \mathbf{v}, W_B) - E(k, \mathbf{v}', W_B) & = \\
+    &= k \cdot t(k, W_B) - \sum_{i \leq k} v_i - \left[k \cdot t(k, W_B) - \sum_{i \leq k} v'_i\right]\\
+    &= \sum_{i \leq k}(v'_i - v_i)
+\end{align*}
+
+Since $v'_i - v_i \leq 0$, the difference in expenditure between moving from a favorable to a less favorable legislature is always non-positive, i.e.
+the government has to spend less resources to pass her preferred policy.
+This holds despite the fact that when these preferences shift there is an increase in the overall size of the coalition.
+This result has a similar flavor to \citet{groseclose_1996_buying}, who motivate their model by stating that it may be optimal to increase the size of the coalition (instead of buying a simple majority) because doing so overall can lead to a reduction in the amount of expenditures by the vote-buyer.
+
+\subsection*{Discussion}
+
+Enacting policy requires the exchange of political currency for votes.
+Whether it be in presidential coalitionism, or in the local city council politics, mayors who wish to govern have to engage in transactions with the legislature.
+What I showed in this section was that political misalignment between the government and the legislature can in fact be counterproductive: more patronage occurs, leading to worse public service outcomes.
+
+The model also highlights a key aspect of clientelism that is often neglected electoral accountability models: voters have a limited voice.
+Ultimately, the exchanges which occur between the legislature and the mayor have little to do with the voter at the end of the pipeline, and more to do with the city councilors.
+The first order requirement for the government is to ensure that it has enough legislative votes in order to enact the very policies that the voter may or may not desire.
+This transactional cost is not illegal: rather, it is necessary for democratic relations between different branches of government.
+
+In the next section, I test whether the empirical implication of the model is correct: does more patronage occur in municipalities with greater political misalignment between the mayors and city councilors.
+I test additional implications of the model, including whether shifts in the resources controlled by the opposition can affect the government's patronage strategy.
 

### 2020-09-08 paper/paper.tex
introduce literature and data

@@
 
-> "I am aware that the position is temporary.
+\begin{quote}
+"I am aware that the position is temporary.
 Especially because it is a political position, decided by the administration.
 If the current administration is out of power, we are automatically dismissed." - Interview with school principal A, August 2019.
+\end{quote}
 
@@
 
-In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009).
+In Brazil, responsibility for primary education is delegated to local governments \citet{paschoal_historia_2009}.
 The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments.
@@
 
-\blue{plots/enrollment_dep.pdf}
+\begin{figure}[h]
+    \label{fig:enrollment}
+    \includegraphics{plots/descriptive_enrollment.pdf}
+    \centering
+\end{figure}
+% \blue{plots/enrollment_dep.pdf}
 
-The local executive branch has sole prerogative over the hiring and firing of teachers and school principals.
-School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (brollo_political_2017).
+% The local executive branch has sole prerogative over the hiring and firing of teachers and school principals.
+School principals are considered positions of trust (\emph{cargos de confiança}) and usually appointed by the mayor or secretary of education \citet{brollo_political_2017}.
 Teachers usually enter the public service through a public exam and are eligible for tenure after two years (gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount.
-Municipal teachers and school principals are overseen by a local department of education (militao_educacao_2014).
+Municipal teachers and school principals are overseen by a local department of education \citet{militao_educacao_2014}.
 
-Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education.
-At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).
+% Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education.
+At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see \href{https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb}{here}.
+While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).
 
-\subsection{Local governance and mayoral coalitions}
+% \subsection{Local governance and mayoral coalitions}
 
-Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council.
+% Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council.
 Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016.
@@
 
-In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013).
+% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013).
 These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
@@
 
-Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions.
+% Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions.
 While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
@@
 
-> School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
+% > School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
 
-> Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]
+% > Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]
 
-These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff.
+% These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff.
 To verify these claims in a broader set of municipalities, as well as outlining the research design, I employ extensive administrative and electoral data collected by the Brazilian federal government.
@@
 
-\section{Theory}
+% \section{Theory}
 
-In this section, I apply a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes.
+% In this section, I apply a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes.
 For any given electoral term, a mayor wants to enact her preferred policy, but to do so needs to pass it through a simple majority in the city council.
@@
 
-In this setting, passing a policy proposal requires outbidding the opposition, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
+% In this setting, passing a policy proposal requires outbidding the opposition, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
 In particular, the amount of resources that the opposition controls can force the mayor's hands and increase the amount of patronage necessary to pass the bill.
@@
 
-The amount of public sector positions the mayor offers to these city councilors also responds to the composition of the city council.
+% The amount of public sector positions the mayor offers to these city councilors also responds to the composition of the city council.
 The intuition is straightforward: if the city councilors are already likely to support the bill, there is no need to offer them additional side payments.
@@
 
-For the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
+% For the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
 I conclude with a discussion of the model and how it applies to the Brazilian municipal context.

### 2020-09-09 paper/paper.tex
incorporate results and estimations

--- 2020-09-08 paper/paper.tex
@@
 \usepackage{amsmath}
+\usepackage{hyperref}
+
+\hypersetup{
+    colorlinks=true,
+    linkcolor=blue,
+    filecolor=blue,      
+    urlcolor=blue,
+    citecolor=blue
+}
+
 \DeclareMathOperator*{\argmax}{arg\,max}
@@
 Previous studies have found that incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
-Often relying on quasi-experimental research designs, these studies provide compelling evidence that actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers these public services and to whom do they respond?
+Often relying on quasi-experimental research designs, these studies provide compelling evidence that actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?
 
-In this paper, I build on extensive fieldwork research across Brazil to argue that public education is captured by local political elites.
-Similar to coalition building in presidential contexts, mayors secure broader support for their policy agenda through the allocation of public sector positions to city councilors and their loyalists \citep{laver_coalitions_1990, power_optimism_2010}.
+In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to capture by local political elites.
+Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}.
 This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}.
@@
 
-\blue{something about coalitions}
-
 In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
@@
 
-In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009).
+In Brazil, responsibility for primary education is delegated to local governments \citet{paschoal_historia_2009}.
 The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments.
-Figure \ref{fig:enrollment} plots the total number of students in primary education per government level, including private schools.
+Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools.
 As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.
 
-\blue{plots/enrollment_dep.pdf}
+\begin{figure}[h]
+    \includegraphics{plots/descriptive_enrollment.pdf}
+    \caption{enrollment}
+    \label{fig:enroll}
+    \centering
+\end{figure}
 
-The local executive branch has sole prerogative over the hiring and firing of teachers and school principals.
-School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (brollo_political_2017).
-Teachers usually enter the public service through a public exam and are eligible for tenure after two years (gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount.
-Municipal teachers and school principals are overseen by a local department of education (militao_educacao_2014).
+The local executive branch has sole jurisdiction over the hiring and firing of teachers and school principals.
+School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education \citet{brollo_victor_2017}.
+Teachers usually enter the public service through a public exam and are eligible for tenure after two years \citet{gatti_formacao_2010}.
+However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount.
+Municipal teachers and school principals are overseen by a local department of education \citet{militao_educacao_2014}.
 
 Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education.
-At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).
+At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.\footnote{For more details on the Fundeb, see \hyperlink{https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb}{here}}.
+While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion \citet{ferraz_corrupting_2012}.
 
@@
 
-Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council.
-Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016.
-These take place ] The executive is responsible for the implementation of social policies such as education, as well as drafting the municipal budget.
+Brazil's local level government is composed by over 5 thousand municipalities, each with their own mayor hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
+Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.\footnote{The last municipal elections were in 2016.
+These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over personnel apppointment.
 The city council, on the other hand, oversees legislation and approves the budget for the fiscal year.
-As noted by souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.
+As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.
 
-In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013).
+In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels \citet{dantas_eleicoes_2013}.
 These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
 Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}).
-While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.^[Interview with C, former chief of staff of municipality J.
-September 2018.]
+While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.\footnote{Interview with C, former chief of staff of municipality J.
+September 2018.}
 
-Interviews with secretaries of education and mayors suggest that city councilors play an important role in staffing decisions.
+Interviews with secretaries of education and mayors provide evidence that city councilors play an important role in staffing decisions.
 While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
@@
 
-> School principal: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
+\begin{quote}
+    \emph{School principal}: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
 
-> Educational counselor: I was also invited to work here, by the city councilor and the department of education.^[Interview with the administrative board of school A., municipality J. August 2019]
+    \emph{Educational counselor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J.
+August 2019.}
+\end{quote}
 
 These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff.
-To verify these claims in a broader set of municipalities, as well as outlining the research design, I employ extensive administrative and electoral data collected by the Brazilian federal government.
+To verify these claims in a broader set of municipalities, as well as outlining the research design, I apply a theory of legislative vote-buying that illustrates the  employ extensive administrative and electoral data collected by the Brazilian federal government.
 In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.
 
+\subsection{Data}
+
+Brazil collects fine-grained data on education and makes it publicly available for research.
+For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools.
+In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese.
+In this study, test scores are the primary metric for assessing the quality of education received by students.
+
+\begin{figure}[h]
+    \includegraphics{plots/saeb_map.pdf}
+    \caption{hello}
+    \label{fig:saeb_map}
+\end{figure}
+
+Figure \ref{fig:saeb_map} presents a map of Brazil with the municipal averages for the SAEB of 2015.
+Warmer colors such as red and orange indicate higher average scores, with the converse denoted by colder, blue shades.
+There is clear variation in average test scores, with the Southeast and Midwest outperforming poorer regions in the North and Northeast.
+Even within regions, however, there is wide heterogeneity.
+In particular, note that in the northern part of Brazil, high-performing municipalities neighbor low-performing ones.
+This indicates that despite spatial clustering, municipal factors can lead to variation in the quality of education.
+
+Employment data on educational staff, including teachers and school principals, are extracted from the RAIS, SAEB and the School Census.\footnote{RAIS data, along with additional Brazilian employment data, can be accessed \hyperlink{http://trabalho.gov.br/dados-abertos}{here}.} The \emph{Relatório Anual de Informações Sociais} (RAIS) is an annual census administered by the now Ministry of Finance to all formal employers in Brazil.
+Employers are mandated to accurately report data on employees, subject to fines for misreporting.
+Subnational governments, including municipalities, also report to the RAIS.
+The dataset therefore contains micro-level information on all municipal personnel, including when they were hired/fired, as well as wages, type of contract and education levels.
+
+Figure \ref{fig:staff_turnover} provides descriptive statistics on bureaucratic personnel in Brazil, segmented by department.
+Educational staff, in this case school principals and teachers, comprise approximately a third of municipal public sector jobs, totaling around 2 million in 2015.
+While comparable to administrative staff, this total significantly exceeds that for healthcare services.
+Focusing on turnover, we note that new hiring and dismissals in education staff is relatively high.
+While lower than that for administrative staff, new hires can represent over 10 percent of extant staff.
+The past two decades was a moment of rapid expansion of municipal staff, and hiring has far exceeded dismissals in that time period.
+
+\begin{figure}[h]
+    \includegraphics{plots/descriptive_staff_turnover.pdf}
+    \label{fig:staff_turnover}
+    \caption{hello}
+\end{figure}
+
+I propose three different specifications for measuring bureaucratic turnover.
+From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year.
+Using school census data, it is possible to track individual teachers across time and schools.
+With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}.
+The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.
+
+$$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$
+
+Information on mayor, city councilors and mayoral coalitions are made available by the Supreme Electoral Court (TSE), the national authority responsible for overseeing elections.\footnote{Available \hyperlink{http://www.tse.jus.br/eleicoes/estatisticas/repositorio-de-dados-eleitorais-1}{here}.} In order to calculate the share of legislative seats held by the mayoral coalition, I match the partisanship of each city councilor to the mayoral electoral coalition.
+The distribution of share of coalition seats is depicted in figure \ref{fig:hist_coalition}.
+Note that although most city councils are controlled by the mayor, in over 40 percent of municipalities the mayoral coalition is a minority government.
+
+\begin{figure}
+    \includegraphics{plots/hist_coalition.pdf}
+    \caption{hello}
+    \label{fig:hist_coalition}
+\end{figure}
+
+Additional data has been collected to supplement the estimation exercise.
+Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\url{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}, (2)\url{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}, (3)\url{https://www.seduc.ce.gov.br/spaece/}.}
+
 \section{Theory}
-
-In this section, I apply a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes.
-For any given electoral term, a mayor wants to enact her preferred policy, but to do so needs to pass it through a simple majority in the city council.
-To ensure victory, the mayor offers appointment positions to voters, in this case, city councilors, who in exchange vote in favor of the bill.
-
-In this setting, passing a policy proposal requires outbidding the opposition, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
-In particular, the amount of resources that the opposition controls can force the mayor's hands and increase the amount of patronage necessary to pass the bill.
-We pursue these empirical implications of the model further in our results section.
-
-The amount of public sector positions the mayor offers to these city councilors also responds to the composition of the city council.
-The intuition is straightforward: if the city councilors are already likely to support the bill, there is no need to offer them additional side payments.
-Note that there is no claim about a dichotomy of clientelism, in which there is or is not patronage \citep{stokes_brokers_2013, grindle_jobs_2012}, but rather a variation in degrees: how many appointments must the mayor give out in order to govern on her preferred policy.
-
-For the remainder of this section, I outline the setting of the model and derive key comparative statics that guide the empirical estimation of this paper.
-I conclude with a discussion of the model and how it applies to the Brazilian municipal context.
+\label{sec:theory}
 
@@
 
-The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.\footnote{Note that I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
+The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.\footnote{Note that for this model, I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
 In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators.
@@
 
-% ## Data
+\section{Empirical results}
 
-% Brazil collects fine-grained data on education and makes it publicly available for research.
-For this study, data on education are collected from two main sources: the SAEB and School Census.^[These can be accessed (here)[http://portal.inep.gov.br/web/guest/dados].] The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools.
-In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese.
-In this study, test scores are the primary metric for assessing the quality of education received by students.
+\subsection*{Model specification}
 
-% \blue{"plots/saeb_map.pdf"}
-
-% Figure \ref{fig:map} presents a map of Brazil with the municipal averages for the SAEB of 2015.
-Warmer colors such as red and orange indicate higher average scores, with the converse denoted by colder, blue shades.
-There is clear variation in average test scores, with the Southeast and Midwest outperforming poorer regions in the North and Northeast.
-Even within regions, however, there is wide heterogeneity.
-In particular, note that in the northern part of Brazil, high-performing municipalities neighbor low-performing ones.
-This indicates that despite spatial clustering, municipal factors can lead to variation in the quality of education.
-
-% Employment data on educational staff, including teachers and school principals, are extracted from the RAIS, SAEB and the School Census.^[RAIS data, along with additional Brazilian employment data, can be accessed (here)[http://trabalho.gov.br/dados-abertos].] The \emph{Relatório Anual de Informações Sociais} (RAIS) is an annual census administered by the now Ministry of Finance to all formal employers in Brazil.
-Employers are mandated to accurately report data on employees, subject to fines for misreporting.
-Subnational governments, including municipalities, also report to the RAIS.
-The dataset therefore contains micro-level information on all municipal personnel, including when they were hired/fired, as well as wages, type of contract and education levels.
-
-% Figure \ref{fig:staff} provides descriptive statistics on bureaucratic personnel in Brazil, segmented by department.
-Educational staff, in this case school principals and teachers, comprise approximately a third of municipal public sector jobs, totaling around 2 million in 2015.
-While comparable to administrative staff, this total significantly exceeds that for healthcare services.
-Focusing on turnover, we note that new hiring and dismissals in education staff is relatively high.
-While lower than that for administrative staff, new hires can represent over 10 percent of extant staff.
-The past two decades was a moment of rapid expansion of municipal staff, and hiring has far exceeded dismissals in that time period.
-
-% \blue{plots/turnover_category.pdf}
-
-% I propose three different specifications for measuring bureaucratic turnover.
-From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year.
-Using school census data, it is possible to track individual teachers across time and schools.
-With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by pereira_junior_indicadores_2016.
-The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.
-
-% $$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$
-
-% Information on mayor, city councilors and mayoral coalitions are made available by the Supreme Electoral Court (TSE), the national authority responsible for overseeing elections.^[Available (here)[http://www.tse.jus.br/eleicoes/estatisticas/repositorio-de-dados-eleitorais-1].] In order to calculate the share of legislative seats held by the mayoral coalition, I match the partisanship of each city councilor to the mayoral electoral coalition.
-The distribution of share of coalition seats is depicted in figure \ref{fig:coalition}.
-Note that although most city councils are controlled by the mayor, in over 40 percent of municipalities the mayoral coalition is a minority government.
-
-% \blue{plots of}
-
-% Additional data has been collected to supplement the estimation exercise.
-Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.^[Respectively, data can be found here: (1)[https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html], (2)[https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf], (3)[https://www.seduc.ce.gov.br/spaece/].]
-
-% # Estimation
-
-% ## Empirical strategy
-
-% This study presents two sets of estimations.
+This study presents two sets of estimations.
 The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education.
@@
 
-% For the first estimation, I employ a hierarchical linear model to estimate the effect of teacher turnover on average test scores for grade $i$ at school $s$, in municipality $j$ and year $t$.
-This estimation strategy has been used in multiple studies to analyze educational, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation (diprete_multilevel_1994, lee_using_2000.) Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise.
+For the first estimation, I test the hypothesis that mayors who control less seats in the city council engage in more patronage.
+To do so I leverage employment data from RAIS and school census data to measure teacher turnover rates.
+I deploy three sets of models.
+For the second and third estimation I turn to the micro, bureaucrat-level and a then to aggregated turnover at the municipal level.
+The first model is a logistic regression where the outcome of interest is whether a particular teacher or school principal is hired/fired for any given year.
+The second model is a linear model with fixed effects, where the outcome of interest is the proportion of educational staff hired/fired for a muncipality at any given year.
+For the second set of estimations, the main specification is:
+
+$$\text{turnover}_{jt} = \gamma \text{coalition seats}_{jst} + \mu P_{jt} + \zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{jt}$$
+
+The share of coalition seats held by the mayoral coalition is the treatment in this set-up.
+We are interested in $\gamma$, the change in bureaucratic turnover associated with variation in the share of legislative seats held by the mayoral coalition.
+Municipal characteristics $W_{jt}$ are similar to the ones used for the first set of estimation.
+I add political variables to the estimation in order to control for differences in mayor partisanship, incumbency status, and individual characteristics of the mayor: professional background, education, and age.
+
+To estimate downstream effects of turnover on education, I employ a hierarchical linear model to estimate the effect of teacher turnover on average test scores for grade $i$ at school $s$, in municipality $j$ and year $t$.
+This estimation strategy has been used in multiple studies to analyze educational outcomes, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation \citep{diprete_multilevel_1994, lee_using_2000} Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise.
 The main specification is as follows:
-% \begin{align*}
-% \text{test score}_{isjt} = \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
-% \psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
-% \end{align*}
-% We are interested in $\beta_1$ and $\beta_3$, the change in test scores associated with teacher or school principal turnover.
+
+\begin{align*}
+\text{test score}_{isjt} = \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
+\psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
+\end{align*}
+
+We are interested in $\beta_1$ and $\beta_3$, the change in test scores associated with teacher or school principal turnover.
 In this set-up, $\beta_3$ is the additional effect of staff turnover on test scores if students are in 9th grade.
@@
 
-% The second set of estimations tests the hypothesis that mayors who control less seats in the city council engage in more patronage.
-To do so I leverage employment data from RAIS and school census data to measure teacher turnover rates.
-I deploy three sets of models.
-For the second and third estimation I turn to the micro, bureaucrat-level and a then to aggregated turnover at the municipal level.
-The first model is a logistic regression where the outcome of interest is whether a particular teacher or school principal is hired/fired for any given year.
-The second model is a linear model with fixed effects, where the outcome of interest is the proportion of educational staff hired/fired for a muncipality at any given year.
-For the second set of estimations, the main specification is:
-% $$\text{turnover}_{jt} = \gamma \text{coalition seats}_{jst} + \mu P_{jt} + \zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{jt}$$
-% The share of coalition seats held by the mayoral coalition is the treatment in this set-up.
-We are interested in $\gamma$, the change in bureaucratic turnover associated with variation in the share of legislative seats held by the mayoral coalition.
-Municipal characteristics $W_{jt}$ are similar to the ones used for the first set of estimation.
-I add political variables to the estimation in order to control for differences in mayor partisanship, incumbency status, and individual characteristics of the mayor: professional background, education, and age.
-
-% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational.
+Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational.
 In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge.
-To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in imai_covariate_2014 which is generalized for continuous treatments.
+To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in \citet{imai_covariate_2014} which is generalized for continuous treatments.
 A covariate balance test for either estimation is presented in the appendix.
 
-% ## Results
+\subsection*{Results}
 
-% Teacher and school principal turnover have significant, negative effects on average test scores, highlighting the potential dangers of bureaucratic turnover on student learning.
+Teacher and school principal turnover have significant, negative effects on average test scores, highlighting the potential dangers of bureaucratic turnover on student learning.
 These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations.
@@
 
-% \begin{landscape}
-% \begin{table}[t]
-%   \centering
-%   \footnotesize
-%   \input{tables/hlm_result}
-%   \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning.
+\begin{figure}[h]
+  \centering
+  \footnotesize
+  \includegraphics{plots/model_turnover_learning.pdf}
+  \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning.
 Models 1 and 2 present results for teacher turnover index constructed at the school level.
@@
 All models include year and state fixed effects.}
-%   \label{tbl:hlm_mods}
-% \end{table}
-% \end{landscape}
+  \label{fig:hlm_mods}
+\end{figure}
 

### 2020-09-09 paper/paper.tex
first edit of final paper

@@
 
+I first turn to testing the main proposition of this paper: that turnover in educational staff stems indeed from coalitional dynamics.
+There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage.
+Figure \ref{tbl:turnover} presents the results of the estimation.
+The first model regresses turnover index on coalition share, where the unit of analysis is a grade-level per school.
+Models 2 and 3 estimate the effect of the share of legislative seats on hiring or dismissal of individual bureaucrats.
+The third set of models (4 and 5) estimate the change in the share of new hires and dismissals at the municipal level.
+All models suggest that an increase in the executive share of legislative seats decrease turnover.
+
+\begin{figure}[h]
+  \includegraphics{plots/plot_coalition_coef_mun.pdf}
+  \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
+Model 1 presents results on the turnover index.
+Models 2 and 3 are logistic regressions at the individual, bureaucrat level.
+Models 4 and 5 are aggregated at the municipal level.
+where the outcome of interest is the proportion of staff either hired or dismissed at a given year.
+Models 1, 3, and 5 include year and state fixed effects.}
+  \label{fig:coalition_turnover}
+\end{figure}
+
+% For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council.
+Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature.
+Overall, these results suggest that weaker executive control over the legislature increases patronage, with potentially negative effects for public service delivery in municipalities in Brazil.
+
+% \begin{figure}[H]
+%     \centering
+%     \includegraphics{plots/plot_pred}
+%     \caption{Predicted values for bureaucratic turnover.
+The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.
+The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
+%     \label{fig:turnover}
+% \end{figure}
+
 Teacher and school principal turnover have significant, negative effects on average test scores, highlighting the potential dangers of bureaucratic turnover on student learning.
 These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations.
-Table \ref{tbl:hlm_mods} presents the results of the estimation on SAEB test scores.
+Table \ref{fig:hlm_mods} presents the results of the estimation on SAEB test scores.
 We present two specifications for turnover.
@@
 
-% Additionally, I present results for the SPAECE standardized exam in table \ref{tbl:spaece}.
-Differently than the SAEB, the SPAECE takes place annually and the average test scores are calculated at the grade, rather than the classroom level.
-The annual frequency of this exam allows us to retrieve more precise estiamtes with respect to the effect of teacher turnover on student test scores, as opposed to the biannual frequency of the SAEB.
-Note that for both estimations, we standardize the outcome of interest.
-Coefficients are therefore interpretable as a standard deviation change in student test with respect to the covariate.
-
-% \begin{table}[t]
-%   \centering
-%   \footnotesize
-%   \input{tables/spaece_result}
-%   \caption{{\bf Turnover and student learning in Ceará.} Teacher turnover have a negative effect on student learning.
-Models 1 and 2 present results for teacher turnover index constructed at the school level.}
-%   \label{tbl:spaece}
-% \end{table}
-
-% I now turn to testing the main proposition of this paper: that turnover in educational staff stems indeed from coalitional dynamics.
-There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage.
-Table \ref{tbl:turnover} presents the results of the estimation.
-The first model regresses turnover index on coalition share, where the unit of analysis is a grade-level per school.
-Models 2 and 3 estimate the effect of the share of legislative seats on hiring or dismissal of individual bureaucrats.
-The third set of models (4 and 5) estimate the change in the share of new hires and dismissals at the municipal level.
-All models suggest that an increase in the executive share of legislative seats decrease turnover.
-
-% \begin{landscape}
-% \begin{table}[t]
-%   \centering
-%   \footnotesize
-%   \input{tables/turnover_result}
-%   \caption{{\bf Executive coalitions and staff turnover.} An incrase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
-Model 1 presents results on the turnover index.
-Models 2 and 3 are logistic regressions at the individual, bureaucrat level.
-Models 4 and 5 are aggregated at the municipal level.
-where the outcome of interest is the proportion of staff either hired or dismissed at a given year.
-Models 1, 3, and 5 include year and state fixed effects.}
-%   \label{tbl:turnover}
-% \end{table}
-% \end{landscape}
-
-% For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council.
-Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature.
-Overall, these results suggest that weaker executive control over the legislature increases patronage, with potentially negative effects for public service delivery in municipalities in Brazil.
-
-% \begin{figure}[H]
-%     \centering
-%     \includegraphics{plots/plot_pred}
-%     \caption{Predicted values for bureaucratic turnover.
-The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.
-The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
-%     \label{fig:turnover}
-% \end{figure}
+\blue{Include some reference coefs in order for the audience to assess the magnitude of the effects}
 
@@
 
-% # Conclusion
+\section{Conclusion}
 
-% Improving the quality of education received by children across the world remains a challenge.
+Improving the quality of education received by children across the world remains a challenge.
 This paper proposes an analytical framework and estimation strategy to understand the decision-making process behind the administration of educational services in Brazil.
@@
 
-% In this study I theorize and demonstrate that staff turnover stems from the executive's need to garner support from legislators in the city council.
+In this study I theorize and demonstrate that staff turnover stems from the executive's need to garner support from legislators in the city council.
 This process of coalition building is consolidated through employment offers to city councilors and their constituencies.
@@
 
-% This bureaucratic turnover has important consequences for the quality of education delivered in municipalities.
+This bureaucratic turnover has important consequences for the quality of education delivered in municipalities.
 I find that turnover has a negative effect on student learning, across a set of specifications for turnover and different evaluation metrics for student learning.
 The evidence therefore seems to point out that mayors with a weaker hold on the city council resort to greater patronage, with negative consequences for student learning.
-This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts (gottlieb_countervailing_2019).
+This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts \citep{gottlieb_countervailing_2019}.
 
-% Future research on patronage and public service delivery would benefit from a clearer treatment of the institutional context in which local political actors operate.
+Future research on patronage and public service delivery would benefit from a clearer treatment of the institutional context in which local political actors operate.
 Important insights have been derived on executive leaders, but these actors seldom govern alone.

### 2020-09-09 source/models/modules/visual_model_coalition.R

@@ -11,7 +11,10 @@ fit_felm <- read_model(
 
 fit_turnover <- read_model(
   "fit_coalition_turnover.rds"
-) # extract model with controls
+) %>%
+  set_names(
+    c("baseline", "controls")
+  )
 
 # turn into debugging
 fit_logit <- fit_logit %>%
@@ -65,19 +68,16 @@ save_fig(
 # ==============================================================================
 # incorporate additional municipal covariates to disentangle effect
 # ==============================================================================
-
-
-# rural_terciles <- fit_logit %>%
-#   pluck(1, "model") %>%
-
 # plot_logit_int <- map(
 #   c(
 #     "rural" = "censo_rural[0.25, 0.5, 0.75]",
 #     # "population" = "censo_pop[1e4, 5e4, 1e5]",
 #     "wage" = "censo_median_wage[172.12, 250.77, 341.896]"
 #   ),
-#   ~ sjPlot::plot_model(fit_logit[[1]], type = "pred", title = NULL, terms = c("coalition_share [all]", .))
+#   ~ sjPlot::plot_model(fit_logit[["controls"]], type = "pred", title = NULL, terms = c("coalition_share [all]", .))
 # )
+
+# sjPlot::plot_model(fit_logit[["controls"]], type = "pred", title = NULL, terms = c("coalition_share [all]", "censo_median_wage[172.12, 250.77, 341.896]"))
 
 # walk2(
 #   plot_logit_int,
@@ -103,9 +103,18 @@ print("visualize association between coalition share and patronage (municipal)")
 coef_mun <- models_table %>%
   map(pluck("controls")) %>%
   set_names(c("turnover", "hired (logit)", "hired (felm)")) %>%
-  map_dfr(broom::tidy, conf.int = T, .id = "model")
+  map_dfr(broom::tidy, .id = "model") %>%
+  mutate(
+    conf.low = estimate - 1.96*std.error,
+    conf.high = estimate + 1.96*std.error
+  )
+
+select_vars <- c("coalition_share", "censo_log_pop", "censo_lit_rate", "censo_median_wage", "chamber_size")
 
 plot_coef_mun <- coef_mun %>%
+  filter(
+    term %in% select_vars
+  ) %>%
   ggplot(
     aes(
       x = term,
@@ -113,17 +122,36 @@ plot_coef_mun <- coef_mun %>%
       color = model
     )
   ) +
-  geom_point() +
+  geom_point(
+    position = position_dodge(.5)
[…]

### 2020-09-09 source/models/modules/visual_model_turnover.R

@@ -70,7 +70,7 @@ plot_model_turnover <- estimate_hlm %>%
   ) +
   geom_point(
     position = ggstance::position_dodgev(height = 0.3),
-    size = 2
+    size = 4
   ) +
   geom_errorbar_tidy +
   theme(
@@ -82,24 +82,24 @@ plot_model_turnover <- estimate_hlm %>%
     limits = c(
       "turnover_index",
       "saeb_principal_experienceless than 2 years",
-      "saeb_teacher_work_schoolless than 2",
-      "saeb_principal_appointmentpublic exam",
-      "saeb_principal_educationhigher education",
-      "education_teacherhigher education",
-      "saeb_wage_teachermore than 3000"
+      "saeb_teacher_work_schoolless than 2"
+      # "saeb_principal_appointmentpublic exam",
+      # "saeb_principal_educationhigher education",
+      # "education_teacherhigher education",
+      # "saeb_wage_teachermore than 3000"
     ),
     labels = c(
       "turnover_index" = "Turnover index",
       "saeb_principal_experienceless than 2 years" = "Principal rotation",
-      "saeb_teacher_work_schoolless than 2" = "Teacher rotation",
-      "saeb_principal_appointmentpublic exam" = "Principal public exam",
-      "saeb_principal_educationhigher education" = "Principal higher education",
-      "education_teacherhigher education" = "Teacher higher education",
-      "saeb_wage_teachermore than 3000" = "Teacher wage more than $750"
+      "saeb_teacher_work_schoolless than 2" = "Teacher rotation"
+      # "saeb_principal_appointmentpublic exam" = "Principal public exam",
+      # "saeb_principal_educationhigher education" = "Principal higher education",
+      # "education_teacherhigher education" = "Teacher higher education",
+      # "saeb_wage_teachermore than 3000" = "Teacher wage more than $750"
     )
   ) +
   coord_flip(
-    xlim = c(-0.2, 0.2)
+    xlim = c(-0.1, 0.05)
   )
   
 save_fig(

### 2020-09-10 paper/paper.tex
add additional plots

@@
 \usepackage{setspace}
-\usepackage[top = 1.25in, bottom = 1.25in, left = 1.25in, right = 1.25in]{geometry}
+\usepackage[top = 1in, bottom = 1in, left = 1in, right = 1in]{geometry}
 \usepackage[utf8]{inputenc}
@@
 \bibliographystyle{unsrtnat}
+
+\title{Governing through patronage: the bargain for education in decentralized Brazil}
+\date{\today}
+
+\author{Galileu Kim}
+\affil{Princeton University}
+
 \begin{document}
 
+\maketitle
+
+\begin{abstract}
+    Why do politicians engage in patronage and what are its consequences for public service delivery?
+I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators.
+Leveraging fieldwork evidence and administrative data, I demonstrate that mayors with a weaker hold on the local legislature resort to patronage to co-opt legislators.
+Educational staff turnover has negative downstream effects on student learning, but limited electoral backlash.
+Weak electoral accountability channels suggests that patronage is primarily a political elite game, with little involvement by the electorate.
+These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
+\end{abstract}
+
+\newpage
 \section{Introduction}
@@
 How can these educational outcomes be improved?
-Previous studies have found that incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
-Often relying on quasi-experimental research designs, these studies provide compelling evidence that actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?
+Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
+Often relying on quasi-experimental research designs, these studies provide compelling evidence that political and bureaucratic actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?
 
@@
 
-I first turn to testing the main proposition of this paper: that turnover in educational staff stems indeed from coalitional dynamics.
+I first turn to testing the main proposition of this paper: that turnover in educational responds to the degree of political alignment between the mayor and politician.
 There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage.
@@
 
-% For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council.
+For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council.
 Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature.
@@
 
-% \begin{figure}[H]
-%     \centering
-%     \includegraphics{plots/plot_pred}
-%     \caption{Predicted values for bureaucratic turnover.
+\begin{figure}[h]
+    \centering
+    \includegraphics{plots/plot_interactions_logit}
+    \caption{Predicted values for bureaucratic turnover.
 The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.
 The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
-%     \label{fig:turnover}
-% \end{figure}
+    \label{fig:turnover}
+\end{figure}
 

### 2020-09-10 paper/paper.tex
continue to redraft paper

@@
 \newcommand{\blue}{\textcolor{blue}}
+% \renewcommand{\includegraphics}{\includegraphics[width=0.6\linewidth]}
 
@@
 
-\subsection{Education in Brazil}
+\subsection{Municipal education in Brazil}
 
@@
 \begin{figure}[h]
-    \includegraphics{plots/descriptive_enrollment.pdf}
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
     \caption{enrollment}
     \label{fig:enroll}
-    \centering
 \end{figure}
@@
 However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount.
-Municipal teachers and school principals are overseen by a local department of education \citet{militao_educacao_2014}.
+Municipal teachers and school principals are overseen by a local department of education \citep{militao_educacao_2014}.
+
+Positions in the educational sector are politically valuable.
+As a department, it comprises over 25 percent of local public sector jobs.
+Due to their relatively high compensation and social status compared to other positions in the local bureaucracy, educational staff positions are particularly valuable for constituencies seeking employment.
+Additionally, anecdotal evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown}
+\end{figure}
 
@@
 \begin{figure}[h]
-    \includegraphics{plots/saeb_map.pdf}
-    \caption{hello}
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
+    \caption{Uneven quality of municipal education.
+Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores.
+Averages are weighted by student participation rate.}
     \label{fig:saeb_map}
@@
 \begin{figure}[h]
-    \includegraphics{plots/descriptive_staff_turnover.pdf}
+    \centering
+    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
     \label{fig:staff_turnover}
-    \caption{hello}
+    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
 \end{figure}
@@
 \begin{figure}
-    \includegraphics{plots/hist_coalition.pdf}
-    \caption{hello}
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
+    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
     \label{fig:hist_coalition}

### 2020-09-10 source/descriptive/modules/visual_electoral.R

@@ -17,7 +17,9 @@ rais <- read_data(
 hist_coalition <- election %>%
     gg_histogram(
         aes(coalition_share * 100),
-        bins = 20
+        bins = 20,
+        color = "#e2735d",
+        fill = "coral3"
     ) +
     labs(
         x = "Share of seats (percentage)",

### 2020-09-10 paper/paper.tex
expand results section

@@
 
-I first turn to testing the main proposition of this paper: that turnover in educational responds to the degree of political alignment between the mayor and politician.
-There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage.
-Figure \ref{tbl:turnover} presents the results of the estimation.
-The first model regresses turnover index on coalition share, where the unit of analysis is a grade-level per school.
-Models 2 and 3 estimate the effect of the share of legislative seats on hiring or dismissal of individual bureaucrats.
-The third set of models (4 and 5) estimate the change in the share of new hires and dismissals at the municipal level.
+I first turn to testing the main proposition of this paper: that turnover in educational staff responds to the degree of political alignment between the mayor and politician.
+In line with theory, there is strong evidence that mayors who are electoral allies with more (less) seats in the city council resort to less (more) patronage.
+This is robust to a set of specifications, including state-year fixed effects and alternative measures of turnover in educational staff.
+
+Table \ref{tbl:turnover} presents the results of the estimation.
+Model 1 regresses turnover index on coalition share, where the unit of analysis is a grade-level per school.
+Models 2 and 3 estimate the effect of the share of legislative seats on hiring of new bureaucrats aggregated at the municipal level.
+The third set of models (5 and 6) estimate the change probability of new bureaucratic hires at the individual level.
 All models suggest that an increase in the executive share of legislative seats decrease turnover.
+
+\begin{landscape}
+    \begin{table}[t]
+      \centering
+      \footnotesize
+      \input{results/turnover_result}
+      \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
+Models 1 and 2 present results on the turnover index at the school level.
+Models 2 and 3 are aggregate hiring rates at the municipal level.
+Models 5 and 6 are logistic regressions at the individual, bureaucrat level.
+where the outcome of interest is the proportion of staff either hired or dismissed at a given year.
+Models 1, 3, and 5 include year and state fixed effects.}
+      \label{tbl:turnover}
+    \end{table}
+\end{landscape}
+
+For visual intuition, I present the estimated coefficients for the models with controls -- models 2, 4, and 6 -- in context with other coefficients for demographic/contextual variables.
+Note that while the share of allied seats is a precise predictor of the degree of turnover in educational staff, other important factors such as the level of economic development -- municipal median wage -- are less informative.
+Literacy rate, although unprecise, is negatively associated with staff turnover, suggesting that a more educated electorate may exert some pressure to retain teachers and school principals.
 
@@
 For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council.
-Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature.
+The reduction in new hires is more pronounced for school principals than teachers, suggesting that exposure to patronage is more concentrated in the leadership positions at the school level.
 Overall, these results suggest that weaker executive control over the legislature increases patronage, with potentially negative effects for public service delivery in municipalities in Brazil.
@@
 
-Teacher and school principal turnover have significant, negative effects on average test scores, highlighting the potential dangers of bureaucratic turnover on student learning.
+Finally, I demonstrate that
+
+Moving on to characterizing the downstream effects of patronage, I find that teacher and school principal turnover have significant, negative effects on average test scores.
 These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations.

### 2020-09-10 source/descriptive/modules/visual_staff_turnover.R

@@ -65,15 +65,13 @@ hist_turnover_index <- censo_school_turnover %>%
     ggplot(
         aes(
             turnover_index,
-            # y = stat(width * density),
-            group = year,
-            color = year
+            ..scaled..,
+            group = as.factor(year),
+            color = as.factor(year)
         )
     ) +
-    geom_density() +
-    scale_y_continuous(
-        labels = scales::percent_format()
-    ) +
+    geom_line(stat="density", size = 1) +
+    # scale_y_continuous(labels = scales::percent, name = "percent") +
     labs(
         x = "Distribution of turnover index",
         y = "Count"
@@ -83,7 +81,6 @@ save_fig(
     hist_turnover_index,
     "hist_turnover_index.pdf"
 )
-
 
 # patronage_category <- patronage %>%
 #     mutate(

### 2020-09-10 paper/paper.tex
include table

@@
 }
+\captionsetup[figure]{font=small,labelfont=small}
 
@@
 \begin{figure}[h]
-  \includegraphics{plots/plot_coalition_coef_mun.pdf}
-  \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
+    \caption{{\bf Political alignment and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
 Model 1 presents results on the turnover index.
@@
     \centering
-    \includegraphics{plots/plot_interactions_logit}
+    \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
     \caption{Predicted values for bureaucratic turnover.
@@
 
-Finally, I demonstrate that
+Finally, I show that the executive-legislative bargain extends to second-term mayors.
+I subset the data used for the above analysis to only mayors who are reelected, and re-run the previous analyses with the same specification.
+I find that the results are similar in magnitude and precision to those presented above.
+This finding suggests that patronage is not motivated solely by reelection concerns.
+Rather, patronage achieves an important policy goal for incumbent mayors who seek to implement their preferred policy, regardless of whether they are just initiating their mandate or concluding it.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun_second.pdf}
+    \caption{\bf{hello}}
+\end{figure}
 
@@
 These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations.
-Table \ref{fig:hlm_mods} presents the results of the estimation on SAEB test scores.
+Table \ref{tbl:student_learning} presents the results of the estimation on SAEB and SPAECE test scores.
 We present two specifications for turnover.
@@
 Personnel who has only recently entered the school recently amass a shorter amount of work experience (less than two years).
+
+\begin{landscape}
+    \begin{table}[t]
+      \centering
+      \footnotesize
+      \input{results/student_learning.tex}
+      \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning.
+Models 1 and 2 present results for teacher turnover index constructed at the school level.
+Models 3 and 4 estimate the effect of new teachers and school principals entering the school (less than two years).
+All models include year and state fixed effects.}
+      \label{tbl:student_learning}
+    \end{table}
+\end{landscape}
 
@@
 
-% As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity.
+As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity.
 It is in these circumstances that validating causal mechanisms through fieldwork and careful data analysis can increase the validity of a causal claim.

### 2020-09-10 source/models/modules/model_turnover_spaece.R

@@ -96,7 +96,7 @@ formulae_spaece <- formulate_models(
   controls
 ) %>%
   map(
-    ~add_felm(., fe = "year", cluster = "0" )
+    ~add_felm(., fe = "year", cluster = "0")
   )
 
 fit_spaece <- map(

### 2020-09-10 source/models/modules/visual_model_turnover.R

@@ -12,14 +12,18 @@ fit_ivreg <- map(
 
 sink(here("figures", "tables", "student_learning.tex"))
 mstar(
-  fit_lmer, fit_spaece, fit_ivreg,
-  keep = c("turnover_index", "saeb_principal_experience", "saeb_teacher_work_school"),
+  fit_lmer, fit_spaece,
+  # type = "text",
+  keep = c("turnover_index", "^saeb_principal_experience", "saeb_teacher_work_school"),
   add.lines = list(c("Controls", rep(c("\\_", "\\checkmark"), 3))),
   dep.var.labels.include = F,
   covariate.labels = c(
-    "Turnover index", "Turnover index $\\times$ Grade 9",
-    "Teacher experience (2 years)", "Teacher experience (10 years)",
-    "School principal experience (2 years)", "School principal experience (10 years)"
+    "Turnover index",
+    "Teacher experience (2-10 years)", "Teacher experience (2 years)",
+    "School principal experience (2-10 years)", "School principal experience (2 years)", "School principal experience (2-10 years) $\\times$ 9th grade",
+    "School principal experience (2 years) $\\times$ 9th grade",
+    "Teacher experience (2-10 years) $\\times$ 9th grade",
+    "Teacher experience (2 years) $\\times$ 9th grade"
   ),
   dep.var.caption = "Student learning",
   column.labels = c("SAEB test score", "SPAECE test score"),

### 2020-09-11 paper/paper.tex
redraft paper

@@
 
-The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations.
+The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations.
 How can these educational outcomes be improved?
 Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
-Often relying on quasi-experimental research designs, these studies provide compelling evidence that political and bureaucratic actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?
+Often relying individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy experiments.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo?
+Who holds power over public education, and whose interests is it meant to serve?
 
-In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to capture by local political elites.
+In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to its capture by local political elites.
 Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}.
 This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}.
-By catering to the city council through patronage appointments, mayors ultimately cause instability in the local educational bureaucracy, with negative downstream consequences for public school students.
+By catering to the city council through patronage appointments, mayors ultimately cause turnover in the local educational bureaucracy, with negative downstream consequences for public school students.
 
 To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council.
-In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage, in particular during the first year of administration.
-Mayors therefore resort to patronage in order to shore up legislative support, with negative consequences for student learning.
+In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage.
+Contrary to electoral accountability frameworks \citep{ferejohn_incumbent_1986,besley_political_2003}, mayors resort to patronage in order to shore up legislative support with limited consequences.
 I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover.
@@
 
-This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015, gulzar_politicians_2017}.
-Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services \citep{ferraz_corrupting_2012}.
-It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy \citep{kohli_state-directed_2004, evans_embedded_1995}.
+This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}.
+I analyze how incentives shape political decision on how to manage local bureaucracies \citep{gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders.
+I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services.
+\citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}.
+I also bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.
 

### 2020-09-11 paper/paper.tex
continue editing patronage_education paper

@@
 
-In Brazil, responsibility for primary education is delegated to local governments \citet{paschoal_historia_2009}.
-The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments.
+In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}.
+In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and substraction.
+The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school strudents attend a public school.
 Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools.
@@
     \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
-    \caption{enrollment}
+    \caption{\textbf{Student enrollment by administration.}}
     \label{fig:enroll}
@@
 $$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$
+
+\blue{Insert more information and descriptive paragraph on turnover index here}
 

### 2020-09-11 paper/paper.tex
redraft intro

@@
 This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}.
-By catering to the city council through patronage appointments, mayors ultimately cause turnover in the local educational bureaucracy, with negative downstream consequences for public school students.
+By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.
 
-To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council.
+To analyze how politics affect staff turnover, I propose a theoretical model of legislative vote-buying, which I apply to the Brazilian context.
+I test the empirical implications of the model using Brazil as a case.
 In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage.
@@
 To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning.
-I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory.
+I construct multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory.
 Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff.

### 2020-09-11 paper/paper.tex
second term

@@
 How can these educational outcomes be improved?
-Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
-Often relying individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy experiments.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo?
+Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
+Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo?
 Who holds power over public education, and whose interests is it meant to serve?
 
-In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to its capture by local political elites.
+To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil.
+In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.
+But the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}.
+In it, the exchange of public sector appointments for political favors -- a canonical example of patronage -- was designed to grant the mayor legislative support to govern according to her policy agenda.
+
+In this paper, I argue that public education in municipalities across Brazil have been captured by political elites, with personnel appointments into public schools and management dominated by bargains between executive and legislative branches.
 Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}.
@@
 
-To analyze how politics affect staff turnover, I propose a theoretical model of legislative vote-buying, which I apply to the Brazilian context.
-I test the empirical implications of the model using Brazil as a case.
-In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage.
-Contrary to electoral accountability frameworks \citep{ferejohn_incumbent_1986,besley_political_2003}, mayors resort to patronage in order to shore up legislative support with limited consequences.
-I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover.
-Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).
+The estimation proceeds in two parts.
+First, I show that political alignment between mayors and city councilors has a direct ffect on patronage.
+Building on a canonical model, I propose a theoretical model of legislative vote-buying from which I derive my main empirical test: mayors who face less opposition in the city council engage in less patronage.
+To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities.
+In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage.
+This results holds across a set of specifications and different measures of turnover.
 
+I then show that this politically induced turnover has a negative impact on student learning.
 To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence.
-Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students.
-To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning.
+Interviews conducted with educational bureaucrats and politicians suggested firm that turnover had a negative impact on teachers' ability to educate students, as time horizons were compressed and bonds between educators and learners were broken.
+To validate these accounts I combine administrative data on education from two separate student learning measures: \emph{Prova Brasil} and the \emph{SPAECE}.
 I construct multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory.
-Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff.
 A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.
@@
 \citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}.
-I also bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.
+I bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.
 
@@
 Due to their relatively high compensation and social status compared to other positions in the local bureaucracy, educational staff positions are particularly valuable for constituencies seeking employment.
-Additionally, anecdotal evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.
+Additionally, qualitative evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.
 
@@
 The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education.
-To do so, I leverage the test scores made available by SAEB and SPAECE to estimate the effect of turnover on student test scores.
+To do so, I leverage the test scores made available by \emph{SAEB} and \emph{SPAECE} to estimate the effect of turnover on student test scores.
 To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample.

### 2020-09-11 source/models/modules/visual_model_coalition.R

@@ -175,23 +175,20 @@ fit_turnover_second_term <- read_model(
 )
 
 models_complete <- list(
-  fit_logit %>% pluck("controls"),
   fit_logit_second_term,
-  fit_felm %>% pluck("controls"),
   fit_felm_second_term,
-  fit_turnover %>% pluck("controls"),
   fit_turnover_second_term
 )
 
 coef_mun_second <- models_complete %>%
-  set_names(c("hired (individual)", "hired (individual - second)", "hired (municipal)", "hired (municipal - second)", "turnover", "turnover (second)")) %>%
+  set_names(c("hired (individual)", "hired (municipal)", "turnover")) %>%
   map_dfr(broom::tidy, .id = "model") %>%
   mutate(
     conf.low = estimate - 1.96*std.error,
     conf.high = estimate + 1.96*std.error
   )
 
-plot_coef_mun_second <- coef_mun %>%
+plot_coef_mun_second <- coef_mun_second %>%
   filter(
     term == "coalition_share"
   ) %>%

### 2020-09-11 paper/paper.tex
final edit paper patronage

@@
     \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
-    \caption{Predicted values for bureaucratic turnover.
-The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.
-The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
+    \caption{\textbf{Differential reductions in patronage in teachers and school principals}.
+Predicted values for bureaucratic turnover plotted against the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.}
     \label{fig:turnover}
@@
     \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun_second.pdf}
-    \caption{\bf{hello}}
+    \caption{\textbf{Patronage continues into the second term.} All models include a full set of controls, as well as year and state fixed effects.}
 \end{figure}
@@
 Work experience for teachers and school principals serve as an alternative measure of turnover.
-Personnel who has only recently entered the school recently amass a shorter amount of work experience (less than two years).
 
@@
 
+I also present the coefficients for the regression exercise above in figure \ref{fig:hlm_mods}.
+Note that the results are consistent across the board, providing strong evidence that staff turnover has negative downstream consequences for student learning.
+
 \begin{figure}[h]
   \centering
-  \footnotesize
-  \includegraphics{plots/model_turnover_learning.pdf}
-  \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning.
-Models 1 and 2 present results for teacher turnover index constructed at the school level.
-Models 3 and 4 estimate the effect of new teachers and school principals entering the school (less than two years).
-All models include year and state fixed effects.}
+  \includegraphics[width=0.6\linewidth]{plots/model_turnover_learning.pdf}
+  \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning, across different sets of exams (\emph{SAEB} and \emph{SPAECE}).
+All models include a full set of controls, as well as year and state fixed effects.}
   \label{fig:hlm_mods}
@@
 
-\blue{Include some reference coefs in order for the audience to assess the magnitude of the effects}
+Finally, I present some evidence of differential returns to patronage, in particular for city councilors.
+For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/reelection_marginal_effect_plots.pdf}
+    \caption{\textbf{Differential returns to patronage}.
+Mayors who engage in patronage do not directly benefit from it.
+Instead, city councilors in the mayoral ally base seem to be benefiting from it at the expense of non-allied city councilors.}
+\end{figure}
 
@@
 
-% \pagebreak
+\pagebreak
 
-% # Appendix
+\section*{Appendix}
 

### 2020-09-11 paper/paper.tex
finalize edits for paper to brazil data lab

@@
 \begin{abstract}
-    Why do politicians engage in patronage and what are its consequences for public service delivery?
-I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators.
-Leveraging fieldwork evidence and administrative data, I demonstrate that mayors with a weaker hold on the local legislature resort to patronage to co-opt legislators.
-Educational staff turnover has negative downstream effects on student learning, but limited electoral backlash.
-Weak electoral accountability channels suggests that patronage is primarily a political elite game, with little involvement by the electorate.
+    How can public education be improved?
+I argue that local politicians govern through patronage, undermining education at the expense of voters.
+Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites.
+Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector.
+The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain.
+Patronage induces turnover in educational staff, with negative downstream effects on student learning.
+Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate.
 These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
@@
 The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations.
-How can these educational outcomes be improved?
+How can public education be improved?
 Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
-Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo?
+Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors relate to one another in local institutions?
 Who holds power over public education, and whose interests is it meant to serve?
@@
 In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.
-But the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}.
-In it, the exchange of public sector appointments for political favors -- a canonical example of patronage -- was designed to grant the mayor legislative support to govern according to her policy agenda.
+However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}.
+In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.
 
@@
 The estimation proceeds in two parts.
-First, I show that political alignment between mayors and city councilors has a direct ffect on patronage.
-Building on a canonical model, I propose a theoretical model of legislative vote-buying from which I derive my main empirical test: mayors who face less opposition in the city council engage in less patronage.
+First, I show that political alignment between mayors and city councilors has a direct effect on patronage.
+Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage.
 To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities.
@@
 The paper is structured as follows.
-Section 1 provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil.
-In section 2 I present the main theoretical argument, with a formal treatment of the subject.
-Section 3 presents the institutional context and data, followed in section 4 by the research design and main results.
+Section \ref{sec:literature} provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil.
+Section \ref{sec:context} presents the context and data.
+In section \ref{sec:theory}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model.
+Section \ref{sec:empirics} presents the research design and main results.
 Section 5 concludes.
@@
 \section{Related literature}
+\label{sec:literature}
 
@@
 \section{Institutional context and data}
+\label{sec:context}
 
@@
 Additional data has been collected to supplement the estimation exercise.
-Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\url{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}, (2)\url{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}, (3)\url{https://www.seduc.ce.gov.br/spaece/}.}
+Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\hyperlink{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}{IBGE}, (2)\hyperlink{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}{Finbra}, (3)\hyperlink{https://www.seduc.ce.gov.br/spaece/}{Spaece}.}
 
@@
 \section{Empirical results}
+\label{sec:empirics}
 
@@
 
-Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational.
+% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational.
 In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge.
@@
 For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education.
+The negative consequences of patronage on student learning, however, do not bite, as nayors who deliver a worse quality of education are no less likely to be reelected in the next term.\footnote{See appendix \ref{app:accountability}.}
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
 
@@
 
-% \pagebreak
+\subsection{Covariate balance}
 
-% ## Covariate balance
+\begin{figure}[h]
+    \centering
+    \includegraphics{plots/covariate_balance.pdf}
+\end{figure}
 
-% ### Education quality vs. staff turnover
+\pagebreak
+
+\subsection{Accountability of mayors}
+\label{app:accountability}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics{plots/accountability_fit.pdf}
+\end{figure}
+
+\pagebreak
+
+\subsection{Regression Discontinuity Design}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.7\linewidth]{plots/visual_rdd_legislative_seat_flip.pdf}
+\end{figure}
+
+\pagebreak
 

### 2020-09-11 source/models/modules/model_accountability.R

@@ -107,10 +107,10 @@ fit_accountability <- map(
     )
 )
 
-plot_accountability <- fit_acc %>% 
+plot_accountability <- fit_accountability %>% 
   map2_df(
     names(.),
-    ~tidy(.x, conf.int = T) %>% 
+    ~broom::tidy(.x, conf.int = T) %>% 
       filter(
         str_detect(
           term,
@@ -137,7 +137,7 @@ plot_accountability <- ggplot(
   ) +
   geom_errorbar_tidy
 
-ggsave(
+save_fig(
   plot_accountability,
-  filename = p_file_here('plots', "accountability_fit.pdf")
+  "accountability_fit.pdf"
 )

### 2020-09-11 paper/paper.tex
finished version paper

@@
     \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
-    \caption{{\bf Political alignment and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
-Model 1 presents results on the turnover index.
-Models 2 and 3 are logistic regressions at the individual, bureaucrat level.
-Models 4 and 5 are aggregated at the municipal level.
-where the outcome of interest is the proportion of staff either hired or dismissed at a given year.
-Models 1, 3, and 5 include year and state fixed effects.}
+    \caption{{\bf Political alignment and staff turnover.} A visual representation of results presented in table \ref{tbl:turnover}. All models include year and state fixed effects.}
     \label{fig:coalition_turnover}
@@
     \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun_second.pdf}
-    \caption{\textbf{Patronage continues into the second term.} All models include a full set of controls, as well as year and state fixed effects.}
+    \caption{\textbf{Patronage continues into the second term.} There is strong evidence that executive-legislative bargains continue into the mayor's second term.
+All models include a full set of controls, as well as year and state fixed effects.}
 \end{figure}

### 2020-09-16 paper/paper.tex
changes in draft

@@
 This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}.
-I analyze how incentives shape political decision on how to manage local bureaucracies \citep{gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders.
+I analyze how incentives shape political decision on how to manage local bureaucracies \citep{duflo_incentives_2012, gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders.
 I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services.

### 2020-09-28 paper/paper.Rmd
incorporate latest changes

--- 2020-07-06 publication/paper_job_market.Rmd
@@
 geometry: margin=1.5in
-abstract: "Why do politicians engage in patronage?
-What are its consequences for public service delivery?
+abstract: "Why do politicians engage in patronage and what are its consequences for public service delivery?
 I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators.

### 2020-09-30 paper/paper.tex
fix bibliography
incorporate latest edits

@@
 \begin{abstract}
-    How can public education be improved?
+    How can governments improve the quality of education?
 I argue that local politicians govern through patronage, undermining education at the expense of voters.
@@
 
-% Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-In many countries, particularly in the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, mathematical operations, elusive.
+Across the world, vulnerable populations rely on governments for access to basic public services such as education.
+In the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, basic mathematical operations.\footnote{The World Bank calls this ``learning poverty".
+See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}.
 Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them.
@@
 How can public education be improved?
+
 Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
@@
 
+\bibliography{dissertation}
+
+\pagebreak
+
 \section*{Appendix}
@@
 
-\pagebreak
-
 % \blue{plots/balance_hlm}
@@
 
-\bibliography{dissertation}
-
 \end{document}

### 2020-09-30 paper/paper.tex
create outline for intro

@@
 Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-In the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, basic mathematical operations.\footnote{The World Bank calls this ``learning poverty".
-See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}.
-Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them.
-\citep{cochrane_policies_1983, falleti_decentralization_2010}.
-How can local governments improve the quality the of public services they deliver?
-In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?
-
-The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations.
-How can public education be improved?
-
-Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
-Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors relate to one another in local institutions?
-Who holds power over public education, and whose interests is it meant to serve?
+In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most.
+This ``learning poverty'' is in  How can local governments improve the quality the of public services they deliver?
+In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning.
 

### 2020-10-01 paper/paper.tex
redraft intro

@@
 In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most.
-This ``learning poverty'' is in  How can local governments improve the quality the of public services they deliver?
+The World Bank refers to as ``learning poverty'': low gains from education even when access to a public school exists, in contrast with earlier focus on access to education.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes stem from an institutional cascade: political institutions shape the street-level bureaucracies responsible for providing these educational services.
+
+How can local governments improve the quality the of public services they deliver?
 In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning.

### 2020-10-01 paper/paper.tex
update angle?

--- 2020-09-30 paper/paper.tex
@@
 Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most.
-This ``learning poverty'' is in  How can local governments improve the quality the of public services they deliver?
-In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning.
+In the developing world, and in particular Latin America, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}.
+The World Bank refers to as ``learning poverty'': low gains from education even when access to a public school exists, in contrast with earlier focus on access to education.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.
 
-To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil.
+How can we intervene in this chain to improve the quality of public services?
+Previous studies have explored how political incentives can improve the quality of public goods provision.
+Some have focused on the direct link between political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}.
+Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce changes  \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate.
+Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.
+
+To shed light on this complexity, over two years I conducted fieldwork research in municipalities across Brazil.
 In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.

### 2020-10-01 paper/paper.tex
fix

@@
 How can we intervene in this chain to improve the quality of public services?
-Previous studies have explored how political incentives can improve the quality of public goods provision.
-Some have focused on the direct link between political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}.
-Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce changes  \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate.
+Previous studies have explored how politicians' incentives can improve the quality of public goods provision.
+Some have focused on the direct link between the political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}.
+Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce positive changes \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate.
 Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.

### 2020-10-02 paper/paper.tex
redraft intro

@@
 Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-In the developing world, and in particular Latin America, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}.
+In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}.
 The World Bank refers to as ``learning poverty'': low gains from education even when access to a public school exists, in contrast with earlier focus on access to education.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.

### 2020-10-05 paper/paper.tex
push plots

@@
 In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}.
-The World Bank refers to as ``learning poverty'': low gains from education even when access to a public school exists, in contrast with earlier focus on access to education.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.
+The World Bank refers to as ``learning poverty'': poor learning when access to a schooling already exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} As scholarly focus shifts from public service provision \citep{kaufman_globalization_2001, } to the quality
+
+Raises two related questions.
+The first one concerns the relationship between the quality of public services and the structure of the bureaucracy \citep{pepinsky_bureaucracy_2017, rasul2018management}.
+
+These outcomes are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.
 

### 2020-10-05 paper/paper.tex
push paper

@@
 
+% what do I need to bring the audience onboard?
+% emphasize the existence of discretion, turnover, political manipulation of public sresources.
+% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns
+
 Across the world, vulnerable populations rely on governments for access to basic public services such as education.
 In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}.
-The World Bank refers to as ``learning poverty'': poor learning when access to a schooling already exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} As scholarly focus shifts from public service provision \citep{kaufman_globalization_2001, } to the quality
+The World Bank refers to as ``learning poverty'': poor learning when access to a schooling already exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} As scholarly focus shifts from public service provision \citep{kaufman_globalization_2001, alesina_public_1999} to the quality of services themselves \citep{}
 
@@
 The first one concerns the relationship between the quality of public services and the structure of the bureaucracy \citep{pepinsky_bureaucracy_2017, rasul2018management}.
+Moving up the causal ladder, the question concerns the relationship between politicians and bureaucratic institutions and how theses personnel decisions \citep{finan_personnel_2015} have downstream consequences for the quality of these services \citep{pepinsky_bureaucracy_2017}
 
@@
 
-\subsection*{Presidential coalitionism and patronage}
+% \subsection*{Presidential coalitionism and patronage}
 
-For every elected mayor in Brazil, a group of legislators are also elected into office.
+% For every elected mayor in Brazil, a group of legislators are also elected into office.
 These political actors have competing claims over the local bureaucracy, with important implications for public service delivery.
@@
 
-In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism.
+% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism.
 Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
@@
 
-In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
+% In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
 Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
@@
 \label{sec:context}
+
+\subsection{Local governance and mayoral coalitions}
+
+Brazil's local level of government is composed of over 5 thousand municipalities, each with their own mayorship (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
+Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, open to reelection for both offices.\footnote{The last municipal elections were in 2016.
+These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over public administration in general, and personnel appointment in particular \citep{akhtari_political_2015,brollo_victor_2017}.
+As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.
+The city council, on the other hand, oversees legislation and approves the budget for the fiscal year, providing a local checks and balances on the executive.
+
+At the federal level, this instituitonal arrangement gives rise to a phenomenon dscribed as presidential coalitionism \citet{raile_executive_2011, power_optimism_2010}.
+Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
+\citep{raile_executive_2011}.
+In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.
+
+In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels \citet{dantas_eleicoes_2013}.
+These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
+Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}).
+While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.\footnote{Interview with C, former chief of staff of municipality J.
+September 2018.}
+
+Interviews with secretaries of education and mayors provide strong evidence that city councilors play an important role in staffing decisions, despite having no \emph{formal} rights to do so.
+While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
+These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council.
+According to accounts by school managers in the municipality of I:
+
+\begin{quote}
+    \emph{School principal}: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
+
+    \emph{Educational councilor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J.
+August 2019.}
+\end{quote}
+
+In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
+Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
+As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.
 
@@
 
-\subsection{Local governance and mayoral coalitions}
-
-Brazil's local level government is composed by over 5 thousand municipalities, each with their own mayor hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
-Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.\footnote{The last municipal elections were in 2016.
-These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over personnel apppointment.
-The city council, on the other hand, oversees legislation and approves the budget for the fiscal year.
-As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.
-
-In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels \citet{dantas_eleicoes_2013}.
-These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
-Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}).
-While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.\footnote{Interview with C, former chief of staff of municipality J.
-September 2018.}
-
-Interviews with secretaries of education and mayors provide evidence that city councilors play an important role in staffing decisions.
-While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
-These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council.
-In the words of a set of school administrators in the municipality of I:
-
-\begin{quote}
-    \emph{School principal}: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
-
-    \emph{Educational counselor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J.
-August 2019.}
-\end{quote}
-
-These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff.
-To verify these claims in a broader set of municipalities, as well as outlining the research design, I apply a theory of legislative vote-buying that illustrates the  employ extensive administrative and electoral data collected by the Brazilian federal government.
-In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.
-
 \subsection{Data}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
+    \includegraphics[]{plots/saeb_map.pdf}
     \caption{Uneven quality of municipal education.
@@
     \centering
-    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
+    \includegraphics[width=0.5\linewidth]{plots/descriptive_staff_turnover.pdf}
+    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
     \label{fig:staff_turnover}
-    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
 \end{figure}

### 2020-10-12 paper/paper.tex
redraft framing

@@
 Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}.
-The World Bank refers to as ``learning poverty'': poor learning when access to a schooling already exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} As scholarly focus shifts from public service provision \citep{kaufman_globalization_2001, alesina_public_1999} to the quality of services themselves \citep{}
+In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
+The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.}
+
+Understanding who benefits from these services and how politicians decide on that allocation is a well-established tradition in political science \citep{boix_democracy_2003, dixit_determinants_1996}.
+But we have less intuition on services that are de facto open to all: public education.
 

### 2020-10-12 paper/paper.tex
redraft intro

@@
 In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
-The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.}
+The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes result from political choices, but a growing body of literature recognizes the important role bureaucratic institutions mediating these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}.
 
-Understanding who benefits from these services and how politicians decide on that allocation is a well-established tradition in political science \citep{boix_democracy_2003, dixit_determinants_1996}.
-But we have less intuition on services that are de facto open to all: public education.
-
-Raises two related questions.
-The first one concerns the relationship between the quality of public services and the structure of the bureaucracy \citep{pepinsky_bureaucracy_2017, rasul2018management}.
-Moving up the causal ladder, the question concerns the relationship between politicians and bureaucratic institutions and how theses personnel decisions \citep{finan_personnel_2015} have downstream consequences for the quality of these services \citep{pepinsky_bureaucracy_2017}
-
-These outcomes are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.
+Understanding how these political decisions cascade down the bureaucracy remains a challenge.
+Public services the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom \citep{}.
 

### 2020-10-13 paper/paper.tex

@@
 Understanding how these political decisions cascade down the bureaucracy remains a challenge.
-Public services the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom \citep{}.
+Public services -- and their quality -- are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.
+This panorama of decisions belie a multiplicity of actors that is often obscured in prevailing models of
 

### 2020-10-13 paper/paper.tex
incorporate new framing for intro

--- 2020-10-12 paper/paper.tex
@@
 % what do I need to bring the audience onboard?
-% emphasize the existence of discretion, turnover, political manipulation of public sresources.
+% emphasize the existence of discretion, turnover, political manipulation of public resources.
 % in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns
@@
 In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
-The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes result from political choices, but a growing body of literature recognizes the important role bureaucratic institutions mediating these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}.
+The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} There is growing recognition of the important role bureaucratic institutions play in shaping these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}, with a renewed focus on how public sector reforms can create positive incentives for bureaucratic performance \citep{duflo_incentives_2012, dal2013strengthening, rasul2018management}.
 
-Understanding how these political decisions cascade down the bureaucracy remains a challenge.
-Public services the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom \citep{}.
+Bureaucratic institutions, however, can only be fully understood in conjunction with their political principals.
+In the developing world, politicians often manipulate bureaucracies to further private agendas, with detrimental consequences for state capacity and public goods provision \citep{evans_bureaucracy_1999,kohli_state-directed_2004,centeno_unpacking_2017}.
+These political decisions cascade down through bureaucratic institutions and into public services \citep{bersch_state_2017}.
 
@@
 Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.
+
+Show you which mechanism operates to hurt state capacity and lead to lower educational outcomes: staff turnover.
 

### 2020-10-15 paper/paper.tex
redraft intro

--- 2020-10-12 paper/paper.tex
@@
 In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
-The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes result from political choices, but a growing body of literature recognizes the important role bureaucratic institutions mediating these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}.
+The World Bank refers to this phenomenon as ``learning poverty'': stunted learning even when access to schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} There is growing recognition of the role bureaucratic institutions play in shaping these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}, but the relationship between politicians, bureaucrats and the quality of public services is less clear.
 
-Understanding how these political decisions cascade down the bureaucracy remains a challenge.
-Public services the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom \citep{}.
+In particular, there is widespread debate whether political actors' discretion over bureaucratic institutions can have positive or negative consequences for public services.
+On the one hand, a large literature on state capacity has argued that political interference has detrimental consequences for state capacity and public goods provision \citep{evans_bureaucracy_1999,kohli_state-directed_2004, akhtari_political_2015}.
+On the other, recent studies have highlighted how greater political control can have a positive effect, as increased political oversight and credit claiming can lead to better developmental outcomes \citep{gulzar_politicians_2017, raffler_does_2016}.
+This dichotomy between politicians and bureaucrats frames our understanding of bureaucratic quality \citep{grindle_jobs_2012}, but unpacking this relationship requires taking a step in and out, recognizing the multiplicity of actors and institutions that frame the political administration of public services \citep{centeno_unpacking_2017}.
 
-How can we intervene in this chain to improve the quality of public services?
-Previous studies have explored how politicians' incentives can improve the quality of public goods provision.
-Some have focused on the direct link between the political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}.
-Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce positive changes \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate.
-Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.
+In much of the developing world, political actors do not govern or manage alone.
+Particularly in developing democracies, the political landscape is composed of multiple elected officials and power-sharing arrangements which ultimately shape the political nexus between elected officials and bureaucratic institutions \citep{francois_how_2015}.
+In the context of a separation of powers, the need for alignment between executive and legislative actors can give rise to political competition and the need to form coalitions that allow executive leaders to govern and implement their policy agenda \citep{cameron_models_2004, gottlieb_countervailing_2019}.
+Where programmatic commitments are weak, this coalition-formation -- known as presidential coalitionism --  manifests itself as transactions in the form of patronage appointments to help secure votes by legislators \citep{power_optimism_2010, amorim_neto_presidential_2000}.
+Appointments into education, due to their high social prestige and potential for electoral mobilization \citep{oliveros_making_2016}, can be particularly valuable.
 
-To shed light on this complexity, over two years I conducted fieldwork research in municipalities across Brazil.
+These patronage transactions between political elites have important consequences for bureaucratic institutions.
+As noted by \citet{evans_embedded_1995}, one feature of high-performing bureaucracies is their cohesion: an ability to retain and select personnel without being exposed to cycles of disruption.
+Patronage appointment
+
+To shed light on this complexity, over the course of two years I conducted fieldwork research in municipalities across Brazil.
 In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.

### 2020-10-21 paper/paper.tex
reframe drafting

--- 2020-10-13 paper/paper.tex
@@
 Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
-The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes result from political choices, but a growing body of literature recognizes the important role bureaucratic institutions mediating these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}.
+Yet in the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
+A growing recognition of the importance of bureaucrats in explaining these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015} has been accompanied by concern over the politician-bureaucrat nexus \citet{raffler_does_2016,gulzar_politicians_2017}.
+These actors, political and bureaucratic, have been extensively modeled under the dyadic framework of principal-agent models \citep{miller2005political}.
+Yet in most developing countries, these political and bureaucratic actors are multiple and the institutional environment under which they operate play a decisive role in how public services are managed and delivered.
 
-Understanding how these political decisions cascade down the bureaucracy remains a challenge.
-Public services -- and their quality -- are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.
-This panorama of decisions belie a multiplicity of actors that is often obscured in prevailing models of
+In developing democracies, and within their political constituencies, politicians are seldom unitary.
+Elected officials, both executive and legislative, have at times overlapping or countervailing roles, with important implications as to how bureaucracies are staffed or the amount of discretion they receive \citep{huber_legislatures_2001,gailmard_formal_2012}.
+Where power-sharing arrangements exist, bureaucratic appointments reflect the relative power of particular constituencies \citep{francois_how_2015}.
+In the context of a separation of powers, the need for alignment between executive and legislative actors may require elected officials, in particular executive leaders who have the power to appoint staff into the bureaucracy, to form coalitions with legislators \citep{groseclose_1996_buying, banks_2000_buying, gottlieb_countervailing_2019}.
+Where programmatic commitments are weak, coalition-formation manifests itself as transactions in the form of patronage appointments to help secure votes by legislators \citep{power_optimism_2010, amorim_neto_presidential_2000}.
 
-How can we intervene in this chain to improve the quality of public services?
-Previous studies have explored how politicians' incentives can improve the quality of public goods provision.
-Some have focused on the direct link between the political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}.
-Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce positive changes \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate.
-Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.
+From an administrative perspective, individual bureaucrats are embedded in an administrative apparatus.
+And a key feature that distinguishes high performing bureaucratic institutions are relatively high levels of cohesion \citet{evans_embedded_1995,evans_bureaucracy_1999}.
+In other words, well-functioning bureaucracies are characterized by an ability to meritocratically recruit and retain talent.
+This is an organizational characteristic, which goes beyond any single bureaucrat and takes into account the bureaucracy as a whole.
+In contrast, bureaucracies exposed to high levels of turnover, in particular when the drivers of staff turnover are political, can display low levels of cohesion and morale.
+\citep{kohli_state-directed_2004, akhtari_political_2015}.
+Patronage transactions between elected officials, in particular when these result into a wave of new political appointees, disrupt established organizational procedures, throwing management into disarray and requiring an adjustment period to the new entries.
 
-To shed light on this complexity, over two years I conducted fieldwork research in municipalities across Brazil.
+To shed light on this complexity, over the course of two years I conducted fieldwork research in municipalities across Brazil.
 In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.

### 2020-11-09 source/descriptive/modules/visual_teacher.R
create descriptive statistics for teachers

@@ -3,8 +3,8 @@
 library(scales)
 
 censo_teacher <- read_data(
-  "censo_escolar",
-  "censo_teacher_dep.rds"
+    "censo_escolar",
+    "censo_teacher_dep.rds"
 )
 
 censo_teacher_latest <- censo_teacher %>%
@@ -12,7 +12,7 @@ censo_teacher_latest <- censo_teacher %>%
 
 # number of teachers per school for latest available data
 plot_n_teacher_per_school <- censo_teacher_latest %>%
-   gg_histogram(
+    gg_histogram(
         aes(n_teacher),
         binwidth = 1
     ) +
@@ -33,5 +33,34 @@ save_fig(
 plot_prop_higher_edu_teachers <- censo_teacher_latest %>%
     gg_histogram(
         aes(prop_higher_edu),
-        binwidth = 0.01
+        binwidth = 0.05
+    )
+
+save_fig(
+    plot_n_teacher_per_school,
+    "plot_number_teachers_per_school.pdf"
+)
+
+# age distribution
+censo_teacher_latest %>%
+    gg_histogram(
+        aes(mean_age),
+        bindwidth = 1
+    )
+
+# create table of descriptive statistics
+censo_teacher_latest %>%
+    select(-cod_ibge_6, -year, -school_id) %>%
+    as.data.frame() %>%
+    stargazer(
+        covariate.labels = c(
+            "Number of teachers per school",
+            "Number of classes per school",
+            "Proportion of teachers w/ higher education",
+            "Average age",
+            "Average nunber of classes per teacher"
+        ),
+        omit.summary.stat = c("p25", "p75"),
+        type = "latex",
+        out = here("figures/results/descriptive_teacher.tex")
     )

### 2020-11-09 paper/paper.tex
initiate descriptives for saeb_principal

@@
 
+Teachers and school principals are responsible for providing all educational services at the local level.
+As of 2016, in the municipal education system there were 3.7 million teachers distributed across the national territory, as well as 148 thousand school principals.
+TableA majority of teachers are hired with a permanent contract (\emph{estatutário}) Over the past decade, there has been an increasing
+
+\begin{table}[h]
+    \centering
+    \footnotesize
+    \input{results/turnover_result}
+    \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
+Models 1 and 2 present results on the turnover index at the school level.
+Models 2 and 3 are aggregate hiring rates at the municipal level.
+Models 5 and 6 are logistic regressions at the individual, bureaucrat level.
+where the outcome of interest is the proportion of staff either hired or dismissed at a given year.
+Models 1, 3, and 5 include year and state fixed effects.}
+    \label{tbl:turnover}
+  \end{table}
+
 The local executive branch has sole jurisdiction over the hiring and firing of teachers and school principals.
@@
 Brazil collects fine-grained data on education and makes it publicly available for research.
-For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools.
+For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute for Educational Studies and Research) to all public schools and a sample of private schools.
 In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese.

### 2020-11-09 source/data/modules/datagen_censo_escolar.R

@@ -13,7 +13,7 @@ censo_school <- read_data(
 )
 
 # keep only active schools and generate municipal table
-censo_school_mun <- censo_school %>%
+censo_school <- censo_school %>%
   filter(
     str_detect(active, "(1|2|ativ)")
   ) %>%
@@ -36,7 +36,7 @@ censo_school_mun <- censo_school %>%
       high_sch_normal,
       na.rm = T
     ),
-     prop_electricity = 1 - mean(
+    prop_electricity = 1 - mean(
       electricity_inexistent,
       na.rm = T
     ),
@@ -55,7 +55,7 @@ censo_school_mun <- censo_school %>%
     prop_kitchen = mean(
       kitchen,
       na.rm = T
-    ),    
+    ),
     prop_rural = mean(
       location == "rural",
       na.rm = T
@@ -88,33 +88,13 @@ censo_school_mun <- censo_school %>%
   ungroup()
 
 # fix entries
-censo_school_mun <- censo_school_mun %>%
+censo_school <- censo_school %>%
   mutate(
     n_pta = if_else(year == 2004, n_pta, NA_integer_),
     n_tchr_unions = if_else(year == 2004, n_tchr_unions, NA_integer_)
   )
 
-censo_school_mun %>%
-  write_data(
-    "censo_escolar", "censo_school_mun.rds"
-  )
-
 censo_school %>%
-  transmute(
-    cod_ibge_6,
-    year,
-    dep,
-    school_id,
-    school_name,
-    active,
-    association_parents_tchrs,
-    access_electricity = 1 - electricity_inexistent,
-    access_water = 1 - water_inexistent,
-    library,
-    lab_info,
-    meal,
-    toilet
-  ) %>%
   write_data(
     "censo_escolar",
     "censo_school.rds"
@@ -127,49 +107,70 @@ reset_env(init_env)
 # missing data for rio grande do sul and santa catarina
 init_env <- ls()
 
-teacher <- read_data(
-  "raw",
-  "censo_escolar",
-  "censo_teacher_sample.csv.gz"
+teacher_files <- c(
+  "censo_teacher_sample.csv.gz",
+  "docente_mun.csv.gz"
+)
+
+teacher <- map(
+  teacher_files,
+  ~ read_data(
+    "raw",
+    "censo_escolar",
+    .
+  )
 )
 
 teacher_summary <- teacher %>%
-  group_by(
-    year,
-    cod_ibge_6,
-    dep,
-    school_id
-  ) %>%
-  summarise(
[…]

### 2020-11-09 source/descriptive/modules/visual_teacher.R

@@ -7,6 +7,10 @@ censo_teacher <- read_data(
     "censo_teacher_dep.rds"
 )
 
+saeb_principal <- read_data(
+    
+)
+
 censo_teacher_latest <- censo_teacher %>%
     filter(dep == "municipal" & year == max(year))
 

### 2020-11-13 paper/paper.tex
fix width of graphics

@@
 \newcommand{\blue}{\textcolor{blue}}
-% \renewcommand{\includegraphics}{\includegraphics[width=0.6\linewidth]}
+% \renewcommand{\includegraphics}{\includegraphics}
 
@@
 
-Brazil's local level of government is composed of over 5 thousand municipalities, each with their own mayorship (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
-Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, open to reelection for both offices.\footnote{The last municipal elections were in 2016.
+Brazil's local level of government is composed of over 5 thousand municipalities, each with their own city hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
+Mayors and city councilors (\textit{vereadores}) are democratically elected every four years, open to reelection for both offices.\footnote{The last municipal elections were in 2016.
 These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over public administration in general, and personnel appointment in particular \citep{akhtari_political_2015,brollo_victor_2017}.
@@
 
-At the federal level, this instituitonal arrangement gives rise to a phenomenon dscribed as presidential coalitionism \citet{raile_executive_2011, power_optimism_2010}.
+At the federal level, this institutional arrangement gives rise to a phenomenon described as presidential coalitionism \citet{raile_executive_2011, power_optimism_2010}.
 Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
+    \includegraphics{plots/descriptive_enrollment.pdf}
     \caption{\textbf{Student enrollment by administration.}}
@@
 TableA majority of teachers are hired with a permanent contract (\emph{estatutário}) Over the past decade, there has been an increasing
-
-\begin{table}[h]
-    \centering
-    \footnotesize
-    \input{results/turnover_result}
-    \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
-Models 1 and 2 present results on the turnover index at the school level.
-Models 2 and 3 are aggregate hiring rates at the municipal level.
-Models 5 and 6 are logistic regressions at the individual, bureaucrat level.
-where the outcome of interest is the proportion of staff either hired or dismissed at a given year.
-Models 1, 3, and 5 include year and state fixed effects.}
-    \label{tbl:turnover}
-  \end{table}
 
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown}
+    \includegraphics{plots/descriptive_staff_breakdown}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.5\linewidth]{plots/descriptive_staff_turnover.pdf}
+    \includegraphics{plots/descriptive_staff_turnover.pdf}
     \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
+    \includegraphics{plots/hist_coalition.pdf}
     \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
+    \includegraphics{plots/plot_coalition_coef_mun.pdf}
     \caption{{\bf Political alignment and staff turnover.} A visual representation of results presented in table \ref{tbl:turnover}. All models include year and state fixed effects.}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
+    \includegraphics{plots/plot_interactions_logit}
     \caption{\textbf{Differential reductions in patronage in teachers and school principals}.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun_second.pdf}
+    \includegraphics{plots/plot_coalition_coef_mun_second.pdf}
     \caption{\textbf{Patronage continues into the second term.} There is strong evidence that executive-legislative bargains continue into the mayor's second term.
@@
   \centering
-  \includegraphics[width=0.6\linewidth]{plots/model_turnover_learning.pdf}
+  \includegraphics{plots/model_turnover_learning.pdf}
   \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning, across different sets of exams (\emph{SAEB} and \emph{SPAECE}).
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/reelection_marginal_effect_plots.pdf}
+    \includegraphics{plots/reelection_marginal_effect_plots.pdf}
     \caption{\textbf{Differential returns to patronage}.
@@
     \centering
-    \includegraphics[width=0.7\linewidth]{plots/visual_rdd_legislative_seat_flip.pdf}
+    \includegraphics{plots/visual_rdd_legislative_seat_flip.pdf}
 \end{figure}

### 2020-11-13 source/descriptive/modules/visual_principal.R

@@ -32,16 +32,25 @@ plot_experience <- saeb_principal %>%
     plot_histogram(
         saeb_principal_experience
     ) +
-    ggtitle("Years of Experience")
+    ggtitle("Years of experience")
 
 plot_appointment <- saeb_principal %>%
     filter(!is.na(saeb_principal_appointment)) %>%
+    mutate(
+        saeb_principal_appointment = if_else(
+            str_detect(saeb_principal_appointment, "selection"),
+            "selection process", saeb_principal_appointment
+        )
+    ) %>%
     plot_histogram(
         saeb_principal_appointment
     ) +
-    ggtitle("Appointment Process")
+    ggtitle("Appointment process")
 
 plot_education <- saeb_principal %>%
+    filter(
+        saeb_principal_education %in% c("higher education", "high school")
+    ) %>%
     plot_histogram(
         saeb_principal_education
     ) +
@@ -53,7 +62,7 @@ plot_age <- saeb_principal %>%
     ) +
     ggtitle("Age")
 
-plot_principal_descriptives <- ggpubr::ggarrange(    
+plot_principal_descriptives <- ggpubr::ggarrange(
         plot_experience,
         plot_appointment,
         plot_education,
@@ -64,5 +73,7 @@ plot_principal_descriptives <- ggpubr::ggarrange(
 
 save_fig(
     plot_principal_descriptives,
-    "plot_principal_descriptive.pdf"
+    "plot_principal_descriptive.pdf",
+    width = 9,
+    height = 5
 )

### 2020-11-16 paper/results/descriptive_teacher.tex
commit changes to saeb exam scores

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Mon, Nov 09, 2020 - 12:09:20 PM
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Number of teachers per school & 72,642 & 4.108 & 4.335 & 1 & 86 \\ 
Number of classes per school & 72,642 & 4.126 & 4.417 & 1 & 173 \\ 
Proportion of teachers w/ higher education & 72,642 & 0.729 & 0.377 & 0 & 1 \\ 
Average age & 72,642 & 0.811 & 0.283 & 0.000 & 1.000 \\ 
Average nunber of classes per teacher & 72,642 & 40.008 & 7.251 & 16.000 & 75.000 \\ 
mean\_n\_class\_per\_teacher & 72,642 & 1.025 & 0.248 & 0 & 7 \\ 
\hline \\[-1.8ex] 
\end{tabular}

### 2020-11-16 source/data/modules/datagen_saeb.R

@@ -44,7 +44,7 @@ saeb_school <- saeb_student %>%
         pc_student,
         age_student,
         brown_black_student,
-      grade_exam
+        grade_exam
       ),
       ~ mean(as.numeric(.x), na.rm = T)
     ),
@@ -121,7 +121,7 @@ saeb_hierarchical <- saeb_hierarchical %>%
       T ~ 9L
     )
   )
-  
+
 saeb_hierarchical %>%
   write_data(
     "saeb",

### 2020-11-16 source/descriptive/modules/visual_teacher.R

@@ -12,7 +12,6 @@ censo_school <- read_data(
     "censo_school.rds"
 )
 
-
 censo_teacher_latest <- censo_teacher %>%
     filter(dep == "municipal" & year == max(year))
 
@@ -64,7 +63,7 @@ censo_teacher_latest %>%
             "Number of classes per school",
             "Proportion of teachers w/ higher education",
             "Average age",
-            "Average nunber of classes per teacher"
+            "Average number of classes per teacher"
         ),
         omit.summary.stat = c("p25", "p75"),
         type = "latex",

### 2020-11-16 source/descriptive/modules/visual_test_scores.R

@@ -7,7 +7,7 @@ saeb_school <- read_data(
 saeb_public <- saeb_school %>%
   filter(
     dep != "",
-    !is.na(grade),
+    !is.na(grade_level),
     dep != "particular",
     year >= 2003
   ) %>% 
@@ -15,16 +15,16 @@ saeb_public <- saeb_school %>%
     year = as.factor(year)
   )
 
-saeb_public %>%
+saeb_school %>%
     gg_summary(
       year,
-      grade_mean
+      grade_exam
     )
 
 saeb_public %>% 
   ggplot(
     aes(
-      year, mean_grade_exam, 
+      year, grade_exam, 
       group = dep, col = dep
     )
   ) + 

### 2020-11-16 source/descriptive/modules/visual_test_scores.R
reformat saeb_school to incorporate school_id

@@ -1,75 +1,76 @@
 # test scores -------------------------------------------------------------
+# note that this data only includes info on muncipal public schools
 saeb_school <- read_data(
-    "saeb",
-    "saeb_school.rds"
+  "saeb",
+  "saeb_school.rds"
 )
 
-saeb_public <- saeb_school %>%
+censo_school_turnover <- read_data(
+  "censo_escolar",
+  "censo_school_turnover.rds"
+)
+
+saeb_school <- saeb_school %>%
   filter(
-    dep != "",
     !is.na(grade_level),
-    dep != "particular",
     year >= 2003
-  ) %>% 
-  mutate(
-    year = as.factor(year)
   )
 
 saeb_school %>%
-    gg_summary(
-      year,
-      grade_exam
-    )
+  gg_summary(
+    year,
+    grade_exam
+  )
 
-saeb_public %>% 
+saeb_school %>%
   ggplot(
     aes(
-      year, grade_exam, 
+      year, grade_exam,
       group = dep, col = dep
     )
-  ) + 
+  ) +
   geom_line() +
   geom_point() +
   labs(
     caption = "Average test scores, by administrative level. Includes math and Portuguese test scores."
-  ) + 
+  ) +
   facet_wrap(
     grade ~ subject
   )
 
-saeb_descriptive <- saeb_exam_mun %>% 
-  join_covariate
+saeb_descriptive <- saeb_school %>%
+  join_covariate()
 
-saeb_descriptive %>% 
+saeb_descriptive %>%
   gg_summary(
-    censo_pop, 
-    mean_grade_exam
+    censo_pop,
+    grade_exam
   ) +
   scale_x_log10()
 
-saeb_descriptive %>% 
+saeb_descriptive %>%
   gg_summary(
-    I(budget_education/censo_pop),
-    mean_grade_exam
+    I(budget_education / censo_pop),
+    grade_exam
   )
 
-saeb_exam_school %>% 
+saeb_school %>%
   inner_join(
     censo_school_turnover,
-    by = c('cod_ibge_6', 'year', 'school_id')
-  ) %>% 
+    by = c("cod_ibge_6", "year", "school_id")
+  ) %>%
   mutate(
     state = str_sub(cod_ibge_6, 1, 2)
-  ) %>% 
+  ) %>%
   group_by(
     state,
     year
-  ) %>% 
+  ) %>%
   mutate(
[…]

### 2020-11-16 paper/paper.tex
commit changes to saeb exam scores
fix descriptive stats for teacher

@@
 As of 2016, in the municipal education system there were 3.7 million teachers distributed across the national territory, as well as 148 thousand school principals.
-TableA majority of teachers are hired with a permanent contract (\emph{estatutário}) Over the past decade, there has been an increasing
+A majority of teachers are hired with a permanent contract (\emph{estatutário}) Over the past decade, there has been an increasing
+
+\begin{table}[t]
+    \input{results/descriptive_teacher}
+    \caption{\textbf{Summary statistics for municipal teachers.} All statistics take as reference year 2017 as reference.}
+\end{table}
 

### 2020-11-16 paper/results/descriptive_teacher.tex
fix descriptive stats for teacher

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Nov 09, 2020 - 12:09:20 PM
-\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
+% Date and time: Mon, Nov 16, 2020 - 10:49:28 AM
+\begin{tabular}{@{\extracolsep{5pt}}lcccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
+Statistic & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Max} \\ 
 \hline \\[-1.8ex] 
-Number of teachers per school & 72,642 & 4.108 & 4.335 & 1 & 86 \\ 
-Number of classes per school & 72,642 & 4.126 & 4.417 & 1 & 173 \\ 
-Proportion of teachers w/ higher education & 72,642 & 0.729 & 0.377 & 0 & 1 \\ 
-Average age & 72,642 & 0.811 & 0.283 & 0.000 & 1.000 \\ 
-Average nunber of classes per teacher & 72,642 & 40.008 & 7.251 & 16.000 & 75.000 \\ 
-mean\_n\_class\_per\_teacher & 72,642 & 1.025 & 0.248 & 0 & 7 \\ 
+Number of teachers per school & 4.108 & 4.335 & 1 & 86 \\ 
+Number of classes per school & 4.126 & 4.417 & 1 & 173 \\ 
+Proportion of teachers w/ higher education & 0.729 & 0.377 & 0 & 1 \\ 
+Average age & 0.811 & 0.283 & 0.000 & 1.000 \\ 
+Average number of classes per teacher & 40.008 & 7.251 & 16.000 & 75.000 \\ 
+mean\_n\_class\_per\_teacher & 1.025 & 0.248 & 0 & 7 \\ 
 \hline \\[-1.8ex] 

### 2020-11-16 source/descriptive/modules/visual_teacher.R

@@ -55,17 +55,26 @@ censo_teacher_latest %>%
 
 # create table of descriptive statistics
 censo_teacher_latest %>%
-    select(-cod_ibge_6, -year, -school_id) %>%
+    select(
+        n_teacher,
+        n_classes,
+        prop_higher_edu,
+        mean_age,
+        prop_female,
+        mean_n_class_per_teacher
+    ) %>%
     as.data.frame() %>%
-    stargazer(
+    stargazer::stargazer(
         covariate.labels = c(
             "Number of teachers per school",
             "Number of classes per school",
             "Proportion of teachers w/ higher education",
             "Average age",
+            "Proportion of women",
             "Average number of classes per teacher"
         ),
-        omit.summary.stat = c("p25", "p75"),
+        float = FALSE,
+        omit.summary.stat = c("p25", "p75", "n"),
         type = "latex",
         out = here("figures/results/descriptive_teacher.tex")
     )

### 2020-11-16 paper/paper.tex
fix principal descriptives labels

--- 2020-11-13 paper/paper.tex
@@
 In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}.
-In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and substraction.
-The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school strudents attend a public school.
+In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in mathematics.
+The municipal educational system has increased in size over the past few decades, and currently over 60 percent of lower school students attend a public school.
 Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools.
@@
 
-Teachers and school principals are responsible for providing all educational services at the local level.
+Teachers and school principals are responsible for providing all educational services at the local level, overseen by a local department of education (\emph{secretaria de educa\c{c}\~{a}o}).
 As of 2016, in the municipal education system there were 3.7 million teachers distributed across the national territory, as well as 148 thousand school principals.
-TableA majority of teachers are hired with a permanent contract (\emph{estatutário}) Over the past decade, there has been an increasing
+A majority of teachers are hired with a permanent contract (\emph{estatutário}), with over 73 percent of teachers enrolled with civil service protection.
+This contrasts with the hiring of school principals, who are politically appointed and therefore have no similar protections set in place.
+
+\begin{table}[t]
+    \input{results/descriptive_teacher}
+    \caption{\textbf{Summary statistics for municipal teachers.} All statistics take as reference year 2017.}
+    \label{tbl:descriptive_teacher}
+\end{table}
+
+Table \ref{tbl:descriptive_teacher} provides some additional descriptive statistics regarding teachers.
+The teacher labor force is highly educated and primarily composed of women, and schools in Brazil are relatively small, with an average of 4 teachers per school.
+As for school principals, while
+
+\begin{figure}[h]
+    \centering
+    \includegraphics{plots/plot_principal_descriptive.pdf}
+\end{figure}
 

### 2020-11-16 paper/results/descriptive_teacher.tex
fix principal descriptives labels

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Nov 16, 2020 - 10:49:28 AM
+% Date and time: Mon, Nov 16, 2020 - 11:05:29 AM
 \begin{tabular}{@{\extracolsep{5pt}}lcccc} 
@@
 Proportion of teachers w/ higher education & 0.729 & 0.377 & 0 & 1 \\ 
-Average age & 0.811 & 0.283 & 0.000 & 1.000 \\ 
-Average number of classes per teacher & 40.008 & 7.251 & 16.000 & 75.000 \\ 
-mean\_n\_class\_per\_teacher & 1.025 & 0.248 & 0 & 7 \\ 
+Average age & 40.008 & 7.251 & 16.000 & 75.000 \\ 
+Proportion of women & 0.811 & 0.283 & 0.000 & 1.000 \\ 
+Average number of classes per teacher & 1.025 & 0.248 & 0 & 7 \\ 
 \hline \\[-1.8ex] 

### 2020-11-16 source/descriptive/modules/visual_principal.R

@@ -32,6 +32,13 @@ plot_experience <- saeb_principal %>%
     plot_histogram(
         saeb_principal_experience
     ) +
+    scale_x_discrete(
+        labels = c(
+            "2 to 10" = "2 to 10\nyears",
+            "less than 2 years" = "less than\n 2 years",
+            "more than 10" = "more than\n10 years"
+        )
+    ) +
     ggtitle("Years of experience")
 
 plot_appointment <- saeb_principal %>%
@@ -44,6 +51,12 @@ plot_appointment <- saeb_principal %>%
     ) %>%
     plot_histogram(
         saeb_principal_appointment
+    ) +
+    scale_x_discrete(
+        labels = c(
+            "political appointment" = "political\nappointment",
+            "selection process" = "selection\nprocess"
+        )
     ) +
     ggtitle("Appointment process")
 
@@ -74,6 +87,6 @@ plot_principal_descriptives <- ggpubr::ggarrange(
 save_fig(
     plot_principal_descriptives,
     "plot_principal_descriptive.pdf",
-    width = 9,
+    width = 6,
     height = 5
 )

### 2020-11-16 source/descriptive/modules/visual_teacher.R

@@ -12,6 +12,11 @@ censo_school <- read_data(
     "censo_school.rds"
 )
 
+rais_edu <- read_data(
+    "rais",
+    "rais_mun_edu.rds"
+)
+
 censo_teacher_latest <- censo_teacher %>%
     filter(dep == "municipal" & year == max(year))
 

### 2020-11-16 paper/paper.tex
fixed descriptive for school principals

@@
 Table \ref{tbl:descriptive_teacher} provides some additional descriptive statistics regarding teachers.
-The teacher labor force is highly educated and primarily composed of women, and schools in Brazil are relatively small, with an average of 4 teachers per school.
+The teacher labor force is highly educated and primarily composed of women, and schools in Brazil are relatively small, withttps://askubuntu.com/questions/487527/give-specific-user-permission-to-write-to-a-folder-using-w-notationh an average of 4 teachers per school.
 As for school principals, while

### 2020-11-16 source/descriptive/modules/visual_principal.R

@@ -39,7 +39,11 @@ plot_experience <- saeb_principal %>%
             "more than 10" = "more than\n10 years"
         )
     ) +
-    ggtitle("Years of experience")
+    ggtitle("Years of experience") +
+    theme(
+        plot.title = element_text(size = 10, face = "bold"), 
+        legend.key.size = unit(1,"line")
+        )
 
 plot_appointment <- saeb_principal %>%
     filter(!is.na(saeb_principal_appointment)) %>%
@@ -58,7 +62,12 @@ plot_appointment <- saeb_principal %>%
             "selection process" = "selection\nprocess"
         )
     ) +
-    ggtitle("Appointment process")
+    ggtitle("Appointment process") +
+    theme(
+        plot.title = element_text(size = 10, face = "bold"), 
+        legend.key.size = unit(1,"line")
+        )
+
 
 plot_education <- saeb_principal %>%
     filter(
@@ -67,13 +76,23 @@ plot_education <- saeb_principal %>%
     plot_histogram(
         saeb_principal_education
     ) +
-    ggtitle("Education")
+    ggtitle("Education") +
+    theme(
+        plot.title = element_text(size = 10, face = "bold"), 
+        legend.key.size = unit(1,"line")
+        )
+
 
 plot_age <- saeb_principal %>%
     plot_histogram(
         saeb_principal_age
     ) +
-    ggtitle("Age")
+    ggtitle("Age") +
+    theme(
+        plot.title = element_text(size = 10, face = "bold"), 
+        legend.key.size = unit(1,"line")
+        )
+
 
 plot_principal_descriptives <- ggpubr::ggarrange(
         plot_experience,
@@ -88,5 +107,5 @@ save_fig(
     plot_principal_descriptives,
     "plot_principal_descriptive.pdf",
     width = 6,
-    height = 5
+    height = 4
 )

### 2020-11-16 paper/paper.tex
create new descriptives by year for rais

--- 2020-11-16 paper/paper.tex
@@
 
-Table \ref{tbl:descriptive_teacher} provides some additional descriptive statistics regarding teachers.
-The teacher labor force is highly educated and primarily composed of women, and schools in Brazil are relatively small, with an average of 4 teachers per school.
-As for school principals, while
+Table \ref{tbl:descriptive_teacher} provides some additional descriptive statistics.
+The teacher labor force is highly educated, with over 70 percent of teachers holding a higher education degree.
+The majority of teachers are women, and schools have a limited amount of teachers, with an average of 4 per school.
 

### 2020-11-16 source/data/modules/datagen_rais.R

@@ -70,3 +70,23 @@ rais_edu_mun %>%
     write_data(
         "rais", "rais_mun_edu.rds"
     )
+
+# descriptive by year
+rais_edu_year <- rais_edu_mun %>%
+    select(-cod_ibge_6, -election_year) %>%
+    group_by(
+        rais_category,
+        year
+    ) %>%
+    summarise(
+        across(
+            everything(),
+            ~ weighted.mean(., w = rais_size, na.rm = T)
+        ),
+        .groups = "drop"
+    )
+
+rais_edu_year %>%
+    write_data(
+        "rais", "rais_year_edu.rds"
+    )

### 2021-03-30 paper/__latexindent_temp.tex
update images

--- 2020-11-16 paper/paper.tex
@@
 }
-\captionsetup[figure]{font=small,labelfont=small}
 
@@
 \newcommand{\blue}{\textcolor{blue}}
-% \renewcommand{\includegraphics}{\includegraphics}
+% \renewcommand{\includegraphics}{\includegraphics[width=0.6\linewidth]}
 
@@
 \begin{abstract}
-    How can governments improve the quality of education?
-I argue that local politicians govern through patronage, undermining education at the expense of voters.
-Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites.
-Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector.
-The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain.
-Patronage induces turnover in educational staff, with negative downstream effects on student learning.
-Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate.
+    Why do politicians engage in patronage and what are its consequences for public service delivery?
+I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators.
+Leveraging fieldwork evidence and administrative data, I demonstrate that mayors with a weaker hold on the local legislature resort to patronage to co-opt legislators.
+Educational staff turnover has negative downstream effects on student learning, but limited electoral backlash.
+Weak electoral accountability channels suggests that patronage is primarily a political elite game, with little involvement by the electorate.
 These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
@@
 
-% what do I need to bring the audience onboard?
-% emphasize the existence of discretion, turnover, political manipulation of public resources.
-% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns
+% Across the world, vulnerable populations rely on governments for access to basic public services such as education.
+In many countries, particularly in the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, mathematical operations, elusive.
+Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them.
+\citep{cochrane_policies_1983, falleti_decentralization_2010}.
+How can local governments improve the quality the of public services they deliver?
+In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?
 
-Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-Yet in the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
-A growing recognition of the importance of bureaucrats in explaining these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015} has been accompanied by concern over the politician-bureaucrat nexus \citet{raffler_does_2016,gulzar_politicians_2017}.
-These actors, political and bureaucratic, have been extensively modeled under the dyadic framework of principal-agent models \citep{miller2005political}.
-Yet in most developing countries, these political and bureaucratic actors are multiple and the institutional environment under which they operate play a decisive role in how public services are managed and delivered.
+The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations.
+How can these educational outcomes be improved?
+Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
+Often relying on quasi-experimental research designs, these studies provide compelling evidence that political and bureaucratic actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?
 
-In developing democracies, and within their political constituencies, politicians are seldom unitary.
-Elected officials, both executive and legislative, have at times overlapping or countervailing roles, with important implications as to how bureaucracies are staffed or the amount of discretion they receive \citep{huber_legislatures_2001,gailmard_formal_2012}.
-Where power-sharing arrangements exist, bureaucratic appointments reflect the relative power of particular constituencies \citep{francois_how_2015}.
-In the context of a separation of powers, the need for alignment between executive and legislative actors may require elected officials, in particular executive leaders who have the power to appoint staff into the bureaucracy, to form coalitions with legislators \citep{groseclose_1996_buying, banks_2000_buying, gottlieb_countervailing_2019}.
-Where programmatic commitments are weak, coalition-formation manifests itself as transactions in the form of patronage appointments to help secure votes by legislators \citep{power_optimism_2010, amorim_neto_presidential_2000}.
-
-From an administrative perspective, individual bureaucrats are embedded in an administrative apparatus.
-And a key feature that distinguishes high performing bureaucratic institutions are relatively high levels of cohesion \citet{evans_embedded_1995,evans_bureaucracy_1999}.
-In other words, well-functioning bureaucracies are characterized by an ability to meritocratically recruit and retain talent.
-This is an organizational characteristic, which goes beyond any single bureaucrat and takes into account the bureaucracy as a whole.
-In contrast, bureaucracies exposed to high levels of turnover, in particular when the drivers of staff turnover are political, can display low levels of cohesion and morale.
-\citep{kohli_state-directed_2004, akhtari_political_2015}.
-Patronage transactions between elected officials, in particular when these result into a wave of new political appointees, disrupt established organizational procedures, throwing management into disarray and requiring an adjustment period to the new entries.
-
-To shed light on this complexity, over the course of two years I conducted fieldwork research in municipalities across Brazil.
-In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.
-However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}.
-In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.
-
-In this paper, I argue that public education in municipalities across Brazil have been captured by political elites, with personnel appointments into public schools and management dominated by bargains between executive and legislative branches.
+In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to capture by local political elites.
 Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}.
 This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}.
-By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.
+By catering to the city council through patronage appointments, mayors ultimately cause instability in the local educational bureaucracy, with negative downstream consequences for public school students.
 
-The estimation proceeds in two parts.
-First, I show that political alignment between mayors and city councilors has a direct effect on patronage.
-Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage.
-To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities.
-In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage.
-This results holds across a set of specifications and different measures of turnover.
+To analyze how politics affect staff turnover, I analyze the relationship between municipal education staff turnover and the share of seats controlled by the executive in the local city council.
+In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage, in particular during the first year of administration.
+Mayors therefore resort to patronage in order to shore up legislative support, with negative consequences for student learning.
+I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover.
+Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).
 
-I then show that this politically induced turnover has a negative impact on student learning.
 To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence.
-Interviews conducted with educational bureaucrats and politicians suggested firm that turnover had a negative impact on teachers' ability to educate students, as time horizons were compressed and bonds between educators and learners were broken.
-To validate these accounts I combine administrative data on education from two separate student learning measures: \emph{Prova Brasil} and the \emph{SPAECE}.
-I construct multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory.
+Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students.
+To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning.
+I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory.
+Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff.
 A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.
 
-This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}.
-I analyze how incentives shape political decision on how to manage local bureaucracies \citep{duflo_incentives_2012, gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders.
-I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services.
-\citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}.
-I bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.
+This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015, gulzar_politicians_2017}.
+Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services \citep{ferraz_corrupting_2012}.
+It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy \citep{kohli_state-directed_2004, evans_embedded_1995}.
 
 The paper is structured as follows.
-Section \ref{sec:literature} provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil.
-Section \ref{sec:context} presents the context and data.
-In section \ref{sec:theory}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model.
-Section \ref{sec:empirics} presents the research design and main results.
+Section 1 provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil.
+In section 2 I present the main theoretical argument, with a formal treatment of the subject.
+Section 3 presents the institutional context and data, followed in section 4 by the research design and main results.
 Section 5 concludes.
@@
 \section{Related literature}
-\label{sec:literature}
 
@@
 
-% \subsection*{Presidential coalitionism and patronage}
+\subsection*{Presidential coalitionism and patronage}
 
-% For every elected mayor in Brazil, a group of legislators are also elected into office.
+For every elected mayor in Brazil, a group of legislators are also elected into office.
 These political actors have competing claims over the local bureaucracy, with important implications for public service delivery.
@@
 
-% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism.
+In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism.
 Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
@@
 
-% In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
+In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
 Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
@@
 \section{Institutional context and data}
-\label{sec:context}
-
-\subsection{Local governance and mayoral coalitions}
-
-Brazil's local level of government is composed of over 5 thousand municipalities, each with their own city hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
-Mayors and city councilors (\textit{vereadores}) are democratically elected every four years, open to reelection for both offices.\footnote{The last municipal elections were in 2016.
-These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over public administration in general, and personnel appointment in particular \citep{akhtari_political_2015,brollo_victor_2017}.
-As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.
-The city council, on the other hand, oversees legislation and approves the budget for the fiscal year, providing a local checks and balances on the executive.
-
-At the federal level, this institutional arrangement gives rise to a phenomenon described as presidential coalitionism \citet{raile_executive_2011, power_optimism_2010}.
-Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
-\citep{raile_executive_2011}.
-In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.
-
-In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels \citet{dantas_eleicoes_2013}.
-These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
-Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}).
-While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.\footnote{Interview with C, former chief of staff of municipality J.
-September 2018.}
-
-Interviews with secretaries of education and mayors provide strong evidence that city councilors play an important role in staffing decisions, despite having no \emph{formal} rights to do so.
-While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
-These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council.
-According to accounts by school managers in the municipality of I:
-
-\begin{quote}
-    \emph{School principal}: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
-
-    \emph{Educational councilor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J.
-August 2019.}
-\end{quote}
-
-In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
-Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
-As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.
 
@@
 
-In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}.
-In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in mathematics.
-The municipal educational system has increased in size over the past few decades, and currently over 60 percent of lower school students attend a public school.
+In Brazil, responsibility for primary education is delegated to local governments \citet{paschoal_historia_2009}.
+The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments.
 Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools.
@@
     \centering
-    \includegraphics{plots/descriptive_enrollment.pdf}
-    \caption{\textbf{Student enrollment by administration.}}
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
+    \caption{enrollment}
     \label{fig:enroll}
-\end{figure}
-
-Teachers and school principals are responsible for providing all educational services at the local level, overseen by a local department of education (\emph{secretaria de educa\c{c}\~{a}o}).
-As of 2016, in the municipal education system there were 3.7 million teachers distributed across the national territory, as well as 148 thousand school principals.
-A majority of teachers are hired with a permanent contract (\emph{estatutário}), with over 73 percent of teachers enrolled with civil service protection.
-This contrasts with the hiring of school principals, who are politically appointed and therefore have no similar protections set in place.
-
-\begin{table}[t]
-    \input{results/descriptive_teacher}
-    \caption{\textbf{Summary statistics for municipal teachers.} All statistics take as reference year 2017.}
-    \label{tbl:descriptive_teacher}
-\end{table}
-
-Table \ref{tbl:descriptive_teacher} provides some additional descriptive statistics regarding teachers.
-The teacher labor force is highly educated and primarily composed of women, and schools in Brazil are relatively small, with an average of 4 teachers per school.
-As for school principals, while
-
-\begin{figure}[h]
-    \centering
-    \includegraphics{plots/plot_principal_descriptive.pdf}
 \end{figure}
@@
 Due to their relatively high compensation and social status compared to other positions in the local bureaucracy, educational staff positions are particularly valuable for constituencies seeking employment.
-Additionally, qualitative evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.
+Additionally, anecdotal evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.
 
@@
     \centering
-    \includegraphics{plots/descriptive_staff_breakdown}
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown}
 \end{figure}
@@
 
+\subsection{Local governance and mayoral coalitions}
+
+Brazil's local level government is composed by over 5 thousand municipalities, each with their own mayor hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
+Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.\footnote{The last municipal elections were in 2016.
+These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over personnel apppointment.
+The city council, on the other hand, oversees legislation and approves the budget for the fiscal year.
+As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.
+
+In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels \citet{dantas_eleicoes_2013}.
+These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
+Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}).
+While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.\footnote{Interview with C, former chief of staff of municipality J.
+September 2018.}
+
+Interviews with secretaries of education and mayors provide evidence that city councilors play an important role in staffing decisions.
+While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
+These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council.
+In the words of a set of school administrators in the municipality of I:
+
+\begin{quote}
+    \emph{School principal}: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
+
+    \emph{Educational counselor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J.
+August 2019.}
+\end{quote}
+
+These accounts, along with previous case studies of municipalities in Brazil, suggest that city councilors play an important role in nominating staff.
+To verify these claims in a broader set of municipalities, as well as outlining the research design, I apply a theory of legislative vote-buying that illustrates the  employ extensive administrative and electoral data collected by the Brazilian federal government.
+In the next section I describe the data, where it is publicly available, as well as the preparation necessary for the set of estimations presented in the study.
+
 \subsection{Data}
@@
 Brazil collects fine-grained data on education and makes it publicly available for research.
-For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute for Educational Studies and Research) to all public schools and a sample of private schools.
+For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools.
 In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese.
@@
     \centering
-    \includegraphics[]{plots/saeb_map.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
     \caption{Uneven quality of municipal education.
@@
     \centering
-    \includegraphics{plots/descriptive_staff_turnover.pdf}
+    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
+    \label{fig:staff_turnover}
     \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
-    \label{fig:staff_turnover}
 \end{figure}
@@
 
-\blue{Insert more information and descriptive paragraph on turnover index here}
-
 Information on mayor, city councilors and mayoral coalitions are made available by the Supreme Electoral Court (TSE), the national authority responsible for overseeing elections.\footnote{Available \hyperlink{http://www.tse.jus.br/eleicoes/estatisticas/repositorio-de-dados-eleitorais-1}{here}.} In order to calculate the share of legislative seats held by the mayoral coalition, I match the partisanship of each city councilor to the mayoral electoral coalition.
@@
     \centering
-    \includegraphics{plots/hist_coalition.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
     \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
@@
 Additional data has been collected to supplement the estimation exercise.
-Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\hyperlink{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}{IBGE}, (2)\hyperlink{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}{Finbra}, (3)\hyperlink{https://www.seduc.ce.gov.br/spaece/}{Spaece}.}
+Sociodemographic data comes from the National Institute for Statistics and Geography (IBGE), budgetary data from (FINBRA) and student test scores from Ceará (SPAEEC) are graciously provided by the Department of Education of Ceará.\footnote{Respectively, data can be found here: (1)\url{https://ces.ibge.gov.br/base-de-dados/links-base-de-dados.html}, (2)\url{https://siconfi.tesouro.gov.br/siconfi/pages/public/consulta_finbra/finbra_list.jsf}, (3)\url{https://www.seduc.ce.gov.br/spaece/}.}
 
@@
 \section{Empirical results}
-\label{sec:empirics}
 
@@
 The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education.
-To do so, I leverage the test scores made available by \emph{SAEB} and \emph{SPAECE} to estimate the effect of turnover on student test scores.
+To do so, I leverage the test scores made available by SAEB and SPAECE to estimate the effect of turnover on student test scores.
 To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample.
@@
 
-% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational.
+Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational.
 In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge.
@@
     \centering
-    \includegraphics{plots/plot_coalition_coef_mun.pdf}
-    \caption{{\bf Political alignment and staff turnover.} A visual representation of results presented in table \ref{tbl:turnover}. All models include year and state fixed effects.}
+  \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
+  \caption{{\bf Political alignment and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
+Model 1 presents results on the turnover index.
+Models 2 and 3 are logistic regressions at the individual, bureaucrat level.
+Models 4 and 5 are aggregated at the municipal level.
+where the outcome of interest is the proportion of staff either hired or dismissed at a given year.
+Models 1, 3, and 5 include year and state fixed effects.}
   \label{fig:coalition_turnover}
@@
     \centering
-    \includegraphics{plots/plot_interactions_logit}
-    \caption{\textbf{Differential reductions in patronage in teachers and school principals}.
-Predicted values for bureaucratic turnover plotted against the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.}
+    \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
+    \caption{Predicted values for bureaucratic turnover.
+The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive.
+The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
     \label{fig:turnover}
@@
 Finally, I show that the executive-legislative bargain extends to second-term mayors.
-I subset the data used for the above analysis to only mayors who are reelected, and re-run the previous analyses with the same specification.
-I find that the results are similar in magnitude and precision to those presented above.
 This finding suggests that patronage is not motivated solely by reelection concerns.
-Rather, patronage achieves an important policy goal for incumbent mayors who seek to implement their preferred policy, regardless of whether they are just initiating their mandate or concluding it.
+Rather, it achieves an important policy goal for incumbent mayors who seek to implement their preferred policy.
 
-\begin{figure}[h]
+\begin{figure}
     \centering
     \includegraphics{plots/plot_coalition_coef_mun_second.pdf}
-    \caption{\textbf{Patronage continues into the second term.} There is strong evidence that executive-legislative bargains continue into the mayor's second term.
-All models include a full set of controls, as well as year and state fixed effects.}
+    \caption{\bf{hello}}
 \end{figure}
@@
 These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations.
-Table \ref{tbl:student_learning} presents the results of the estimation on SAEB and SPAECE test scores.
+Table \ref{fig:hlm_mods} presents the results of the estimation on SAEB test scores.
 We present two specifications for turnover.
@@
 Work experience for teachers and school principals serve as an alternative measure of turnover.
+Personnel who has only recently entered the school recently amass a shorter amount of work experience (less than two years).
 
-\begin{landscape}
-    \begin{table}[t]
+\begin{figure}[h]
   \centering
   \footnotesize
-      \input{results/student_learning.tex}
+  \includegraphics{plots/model_turnover_learning.pdf}
   \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning.
@@
 All models include year and state fixed effects.}
-      \label{tbl:student_learning}
-    \end{table}
-\end{landscape}
-
-I also present the coefficients for the regression exercise above in figure \ref{fig:hlm_mods}.
-Note that the results are consistent across the board, providing strong evidence that staff turnover has negative downstream consequences for student learning.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics{plots/model_turnover_learning.pdf}
-  \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning, across different sets of exams (\emph{SAEB} and \emph{SPAECE}).
-All models include a full set of controls, as well as year and state fixed effects.}
   \label{fig:hlm_mods}
@@
 
-Finally, I present some evidence of differential returns to patronage, in particular for city councilors.
-For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education.
-The negative consequences of patronage on student learning, however, do not bite, as nayors who deliver a worse quality of education are no less likely to be reelected in the next term.\footnote{See appendix \ref{app:accountability}.}
+\blue{Include some reference coefs in order for the audience to assess the magnitude of the effects}
 
-\begin{figure}[h]
-    \centering
-    \includegraphics{plots/reelection_marginal_effect_plots.pdf}
-    \caption{\textbf{Differential returns to patronage}.
-Mayors who engage in patronage do not directly benefit from it.
-Instead, city councilors in the mayoral ally base seem to be benefiting from it at the expense of non-allied city councilors.}
-\end{figure}
-
-As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity.
+% As a final note, this set of estimations are based on observational data and therefore suffer from well-known concerns of endogeneity.
 It is in these circumstances that validating causal mechanisms through fieldwork and careful data analysis can increase the validity of a causal claim.
@@
 \section{Conclusion}
-\label{sec:conclusion}
 
@@
 
-\pagebreak
+% \pagebreak
 
-\bibliography{dissertation}
+% # Appendix
 
-\pagebreak
+% \pagebreak
 
-\section*{Appendix}
+% ## Covariate balance
 
-\subsection{Covariate balance}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics{plots/covariate_balance.pdf}
-\end{figure}
-
-\pagebreak
-
-\subsection{Accountability of mayors}
-\label{app:accountability}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics{plots/accountability_fit.pdf}
-\end{figure}
-
-\pagebreak
-
-\subsection{Regression Discontinuity Design}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics{plots/visual_rdd_legislative_seat_flip.pdf}
-\end{figure}
+% ### Education quality vs. staff turnover
 
@@
 
+\bibliography{dissertation}
+
 \end{document}

### 2021-03-30 paper/paper.tex
update images

--- 2020-11-16 paper/paper.tex
@@
 \usepackage{setspace}
-\usepackage[top = 1in, bottom = 1in, left = 1in, right = 1in]{geometry}
+\usepackage[top = 1.5in, bottom = 1.5in, left = 1.5in, right = 1.5in]{geometry}
 \usepackage[utf8]{inputenc}
@@
 
-% what do I need to bring the audience onboard?
-% emphasize the existence of discretion, turnover, political manipulation of public resources.
-% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns
+Across the world, vulnerable populations rely on governments for access to basic public services such as education.
+In the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, basic mathematical operations.\footnote{The World Bank calls this ``learning poverty".
+See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}.
+Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them.
+\citep{cochrane_policies_1983, falleti_decentralization_2010}.
+How can local governments improve the quality the of public services they deliver?
+In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?
 
-Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-Yet in the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}.
-A growing recognition of the importance of bureaucrats in explaining these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015} has been accompanied by concern over the politician-bureaucrat nexus \citet{raffler_does_2016,gulzar_politicians_2017}.
-These actors, political and bureaucratic, have been extensively modeled under the dyadic framework of principal-agent models \citep{miller2005political}.
-Yet in most developing countries, these political and bureaucratic actors are multiple and the institutional environment under which they operate play a decisive role in how public services are managed and delivered.
+The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations.
+How can public education be improved?
 
-In developing democracies, and within their political constituencies, politicians are seldom unitary.
-Elected officials, both executive and legislative, have at times overlapping or countervailing roles, with important implications as to how bureaucracies are staffed or the amount of discretion they receive \citep{huber_legislatures_2001,gailmard_formal_2012}.
-Where power-sharing arrangements exist, bureaucratic appointments reflect the relative power of particular constituencies \citep{francois_how_2015}.
-In the context of a separation of powers, the need for alignment between executive and legislative actors may require elected officials, in particular executive leaders who have the power to appoint staff into the bureaucracy, to form coalitions with legislators \citep{groseclose_1996_buying, banks_2000_buying, gottlieb_countervailing_2019}.
-Where programmatic commitments are weak, coalition-formation manifests itself as transactions in the form of patronage appointments to help secure votes by legislators \citep{power_optimism_2010, amorim_neto_presidential_2000}.
+Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
+Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors relate to one another in local institutions?
+Who holds power over public education, and whose interests is it meant to serve?
 
-From an administrative perspective, individual bureaucrats are embedded in an administrative apparatus.
-And a key feature that distinguishes high performing bureaucratic institutions are relatively high levels of cohesion \citet{evans_embedded_1995,evans_bureaucracy_1999}.
-In other words, well-functioning bureaucracies are characterized by an ability to meritocratically recruit and retain talent.
-This is an organizational characteristic, which goes beyond any single bureaucrat and takes into account the bureaucracy as a whole.
-In contrast, bureaucracies exposed to high levels of turnover, in particular when the drivers of staff turnover are political, can display low levels of cohesion and morale.
-\citep{kohli_state-directed_2004, akhtari_political_2015}.
-Patronage transactions between elected officials, in particular when these result into a wave of new political appointees, disrupt established organizational procedures, throwing management into disarray and requiring an adjustment period to the new entries.
-
-To shed light on this complexity, over the course of two years I conducted fieldwork research in municipalities across Brazil.
+To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil.
 In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.
@@
 
-\subsection{Local governance and mayoral coalitions}
-
-Brazil's local level of government is composed of over 5 thousand municipalities, each with their own city hall (\emph{prefeitura}) and city council (\emph{c\^{a}mara dos vereadores}).
-Mayors and city councilors (\textit{vereadores}) are democratically elected every four years, open to reelection for both offices.\footnote{The last municipal elections were in 2016.
-These take place every 4 years.} The executive is responsible for the management of public services such as education, with exclusive rights over public administration in general, and personnel appointment in particular \citep{akhtari_political_2015,brollo_victor_2017}.
-As noted by \citet{souza_reforma_2004}, decisions over whom to appoint into the educational sector are primarily in the hands of the mayor and her secretary of education.
-The city council, on the other hand, oversees legislation and approves the budget for the fiscal year, providing a local checks and balances on the executive.
-
-At the federal level, this institutional arrangement gives rise to a phenomenon described as presidential coalitionism \citet{raile_executive_2011, power_optimism_2010}.
-Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
-\citep{raile_executive_2011}.
-In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.
-
-In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels \citet{dantas_eleicoes_2013}.
-These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber.
-Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}).
-While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.\footnote{Interview with C, former chief of staff of municipality J.
-September 2018.}
-
-Interviews with secretaries of education and mayors provide strong evidence that city councilors play an important role in staffing decisions, despite having no \emph{formal} rights to do so.
-While mayors hold jurisdiction over personnel decisions, there are extensive consultations between mayors and city councilors to decide who becomes the principal of a school, or which teacher remains in a school or leaves.
-These executive-legislative consultations ultimately determine the allocation (\textit{alotação}) of educational staff, serving as the primary tool through which mayors reward or coopt legislators into supporting them in the city council.
-According to accounts by school managers in the municipality of I:
-
-\begin{quote}
-    \emph{School principal}: Here, we are invited to work at the school by the department of education, with the [political] candidate, the city councilor...deciding which are the positions they are searching for and appointing people they think have the necessary qualifications.
-
-    \emph{Educational councilor}: I was also invited to work here, by the city councilor and the department of education.\footnote{Interview with the administrative board of school A., municipality J.
-August 2019.}
-\end{quote}
-
-In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
-Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
-As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.
-
 \subsection{Municipal education in Brazil}
@@
 In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}.
-In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in mathematics.
-The municipal educational system has increased in size over the past few decades, and currently over 60 percent of lower school students attend a public school.
+In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and substraction.
+The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school strudents attend a public school.
 Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools.
@@
     \centering
-    \includegraphics{plots/descriptive_enrollment.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
     \caption{\textbf{Student enrollment by administration.}}
     \label{fig:enroll}
-\end{figure}
-
-Teachers and school principals are responsible for providing all educational services at the local level, overseen by a local department of education (\emph{secretaria de educa\c{c}\~{a}o}).
-As of 2016, in the municipal education system there were 3.7 million teachers distributed across the national territory, as well as 148 thousand school principals.
-A majority of teachers are hired with a permanent contract (\emph{estatutário}), with over 73 percent of teachers enrolled with civil service protection.
-This contrasts with the hiring of school principals, who are politically appointed and therefore have no similar protections set in place.
-
-\begin{table}[t]
-    \input{results/descriptive_teacher}
-    \caption{\textbf{Summary statistics for municipal teachers.} All statistics take as reference year 2017.}
-    \label{tbl:descriptive_teacher}
-\end{table}
-
-Table \ref{tbl:descriptive_teacher} provides some additional descriptive statistics regarding teachers.
-The teacher labor force is highly educated and primarily composed of women, and schools in Brazil are relatively small, with an average of 4 teachers per school.
-As for school principals, while
-
-\begin{figure}[h]
-    \centering
-    \includegraphics{plots/plot_principal_descriptive.pdf}
 \end{figure}
@@
     \centering
-    \includegraphics{plots/plot_coalition_coef_mun.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
     \caption{{\bf Political alignment and staff turnover.} A visual representation of results presented in table \ref{tbl:turnover}. All models include year and state fixed effects.}
@@
     \centering
-    \includegraphics{plots/plot_interactions_logit}
+    \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
     \caption{\textbf{Differential reductions in patronage in teachers and school principals}.
@@
     \centering
-    \includegraphics{plots/plot_coalition_coef_mun_second.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun_second.pdf}
     \caption{\textbf{Patronage continues into the second term.} There is strong evidence that executive-legislative bargains continue into the mayor's second term.
@@
   \centering
-  \includegraphics{plots/model_turnover_learning.pdf}
+  \includegraphics[width=0.6\linewidth]{plots/model_turnover_learning.pdf}
   \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning, across different sets of exams (\emph{SAEB} and \emph{SPAECE}).
@@
     \centering
-    \includegraphics{plots/reelection_marginal_effect_plots.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/reelection_marginal_effect_plots.pdf}
     \caption{\textbf{Differential returns to patronage}.
@@
     \centering
-    \includegraphics{plots/visual_rdd_legislative_seat_flip.pdf}
+    \includegraphics[width=0.7\linewidth]{plots/visual_rdd_legislative_seat_flip.pdf}
 \end{figure}

### 2021-03-30 paper/theory/model_executive_bargain.tex
update images

@@
 
-In this section, I use a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes.
+In this section, I apply a canonical vote-buying model in legislatures \citep{groseclose_1996_buying, banks_2000_buying} to model mayor's strategy to use patronage to buy legislative votes.
 For any given electoral term, a mayor wants to enact her preferred policy, but to do so needs to pass it through a simple majority in the city council.
@@
 
-An important feature of the model is the presence of an opposition party that competes with the mayor for legislative votes.
 In this setting, passing a policy proposal requires outbidding the opposition, which generates interesting results regarding the optimal size of the coalition and conditions under which we should observe minimal, super and universal coalitions.
@@
 
-The amount of public sector positions the mayor offers to these city councilors responds to the composition of the city council.
+The amount of public sector positions the mayor offers to these city councilors also responds to the composition of the city council.
 The intuition is straightforward: if the city councilors are already likely to support the bill, there is no need to offer them additional side payments.
@@
 
-Each city councilor is characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ entails that the mayor's proposal $x$ is preferred.
-Let $\mathbf{v} = (v_1, ..., v_n)$ describe the preference profile for the city council.
+Each city councilor is characterized by a publicly observed policy preference $v_i$ for all $i \in N$, where $v_i > 0$ denotes that the mayor's proposal $x$ is preferred by legislator $i$.
+Let $\mathbf{v} = (v_1, ..., v_n)$ describe the preference profile for the city councilors.
 In this setting, $v_i$ measures the degree to which an individual city councilor supports the mayor, with higher values of $v$ denoting stronger support for the mayor and vice versa.
 Payoff are realized when city councilor $i$ votes, independent of the outcome of the voting procedure.
-This sincere voting preference forecloses the possibility of general equilibria in which $i$'s voting behavior affect $j$.
+This sincere voting preference closes the possibility of general equilibria in which $i$'s voting behavior affect $j$.
 
@@
 
-These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{Details as to why can be found in the appendix.} We can simplify the total expenditure on patronage by the government, $S(a)$, as
+These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{A detailed explanation can be found in the appendix.} We can simplify the total expenditure on patronage by the government, $S(a)$, as
 

### 2021-04-01 source/descriptive/modules/visual_staff_turnover.R
incorporate changes to turnover index

@@ -59,25 +59,24 @@ save_fig(
 # distribution of teacher turnover index
 hist_turnover_index <- censo_school_turnover %>%
     filter(
-        between(year, 2007, 2015) &
+        between(year, 2008, 2015) &
         n >= 5
     ) %>%
     ggplot(
         aes(
             turnover_index,
-            # y = stat(width * density),
-            group = year,
-            color = year
+            y = stat(width * density)
         )
     ) +
-    geom_density() +
+    geom_histogram() +
     scale_y_continuous(
         labels = scales::percent_format()
     ) +
     labs(
-        x = "Distribution of turnover index",
+        x = "Distribution` of turnover index",
         y = "Count"
-    )
+    ) +
+    facet_wrap(year ~ .)
 
 save_fig(
     hist_turnover_index,

### 2022-08-29 paper/chapter_education.tex
fix: incorporate latest version of chapter

--- 2021-03-30 paper/paper.tex
@@
-\documentclass[12pt,a4paper]{article}
+\documentclass[12pt]{report}
 \usepackage{setspace}
-\usepackage[top = 1.5in, bottom = 1.5in, left = 1.5in, right = 1.5in]{geometry}
+\usepackage[top = 1in, bottom = 1in, left = 1.5in, right = 1.5in]{geometry}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{titletoc}
-\setlength\parindent{20pt}
-\usepackage{setspace}
 \usepackage{multirow}
-\usepackage{caption}
 \usepackage{titling}
-\usepackage{amsmath}
+\usepackage[utf8]{inputenc}
 \usepackage{hyperref}
+\usepackage{kbordermatrix}
+\usepackage{kotex}
+\usepackage{epigraph}
+\usepackage{sectsty}
+\allsectionsfont{\singlespacing}
 
@@
 \DeclareMathOperator*{\argmax}{arg\,max}
-\onehalfspacing
-\usepackage[hang,flushmargin]{footmisc}
 
@@
 \newtheorem{corollary}{Corollary}
+\theoremstyle{definition}
+\newtheorem{hypothesis}{Hypothesis}
+\newtheorem{definition}{Definition}
+
+\newcommand{\wb}{\overline{w}}
+\newcommand{\pb}{\overline{p}}
+\newcommand{\bb}{\overline{b}}
+\newcommand{\de}{\delta}
+\newcommand{\St}{\mathcal{S}}
+\newcommand{\Sb}{\overline{\mathcal{S}}}
+\newcommand{\A}{\mathcal{A}}
+\newcommand{\note}{\textcolor{blue}}
 \renewcommand{\theproposition}{\arabic{proposition}}
@@
 
-\bibliographystyle{unsrtnat}
+\bibliographystyle{apsr}
 
-\title{Governing through patronage: the bargain for education in decentralized Brazil}
-\date{\today}
+\section*{Abstract:}
 
-\author{Galileu Kim}
-\affil{Princeton University}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-    How can governments improve the quality of education?
-I argue that local politicians govern through patronage, undermining education at the expense of voters.
-Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites.
-Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector.
+\emph{How do political institutions shape decisions over bureaucratic personnel?
+And what are the implications of these management decisions on public service delivery?
+I argue that local mayors govern through patronage, lowering the quality of educational services.
+Qualitative evidence gathered in fieldwork in Brazil, as well as large-scale administrative data suggest that public education is captured by local political elites.
+Mayors buy off local legislators to enact policy agendas by offering positions in the department of education.
 The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain.
 Patronage induces turnover in educational staff, with negative downstream effects on student learning.
-Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate.
-These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
-\end{abstract}
+These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.}
 
-\newpage
 \section{Introduction}
@@
 Across the world, vulnerable populations rely on governments for access to basic public services such as education.
-In the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, basic mathematical operations.\footnote{The World Bank calls this ``learning poverty".
+In the developing world, the quality of public education remains low: functional illiteracy remains widespread, basic mathematical operations inaccessible.\footnote{The World Bank calls this ``learning poverty".
 See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}.
-Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them.
-\citep{cochrane_policies_1983, falleti_decentralization_2010}.
-How can local governments improve the quality the of public services they deliver?
-In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?
+Politicians play an important role in deciding whether or not to deliver these services \citep{gulzar_politicians_2017,besley_political_2003}, but alongside elected officials, bureaucratic institutions shape how these services are delivered \citep{pepinsky_bureaucracy_2017, finan2017personnel}.
+How do politicians manage bureaucratic institutions, and what are their downstream consequences for public services?
 
-The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations.
-How can public education be improved?
+Previous studies have found that individual-level incentives, electoral or performance related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
+Often relying on targeted interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But across the developing world, political and bureaucratic actors do not govern alone.
+Public officials are embedded in institutions that shape and constrain their actions \citep{thelen_historical_1999}.
+Moreover, it is common that instead of a single political actor, multiple principals compete over the same pool of resources, requiring coalitional bargains to effectively govern \citep{francois_how_2015}.
+How do these coalitional bargains affect patronage appointments, and what are its implications for the population they are meant to serve?
 
-Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}.
-Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors relate to one another in local institutions?
-Who holds power over public education, and whose interests is it meant to serve?
+To answer these questions, I conducted fieldwork research in municipalities across Brazil over the course of two years.
+In interviews with local politicians, school principals and teachers, I uncovered the pervasive influence of clientelism, and how it extends to public education with negative consequences on educational outcomes \citep{stokes2013brokers}.
+Clientelism in the educational sector was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government that occur between elections \citep{raile_executive_2011}.
+In Brazilian municipalities, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.
 
-To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil.
-In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}.
-However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}.
-In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.
-
-In this paper, I argue that public education in municipalities across Brazil have been captured by political elites, with personnel appointments into public schools and management dominated by bargains between executive and legislative branches.
+In this paper, I argue that public education in municipalities across Brazil is a bargaining chip between political elites, with personnel appointments into public schools and management dominated by transactions occurring between executive and legislative branches.
 Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}.
-This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}.
+This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over constituency welfare \citep{ferejohn_incumbent_1986}.
 By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.
 
-The estimation proceeds in two parts.
+The empirical identification of these theoretical mechanisms proceeds in two parts.
 First, I show that political alignment between mayors and city councilors has a direct effect on patronage.
@@
 In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage.
-This results holds across a set of specifications and different measures of turnover.
+This results holds across a set of model specifications and different measures of turnover.
 
@@
 This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}.
-I analyze how incentives shape political decision on how to manage local bureaucracies \citep{duflo_incentives_2012, gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders.
+I analyze how incentives shape political decision on how to manage local bureaucracies \citep{duflo_incentives_2012, gulzar_politicians_2017}, highlighting the policy goals of executive leaders and how these require bargains that leave public service beneficiaries worse off.
 I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services.
@@
 Section \ref{sec:literature} provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil.
-Section \ref{sec:context} presents the context and data.
-In section \ref{sec:theory}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model.
-Section \ref{sec:empirics} presents the research design and main results.
-Section 5 concludes.
+Section \ref{sec:context_edu} presents the context and data.
+In section \ref{sec:theory_bargain}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model.
+Section \ref{sec:empirics_edu} presents the research design and main results.
+Section \ref{sec:conclusion_edu} concludes.
 
@@
 \label{sec:literature}
-
-In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions.
-I also highlight how my research incorporates multiplicity in political actors and how this affects bargaining over public sector jobs.
-I address this gap by adapting previous analyses of executive-legislative bargaining to bureaucratic control at the local level.
 
@@
 Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference.
-Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores.
+\citet{toral2019benefits} finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores.
 \citet{gulzar_politicians_2017} show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment.
-\citet{akhtari_political_2015}, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.
+\citet{akhtari_political_2015}, on the other hand, highlight the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.
 
@@
 These political actors have competing claims over the local bureaucracy, with important implications for public service delivery.
-This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}.
+This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle2012jobs, mccarty_appointments_2004}.
 Divergent political interests can lead to strategic interaction between executive and legislative actors.
-A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy.
-\citet{raile_executive_2011, power_optimism_2010}.
+A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy \citep{raile_executive_2011, power_optimism_2010}.
 
-In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism.
-Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
+In the Brazilian federal context, executive-legislative relations are predominantly analyzed under the prism of presidential coalitionism \citep{neto_presidential_2006}.
+Due to a multiparty system, that is common across federal, state and municipal level, rarely do presidents, governors or mayors secure a majority in the legislative chamber.
+Executive leaders garner legislative support from the legislature by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions.
 \citep{raile_executive_2011}.
-In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.
+In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, figueiredo_executivo_1999}.
 
-In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies.
+In municipalities, mayors have to, in a similar fashion, garner legislative support from city councilors to secure budgetary approval and implement desired policies.
 Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}.
-As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.
+As noted by a former mayor of the municipality of Sobral, ``city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.
 
@@
 
-Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects.
-As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population \citet{gailmard_slackers_2007}.
-Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher \citet{clotfelter_teacher_2007}.
-\citet{akhtari_political_2015} finds that students attending a school with a recently appointed school principal perform worse in standardized test scores.
-When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones \citet{colonnelli_patronage_2017}.
+Bureaucracies exposed to turnover experience performance shocks, often with detrimental effects.
+As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population \citep{gailmard_slackers_2007}.
+\citet{clotfelter_teacher_2007} finds that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher.
+\citet{akhtari_political_2015} show, in municipalities in Brazil, that students attending a school with a recently appointed school principal perform worse in standardized test scores.
+When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones \citep{colonnelli_patronage_2017}.
 
-\begin{quote}
-"I am aware that the position is temporary.
-Especially because it is a political position, decided by the administration.
-If the current administration is out of power, we are automatically dismissed." - Interview with school principal A, August 2019.
-\end{quote}
+This empirical record suggests that, as in any organization, bureaucratic turnover is often costly, especially in the initial phases of change.
+While the benefits of changes have been extolled by the embrace of New Public Management in more recent years \citep{osborne_reinventing_1993}, an attention to continuity and stability in bureaucratic organizations has been the subject of constant attention in scholarly literature on public administration \citep{o2003plus,suleiman_dismantling_2013}.
+Bureaucratic organizations are complex, and have to fulfill mandates that require familiarity with both the population to be served and the procedures set in place to comply both with external and internal regulations.
 
-Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level.
-In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers.
-In the following section I describe the institutional context for public education in Brazil, as well as the data employed for the the estimation.
+If disruption is to be beneficial, there must be time for these turnover induced changes to mature and bear fruits.
+This extended time horizon has to be coupled with an identification of organizational priorities, rather than political ones \citep{nunes_gramatica_1997}.
+If cycles of disruption are associated with political imperatives rather than bureaucratic performance ones, an additional misalignment of incentives takes hold of bureaucratic institutions.
+\citet{evans_embedded_1995} posits that high bureaucratic performance is premised on embedded autonomy.
+What that theoretical and empirical claim implies is that while bureaucracies must respond to changes in their environment -- and alter themselves in the process -- they also require a degree of autonomy to effect these gradual changes.
 
 \section{Institutional context and data}
-\label{sec:context}
+\label{sec:context_edu}
 
@@
 In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}.
-In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and substraction.
-The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school strudents attend a public school.
+In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and subtraction.
+The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school students attend a public school.
 Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_enrollment.pdf}
     \caption{\textbf{Student enrollment by administration.}}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_staff_breakdown}
 \end{figure}
@@
 While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion \citet{ferraz_corrupting_2012}.
+
+\subsection{Bureaucratic turnover}
+
+In interviews with local education officials, it became clear the extent to which turnover was detrimental to the provision of public services.
+In conversations with the \emph{n\'{u}cleo gestor} (administrative board) of a school in the municipality of I, administrators complained of the need to learn the tasks and processes required to run a school.
+This process, according to them, took over a year or even two, and if the administrative board or teachers were replaced, the administrative expertise of the school was reset and the new administration had to go through the entire process once again.
+
+In the municipality of I, one of the poorest run educational systems in the state of Cear\'{a}, bureaucratic turnover was the rule, not the exception. As put
+
+\begin{quote}
+"I am aware that the position is temporary.
+Especially because it is a political position, decided by the administration.
+If the current administration is out of power, we are automatically dismissed." - Interview with school principal A, August 2019.
+\end{quote}
+
+Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level.
+In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers.
 
@@
 Brazil collects fine-grained data on education and makes it publicly available for research.
-For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools.
-In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese.
+Data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute for Educational Studies and Research) to all public schools and a sample of private schools.
+In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both Mathematics and Portuguese.
 In this study, test scores are the primary metric for assessing the quality of education received by students.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/saeb_map.pdf}
     \caption{Uneven quality of municipal education.
@@
     \centering
-    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_staff_breakdown.pdf}
+    \caption{Proportion of public sector jobs by department}
     \label{fig:staff_turnover}
-    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
 \end{figure}
@@
 I propose three different specifications for measuring bureaucratic turnover.
-From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year.
+From the RAIS I calculate the municipal-level percentage of teachers and principals who are dismissed and hired at any given year.
 Using school census data, it is possible to track individual teachers across time and schools.
-With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}.
+With that data I calculate a turnover index for school $s$, grade $g$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}.
 The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.
 
-$$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$
+$$\text{turnover}_{sgjt} = \frac{\text{exit}_{sgjt} + \text{entry}_{sgjt}}{N_{sgjt} + N_{sgjt-1}}$$
 
-\blue{Insert more information and descriptive paragraph on turnover index here}
+The following example illustrates how the index operates.
+Suppose that in school $s$ for the year 2010, there are three teachers working.
+In the following year, one new teacher arrives and two leave.
+Our index would reflect these changes: $\text{turnover} = \frac{1 + 2}{3 + 2} = 0.6$.
+If, in contrast, no teachers left or entered the school, we would instead have a turnover index of 0 for the school.
+The added advantage of this measure is that it simultaneously captures a broader definition of bureaucratic turnover, that includes both exits and entries into the local school.
+Figure \ref{fig:hist_turnover} presents the distribution of turnover index over time.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/hist_turnover_index.pdf}
+    \caption{Distribution of turnover index by year, pooled across schools. Note that for robustness, only schools with more than 5 teachers are included.}
+    \label{fig:hist_turnover}
+\end{figure}
+
+Note that there is not significant variation over time of the turnover index, indicating that there are no visible time trends or reduction in teacher turnover.
+This indicates that turnover is a chronic challenge for schools in Brazil and has not reduced significantly over time, despite efforts by the Ministry of Education to increase professionalization of eduactional staff through more strict public exam requirements.
+It is important to highlight that entry or exit in this context does not indicate that a teacher is necessarily leaving or entering the educational system, but rather being relocated to another school within the same system.
 
@@
 
-\begin{figure}
+\begin{figure}[h]
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
-    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/hist_coalition.pdf}
+    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the eletcoral period of 2002-2016.}
     \label{fig:hist_coalition}
@@
 \section{Theory}
-\label{sec:theory}
+\label{sec:theory_bargain}
+
+In this section, I present an abridged version of the theoretical framework presented by \citet{banks_2000_buying} and \citet{groseclose_1996_buying}.
+In this framework, two players -- government and opposition -- compete over legislative votes in order to pass their preferred policy proposals.
+Each legislative member casts an aye or nay vote depending on both their a priori preferences over the competing proposals as well as the bribe schedule induced by the competition between the two players.
+There is perfect information: both parties can fully observe the actions of each other and there is no uncertainty in the model.
+
+This simplification costs us a degree of realism -- one could imagine that bribes are actually occurring in the backstage -- but gains us parsimony and tractability.
+The key comparative static of interest is how exogenous changes in the preference profile of legislative members induce changes in the bribing decisions of the government.
+Intuitively, if a legislative member opposes a proposal, they require additional bribes to vote in favor.
+Generalizing, as the preference profile of members of the legislature are more favorable (opposed) towards the policy proposal of the politician, the less (more) resources must the politician spend to garner their vote.
 
@@
 
-The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.\footnote{Note that for this model, I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
+In this setting, two players, the government $G$ and opposition $O$, compete over legislative votes to enact their preferred policies.\footnote{Note that for this model, I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition.
 In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators.
-The total amount of political resources available is $W_G$ and $W_O$, which for the mayor includes public sector appointments.
+The total amount of political resources available is $W_G$ and $W_O$, which in our context are simply public sector appointments for the mayor, and other kinds of resources -- monetary, limited appointments -- for the opposition.
 
@@
 
-We solve the game through backward induction. The timing of the game is as follows:
+The game is solved through backward induction. The timing of the game is as follows:
 
@@
 
-Following Groseclose and Snyder (1996) and Banks (2000) we focus our analysis on the set of equilibria in which the mayor wins.\footnote{Since strategies for both players are symmetrical, any set of equilibria in which the mayor loses can be modeled as cases in which the the opposition loses.} In this context, the amount of patronage resources $W_G$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$.
+Following \citet{groseclose_1996_buying} and \citet{banks_2000_buying} we focus our analysis on the set of equilibria in which the mayor wins.\footnote{Since strategies for both players are symmetrical, any set of equilibria in which the mayor loses can be modeled as cases in which the the opposition loses.} In this context, the amount of patronage resources $W_G$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$.
 Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(a) = \sum_{i = 1}^n a_i$ be the total amount of patronage disbursed.
@@
 
-I impose the following two assumptions:
+Following \citet{banks_2000_buying}, I impose the following two assumptions:
 \begin{align*}
@@
 
-% For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$.
+These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$.
 Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$.
-Then, under $A2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$: $a'_i = t(a) - v_i$, $a'_j = 0$ and $a'_m = a_m$ for all $m \notin i, j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
+Then, under $A2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$: $a'_i = t(a) - v_i$, $a'_j = 0$ and $a'_m = a_m$ for all $m \notin i, j$.
+Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e.
 $a'_i \leq a_j$.
-Also, because of $A2$ $a'_j \leq a_j$, since it guarantees $a_j$ is non-negative.} Generalizing, and recalling that $v_1 \geq ...
+Also, because of $A2$ $a'_j \leq a_j$, since it guarantees $a_j$ is non-negative.
+Generalizing, and recalling that $v_1 \geq ...
 \geq v_n$, we see that for all $a \in U^l(\mathbf{v}, W_b)$ there exists a bribe schedule $a' \in U^l(\mathbf{v}, W_b)$ such that $S(a') \leq S(a)$ and $C(a') = \{1, ..., k(a)\}$.
 Therefore, we can without loss of generality restrict attention to schedules $a$ by $A$ which bribe the first $k(a)$ voters.
-Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.
-
-These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{A detailed explanation can be found in the appendix.} We can simplify the total expenditure on patronage by the government, $S(a)$, as
+Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.} We can simplify the total expenditure on patronage by the government, $S(a)$, as
 
@@
 
-Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.
+% Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.
 
-\begin{proposition}
-    (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
-\end{proposition}
+% \begin{proposition}
+%     (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
+% \end{proposition}
 
-Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity.
+% Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity.
 Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v'}, W_B)$.
@@
 
-From equation \ref{eqn:optimal_k} we infer that $\Delta(k' - 1, \mathbf{v}', W_B) < 0$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$.
+% From equation \ref{eqn:optimal_k} we infer that $\Delta(k' - 1, \mathbf{v}', W_B) < 0$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$.
 Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$.
 Then, $v_{k'} > T(k' - 1, W_B)$, and hence $\Delta(k' - 1, \mathbf{v}, W_B) < 0$.
-But then from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$.
-Therefore, the following holds:
+Therefore, from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$.
+The following proposition thus holds:
 
-\begin{proposition}
-    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
-\end{proposition}
+% \begin{proposition}
+%     For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
+% \end{proposition}
 
-In words, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases.
+% Substantively, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases.
 Substantively, the number of voters bribed by $A$ weakly increases as the voter who receives the largest bribe finds $A$'s preferred alternative, $x$, more attractive.
-Similarly
 
-\begin{proposition}
-    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k' + 1} \leq v'_{k' + 1}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \leq k^*(\mathbf{v}', W_B)$
-\end{proposition}
-
-The ``convexity" of $E$ guarantees that local information is sufficient to generate comparative statistics regarding changes in preferences $\mathbf{v}' \rightarrow \mathbf{v}$.
-We can characterize the change in total expenditures as a result of a shift in voter preferences.
+With this theoretical framework, we can characterize change in total expenditures as a result of a shift in voter preferences.
 Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$.
@@
 Whether it be in presidential coalitionism, or in the local city council politics, mayors who wish to govern have to engage in transactions with the legislature.
-What I showed in this section was that political misalignment between the government and the legislature can in fact be counterproductive: more patronage occurs, leading to worse public service outcomes.
+What I showed in this section was that political misalignment between the government and the legislature can in fact be counterproductive: the more opposition in the legislature a mayor faces, the more patronage she has to engage in to coopt them.
 
 The model also highlights a key aspect of clientelism that is often neglected electoral accountability models: voters have a limited voice.
-Ultimately, the exchanges which occur between the legislature and the mayor have little to do with the voter at the end of the pipeline, and more to do with the city councilors.
-The first order requirement for the government is to ensure that it has enough legislative votes in order to enact the very policies that the voter may or may not desire.
-This transactional cost is not illegal: rather, it is necessary for democratic relations between different branches of government.
+Ultimately, the exchanges which occur between the legislature and the mayor have little to do with the voter at the end of the pipeline, and far more to do with the city councilors.
+The first order requirement for the mayor is to ensure that she has enough legislative votes in order to enact the very policies that the voter may or may not desire.
+This transactional cost is not illegal by design.
+It is often the cost for democratic relations between different branches of government.
 
@@
 \section{Empirical results}
-\label{sec:empirics}
+\label{sec:empirics_edu}
 
@@
 This study presents two sets of estimations.
-The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education.
+The first one estimates the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education.
 To do so, I leverage the test scores made available by \emph{SAEB} and \emph{SPAECE} to estimate the effect of turnover on student test scores.
-To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample.
+To avoid interference across units of analysis, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample.
 The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.
 
-For the first estimation, I test the hypothesis that mayors who control less seats in the city council engage in more patronage.
+For the second estimation, I test the hypothesis that mayors who control less seats in the city council engage in more patronage.
 To do so I leverage employment data from RAIS and school census data to measure teacher turnover rates.
 I deploy three sets of models.
-For the second and third estimation I turn to the micro, bureaucrat-level and a then to aggregated turnover at the municipal level.
+For the second and third estimation I turn to the micro, bureaucrat-level and then to aggregated turnover at the municipal level.
 The first model is a logistic regression where the outcome of interest is whether a particular teacher or school principal is hired/fired for any given year.
@@
 
-I first turn to testing the main proposition of this paper: that turnover in educational staff responds to the degree of political alignment between the mayor and politician.
+I first turn to testing the main proposition of this paper: that turnover in educational staff responds to the degree of political alignment between the mayor and the legislature.
 In line with theory, there is strong evidence that mayors who are electoral allies with more (less) seats in the city council resort to less (more) patronage.
@@
       \footnotesize
-      \input{results/turnover_result}
+      \input{chapters/chapter_1/results/turnover_result}
       \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/plot_coalition_coef_mun.pdf}
     \caption{{\bf Political alignment and staff turnover.} A visual representation of results presented in table \ref{tbl:turnover}. All models include year and state fixed effects.}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/plot_interactions_logit}
     \caption{\textbf{Differential reductions in patronage in teachers and school principals}.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun_second.pdf}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/plot_coalition_coef_mun_second.pdf}
     \caption{\textbf{Patronage continues into the second term.} There is strong evidence that executive-legislative bargains continue into the mayor's second term.
@@
 Moving on to characterizing the downstream effects of patronage, I find that teacher and school principal turnover have significant, negative effects on average test scores.
-These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations.
+These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE).
 Table \ref{tbl:student_learning} presents the results of the estimation on SAEB and SPAECE test scores.
-We present two specifications for turnover.
+I present two specifications for turnover.
 Turnover index measures the amount of turnover in teachers at the school level.
@@
       \footnotesize
-      \input{results/student_learning.tex}
+      \input{chapters/chapter_1/results/student_learning.tex}
       \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning.
@@
   \centering
-  \includegraphics[width=0.6\linewidth]{plots/model_turnover_learning.pdf}
+  \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/model_turnover_learning.pdf}
   \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning, across different sets of exams (\emph{SAEB} and \emph{SPAECE}).
@@
 Finally, I present some evidence of differential returns to patronage, in particular for city councilors.
-For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education.
+For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term mandate.
+While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education.
 The negative consequences of patronage on student learning, however, do not bite, as nayors who deliver a worse quality of education are no less likely to be reelected in the next term.\footnote{See appendix \ref{app:accountability}.}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{plots/reelection_marginal_effect_plots.pdf}
+    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/reelection_marginal_effect_plots.pdf}
     \caption{\textbf{Differential returns to patronage}.
@@
 \section{Conclusion}
-\label{sec:conclusion}
+\label{sec:conclusion_edu}
 
@@
 
-\bibliography{dissertation}
-
-\pagebreak
-
 \section*{Appendix}
@@
     \centering
-    \includegraphics{plots/covariate_balance.pdf}
+    \includegraphics{chapters/chapter_1/plots/covariate_balance.pdf}
 \end{figure}
@@
     \centering
-    \includegraphics{plots/accountability_fit.pdf}
+    \includegraphics{chapters/chapter_1/plots/accountability_fit.pdf}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.7\linewidth]{plots/visual_rdd_legislative_seat_flip.pdf}
+    \includegraphics[width=0.7\linewidth]{chapters/chapter_1/plots/visual_rdd_legislative_seat_flip.pdf}
 \end{figure}
 
-% \blue{plots/balance_hlm}
+% \blue{chapters/chapter_1/plots/balance_hlm}
 
@@
 
-% \blue{plots/balance_felm.pdf}
+% \blue{chapters/chapter_1/plots/balance_felm.pdf}
 
 % \pagebreak
-
-\end{document}

### 2022-08-31 paper/chapter_education.tex
feat: add literature

@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_enrollment.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
     \caption{\textbf{Student enrollment by administration.}}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_staff_breakdown}
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/saeb_map.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
     \caption{Uneven quality of municipal education.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_staff_breakdown.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown.pdf}
     \caption{Proportion of public sector jobs by department}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/hist_turnover_index.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/hist_turnover_index.pdf}
     \caption{Distribution of turnover index by year, pooled across schools. Note that for robustness, only schools with more than 5 teachers are included.}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/hist_coalition.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
     \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the eletcoral period of 2002-2016.}
@@
       \footnotesize
-      \input{chapters/chapter_1/results/turnover_result}
+      \input{results/turnover_result}
       \caption{{\bf Executive coalitions and staff turnover.} An increase in the share of legislative seats held by the mayoral coalition decrease the amount of turnover for teachers and school principals, including hires or dismissals.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/plot_coalition_coef_mun.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun.pdf}
     \caption{{\bf Political alignment and staff turnover.} A visual representation of results presented in table \ref{tbl:turnover}. All models include year and state fixed effects.}
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/plot_interactions_logit}
+    \includegraphics[width=0.6\linewidth]{plots/plot_interactions_logit}
     \caption{\textbf{Differential reductions in patronage in teachers and school principals}.
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/plot_coalition_coef_mun_second.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/plot_coalition_coef_mun_second.pdf}
     \caption{\textbf{Patronage continues into the second term.} There is strong evidence that executive-legislative bargains continue into the mayor's second term.
@@
       \footnotesize
-      \input{chapters/chapter_1/results/student_learning.tex}
+      \input{results/student_learning.tex}
       \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning.
@@
   \centering
-  \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/model_turnover_learning.pdf}
+  \includegraphics[width=0.6\linewidth]{plots/model_turnover_learning.pdf}
   \caption{{\bf Bureaucratic turnover and student learning} Teacher and school principal turnover have a negative effect on student learning, across different sets of exams (\emph{SAEB} and \emph{SPAECE}).
@@
     \centering
-    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/reelection_marginal_effect_plots.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/reelection_marginal_effect_plots.pdf}
     \caption{\textbf{Differential returns to patronage}.
@@
     \centering
-    \includegraphics{chapters/chapter_1/plots/covariate_balance.pdf}
+    \includegraphics{plots/covariate_balance.pdf}
 \end{figure}
@@
     \centering
-    \includegraphics{chapters/chapter_1/plots/accountability_fit.pdf}
+    \includegraphics{plots/accountability_fit.pdf}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.7\linewidth]{chapters/chapter_1/plots/visual_rdd_legislative_seat_flip.pdf}
+    \includegraphics[width=0.7\linewidth]{plots/visual_rdd_legislative_seat_flip.pdf}
 \end{figure}
 
-% \blue{chapters/chapter_1/plots/balance_hlm}
+% \blue{plots/balance_hlm}
 
@@
 
-% \blue{chapters/chapter_1/plots/balance_felm.pdf}
+% \blue{plots/balance_felm.pdf}
 

### 2022-08-31 paper/paper.tex
fix: fix compiling of paper

@@
+% !TEX root = paper.tex
 \documentclass[12pt,a4paper]{article}
@@
 
-\title{Governing through patronage: the bargain for education in decentralized Brazil}
+\title{Governing through patronage: the bargain for education in municipalities of Brazil}
 \date{\today}
@@
 \author{Galileu Kim}
-\affil{Princeton University}
+\affil{DIME, World Bank}
 
@@
 \begin{abstract}
-    How can governments improve the quality of education?
-I argue that local politicians govern through patronage, undermining education at the expense of voters.
-Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites.
-Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector.
+    How do political institutions shape decisions over bureaucratic personnel?
+And what are the implications of these management decisions on public service delivery?
+I argue that local mayors govern through patronage, lowering the quality of educational services.
+Qualitative evidence gathered in fieldwork in Brazil, as well as large-scale administrative data suggest that public education is captured by local political elites.
+Mayors buy off local legislators to enact policy agendas by offering positions in the department of education.
 The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain.
 Patronage induces turnover in educational staff, with negative downstream effects on student learning.
-Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate.
 These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
@@
 
-\newpage
 \section{Introduction}
@@
     \centering
-    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
+    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown.pdf}
+    \caption{Proportion of public sector jobs by department}
     \label{fig:staff_turnover}
-    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
 \end{figure}
@@
 I propose three different specifications for measuring bureaucratic turnover.
-From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year.
+From the RAIS I calculate the municipal-level percentage of teachers and principals who are dismissed and hired at any given year.
 Using school census data, it is possible to track individual teachers across time and schools.
-With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}.
+With that data I calculate a turnover index for school $s$, grade $g$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}.
 The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.
 
-$$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$
+$$\text{turnover}_{sgjt} = \frac{\text{exit}_{sgjt} + \text{entry}_{sgjt}}{N_{sgjt} + N_{sgjt-1}}$$
 
-\blue{Insert more information and descriptive paragraph on turnover index here}
+The following example illustrates how the index operates.
+Suppose that in school $s$ for the year 2010, there are three teachers working.
+In the following year, one new teacher arrives and two leave.
+Our index would reflect these changes: $\text{turnover} = \frac{1 + 2}{3 + 2} = 0.6$.
+If, in contrast, no teachers left or entered the school, we would instead have a turnover index of 0 for the school.
+The added advantage of this measure is that it simultaneously captures a broader definition of bureaucratic turnover, that includes both exits and entries into the local school.
+Figure \ref{fig:hist_turnover} presents the distribution of turnover index over time.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.6\linewidth]{plots/hist_turnover_index.pdf}
+    \caption{Distribution of turnover index by year, pooled across schools. Note that for robustness, only schools with more than 5 teachers are included.}
+    \label{fig:hist_turnover}
+\end{figure}
+
+Note that there is not significant variation over time of the turnover index, indicating that there are no visible time trends or reduction in teacher turnover.
+This indicates that turnover is a chronic challenge for schools in Brazil and has not reduced significantly over time, despite efforts by the Ministry of Education to increase professionalization of eduactional staff through more strict public exam requirements.
+It is important to highlight that entry or exit in this context does not indicate that a teacher is necessarily leaving or entering the educational system, but rather being relocated to another school within the same system.
 
@@
 
-\begin{figure}
+\begin{figure}[h]
     \centering
     \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
-    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
+    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the eletcoral period of 2002-2016.}
     \label{fig:hist_coalition}
@@
 \pagebreak
-
 \section*{Appendix}

### 2022-08-31 source/models/lmer/multilevel.R

@@ -1,29 +1,0 @@
-library(foreign)
-library(tidyverse)
-
-
-snijders <- read.dta("https://data.princeton.edu/pop510/snijders.dta")
-
-snijders <- snijders %>%
-    mutate(
-        iqvc = iq_verb - mean(iq_verb)
-    )
-
-
-fits <- group_by(snijders, schoolnr) %>% 
-    do( lf = lm(langpost ~ iqvc, data = .))
-
-ols <- data.frame(id = fits[[1]], t(sapply(fits[[2]],coef)))
-
-names(ols) <- c("schoolnr", "sa", "sb")
-
-snijders <- snijders %>%
-    left_join(
-        ols,
-        by = "schoolnr"
-    ) %>%
-    mutate(
-        fv = sa + sb * iqvc
-    )
-
-ggplot(snijders, aes(iqvc, fv, group = schoolnr)) + geom_line()
